### 2018-07-25 phonon_polaritons_manuscript/paper.tex

\documentclass[superscriptaddress,reprint,prb]{revtex4-1}
\usepackage{graphicx}
\usepackage[margin=0.7in]{geometry}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{multirow}

\renewcommand{\Im}{\operatorname{Im}}
\renewcommand{\Re}{\operatorname{Re}}
\newcommand{\sub}[1]{\ensuremath{_{\textrm{#1}}}} %Upright multi-character subscript
\newcommand{\super}[1]{\ensuremath{^{\textrm{#1}}}} %Upright multi-character superscript

\newcommand{\HarvardSEAS}{John A. Paulson School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, USA}
\newcommand{\MITPhy}{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA, USA}

\usepackage[usenames]{color}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{TODO(#1)}}}
\newcommand{\edited}[1]{{\color{red} #1}}

\begin{document}

\title{Theory of phonon polaritons in polar dielectrics}

\author{Nicholas Rivera*}\email{nrivera@seas.harvard.edu}\affiliation{\HarvardSEAS}\affiliation{\MITPhy}
\author{Jennifer Coulter*}\email{jcoulter@g.harvard.edu}\affiliation{\HarvardSEAS}
\author{Prineha Narang}\email{prineha@seas.harvard.edu}\affiliation{\HarvardSEAS}

\date{\today}

\begin{abstract}
The ability to use photonic quasiparticles to control electromagnetic energy far below the diffraction limit is a defining paradigm in nanophotonics.  Largely, the manifestation of this idea has been plasmonics, which have been demonstrated to confine electromagnetic fields on the scale of nanometers. A more recent development is the measurement and manipulation of extremely confined phonon polariton modes in polar dielectrics such as silicon carbide and hexagonal boron nitride. These modes pave the way for nanophotonics and extreme light-matter interactions in the mid-IR to THz frequency band. To further advance this promising field, it is of great interest to compute the optical response of polar dielectrics from first principles. In this work, we develop linear response theory for an ionic lattice coupled to the electromagnetic field in arbitrary dimensions and derive expressions for the nonlocal dielectric function. While in some limits, our results qualitatively match the famous Lorentz oscillator model, we are able to specify its parameters in terms of the masses of the atoms in the unit cell, the tensor of Born effective charges of each atom in the unit cell,  phonon band-structure, and the phonon polarizations.  Moreover, our results allow us to point to a regime where quantum nonlocal effects may correct the phonon polariton dispersion, which may be relevant in experiments which confine electromagnetic fields to the scale of 1 nm. Our results also allow us for the first time to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of ab-initio-derived parameters of the underlying lattice. We derive the electrodynamics of 2D phonon polaritons, finding their fields and dispersion. Finally, to show the practical usage of the framework derived, we provide ab-initio calculations for silicon carbide, and show that they agree well with experimental data.
\end{abstract}

\maketitle

Phonon polaritons, hybrid quasiparticles of photons and optical phonons, offer great promise for deeply sub-diffractional control of electromagnetic fields at mid-IR and THz frequencies. These polaritons share many features in common with plasmon-polaritons in conductors, a heavily well-studied topic. Namely, in recent years, it has been shown that phonon polaritons offer the ability to confine light to volumes over $10^6$ times smaller than that of a photon in free-space at the diffraction limit \cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra}. Because of this, and their relatively high lifetimes of around picoseconds, they allow new opportunities for vibrational spectroscopy, for radiative heat transfer \cite{hillenbrand2002phonon}, and for control of dynamics in quantum emitters \cite{kumar2015tunable,rivera2017making}. Largely, the core features of phonon polaritons, such as mode shape, confinement, and propagation characteristics, can be understood through the properties of the Lorentz oscillator dielectric function, which has been successfully applied to model experiments involving the launching and propagation of these polaritons.

There are nevertheless some outstanding questions regarding the behavior of phonon polaritons. Here, we present a few. 
\begin{enumerate}
\item{Can the very successful Lorentz oscillator model ever break down? One particular mechanism of breakdown, which could be present in the case of extremely confined polaritons, concerns effects coming from spatial dispersion. Such short phonon polariton wavelengths are indeed predicted by classical electrodynamics of a polar dielectric, especially in the case of a very thin film (nm-thick).}
\item{What are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators? Can they support extremely confined polaritonic modes, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene?}
\item{In general, are the properties of optical phonons inherited by the phonon polariton? What is the influence of concepts such as interface phonons and phonon topology on the polariton? The ability to design new phonon polariton materials by designing the optical phonon band structure allows a predictive way to generate polaritonic materials.}
\end{enumerate}

Indeed, the way to proceed on all of these questions is an ab-initio framework to predict the phonon contribution to the dielectric tensor in arbitrary settings. Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in software packages such as ABINIT \cite{abinit1,abinit2,abinit3}. Here, we develop a unified 
theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality. Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in two- and three-dimensional materials. We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles. From this theory, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid  but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.

\section{Linear response of phonon polaritons in three dimensions}

\textcolor{blue}{Pri: one place we can get criticized here by an experimentalist (or theorist) is that at least for the 3D spatially-local case, there's this thing that Caldwell called the TO-LO formalism in his proposal. It's not actually a formalism... it's just an expression for the dielectric function in terms of the high-frequency electronic part, and the various TO and LO phonon frequencies, as well as the damping. It features some fairly simple-to-get experimental parameters (though actually I don't know how they get the LO frequency since only TO comes up in Raman I think), and reproduces the data generally pretty well. I think actually this is what they did in that arXiv paper that worried us about a month ago. Anyway, when we do nonlocal and reduced-dimensional stuff, I think this is where we can really be different. In any event, it's good to anticipate this criticism in advance so we can potentially get in front of it. What do you think?}

\textcolor{blue}{To all: I'm new to phonons and am likely making some non-standard notational choices? If so, feel free to point out where and I'll try to standardize it.} We start by reviewing the theory of phonon polaritons in a 3D bulk, developing the theory in a way that is parallel to the linear response theory of electronic contributions to the permittivity and plasmonic properties. Here, the matter subsystem which couples to light is the ionic lattice of the crystal. Thus, we develop the linear response theory of an ionic lattice to an electromagnetic field. The interaction Hamiltonian of the lattice with the electromagnetic field is taken as
\begin{equation}
H_{int} = -\int d^3x ~\mathbf{E}(\mathbf{x},t)\cdot\mathbf{P}(\mathbf{x},t),
\end{equation}
where $\mathbf{E}$ is a driving electric field, and $\mathbf{P}$ is the polarization associated with the lattice. We write the polarization as a sum of contributions from all unit cells in the system. Namely
\begin{equation}
\mathbf{P}(\mathbf{r},t) = \sum\limits_{\text{sites, }\mathbf{R}}\mathbf{d}(\mathbf{R},t)\delta(\mathbf{r}-\mathbf{R}),
\end{equation}
where $\mathbf{d}(\mathbf{R},t)$ is the time-dependent dipole moment associated with lattice site $\mathbf{R}$. In expressing the polarization in this fashion, we are assuming that the displacement of the ions is far smaller than the size of the unit cell, which is, even in a deeply anharmonic regime, a good approximation. This said, we operate in the harmonic regime.  The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa} \mathbf{b}_{\kappa} \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$. Note that the Born effective charges are both tensorial and site-dependent. \textcolor{blue}{I still want to give a decent intuition for these Born effective charges as I find them a bit mysterious and so will a reader who is from photonics.} This displacement can be expanded in phonon modes as
\begin{equation}
\mathbf{u}(\mathbf{R}_{\kappa},t) = \sum\limits_{\mathbf{q}\sigma}\sqrt{\frac{\hbar}{2M_{\kappa}N\omega_{\mathbf{q}\sigma}}}\left( e^{i\mathbf{q}\cdot\mathbf{R}_{\kappa}-i\omega_{\mathbf{q}\sigma}t}\hat{e}_{\mathbf{q}\sigma}(\mathbf{b}_{\kappa})a_{\mathbf{q}\sigma} + \text{h.c.} \right),
\end{equation}
where $M_{\kappa}$ is the mass of atom $\kappa$, $\omega_{qs}$ is the phonon frequency at wavevector $\mathbf{q}$ and branch $s$, and $N$ is the number of unit cells.
Assuming that the field is weak, the expectation value of the polarization in the lattice will depend linearly on the total field acting on the ions: this field is the sum of the external field with the field resulting from induced electronic and ionic polarization, and it is this field which appears in the interaction Hamiltonian of Equation (1). We define the linear-response relation for a bulk crystal in Fourier space as $\langle\mathbf{P}(\mathbf{q},\omega)\rangle = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, where we refer to $\boldsymbol{\Pi}(\mathbf{q},\omega)$ as the polarization-polarization response function.

To connect this polarization-polarization response function to the observed dielectric function, we write down Maxwell's equations for the electric field sourced by a polarization density. We have that
\begin{equation}
\left(-\mathbf{q}\times\mathbf{q}\times - \epsilon_{\infty}\frac{\omega^2}{c^2} \right)\mathbf{E} = \omega^2\mu_0\mathbf{P}(\mathbf{q},\omega),
\end{equation}
where $\omega$ is the frequency of light under consideration $\mathbf{P}$ is the polarization at that frequency, and $\epsilon_{\infty}$ is the high-frequency dielectric function associated with the electrons. Now, we identify the average polarization $\langle\mathbf{P}\rangle$ with the classical polarization $\mathbf{P}$ in Maxwell's equations. We note that an analogous procedure is also employed in the linear-response theory of the longitudinal response of electrons, where instead of the polarization density being averaged, the charge density is averaged. And instead of the solving Maxwell's equations for the field, one solves Laplace's equation for the potential.  Using the linear response relation $\mathbf{P} = \epsilon_0\boldsymbol{\Pi}\mathbf{E}$, we have simply that
\begin{equation}
\left(-\mathbf{q}\times\mathbf{q}\times - \left(\epsilon_{\infty} + \boldsymbol{\Pi} \right)\frac{\omega^2}{c^2}\right)\mathbf{E} = 0,
\end{equation}
leading to the identification of of the dielectric tensor $\boldsymbol{\epsilon} = \epsilon_{\infty} + \boldsymbol{\Pi}$.

Now, we determine the polarization-polarization response function. It is given by the Kubo formula at finite temperature via
\begin{equation}
\boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 V}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i\frac{\Gamma}{2}}\left(e^{\beta E_m}-e^{\beta E_n} \right),
\end{equation}
%where $\Gamma$ is a positive infinitesimal, and it is to be understood that rigorously, the limit $\Gamma \rightarrow 0$ is to be taken. In a situation where there is phonon dissipation, typically due to electron-phonon and phonon-phonon scattering, we make take a relaxation-time approximation and interpret $\Gamma$ as the finite relaxation time. 
where $m$ and $n$ are taken to be eigenstates of the lattice, $V$ is a normalization volume, and we have written the polarization-polarization susceptibility in Fourier space. It is  understood that the polarization operator is now the Schrodinger operator. $\Gamma$ is a positive infinitesimal, and it is to be understood that rigorously, the limit $\Gamma \rightarrow 0$ is to be taken. In a situation where there is phonon dissipation, typically due to electron-phonon and phonon-phonon scattering, we make take a relaxation-time approximation and interpret $\Gamma$ as the finite relaxation time.   The eigenstates span the phononic Fock space. For simplicity, we will work at zero temperature from here onwards. In this case, in the first term, the states $m$ must be zero-phonon states, while in the second term, the states $n$ must be zero-phonon states. Due to the displacement being linear in creation and annihilation operators, the states $n$ and $m$ in the first and second terms (resp.) must be one-phonon states. The Fourier transformed polarization operators are found to be
\begin{equation}
\mathbf{P}(\mathbf{q}) = \sum\limits_{\mathbf{R}}\mathbf{d}(\mathbf{R})e^{-i\mathbf{q}\cdot\mathbf{R}} \equiv \mathbf{d}(\mathbf{q}),
\end{equation}
where it is to be understood that $\mathbf{P}(\mathbf{q})$ refers to a continuous Fourier transform while $\mathbf{d}(\mathbf{q})$ refers to a discrete Fourier transform. Plugging in the displacement operator of Equation (3), one finds that $\mathbf{d}(\mathbf{q})$ is given by
\begin{equation}
\mathbf{d}(\mathbf{q}) = \sum\limits_{\sigma}\sqrt{\frac{\hbar N}{2\omega_{\mathbf{q}\sigma}}} \left(\mathbf{S}_{\mathbf{q}\sigma}a_{\mathbf{q}\sigma} + \text{h.c.}\right),
\end{equation}
where 
\begin{equation}
\mathbf{S}_{\mathbf{q}\sigma} = \sum\limits_{\kappa} \frac{1}{\sqrt{M_{\kappa}}}\mathbf{Z}_{\kappa}\hat{e}_{\mathbf{q}\sigma}(\mathbf{R}_{\kappa}).
\end{equation}
This can be expressed in terms of more standard outputs of ab-initio software that calculates phonon properties. In particular, we can take $\hat{e}_{\mathbf{q}\sigma}$, the eigenvectors of the dynamical matrix, and define eigendisplacements via $\boldsymbol{\eta}_{\mathbf{q}\sigma\kappa}=\frac{\hat{e}_{\mathbf{q}\sigma}}{\sqrt{M_{\kappa}}}$. In which case, we may cast Equation (9) into the form:
\begin{equation}
\mathbf{S}_{\mathbf{q}\sigma} = \sum\limits_{\kappa} \mathbf{Z}_{\kappa}\boldsymbol{\eta}_{\mathbf{q}\sigma\kappa}.
\end{equation}
We may now use this in concurrence with Equation (6) to evaluate the dielectric function. Consider $\boldsymbol{\Pi}(\mathbf{q},\omega)$. From the form of Equation (8), we see that at zero temperature, the first term only sees contributions from $m=|0_{\mathbf{q}\sigma}\rangle$ and $n=|1_{\mathbf{q}\sigma}\rangle$, where $0_{\mathbf{q}\sigma}$ and $1_{\mathbf{q}\sigma}$ refer to the number of phonons in mode $\mathbf{q}\sigma$. Here $\sigma$ can be any branch of the phonon band structure. In the second term, $m$ and $n$ switch. In the limit where the phonons are very weakly damped, $\Gamma_{\mathbf{q}\sigma} \\ \omega_{\mathbf{q}\sigma}$. this leads to:
\begin{equation}
\boldsymbol{\Pi}(\mathbf{q},\omega) = \frac{1}{\Omega}\sum\limits_{\sigma}\frac{\mathbf{S}_{\mathbf{q}\sigma}\otimes \mathbf{S}_{\mathbf{q}\sigma}^*}{\omega^2_{\mathbf{q}\sigma}-\omega^2-i\Gamma_{\mathbf{q}\sigma}\omega},
\end{equation}
where $\Omega=N/V$ is the volume of the unit cell. Rigorously, the phonon frequencies in this equations are real. However, in the realistic situation where there is phonon dissipation, typically due to electron-phonon and phonon-phonon scattering, we may make take a "relaxation-time approximation" which effectively results in expressing $\omega_{\mathbf{q}\sigma} \rightarrow \omega_{\mathbf{q}\sigma} - i\frac{\Gamma_{\mathbf{q}\sigma}}{2}$, where $\Gamma_{\mathbf{q}\sigma}$ is the dissipation rate of the phonon. Finally, the dielectric function takes the form
\begin{equation}
\epsilon(\mathbf{q},\omega) = \epsilon_{\infty} + \frac{1}{\Omega}\sum\limits_{\sigma}\frac{\mathbf{S}_{\mathbf{q}\sigma}\otimes \mathbf{S}_{\mathbf{q}\sigma}^*}{\omega^2_{\mathbf{q}\sigma}-\omega^2-i\Gamma_{\mathbf{q}\sigma}\omega}.
\end{equation}
From this equation, let us consider an extremely simplified limit. Assume that the permittivity is isotropic. Further assume that the site-specific masses collapse to a single effective mass $M^*$, and that the site-specific and tensorial Born effective charges collapse to a single effective charge $Q^*$. Finally, assume that the only phonon with oscillator strength (non-zero $\mathbf{S}$) is the transverse optical phonon at frequency $\omega_{TO}$. In this case, one arises at a dielectric function 
\begin{equation}
\epsilon(\omega) = \epsilon_{\infty} + \frac{NQ^{*2}}{\epsilon_0 M^*V}\frac{1}{\omega^2_{TO}-\omega^2-i\Gamma_{TO}\omega},
\end{equation}
which coincides precisely with the phenomenological Lorentz oscillator model, showing the consistency, of our approach. That said, our approach takes into account complications arising from the tensorial and site-dependent nature of the Born charges, the effects of a complex unit-cell, as well as a detailed understanding of the phonon band-structure and how it affects the features of the predicted dielectric function.

Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function. 
\textcolor{blue}{Should this part below be moved to the outlook or does it make sense to be here instead?} Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion. As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components. However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$. In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.  Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone. It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons. However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet. Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields. These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.

\section{Application to silicon carbide}
% calculation basics
To demonstrate the practical use of this theoretical framework, we present an example bulk calculation for the 4H polytype of silicon carbide (SiC), whose unit cell is shown in Figure 1(a).  %Calculations of the frequency-dependent dielectric tensor 
Phonon bandstructure of SiC were performed within density functional perturbation
theory (DFPT) as implemented in the ABINIT package\cite{abinit1,abinit2,abinit3,Gonze1997dynamical,Hamann2005metric}, with electric field response included to appropriately capture LO-TO splitting. Initial ground state density functional theory (DFT) properties were predicted using the PBEsol parameterization of the norm-conserving Vanderbilt pseudopotentials \cite{PBEsol,ONCV_PPs,pseudodojo}. We calculated the phonon modes, their frequencies (shown in Figure 1(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation (12), using experimental values for the dissipation rates $\Gamma$. We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained ab-initio by calculating lifetimes associated with electron-phonon scattering and three-phonon decay processes.

\begin{figure}[t]
\includegraphics[width=8cm]{structure.png}
\caption{\textbf{Crystal structure and phonon dispersion of 4H silicon carbide.} a) The hexagonal unit cell structure of silicon carbide allows different optical response in the a-b plane and along the c axis. b) The full phonon bandstructure of silicon carbide, from which the phonon eigendisplacements associated with the gamma point were used to calculate the frequency-dependent dielectric function.}
\label{fig:phonons}
\end{figure}

% about the scissor shift -- and a defense sentence in case someone hits us with "semi-empirical" later on
Additionally, a scissor shift of 1.01 eV was applied to shift the calculated band gap of 2.26 eV to match the experimental gap of $\sim$3.27 eV \cite{band_gap_expt1,band_gap_expt2}. This is necessary to correct for the well-known underestimation of the band gap in ground state density functional theory calculations and the corresponding effects to the dielectric response.
The high-frequency dielectric tensor was found to be $\epsilon^{\infty}_{\parallel}$ = 6.986 and $\epsilon^{\infty}_{\perp}$ = 7.307.  This also leads to a modification of the phonon spectra. % For a material in which the band gap is not as well known as in SiC, a more computationally intensive GW calculation could be used for an entirely \it{ab initio} implementation of our theory. If excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter prediction might be used to further improve predictive accuracy. 
The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier}, is shown in Figure \ref{fig:epsilon}. As can be seen, the agreement between the theoretical calculation and the experimental results is quite good.

\textcolor{red}{ Should we mention here that while we take the dissipation rate and bandgap from experiment, that these quantities can be obtained ab initio as well? We could cite packages that would allow someone to do this for a material where properties aren't well known, just to keep ourselves in the "first principles" designation. Maybe a note in the supplement or a footnote with this to prevent "semi-empirical" finger pointing?}

\textcolor{blue}{I included a sentence addressing this important point after the sentence where we say that we take the lifetimes from experiment. }

% can also put this table just by listing eps^inf_perp and parallel in the main text.  
\begin{figure}[t]
\includegraphics[width=8.5cm]{SiC_comparison_thy_vs_expt.pdf}
\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk 4H silicon carbide.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk 4H silicon carbide from using Equation (12). The phonon properties and $\epsilon_{\infty}$ calculated from the ABINIT package. Overlaid in blue dots are the experimental values for the dielectric function, taken from \cite{tiwald1999carrier}. The real part is shown in (a) while the imaginary part is shown in (b). }
\label{fig:epsilon}
\end{figure}

%Thus, to evaluate it, we must know the microscopic states of the lattice, in addition to matrix elements of displacement. Working in the harmonic approximation, the states are simply phonon-occupation states $|\{n_{qs}\}\rangle$, labeled by their wavevector $\mathbf{q}$ in the Brillouin zone and their band $s$ which encodes transverse/longitudinal polarization of acoustic/optical branches. The phonons are also labeled by their frequency $\omega_{qs}$ and the polarization vector at each basis site $b$, $\hat{e}_{\mathbf{q}s}(\mathbf{b})$. The displacement of each ion in the lattice can be expressed as an expansion over phonon modes as:
% \begin{equation}
% \mathbf{u}(\mathbf{R},\mathbf{b},t) = .
% \end{equation}
% From the displacement, we then must construct the dipole moment associated with each unit cell in the lattice. This is given by:
% \begin{equation}
% \mathbf{d}(\mathbf{R}) = e\sum\limits_{\mathbf{b}} Z_{ij}u_{j},
% \end{equation}
% where $\mathbf{Z}$ is a tensor of Born effective charges. These charges formally are defined as the derivative of the lattice polarization with respect to charge displacement. They act as effective charges for the ions, taking into account the motion of electrons and other ions in the system. Once we have the dipole moment in each unit cell, we can now complete the formal evaluation of the polarization-polarization response function. Assuming that we are operating at zero temperature, it is given by:

\section{Linear response of phonon polaritons in two dimensions}
We now move to generalize the derivation of the previous section to describe the optical phonon contribution to Maxwell's equations coming from a two-dimensional material. Because of the change of symmetry in the system, we will not be able to perform the analysis by Fourier transforming Maxwell's equations in three dimensions. However, we will apply a technique which is often used to analyze electromagnetic modes in 2D systems such as plasmons in 2D conductors. An example of the technique is shown in detail in \cite{jablan2009plasmonics}, but we summarize the essential details here.

Consider a two-dimensional ionic lattice located in the $xy$ plane ($z=0$) surrounded by a homogeneous environment of permittivity $\epsilon$. We consider the simplified case where the lattice is free-standing, to show the essential concept. We search for a mode whose magnetic field $\mathbf{H}$ is of the form $\mathbf{H}(z)e^{i\mathbf{q}\cdot\mathbf{r}_{||}-i\omega  t}$, where $\mathbf{r}_{||}=(x,y,0)$ is the position in the plane, $\mathbf{q}$ is the wavevector and $\omega_{\mathbf{q}}$ is the frequency. To simplify the discussion even further and furnish analytical forms for the phonon-polariton modes in 2D, we assume this material has mirror symmetry along the axis of $\mathbf{q}$. In this case, the electromagnetic modes can be decomposed into TE and TM-polarized modes. As is typical in the study of highly-confined polaritons, whether they are plasmon- or phonon- polaritons, the TM mode is the highly confined one, and occurs when $\epsilon < 0$. In our convention, the TM mode is such that $\mathbf{H}$ is in the plane of the lattice and perpendicular to $\mathbf{q}$. Without loss of generality, define the $\mathbf{q}$ direction to be the $x$ direction, making the magnetic field in the $y$ direction. Moreover, the magnetic field in this mode flips sign under $z \rightarrow -z$.

For $z \neq 0$, the Maxwell equation for the magnetic field is $(\nabla\times\nabla\times - \epsilon\frac{\omega^2}{c^2})\mathbf{H} = 0$. This equation, for the TM mode described above, becomes 
\begin{equation}
\left(-\frac{d^2}{dz^2}+\mathbf{q}^2-\epsilon\frac{\omega^2}{c^2} \right)H = 0,
\end{equation}
where we have written $\mathbf{H} = H\hat{y}$. The solution in the region $z > 0$ is simply \begin{align}
H &= e^{i\mathbf{q}\cdot\mathbf{r}_{||}-\kappa z}, ~z>0 \nonumber \\
H &= -e^{i\mathbf{q}\cdot\mathbf{r}_{||}-\kappa z}, ~z>0 \nonumber \\
\end{align}
where $\kappa = \sqrt{\mathbf{q}^2-\epsilon\frac{\omega^2}{c^2}}$ and we have implicitly assumed $q > \epsilon\frac{\omega}{c}$ as we are interested in confined evanescent modes. To join these two solutions consistently, we apply the interface conditions for the magnetic field. They are $\hat{z}\times(\mathbf{H}^{(+)}-\mathbf{H}^{(-)}) = \mathbf{K} = -i\omega\mathbf{P}_s$ where $\mathbf{K}$ is the surface current density, expressed through the surface polarization density $\mathbf{P}_s$. 
Now we use the linear response relation $\mathbf{P}(\mathbf{q},\omega) = \boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$ in addition to Ampere's law $\nabla\times\mathbf{H} = -i\omega\epsilon\mathbf{E}$. Plugging these relations in yields an implicit equation connecting the polarization response function, the frequency of the 2D phonon polariton, and its frequency. Taking the case where $x$ coincides with a principal axis of the system, we have that 
\begin{equation}
q\Pi_{xx}(q,\omega)  = -2\epsilon.
\end{equation}

To connect this relation to the discussion of the three-dimensional case, we write a microscopic form for the polarization response function of the ionic lattice. It is simply
\begin{equation}
\boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 A}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i0^+}\left(e^{\beta E_m}-e^{\beta E_n} \right),
\end{equation}
where the only change is to replace a normalization volume with a normalization area $A$. Here, $\mathbf{P}(\mathbf{q})$ is the 2D Fourier transform of the polarization density. As the form of the displacement of the ions is still given by Equation (3), $\mathbf{P}(\mathbf{q})$ is still given by Equations (7-10), but with the Born charges and eigendisplacements appropriate to the 2D system of interest. Note that although the lattice is two dimensional, in general, the ions may be displaced in any three directions and thus $\boldsymbol{\Pi}$ is still a $3\times3$ matrix. Following the same kind of reasoning that lead to Equation (13), if we parameterize the effect of the different masses and Born charges in the unit cell by $M^*$ and $Q^*$, we will find a form for the polarization response function which is given by the second term of Equation (13) but with $V$ replaced by $A$. Defining $N/A \equiv n_s$, Equation (14) for the dispersion relation of 2D phonon polaritons tells us that
\begin{equation}
\omega = \sqrt{\omega_{TO}^2+\frac{n_sQ^{*2}}{2M^*\epsilon_0\epsilon}q}.
\end{equation}
Note that this dispersion has a strong resemblance to the plasmon dispersion relation of two-dimensional electron gases, which is normally of the form $\omega = \sqrt{\frac{n_s e^2  }{2m\epsilon_0\epsilon}q}$ \cite{stern1967polarizability}. The only difference is the replacement of electron parameters (density, charge, mass) by the corresponding lattice parameters in the phononic case, and a shift in the minimum frequency by the transverse optical phonon frequency.

\textcolor{blue}{Still looking for more to say about this. This probably requires some brainstorming discussion between us. For now, what I'll do is prepare a plot of the Fourier transform of the green's function of the Maxwell equations for the 2D sheet (p-polarized reflectivity, as was done in Argentene).}

% We now move to generalize the derivation of the previous section to describe the optical phonon contribution to Maxwell's equations coming from a two-dimensional material. Because of the change of symmetry in the system, we will not be able to perform the analysis by Fourier transforming Maxwell's equations in three dimensions. Nevertheless, we may proceed as follows. Consider Maxwell's equations with a source density given by
% \begin{equation}
% \mathbf{P}(\mathbf{r}) = \mathbf{P}(\mathbf{r}_{||})\delta(z) = \sum\limits_{\mathbf{R}}\mathbf{d}(\mathbf{R})\delta(\mathbf{R}-\mathbf{r})\delta(z).
% \end{equation}
% Taking the fourier transform in two dimensions gives:
% \begin{equation}
% \mathbf{P}(\mathbf{q},\omega) = \delta(z)\sum\limits_{\mathbf{R}}\mathbf{d}(\mathbf{R})e^{-i\mathbf{q}\cdot\mathbf{R}}.
% \end{equation}
% Fourier transforming the Maxwell equations in two dimensions, leaving the third unchanged connects the fourier transformed polarization density to the  
% \begin{equation}
% \left(-\frac{d^2}{dz^2} + \mathbf{q}^2 - \frac{\omega^2}{c^2} \right)\mathbf{E}(\mathbf{q},\omega) = \mu_0\omega^2\mathbf{P}(\mathbf{q},\omega)\delta(z)
% \end{equation}
% To connect to response functions, we note that: the Hamiltonian is given by Equation (1), integrated over two dimensions, the linear response relation is $\langle \mathbf{P}(\mathbf{q},\omega) = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, and that the susceptibility is still defined by Equation (6), except that the units of the polarization density are different, and the volume factor becomes an area factor $A$. Given this 2D polarizability of the ionic lattice, we may solve Maxwell's equations for the resultant fields, as well as the dispersion relation of the phonon-polariton modes. We consider for simplicity an isotropic case

%While we could work in Fourier space in two dimensions and real-space in the third, we employ a simpler argument for the phonon-polariton modes of a 2D structure and the response function that captures such phonon-polariton response.

%In particular, what we will do is, in the spirit of the self-consistent mean-field approximation, find an expression for the expectation value of the dipole moment at each site in the two-dimensional lattice. Then, using Maxwell's equations, write the electromagnetic field as a superposition of fields from the various dipoles. In particular, suppose we have a two-dimensional lattice with sites $\mathbf{R}$ in the xy-plane (i.e., the two-dimensional layer is located at $z=0$). The polarization density can then be expressed as 
%\begin{equation}
%\mathbf{P}(\mathbf{r}) = \sum\limits_{\mathbf{R}}\mathbf{d}(\mathbf{R})\delta(\mathbf{r}-\mathbf{R})\delta(z)
%\end{equation}

%The dipole moment associated with each unit cell is proportional to this total field, and then we may solve a self-consistent equation based on the polarization to find the

% Anything else we need?

\section{Outlook}

\textcolor{blue}{Pri: blue text below means this particular suggestion is heavily inspired from what Caldwell sent us.} In summary, we have provided a theoretical framework based on linear response theory to calculate the phonon contribution to the dielectric function from first principles. Our framework is rather versatile, allowing us to: use first principles calculations to get the dielectric function, predict how nonlocality enters the dielectric function, and predict the effect of reduced dimensionality. A particularly interesting case to consider in future work would be to find a system where the optical phonons are drastically different from 3D to 2D. Perhaps it is possible that there are some materials in which the 2D optical phonons experience lower losses due to a reduced scattering phase space. The formalism we provide here may also be extended to understand phonon-polaritons in other more atypical reduced-dimensional settings, such as zero-dimensional settings in single emitters, i.e., "molecular phonon polaritons", in analogy to recent work on "molecular plasmons" \cite{manjavacas2013tunable,lauchner2015molecular}. Notably, we go beyond oversimplifications of the Lorentz oscillator model in which the Born charges are treated as a single scalar quantity and can treat the influence and interplay of many phonon modes that contribute to the dielectric function. We corroborated our approach through density functional theoretic calculations.

In the long run, besides furnishing an ab-initio description of nonlocality and reduced-dimension phonon polaritons, there are a number of interesting questions that can be addressed by the framework provided here. For example, we may consider situations such as \textcolor{blue}{the phonon polaritons associated with optical phonons at the interface of two materials. Another situation of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other.} In that case, it would be of interest to see how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.

\bibliographystyle{apsrev4-1}
\bibliography{references}

\end{document}

### 2018-08-02 paper.tex
update manuscript
made edits discussed yesterday

--- 2018-07-25 phonon_polaritons_manuscript/paper.tex
@@
 The ability to use photonic quasiparticles to control electromagnetic energy far below the diffraction limit is a defining paradigm in nanophotonics.
-Largely, the manifestation of this idea has been plasmonics, which have been demonstrated to confine electromagnetic fields on the scale of nanometers.
-A more recent development is the measurement and manipulation of extremely confined phonon polariton modes in polar dielectrics such as silicon carbide and hexagonal boron nitride.
-These modes pave the way for nanophotonics and extreme light-matter interactions in the mid-IR to THz frequency band.
-To further advance this promising field, it is of great interest to compute the optical response of polar dielectrics from first principles.
-In this work, we develop linear response theory for an ionic lattice coupled to the electromagnetic field in arbitrary dimensions and derive expressions for the nonlocal dielectric function.
-While in some limits, our results qualitatively match the famous Lorentz oscillator model, we are able to specify its parameters in terms of the masses of the atoms in the unit cell, the tensor of Born effective charges of each atom in the unit cell,  phonon band-structure, and the phonon polarizations.
+An important recent development along this paradigm is the measurement and manipulation of extremely confined phonon polariton modes in polar dielectrics such as silicon carbide and hexagonal boron nitride, which pave the way for nanophotonics and extreme light-matter interactions in the mid-IR to THz frequency band.
+To further advance this promising field, it is of great interest to predict the optical response of polar dielectrics from first principles.
+In this work, we develop linear response theory for an ionic lattice coupled to the electromagnetic field in arbitrary dimensions and derive expressions for the dielectric function.
+For a 3D bulk in the spatially local regime, our results qualitatively match the famous Lorentz oscillator model.
+However, we are able to specify its parameters in terms of the masses of the atoms in the unit cell, the tensor of Born effective charges of each atom in the unit cell,  and the phonon-bandstructure.
 Moreover, our results allow us to point to a regime where quantum nonlocal effects may correct the phonon polariton dispersion, which may be relevant in experiments which confine electromagnetic fields to the scale of 1 nm.
-Our results also allow us for the first time to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of ab-initio-derived parameters of the underlying lattice.
+Our results also allow us for the first time to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of the same ab-initio-derived parameters of the underlying lattice.
 We derive the electrodynamics of 2D phonon polaritons, finding their fields and dispersion.
 Finally, to show the practical usage of the framework derived, we provide ab-initio calculations for silicon carbide, and show that they agree well with experimental data.
+Our results provide a stepping stone for eventually considering from first principles the effects of interface phonons, phonon strong coupling, and chiral phonons on the properties of phonon polaritons.
 \end{abstract}
@@
 
-There are nevertheless some outstanding questions regarding the behavior of phonon polaritons. Here, we present a few. 
-\begin{enumerate}
-\item{Can the very successful Lorentz oscillator model ever break down?
+There are nevertheless some outstanding questions regarding the behavior of phonon polaritons.
+Here, we present a few.
+First, can the very successful Lorentz oscillator model ever break down?
 One particular mechanism of breakdown, which could be present in the case of extremely confined polaritons, concerns effects coming from spatial dispersion.
-Such short phonon polariton wavelengths are indeed predicted by classical electrodynamics of a polar dielectric, especially in the case of a very thin film (nm-thick).}
-\item{What are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators?
-Can they support extremely confined polaritonic modes, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene?}
-\item{In general, are the properties of optical phonons inherited by the phonon polariton?
-What is the influence of concepts such as interface phonons and phonon topology on the polariton?
-The ability to design new phonon polariton materials by designing the optical phonon band structure allows a predictive way to generate polaritonic materials.}
-\end{enumerate}
+Such short phonon polariton wavelengths are indeed predicted by classical electrodynamics of a polar dielectric, especially in the case of a very thin film (nm-thick).
+Second, what are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators?
+Can they support extremely confined polaritonic modes, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene?
+In general, are the properties of optical phonons inherited by the phonon polariton?
+Finally, what is the influence of non-trivial spatial shaping of phonons on the resulting polariton?
+Examples of non-trivial spatial shaping include chiral phonons, interface phonons, and mechanical heterostructurs.
+The ability to design new phonon polariton materials by designing the optical phonon band structure allows a predictive way to generate polaritonic nano-materials.
 
-Indeed, the way to proceed on all of these questions is an ab-initio framework to predict the phonon contribution to the dielectric tensor in arbitrary settings.
+The way to proceed on all of these questions is an ab-initio framework to predict the phonon contribution to the dielectric tensor in arbitrary settings.
 Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in software packages such as ABINIT \cite{abinit1,abinit2,abinit3}.
@@
 
-\textcolor{blue}{Pri: one place we can get criticized here by an experimentalist (or theorist) is that at least for the 3D spatially-local case, there's this thing that Caldwell called the TO-LO formalism in his proposal.
-It's not actually a formalism...
-it's just an expression for the dielectric function in terms of the high-frequency electronic part, and the various TO and LO phonon frequencies, as well as the damping.
-It features some fairly simple-to-get experimental parameters (though actually I don't know how they get the LO frequency since only TO comes up in Raman I think), and reproduces the data generally pretty well.
-I think actually this is what they did in that arXiv paper that worried us about a month ago.
-Anyway, when we do nonlocal and reduced-dimensional stuff, I think this is where we can really be different.
-In any event, it's good to anticipate this criticism in advance so we can potentially get in front of it.
-What do you think?}
-
-\textcolor{blue}{To all: I'm new to phonons and am likely making some non-standard notational choices?
-If so, feel free to point out where and I'll try to standardize it.} We start by reviewing the theory of phonon polaritons in a 3D bulk, developing the theory in a way that is parallel to the linear response theory of electronic contributions to the permittivity and plasmonic properties.
+We start by developing the theory of phonon polaritons in a 3D bulk in a way that is parallel to the linear response theory of electronic contributions to the permittivity and plasmonic properties.
 Here, the matter subsystem which couples to light is the ionic lattice of the crystal.
@@
 Note that the Born effective charges are both tensorial and site-dependent.
-\textcolor{blue}{I still want to give a decent intuition for these Born effective charges as I find them a bit mysterious and so will a reader who is from photonics.} This displacement can be expanded in phonon modes as
+This displacement can be expanded in phonon modes as
 \begin{equation}
@@
 Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function.
-\textcolor{blue}{Should this part below be moved to the outlook or does it make sense to be here instead?} Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.
+Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.
 As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components.
@@
 As can be seen, the agreement between the theoretical calculation and the experimental results is quite good.
-
-\textcolor{red}{ Should we mention here that while we take the dissipation rate and bandgap from experiment, that these quantities can be obtained ab initio as well?
-We could cite packages that would allow someone to do this for a material where properties aren't well known, just to keep ourselves in the "first principles" designation.
-Maybe a note in the supplement or a footnote with this to prevent "semi-empirical" finger pointing?}
-
-\textcolor{blue}{I included a sentence addressing this important point after the sentence where we say that we take the lifetimes from experiment. }
 
@@
 
-\textcolor{blue}{Pri: blue text below means this particular suggestion is heavily inspired from what Caldwell sent us.} In summary, we have provided a theoretical framework based on linear response theory to calculate the phonon contribution to the dielectric function from first principles.
+In summary, we have provided a theoretical framework based on linear response theory to calculate the phonon contribution to the dielectric function from first principles.
 Our framework is rather versatile, allowing us to: use first principles calculations to get the dielectric function, predict how nonlocality enters the dielectric function, and predict the effect of reduced dimensionality.
@@
 In the long run, besides furnishing an ab-initio description of nonlocality and reduced-dimension phonon polaritons, there are a number of interesting questions that can be addressed by the framework provided here.
-For example, we may consider situations such as \textcolor{blue}{the phonon polaritons associated with optical phonons at the interface of two materials.
-Another situation of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other.} In that case, it would be of interest to see how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.
+For example, we may consider situations such as the phonon polaritons associated with optical phonons at the interface of two materials.
+Another situation of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other.
+In that case, it would be of interest to see how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.
 

### 2018-08-02 paper.tex
some paper edits on 2d and nonlocal

@@
 Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.
-As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components.
-However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.
+Therefore, it is possible that the phonon polariton resonances red-shift or blue-shift, depending on how the relative position of the phonons at finite-wavevector versus zero wavevector.
+For example, in 4H-SiC, upon moving along the $\Gamma-L$ direction of the Brillouin zone, the optical phonon which determines the phonon polariton, calculated to be at 764 cm$^{-1}$ at $\Gamma$, red-shifts by 2  cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, by 5 cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, and 13 cm$^{-1}$ at 1/2 of the way along the $\Gamma-L$ direction.
+This will lead to a red-shift of the Reststrahlen bands by these amounts.
+
+%$\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.  I
 In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.
@@
 These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.
+
+%This is exemplified in Figure 3, where we show the ionic contribution to the permittivity (i.e.
+we do not consider the electronic part).
+We consider wavevector directions along $\Gamma-L$ and consider wavevectors that are $1/10$ and $1/4$ of the way along the Brillouin zone.
+We see that the major effect is to red-shift the phonon resonance by a few inverse centimeters.
+However, the effect "downstream", in the region where phonon polaritons are, is much smaller, with the permittivity being almost the same as in the local case.
+In making this plot, we make the assumption that the variation of the Born effective charges is negligible upon moving to finite-wavevector optical phonons.
+
+%However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.
 
@@
 We now move to generalize the derivation of the previous section to describe the optical phonon contribution to Maxwell's equations coming from a two-dimensional material.
+Unlike the previous section, we will not demonstrate the results of the calculation on a particular material.
+Here, we are only interested in finding an expression for the optical response of the surface in terms of ab-initio-derivable parameters, as well as the general form of the dispersion relation of phonon-polaritons in 2D.
 Because of the change of symmetry in the system, we will not be able to perform the analysis by Fourier transforming Maxwell's equations in three dimensions.
@@
 \end{equation}
-Note that this dispersion has a strong resemblance to the plasmon dispersion relation of two-dimensional electron gases, which is normally of the form $\omega = \sqrt{\frac{n_s e^2  }{2m\epsilon_0\epsilon}q}$ \cite{stern1967polarizability}.
+Note that this dispersion has a strong resemblance to the plasmon dispersion relation of two-dimensional electron gases, which is normally of the form\cite{stern1967polarizability} $\omega = \sqrt{\frac{n_s e^2  }{2m\epsilon_0\epsilon}q}$ .
 The only difference is the replacement of electron parameters (density, charge, mass) by the corresponding lattice parameters in the phononic case, and a shift in the minimum frequency by the transverse optical phonon frequency.
 
-\textcolor{blue}{Still looking for more to say about this.
+We note that the dispersion relation of Equation (8) can in fact be obtained by taking the infinitely thin-limit of a bulk polar dielectric, and finding the dispersion of the resulting phonon polaritons in the quasi-electrostatic limit.
+In particular, for an isotropic polar dielectric of thickness $2d$ with boundaries at $z=\pm d$, solving Laplace's equations yields as the dispersion of the even parity mode:
+\begin{equation}
+\tanh(qd) = -\frac{\epsilon}{\epsilon(\omega)} \rightarrow qd\epsilon(\omega) = -\epsilon,
+\end{equation}
+with $\epsilon$ the dielectric constant of the surrounding background and $\epsilon(\omega)$ that of the polar dielectric.
+The right-hand side is obtained in the limit of $d\rightarrow 0$.
+Taking Equation (13) for the dielectric function in bulk, we have that (in the absence of losses) $\epsilon + qd\epsilon_{\infty} + q\frac{NQ^{*2}}{2\epsilon_0 M^*A}\frac{1}{\omega_{TO}^2-\omega^2} = 0$, where we have identified $V = 2Ad$.
+In the limit of small $d$, i.e., $qd \ll 1$, we recover Equations (16) and (18) with $\Pi = \frac{NQ^{*2}}{\epsilon_0 MA}\frac{1}{\omega^2_{TO}-\omega^2}$.
+A useful result which follows is that we may also derive another important figure of merit, the $p-$polarized reflectivity of a 2D polar dielectric, which is related to the local density of states of the polar dielectric, which quantifies the strength of coupling to external probes such as a nano-antenna or a quantum emitter.
+By considering the infinitely thin limit of a polar dielectric with dielectric function $\epsilon(\omega)$ and defining $\epsilon(\omega) = \frac{\Pi}{2d}$, we immediately find that the p-polarized reflectivity is given by
+\begin{equation}
+r_p = \frac{1}{1+\frac{2\epsilon}{q\Pi}}.
+\end{equation}
+
+Before concluding, we note that in general, when considering bulk and 2D versions of a material, the effective parameters $Q^*$ and $M$ will change as a result of changes to the phonon band structure.
+The losses may also change drastically in the reduction of dimensions.
+For example, it was found in several experimental works that the damping rate of the TO phonon in hexagonal boron nitride nearly doubles upon going from bulk to two-dimensions \cite{gorbachev2011hunting,tran2016quantum}, which may strongly suppress the ability to image the polariton with near-field probes.
+This is because due to the small thickness, the dispersion of the phonon polariton is quite flat, and very high wavevector modes, corresponding to the maximum wavevector detectable by a near-field tip, occur at frequencies very close to the TO phonon frequency.
+Near the TO phonon-frequency however, the ratio $\frac{\text{Im }\Pi}{\text{Re }\Pi}$ is the highest, corresponding to very low quality-factor modes.
+In addition to these considerations, the values of the parameters may depend on the background material, which may change the phonon band structure.
+
+We provided this thin-film derivation as a simple way to understand the content of Equation (16).
+In general Equation (17) should be used to calculate $\Pi$ and the corresponding dispersion.
+We note that a similar situation appears in the treatment of 2D plasmons: namely that a relation like (16) appears, except with $\Pi$ the polarization-polarization response of a 2D electron gas.
+And furthermore that one can also obtain the same equation by considering the ultra-thin limit of a 3D plasmonic material \cite{jablan2009plasmonics,jablan2013plasmons}.
+
+%\textcolor{blue}{Still looking for more to say about this.
 This probably requires some brainstorming discussion between us.

### 2018-08-03 paper.tex
addition of hBN calculation information

@@
 
+Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function.
+Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.
+As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components.
+However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.
+In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.
+Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone.
+It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons.
+However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet.
+Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields.
+These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.
+
 \section{Application to silicon carbide}
 % calculation basics
-To demonstrate the practical use of this theoretical framework, we present an example bulk calculation for the 4H polytype of silicon carbide (SiC), whose unit cell is shown in Figure 1(a).
-%Calculations of the frequency-dependent dielectric tensor
-Phonon bandstructure of SiC were performed within density functional perturbation
+To demonstrate the practical use of this theoretical framework, we present an example bulk calculation for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in Figure 1(a). 
+Phonon bandstructures for SiC and hBN were performed within density functional perturbation
 theory (DFPT) as implemented in the ABINIT package\cite{abinit1,abinit2,abinit3,Gonze1997dynamical,Hamann2005metric}, with electric field response included to appropriately capture LO-TO splitting.
-Initial ground state density functional theory (DFT) properties were predicted using the PBEsol parameterization of the norm-conserving Vanderbilt pseudopotentials \cite{PBEsol,ONCV_PPs,pseudodojo}.
+Initial ground state density functional theory (DFT) properties were predicted using norm-conserving Vanderbilt pseudopotentials of the PBEsol parameterization for SiC, and PBE with a D2 (Grimme) van der Waals correction for hBN \cite{PBEsol,PBE,ONCV_PPs,pseudodojo,grimme}.
 We calculated the phonon modes, their frequencies (shown in Figure 1(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation (12), using experimental values for the dissipation rates $\Gamma$.
@@
 % about the scissor shift -- and a defense sentence in case someone hits us with "semi-empirical" later on
-Additionally, a scissor shift of 1.01 eV was applied to shift the calculated band gap of 2.26 eV to match the experimental gap of $\sim$3.27 eV \cite{band_gap_expt1,band_gap_expt2}.
+Additionally, for SiC a scissor shift of 1.01 eV was applied to shift the calculated band gap of 2.26 eV to match the experimental gap of $\sim$3.27 eV \cite{band_gap_expt1,band_gap_expt2}.
 This is necessary to correct for the well-known underestimation of the band gap in ground state density functional theory calculations and the corresponding effects to the dielectric response.
-The high-frequency dielectric tensor was found to be $\epsilon^{\infty}_{\parallel}$ = 6.986 and $\epsilon^{\infty}_{\perp}$ = 7.307.
+In materials like hBN where excitonic effects play a significant role, the optical properties of the system are not determined solely by the energy associated with band to band recombination, and instead must also account for the exciton binding energy\cite{robert2016excitonic}.
+Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{bandgap_hbn} and E$_{b}$ = 0.67 eV from recent ab-initio calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV.
+As the bandgap we compute from DFT is 4.48 eV, we did not require a correction to the gap in hBN.
+From these calculations, the high-frequency dielectric tensor for SiC was found to be $\epsilon^{\infty}_{\parallel}$ = 6.986, $\epsilon^{\infty}_{\perp}$ = 7.307 and for hBN, $\epsilon^{\infty}_{\parallel}$ = 4.634, $\epsilon^{\infty}_{\perp}$ = 2.593.
 This also leads to a modification of the phonon spectra.
@@
 If excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter prediction might be used to further improve predictive accuracy.
-The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier}, is shown in Figure \ref{fig:epsilon}.
+The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figure \ref{fig:epsilon}.
 As can be seen, the agreement between the theoretical calculation and the experimental results is quite good.

### 2018-08-03 paper.tex
commit

--- 2018-08-02 paper.tex
@@
 Can they support extremely confined polaritonic modes, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene?
-In general, are the properties of optical phonons inherited by the phonon polariton?
 Finally, what is the influence of non-trivial spatial shaping of phonons on the resulting polariton?
@@
 Here, we develop  a unified
-theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality.
+theoretical framework based on linear response to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality.
 Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in two- and three-dimensional materials.
-We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles.
-From this theory, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid  but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.
+We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles, and allows us to point to the distinct kinds of non-localities that appear in phonons.
+Using this theory, we first demonstrate that we may accurately predict the permittivity of bulk polar dielectrics such as silicon carbide.
+Using this framework, we show how to extend it to predict the properties of phonon polaritons in monolayers.
+Although we do not provide an ab-initio example of the phonon polariton dispersion in 2D, we, using reasonable simplifying assumptions, predict a universal form for the dispersion relation of phonon polaritons in two dimensions.
 
@@
 This said, we operate in the harmonic regime.
-The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa} \mathbf{b}_{\kappa} \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$.
+The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa}  \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$.
 Note that the Born effective charges are both tensorial and site-dependent.
@@
 
-Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function.
+We now briefly comment on spatial nonlocality in the phonon contribution to the dielectric function.
 Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.
@@
 For example, in 4H-SiC, upon moving along the $\Gamma-L$ direction of the Brillouin zone, the optical phonon which determines the phonon polariton, calculated to be at 764 cm$^{-1}$ at $\Gamma$, red-shifts by 2  cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, by 5 cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, and 13 cm$^{-1}$ at 1/2 of the way along the $\Gamma-L$ direction.
-This will lead to a red-shift of the Reststrahlen bands by these amounts.
+This will lead to a red-shift of the lower edge of the Reststrahlen band by these amounts.
+Beyond the shifts in the frequencies, the oscillator strengths ($\mathbf{S}_{\mathbf{q}\sigma}\otimes\mathbf{S}_{\mathbf{q}\sigma}$) of the phonon modes will also change as a function of frequency, and this will affect the strength of the phonon contribution to the permittivity.
 
@@
 %However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.
+
+%Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function.
+Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.
+As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components.
+However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.
+In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.
+Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone.
+It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons.
+However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet.
+Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields.
+These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.
+%>>>>>>> 3608b47a3ff25448fce9243424e9cbe4bd5888a3
 
@@
 We now move to generalize the derivation of the previous section to describe the optical phonon contribution to Maxwell's equations coming from a two-dimensional material.
-Unlike the previous section, we will not demonstrate the results of the calculation on a particular material.
+Unlike the previous section, we will not demonstrate the formalism on a particular material.
 Here, we are only interested in finding an expression for the optical response of the surface in terms of ab-initio-derivable parameters, as well as the general form of the dispersion relation of phonon-polaritons in 2D.
@@
 Note that although the lattice is two dimensional, in general, the ions may be displaced in any three directions and thus $\boldsymbol{\Pi}$ is still a $3\times3$ matrix.
+
 Following the same kind of reasoning that lead to Equation (13), if we parameterize the effect of the different masses and Born charges in the unit cell by $M^*$ and $Q^*$, we will find a form for the polarization response function which is given by the second term of Equation (13) but with $V$ replaced by $A$.
@@
 The only difference is the replacement of electron parameters (density, charge, mass) by the corresponding lattice parameters in the phononic case, and a shift in the minimum frequency by the transverse optical phonon frequency.
+We note that in principle, there may be an additional background dielectric screening from the 2D sheet itself $\epsilon_{\infty}$, in which case, it would be added to $\boldsymbol{\Pi}$, leading to a shift in the dispersion with the overall form remaining the same.
 
@@
 \begin{equation}
-r_p = \frac{1}{1+\frac{2\epsilon}{q\Pi}}.
+r_p = \frac{1}{1+\frac{2\epsilon}{q\Pi}},
 \end{equation}
+whose poles as expected correspond to the dispersion relation of Equation (16).
 

### 2018-08-03 paper.tex
update text

--- 2018-08-03 paper.tex
@@
 We derive the electrodynamics of 2D phonon polaritons, finding their fields and dispersion.
-Finally, to show the practical usage of the framework derived, we provide ab-initio calculations for silicon carbide, and show that they agree well with experimental data.
+Finally, to show the practical usage of the framework derived, we provide ab-initio calculations for silicon carbide and hexagonal boron nitride, and show that they in general agree well with experimental data.
 Our results provide a stepping stone for eventually considering from first principles the effects of interface phonons, phonon strong coupling, and chiral phonons on the properties of phonon polaritons.
@@
 
-\section{Application to silicon carbide}
+\section{Application to bulk materials}
 % calculation basics
@@
 \label{fig:epsilon}
+\end{figure}
+
+\begin{figure}[t]
+\includegraphics[width=8.5cm]{hBN_comparison_thy_vs_expt.pdf}
+\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk hexagonal boron nitride.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk hexagonal boron nitride from using Equation (12).
+The phonon properties and $\epsilon_{\infty}$ calculated from the ABINIT package.
+Overlaid in blue dots are the experimental values for the dielectric function, taken from \cite{caldwell2014sub}.
+The component of the real part of the permittivity parallel to the c-axis is shown in (a) component parallel is shown in (b).}
+\label{fig:phonons}
 \end{figure}

### 2018-08-03 paper.tex
update to structure figures

@@
 % calculation basics
-To demonstrate the practical use of this theoretical framework, we present an example bulk calculation for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in Figure 1(a). 
-Phonon bandstructures for SiC and hBN were performed within density functional perturbation
+To demonstrate the practical use of this theoretical framework, we present an example bulk computation for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in Figure 1(a). 
+Phonon bandstructures for SiC and hBN were calculated within density functional perturbation
 theory (DFPT) as implemented in the ABINIT package\cite{abinit1,abinit2,abinit3,Gonze1997dynamical,Hamann2005metric}, with electric field response included to appropriately capture LO-TO splitting.
@@
 \begin{figure}[t]
-\includegraphics[width=8cm]{structure.png}
+\includegraphics[width=8cm]{sic-structure.pdf}
 \caption{\textbf{Crystal structure and phonon dispersion of 4H silicon carbide.} a) The hexagonal unit cell structure of silicon carbide allows different optical response in the a-b plane and along the c axis.

### 2018-08-03 paper.tex
combined figures, added a ref, draft is ready

@@
 
-\author{Nicholas Rivera*}\email{nrivera@seas.harvard.edu}\affiliation{\HarvardSEAS}\affiliation{\MITPhy}
-\author{Jennifer Coulter*}\email{jcoulter@g.harvard.edu}\affiliation{\HarvardSEAS}
+\author{Nicholas Rivera$^\perp$}\email{nrivera@seas.harvard.edu}\affiliation{\HarvardSEAS}\affiliation{\MITPhy}
+\author{Jennifer Coulter$^\perp$}\email{jcoulter@g.harvard.edu}\affiliation{\HarvardSEAS}
 \author{Prineha Narang}\email{prineha@seas.harvard.edu}\affiliation{\HarvardSEAS}
@@
 Second, what are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators?
-Can they support extremely confined polaritonic modes, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene?
+Can they support extremely confined polaritonic modes which allow for extreme light-matter interactions, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene \cite{jablan2009plasmonics, koppens2011graphene, fei2012gate, chen2012optical,rivera2016shrinking,lundeberg2017tuning}?
 In general, are the properties of optical phonons inherited by the phonon polariton?
 Finally, what is the influence of non-trivial spatial shaping of phonons on the resulting polariton?
-Examples of non-trivial spatial shaping include chiral phonons, interface phonons, and mechanical heterostructurs.
+Examples of non-trivial spatial shaping include but are not limited to chiral phonons \cite{zhu2018observation} and mechanical heterostructures \cite{ratchford2018controlling}.
 The ability to design new phonon polariton materials by designing the optical phonon band structure allows a predictive way to generate polaritonic nano-materials.
@@
 This will lead to a red-shift of the Reststrahlen bands by these amounts.
+\begin{figure*}[t]
+\includegraphics[width=16cm]{combined_structure.pdf}
+\caption{\textbf{Crystal structure and phonon dispersion of bulk 4H silicon carbide and hexagonal boron nitride.} (a) The hexagonal unit cell structure of silicon carbide allows different optical response in the a-b plane and along the c axis.
+(b) The full phonon bandstructure of silicon carbide, from which the phonon eigendisplacements associated with the gamma point were used to calculate the frequency-dependent dielectric function.
+Structure (looking from stacking plane) (c) and phonon bandstructure (d) as before but for bulk hexagonal boron nitride.}
+\label{fig:phonons}
+\end{figure*}
 
@@
 % calculation basics
-To demonstrate the practical use of this theoretical framework, we present an example bulk computation for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in Figure 1(a). 
+To demonstrate the practical use of this theoretical framework, we present an example calculation of the dielectric function for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in Figure 1(a).
 Phonon bandstructures for SiC and hBN were calculated within density functional perturbation
@@
 
-\begin{figure}[t]
-\includegraphics[width=8cm]{sic-structure.pdf}
-\caption{\textbf{Crystal structure and phonon dispersion of 4H silicon carbide.} a) The hexagonal unit cell structure of silicon carbide allows different optical response in the a-b plane and along the c axis.
-b) The full phonon bandstructure of silicon carbide, from which the phonon eigendisplacements associated with the gamma point were used to calculate the frequency-dependent dielectric function.}
-\label{fig:phonons}
-\end{figure}
-
 % about the scissor shift -- and a defense sentence in case someone hits us with "semi-empirical" later on
@@
 In materials like hBN where excitonic effects play a significant role, the optical properties of the system are not determined solely by the energy associated with band to band recombination, and instead must also account for the exciton binding energy\cite{robert2016excitonic}.
-Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{bandgap_hbn} and E$_{b}$ = 0.67 eV from recent ab-initio calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV.
+Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{levinshtein2001properties} and E$_{b}$ = 0.67 eV from recent ab-initio calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV.
 As the bandgap we compute from DFT is 4.48 eV, we did not require a correction to the gap in hBN.

### 2018-08-03 paper.tex
minor tweak... now i'm done :)

@@
 Our results also allow us for the first time to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of the same ab-initio-derived parameters of the underlying lattice.
-We derive the electrodynamics of 2D phonon polaritons, finding their fields and dispersion.
-Finally, to show the practical usage of the framework derived, we provide ab-initio calculations for silicon carbide and hexagonal boron nitride, and show that they in general agree well with experimental data.
+To show the practical usage of the framework derived, we provide ab-initio calculations for bulk silicon carbide and bulk hexagonal boron nitride, and show that they in general agree well with the measured values of their dielectric functions.
+Finally, to show the generality of the framework developed here, we derive the electrodynamics of 2D phonon polaritons, finding a universal form for their electromagnetic modes and dispersion relation, which can be specified in terms of ab-initio parameters.
 Our results provide a stepping stone for eventually considering from first principles the effects of interface phonons, phonon strong coupling, and chiral phonons on the properties of phonon polaritons.
@@
 theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality.
-Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in two- and three-dimensional materials.
+Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in polar materials of arbitrary dimensionality.
 We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles.
 From this theory, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid  but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.
+We apply this formalism to calculate the dielectric function of silicon carbide and hexagonal boron nitride from first principles, and show that the results agree well with measurements.
+Finally, we develop the theory of phonon polaritons in a 2D polar material, specifying the dielectric function in terms of ab-initio parameters, and then finding a universal form for the dispersion relation of 2D phonon polaritons.
 
@@
 Now, we identify the average polarization $\langle\mathbf{P}\rangle$ with the classical polarization $\mathbf{P}$ in Maxwell's equations.
+This is essentially the random-phase approximation (RPA).
 We note that an analogous procedure is also employed in the linear-response theory of the longitudinal response of electrons, where instead of the polarization density being averaged, the charge density is averaged.
@@
 The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figure \ref{fig:epsilon}.
-As can be seen, the agreement between the theoretical calculation and the experimental results is quite good.
+As can be seen, the agreement between the theoretical calculation and the experimental results is quite good for silicon carbide, and good for the upper Reststrahlen band of hBN.
+For the lower Reststrahlen band, the major discrepancy results from a roughly 5\% deviation between calculated and measured transverse optical phonon frequencies.
 

### 2018-08-05 paper.tex
Edits.

@@
-\documentclass[superscriptaddress,reprint,prb]{revtex4-1}
+\documentclass[aps,prb,twocolumn,
+	groupedaddress,superscriptaddress,
+	amsfonts,amssymb,amsmath,floatfix,
+	citeautoscript]{revtex4-1}
+
 \usepackage{graphicx}
-\usepackage[margin=0.7in]{geometry}
-\usepackage{amsmath}
+\usepackage[centering,hmargin=20mm,tmargin=30mm,bmargin=25mm]{geometry}
+\usepackage{multirow}
+\usepackage{newtxtext}
+\usepackage[cmintegrals]{newtxmath}
+
+%----- References -----
+\usepackage{xcolor}
 \usepackage{hyperref}
-\usepackage{multirow}
+\hypersetup{colorlinks,
+	linkcolor={blue!75!black!80!yellow},
+	citecolor={blue!75!black!80!yellow},
+	urlcolor={blue!75!black!80!yellow}
+}
+
+%----- Captions in sans font -----
+\makeatletter
+\renewcommand\@make@capt@title[2]{%
+	\@ifx@empty\float@link{\@firstofone}{\expandafter\href\expandafter{\float@link}}%
+	\sffamily{\textbf{#1}}\@caption@fignum@sep#2
+}%
+\renewcommand\figurename{Fig.}
+\makeatother
+
+\thickmuskip=5mu plus 2mu minus 1mu  %binary relations (default, 5mu plus 5mu)
+\medmuskip=4mu plus 2mu minus 2mu    %binary operations (default, 4mu plus 2mu minus 4mu)
+
+\frenchspacing %Ensure that revTeX does not do "double spaces" after punctuation
 
@@
 \usepackage[usenames]{color}
-\newcommand{\TODO}[1]{\textcolor{red}{\textbf{TODO(#1)}}}
 \newcommand{\edited}[1]{{\color{red} #1}}
@@
 
-\title{Theory of phonon polaritons in polar dielectrics}
+\title{Theory of Phonon Polaritons in Polar Dielectrics}
 
@@
 The ability to use photonic quasiparticles to control electromagnetic energy far below the diffraction limit is a defining paradigm in nanophotonics.
-An important recent development along this paradigm is the measurement and manipulation of extremely confined phonon polariton modes in polar dielectrics such as silicon carbide and hexagonal boron nitride, which pave the way for nanophotonics and extreme light-matter interactions in the mid-IR to THz frequency band.
+An important recent development in this field is the measurement and manipulation of extremely confined phonon-polariton modes in polar dielectrics such as silicon carbide and hexagonal boron nitride, which pave the way for nanophotonics and extreme light-matter interactions in the mid-IR to THz frequency band.
 To further advance this promising field, it is of great interest to predict the optical response of polar dielectrics from first principles.
-In this work, we develop linear response theory for an ionic lattice coupled to the electromagnetic field in arbitrary dimensions and derive expressions for the dielectric function.
+Here we develop a linear response theory for an ionic lattice coupled to the electromagnetic field in arbitrary dimensions and derive expressions for the dielectric function.
 For a 3D bulk in the spatially local regime, our results qualitatively match the famous Lorentz oscillator model.
-However, we are able to specify its parameters in terms of the masses of the atoms in the unit cell, the tensor of Born effective charges of each atom in the unit cell,  and the phonon-bandstructure.
-Moreover, our results allow us to point to a regime where quantum nonlocal effects may correct the phonon polariton dispersion, which may be relevant in experiments which confine electromagnetic fields to the scale of 1 nm.
-Our results also allow us for the first time to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of the same ab-initio-derived parameters of the underlying lattice.
-To show the practical usage of the framework derived, we provide ab-initio calculations for bulk silicon carbide and bulk hexagonal boron nitride, and show that they in general agree well with the measured values of their dielectric functions.
-Finally, to show the generality of the framework developed here, we derive the electrodynamics of 2D phonon polaritons, finding a universal form for their electromagnetic modes and dispersion relation, which can be specified in terms of ab-initio parameters.
-Our results provide a stepping stone for eventually considering from first principles the effects of interface phonons, phonon strong coupling, and chiral phonons on the properties of phonon polaritons.
+However, unlike previous empirical descriptions, we are able to specify Lorentz oscillator parameters in terms of \emph{ab initio} material properties namely the masses of the atoms in the unit cell, the tensor of Born effective charges of each atom in the unit cell, and the phonon-dispersion (bandstructure).
+Moreover, our results allow us to identify a regime where quantum nonlocal effects may correct the phonon polariton dispersion, extremely relevant in atomic-scale experiments which confine electromagnetic fields to the scale of 1 nm.
+Our results also allow us, for the first time, to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of the same \emph{ab initio}-derived parameters of the underlying lattice.
+To demonstrate the framework derived, we provide \emph{ab initio} calculations for bulk silicon carbide and bulk hexagonal boron nitride, and show that they agree well with the measured values of their dielectric functions.
+Finally, to show the generality of the framework developed here, we derive the electrodynamics of 2D phonon polaritons, finding a universal form for their electromagnetic modes and dispersion relation, which can be specified in terms of \emph{ab initio} parameters.
+Our results point the way towards first principles descriptions of the effect of interface phonons, phonon strong coupling, and chiral phonons on the properties of phonon polaritons.
 \end{abstract}
@@
 These polaritons share many features in common with plasmon-polaritons in conductors, a heavily well-studied topic.
-Namely, in recent years, it has been shown that phonon polaritons offer the ability to confine light to volumes over $10^6$ times smaller than that of a photon in free-space at the diffraction limit \cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra}.
-Because of this, and their relatively high lifetimes of around picoseconds, they allow new opportunities for vibrational spectroscopy, for radiative heat transfer \cite{hillenbrand2002phonon}, and for control of dynamics in quantum emitters \cite{kumar2015tunable,rivera2017making}.
-Largely, the core features of phonon polaritons, such as mode shape, confinement, and propagation characteristics, can be understood through the properties of the Lorentz oscillator dielectric function, which has been successfully applied to model experiments involving the launching and propagation of these polaritons.
+In recent years, it has been shown that phonon polaritons offer the ability to confine light to volumes over $10^6$ times smaller than that of a photon in free-space at the diffraction limit \cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra}.
+Due to this remarkable confinement and their relatively high lifetimes of around picoseconds, phonon polaritons open new opportunities for vibrational spectroscopy, for radiative heat transfer \cite{hillenbrand2002phonon}, and for control of dynamics in quantum emitters \cite{kumar2015tunable,rivera2017making}.
+Typically, the core features of phonon polaritons, such as mode shape, confinement, and propagation characteristics, can be understood through the properties of the Lorentz oscillator dielectric function, which has been successfully applied to model experiments involving the launching and propagation of these polaritons.
 
-There are nevertheless some outstanding questions regarding the behavior of phonon polaritons.
+There are nevertheless some critical questions regarding the behavior of phonon polaritons.
 Here, we present a few.
-First, can the very successful Lorentz oscillator model ever break down?
+First, would the very successful Lorentz oscillator model ever break down and how?
 One particular mechanism of breakdown, which could be present in the case of extremely confined polaritons, concerns effects coming from spatial dispersion.
 Such short phonon polariton wavelengths are indeed predicted by classical electrodynamics of a polar dielectric, especially in the case of a very thin film (nm-thick).
-Second, what are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators?
-Can they support extremely confined polaritonic modes which allow for extreme light-matter interactions, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene \cite{jablan2009plasmonics, koppens2011graphene, fei2012gate, chen2012optical,rivera2016shrinking,lundeberg2017tuning}?
-In general, are the properties of optical phonons inherited by the phonon polariton?
+Second, what are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators and can they support extremely confined polaritonic modes which allow for extreme light-matter interactions, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene \cite{jablan2009plasmonics, koppens2011graphene, fei2012gate, chen2012optical,rivera2016shrinking,lundeberg2017tuning}?
+In general, how would the properties of optical phonons inherited by the phonon polariton?
 Finally, what is the influence of non-trivial spatial shaping of phonons on the resulting polariton?
@@
 
-The way to proceed on all of these questions is an ab-initio framework to predict the phonon contribution to the dielectric tensor in arbitrary settings.
-Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in software packages such as ABINIT \cite{abinit1,abinit2,abinit3}.
+The way to proceed on all of these questions is an \emph{ab initio} framework to predict the phonon contribution to the dielectric tensor in arbitrary settings.
+Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in \emph{ab initio} materials physics software packages such as \cite{abinit1,abinit2,abinit3}.
 Here, we develop a unified
@@
 We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles.
-From this theory, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid  but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.
+From this theoretical framework, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.
 We apply this formalism to calculate the dielectric function of silicon carbide and hexagonal boron nitride from first principles, and show that the results agree well with measurements.
-Finally, we develop the theory of phonon polaritons in a 2D polar material, specifying the dielectric function in terms of ab-initio parameters, and then finding a universal form for the dispersion relation of 2D phonon polaritons.
+Finally, we develop the theory of phonon polaritons in a 2D polar material, specifying the dielectric function in terms of \emph{ab initio} parameters, and then finding a universal form for the dispersion relation of 2D phonon polaritons.
 
@@
 Here, the matter subsystem which couples to light is the ionic lattice of the crystal.
-Thus, we develop the linear response theory of an ionic lattice to an electromagnetic field.
-The interaction Hamiltonian of the lattice with the electromagnetic field is taken as
+Thus, we develop the linear response theory of an ionic lattice coupled to an electromagnetic field.
+The interaction Hamiltonian of the lattice with the electromagnetic field is taken as:
 \begin{equation}
@@
 We note that an analogous procedure is also employed in the linear-response theory of the longitudinal response of electrons, where instead of the polarization density being averaged, the charge density is averaged.
-And instead of the solving Maxwell's equations for the field, one solves Laplace's equation for the potential.
+And instead of the solving Maxwell's equations for the field, we solve Laplace's equation for the potential.
 Using the linear response relation $\mathbf{P} = \epsilon_0\boldsymbol{\Pi}\mathbf{E}$, we have simply that
@@
 $\Gamma$ is a positive infinitesimal, and it is to be understood that rigorously, the limit $\Gamma \rightarrow 0$ is to be taken.
-In a situation where there is phonon dissipation, typically due to electron-phonon and phonon-phonon scattering, we make take a relaxation-time approximation and interpret $\Gamma$ as the finite relaxation time.
+In cases where there is phonon dissipation, typically due to electron-phonon and phonon-phonon scattering, we make take a relaxation-time approximation and interpret $\Gamma$ as the finite relaxation time.
 The eigenstates span the phononic Fock space.
@@
 where it is to be understood that $\mathbf{P}(\mathbf{q})$ refers to a continuous Fourier transform while $\mathbf{d}(\mathbf{q})$ refers to a discrete Fourier transform.
-Plugging in the displacement operator of Equation (3), one finds that $\mathbf{d}(\mathbf{q})$ is given by
+Substituting in the displacement operator of Equation (3), we find that $\mathbf{d}(\mathbf{q})$ is given by
 \begin{equation}
@@
 \end{equation}
-This can be expressed in terms of more standard outputs of ab-initio software that calculates phonon properties.
+This can be expressed in terms of more standard outputs of \emph{ab initio} materials physics methods that calculate phonon properties.
 In particular, we can take $\hat{e}_{\mathbf{q}\sigma}$, the eigenvectors of the dynamical matrix, and define eigendisplacements via $\boldsymbol{\eta}_{\mathbf{q}\sigma\kappa}=\frac{\hat{e}_{\mathbf{q}\sigma}}{\sqrt{M_{\kappa}}}$.
@@
 
-Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function.
+Before applying this formalism to a real material and discussion of the two-dimensional case, we briefly consider and comment on `nonlocality' in the phonon contribution to the dielectric function.
 Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.
@@
 Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone.
-It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons.
+It would then seem extremely challenging to observe nonlocal behavior of phonon-polaritons.
 However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet.
-Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields.
+Additionally, another recent experiment \cite{benz2016single} has leveraged so-called `picocavities', sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields.
 These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.
@@
 We calculated the phonon modes, their frequencies (shown in Figure 1(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation (12), using experimental values for the dissipation rates $\Gamma$.
-We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained ab-initio by calculating lifetimes associated with electron-phonon scattering and three-phonon decay processes.
+We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering and three-phonon decay processes.
 
@@
 In materials like hBN where excitonic effects play a significant role, the optical properties of the system are not determined solely by the energy associated with band to band recombination, and instead must also account for the exciton binding energy\cite{robert2016excitonic}.
-Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{levinshtein2001properties} and E$_{b}$ = 0.67 eV from recent ab-initio calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV.
+Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{levinshtein2001properties} and E$_{b}$ = 0.67 eV from recent \emph{ab initio} calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV.
 As the bandgap we compute from DFT is 4.48 eV, we did not require a correction to the gap in hBN.
@@
 Unlike the previous section, we will not demonstrate the results of the calculation on a particular material.
-Here, we are only interested in finding an expression for the optical response of the surface in terms of ab-initio-derivable parameters, as well as the general form of the dispersion relation of phonon-polaritons in 2D.
+Here, we are only interested in finding an expression for the optical response of the surface in terms of \emph{ab initio}-derivable parameters, as well as the general form of the dispersion relation of phonon-polaritons in 2D.
 Because of the change of symmetry in the system, we will not be able to perform the analysis by Fourier transforming Maxwell's equations in three dimensions.
@@
 
-Before concluding, we note that in general, when considering bulk and 2D versions of a material, the effective parameters $Q^*$ and $M$ will change as a result of changes to the phonon band structure.
+ We note that, in general, when considering bulk and 2D versions of a material, the effective parameters $Q^*$ and $M$ will change as a result of changes to the phonon band structure.
 The losses may also change drastically in the reduction of dimensions.
-For example, it was found in several experimental works that the damping rate of the TO phonon in hexagonal boron nitride nearly doubles upon going from bulk to two-dimensions \cite{gorbachev2011hunting,tran2016quantum}, which may strongly suppress the ability to image the polariton with near-field probes.
-This is because due to the small thickness, the dispersion of the phonon polariton is quite flat, and very high wavevector modes, corresponding to the maximum wavevector detectable by a near-field tip, occur at frequencies very close to the TO phonon frequency.
+For example, it was found in several experiments that the damping rate of the TO phonon in hexagonal boron nitride nearly doubles upon going from bulk to 2D\cite{gorbachev2011hunting,tran2016quantum}, which may strongly suppress the ability to image the polariton with near-field probes.
+This is due to the small thickness where the dispersion of the phonon polariton is quite flat, and very high wavevector modes, corresponding to the maximum wavevector detectable by a near-field tip, occur at frequencies very close to the TO phonon frequency.
 Near the TO phonon-frequency however, the ratio $\frac{\text{Im }\Pi}{\text{Re }\Pi}$ is the highest, corresponding to very low quality-factor modes.
@@
 And furthermore that one can also obtain the same equation by considering the ultra-thin limit of a 3D plasmonic material \cite{jablan2009plasmonics,jablan2013plasmons}.
-
-%\textcolor{blue}{Still looking for more to say about this.
-This probably requires some brainstorming discussion between us.
-For now, what I'll do is prepare a plot of the Fourier transform of the green's function of the Maxwell equations for the 2D sheet (p-polarized reflectivity, as was done in Argentene).}
 
@@
 
-%The dipole moment associated with each unit cell is proportional to this total field, and then we may solve a self-consistent equation based on the polarization to find the
-
-% Anything else we need?
-
-\section{Outlook}
+\section{Summary and Outlook}
 
@@
 Perhaps it is possible that there are some materials in which the 2D optical phonons experience lower losses due to a reduced scattering phase space.
-The formalism we provide here may also be extended to understand phonon-polaritons in other more atypical reduced-dimensional settings, such as zero-dimensional settings in single emitters, i.e., "molecular phonon polaritons", in analogy to recent work on "molecular plasmons" \cite{manjavacas2013tunable,lauchner2015molecular}.
+The formalism we provide here may also be extended to understand phonon-polaritons in other more atypical reduced-dimensional settings, such as zero-dimensional settings in single emitters, i.e., `molecular phonon polaritons', in analogy to recent work on `molecular plasmons' \cite{manjavacas2013tunable,lauchner2015molecular}.
 Notably, we go beyond oversimplifications of the Lorentz oscillator model in which the Born charges are treated as a single scalar quantity and can treat the influence and interplay of many phonon modes that contribute to the dielectric function.
@@
 
-In the long run, besides furnishing an ab-initio description of nonlocality and reduced-dimension phonon polaritons, there are a number of interesting questions that can be addressed by the framework provided here.
+In future work, besides furnishing an \emph{ab initio} description of nonlocality and reduced-dimension phonon polaritons, there are a number of interesting questions that can be addressed by the framework provided here.
 For example, we may consider situations such as the phonon polaritons associated with optical phonons at the interface of two materials.
-Another situation of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other.
-In that case, it would be of interest to see how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.
+Another system of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other.
+In that case, it would be relevant to see how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.
 

### 2018-08-05 paper.tex
More edits.

@@
 
-The way to proceed on all of these questions is an \emph{ab initio} framework to predict the phonon contribution to the dielectric tensor in arbitrary settings.
-Frameworks exist for calculating the local permittivity coming from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and implemented in \emph{ab initio} materials physics software packages such as \cite{abinit1,abinit2,abinit3}.
+To address these questions an \emph{ab initio} framework to predict the phonon contribution to the dielectric tensor in arbitrary settings is essential.
+Methodologies exist to calculate the local permittivity contribution from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and are implemented in \emph{ab initio} materials physics software packages\cite{abinit1,abinit2,abinit3}.
 Here, we develop a unified
-theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects coming from arbitrary dimensionality.
+theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects that emerge from arbitrary dimensionality.
 Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in polar materials of arbitrary dimensionality.
 We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles.
-From this theoretical framework, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.
+Importantly, from this theoretical framework, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.
 We apply this formalism to calculate the dielectric function of silicon carbide and hexagonal boron nitride from first principles, and show that the results agree well with measurements.

### 2018-08-05 paper.tex
I found some repeated sections, commented. Made some other edits. I noticed that equations and figures have explicit callouts rather than using ~\ref and ref labels. Could either Nick or Jenny fix this?

@@
 
-Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on "nonlocality" in the phonon contribution to the dielectric function.
+%section below appears twice - commenting it out.
+
+%Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on `nonlocality' in the phonon contribution to the dielectric function.
 Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.
@@
 Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone.
-It would then seem almost hopeless to see nonlocal behavior for the phonon-polaritons.
+It would then appear  to see nonlocal behavior for the phonon-polaritons.
 However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet.
@@
 
-\section{Application to bulk materials}
+\section{Application to bulk semiconductor systems}
 % calculation basics
@@
 This also leads to a modification of the phonon spectra.
-% For a material in which the band gap is not as well known as in SiC, a more computationally intensive GW calculation could be used for an entirely \it{ab initio} implementation of our theory.
-If excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter prediction might be used to further improve predictive accuracy.
+Note that for a materials where the optical band gap is not as well known as in SiC, a more computationally intensive GW calculation could be used while retaining the \emph{ab initio} implementation of our theory.
+Further, if excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter Equation prediction might be used to further improve predictive accuracy within the same framework presented here.
 The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figure \ref{fig:epsilon}.
@@
 Another system of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other.
-In that case, it would be relevant to see how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.
+In that case, it would be relevant to evaluate how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.
 

### 2018-08-05 paper.tex
added an acknowledgements sectiokn

@@
 
+\section{Acknowledgements}
+This research used resources of the National Energy Research Scientific Computing Center, a DOE Office of Science User Facility supported by the Office of Science
+of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231.
+NR and JC recognize the support of the DOE Computational Science Graduate Fellowship (CSGF).
+PN acknowledges start-up funding from the Harvard John A.
+Paulson School of Engineering and Applied Sciences.
+%The authors thank XXXX for helpful discussions.
+
 \bibliographystyle{apsrev4-1}

### 2018-08-06 paper.tex
eq and fig refs and addedCSGF fellow number (jenny i think yours might be different)

@@
 
-\usepackage[usenames]{color}
-\newcommand{\edited}[1]{{\color{red} #1}}
+%\usepackage[usenames]{color}
+%\newcommand{\edited}[1]{{\color{red} #1}}
 
@@
 The interaction Hamiltonian of the lattice with the electromagnetic field is taken as:
-\begin{equation}
-H_{int} = -\int d^3x ~\mathbf{E}(\mathbf{x},t)\cdot\mathbf{P}(\mathbf{x},t),
+\begin{equation}\label{eq:interaction}
+H_{\text{int}} = -\int d^3x ~\mathbf{E}(\mathbf{x},t)\cdot\mathbf{P}(\mathbf{x},t),
 \end{equation}
@@
 Namely
-\begin{equation}
+\begin{equation}\label{eq:dipolesum}
 \mathbf{P}(\mathbf{r},t) = \sum\limits_{\text{sites, }\mathbf{R}}\mathbf{d}(\mathbf{R},t)\delta(\mathbf{r}-\mathbf{R}),
@@
 This said, we operate in the harmonic regime.
-The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa} \mathbf{b}_{\kappa} \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$.
+The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa}  \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$.
 Note that the Born effective charges are both tensorial and site-dependent.
 This displacement can be expanded in phonon modes as
-\begin{equation}
+\begin{equation}\label{eq:phononfield}
 \mathbf{u}(\mathbf{R}_{\kappa},t) = \sum\limits_{\mathbf{q}\sigma}\sqrt{\frac{\hbar}{2M_{\kappa}N\omega_{\mathbf{q}\sigma}}}\left( e^{i\mathbf{q}\cdot\mathbf{R}_{\kappa}-i\omega_{\mathbf{q}\sigma}t}\hat{e}_{\mathbf{q}\sigma}(\mathbf{b}_{\kappa})a_{\mathbf{q}\sigma} + \text{h.c.} \right),
@@
 where $M_{\kappa}$ is the mass of atom $\kappa$, $\omega_{qs}$ is the phonon frequency at wavevector $\mathbf{q}$ and branch $s$, and $N$ is the number of unit cells.
-Assuming that the field is weak, the expectation value of the polarization in the lattice will depend linearly on the total field acting on the ions: this field is the sum of the external field with the field resulting from induced electronic and ionic polarization, and it is this field which appears in the interaction Hamiltonian of Equation (1).
+Assuming that the field is weak, the expectation value of the polarization in the lattice will depend linearly on the total field acting on the ions: this field is the sum of the external field with the field resulting from induced electronic and ionic polarization, and it is this field which appears in the interaction Hamiltonian of Equation \ref{eq:interaction}.
 We define the linear-response relation for a bulk crystal in Fourier space as $\langle\mathbf{P}(\mathbf{q},\omega)\rangle = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, where we refer to $\boldsymbol{\Pi}(\mathbf{q},\omega)$ as the polarization-polarization response function.
@@
 We have that
-\begin{equation}
+\begin{equation}\label{eq:maxwellsourced}
 \left(-\mathbf{q}\times\mathbf{q}\times - \epsilon_{\infty}\frac{\omega^2}{c^2} \right)\mathbf{E} = \omega^2\mu_0\mathbf{P}(\mathbf{q},\omega),
@@
 Using the linear response relation $\mathbf{P} = \epsilon_0\boldsymbol{\Pi}\mathbf{E}$, we have simply that
-\begin{equation}
+\begin{equation}\label{eq:maxwellepsilon}
 \left(-\mathbf{q}\times\mathbf{q}\times - \left(\epsilon_{\infty} + \boldsymbol{\Pi} \right)\frac{\omega^2}{c^2}\right)\mathbf{E} = 0,
@@
 Now, we determine the polarization-polarization response function. It is given by the Kubo formula at finite temperature via
-\begin{equation}
+\begin{equation}\label{eq:polarizationkubo}
 \boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 V}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i\frac{\Gamma}{2}}\left(e^{\beta E_m}-e^{\beta E_n} \right),
@@
 The Fourier transformed polarization operators are found to be
-\begin{equation}
+\begin{equation}\label{eq:fourierpolarization}
 \mathbf{P}(\mathbf{q}) = \sum\limits_{\mathbf{R}}\mathbf{d}(\mathbf{R})e^{-i\mathbf{q}\cdot\mathbf{R}} \equiv \mathbf{d}(\mathbf{q}),
@@
 where it is to be understood that $\mathbf{P}(\mathbf{q})$ refers to a continuous Fourier transform while $\mathbf{d}(\mathbf{q})$ refers to a discrete Fourier transform.
-Substituting in the displacement operator of Equation (3), we find that $\mathbf{d}(\mathbf{q})$ is given by
-\begin{equation}
+Substituting in the displacement operator of Equation \ref{eq:phononfield}, we find that $\mathbf{d}(\mathbf{q})$ is given by
+\begin{equation}\label{eq:phononstodipole}
 \mathbf{d}(\mathbf{q}) = \sum\limits_{\sigma}\sqrt{\frac{\hbar N}{2\omega_{\mathbf{q}\sigma}}} \left(\mathbf{S}_{\mathbf{q}\sigma}a_{\mathbf{q}\sigma} + \text{h.c.}\right),
@@
 where 
-\begin{equation}
+\begin{equation}\label{eq:oscillatordefinition}
 \mathbf{S}_{\mathbf{q}\sigma} = \sum\limits_{\kappa} \frac{1}{\sqrt{M_{\kappa}}}\mathbf{Z}_{\kappa}\hat{e}_{\mathbf{q}\sigma}(\mathbf{R}_{\kappa}).
@@
 In particular, we can take $\hat{e}_{\mathbf{q}\sigma}$, the eigenvectors of the dynamical matrix, and define eigendisplacements via $\boldsymbol{\eta}_{\mathbf{q}\sigma\kappa}=\frac{\hat{e}_{\mathbf{q}\sigma}}{\sqrt{M_{\kappa}}}$.
-In which case, we may cast Equation (9) into the form:
-\begin{equation}
+In which case, we may cast Equation \ref{eq:oscillatordefinition} into the form:
+\begin{equation}\label{eq:oscillatorstrength}
 \mathbf{S}_{\mathbf{q}\sigma} = \sum\limits_{\kappa} \mathbf{Z}_{\kappa}\boldsymbol{\eta}_{\mathbf{q}\sigma\kappa}.
 \end{equation}
-We may now use this in concurrence with Equation (6) to evaluate the dielectric function.
+We may now use this in concurrence with Equation \cite{eq:polarizationkubo} to evaluate the dielectric function.
 Consider $\boldsymbol{\Pi}(\mathbf{q},\omega)$.
-From the form of Equation (8), we see that at zero temperature, the first term only sees contributions from $m=|0_{\mathbf{q}\sigma}\rangle$ and $n=|1_{\mathbf{q}\sigma}\rangle$, where $0_{\mathbf{q}\sigma}$ and $1_{\mathbf{q}\sigma}$ refer to the number of phonons in mode $\mathbf{q}\sigma$.
+From the form of Equation \ref{eq:phononstodipole}, we see that at zero temperature, the first term only sees contributions from $m=|0_{\mathbf{q}\sigma}\rangle$ and $n=|1_{\mathbf{q}\sigma}\rangle$, where $0_{\mathbf{q}\sigma}$ and $1_{\mathbf{q}\sigma}$ refer to the number of phonons in mode $\mathbf{q}\sigma$.
 Here $\sigma$ can be any branch of the phonon band structure.
@@
 this leads to:
-\begin{equation}
+\begin{equation}\label{eq:fullsusceptibility}
 \boldsymbol{\Pi}(\mathbf{q},\omega) = \frac{1}{\Omega}\sum\limits_{\sigma}\frac{\mathbf{S}_{\mathbf{q}\sigma}\otimes \mathbf{S}_{\mathbf{q}\sigma}^*}{\omega^2_{\mathbf{q}\sigma}-\omega^2-i\Gamma_{\mathbf{q}\sigma}\omega},
@@
 Finally, the dielectric function takes the form
-\begin{equation}
+\begin{equation}\label{eq:pitoepsilon}
 \epsilon(\mathbf{q},\omega) = \epsilon_{\infty} + \frac{1}{\Omega}\sum\limits_{\sigma}\frac{\mathbf{S}_{\mathbf{q}\sigma}\otimes \mathbf{S}_{\mathbf{q}\sigma}^*}{\omega^2_{\mathbf{q}\sigma}-\omega^2-i\Gamma_{\mathbf{q}\sigma}\omega}.
@@
 In this case, one arises at a dielectric function
-\begin{equation}
+\begin{equation}\label{eq:lorentzoscillator}
 \epsilon(\omega) = \epsilon_{\infty} + \frac{NQ^{*2}}{\epsilon_0 M^*V}\frac{1}{\omega^2_{TO}-\omega^2-i\Gamma_{TO}\omega},
@@
 Before applying this formalism to a real material and discussion of the two-dimensional case, we briefly consider and comment on `nonlocality' in the phonon contribution to the dielectric function.
-Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.
+Note that Equation \ref{eq:pitoepsilon} has a wavevector dependence which is implicit through the phonon dispersion.
 Therefore, it is possible that the phonon polariton resonances red-shift or blue-shift, depending on how the relative position of the phonons at finite-wavevector versus zero wavevector.
@@
 
-%This is exemplified in Figure 3, where we show the ionic contribution to the permittivity (i.e.
-we do not consider the electronic part).
-We consider wavevector directions along $\Gamma-L$ and consider wavevectors that are $1/10$ and $1/4$ of the way along the Brillouin zone.
-We see that the major effect is to red-shift the phonon resonance by a few inverse centimeters.
-However, the effect "downstream", in the region where phonon polaritons are, is much smaller, with the permittivity being almost the same as in the local case.
-In making this plot, we make the assumption that the variation of the Born effective charges is negligible upon moving to finite-wavevector optical phonons.
-
-%However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.
-
-%section below appears twice - commenting it out.
-
-%Before applying this formalism to a real material and also before discussion of the two-dimensional case, we briefly comment on `nonlocality' in the phonon contribution to the dielectric function.
-Note that Equation (12) has a wavevector dependence which is implicit through the phonon dispersion.
-As a result, the phonon contribution to the dielectric function sees contributions from the entire phonon band structure, including the large wavevector components.
-However, $\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.
-In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.
-Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone.
-It would then appear  to see nonlocal behavior for the phonon-polaritons.
-However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet.
-Additionally, another recent experiment \cite{benz2016single} has leveraged so-called "picocavities", sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields.
-These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.
-
 \section{Application to bulk semiconductor systems}
 % calculation basics
-To demonstrate the practical use of this theoretical framework, we present an example calculation of the dielectric function for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in Figure 1(a).
+To demonstrate the practical use of this theoretical framework, we present an example calculation of the dielectric function for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in \ref{fig:phonons}(a).
 Phonon bandstructures for SiC and hBN were calculated within density functional perturbation
@@
 Initial ground state density functional theory (DFT) properties were predicted using norm-conserving Vanderbilt pseudopotentials of the PBEsol parameterization for SiC, and PBE with a D2 (Grimme) van der Waals correction for hBN \cite{PBEsol,PBE,ONCV_PPs,pseudodojo,grimme}.
-We calculated the phonon modes, their frequencies (shown in Figure 1(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation (12), using experimental values for the dissipation rates $\Gamma$.
+We calculated the phonon modes, their frequencies (shown in Figure \ref{fig:phonons}(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation \ref{eq:pitoepsilon}, using experimental values for the dissipation rates $\Gamma$.
 We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering and three-phonon decay processes.
@@
 Further, if excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter Equation prediction might be used to further improve predictive accuracy within the same framework presented here.
-The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figure \ref{fig:epsilon}.
+The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figures \ref{fig:sicepsilon} and \ref{fig:hbnepsilon}.
 As can be seen, the agreement between the theoretical calculation and the experimental results is quite good for silicon carbide, and good for the upper Reststrahlen band of hBN.
@@
 \includegraphics[width=8.5cm]{SiC_comparison_thy_vs_expt.pdf}
-\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk 4H silicon carbide.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk 4H silicon carbide from using Equation (12).
+\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk 4H silicon carbide.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk 4H silicon carbide from using Equation \ref{eq:pitoepsilon}.
 The phonon properties and $\epsilon_{\infty}$ calculated from the ABINIT package.
@@
 }
-\label{fig:epsilon}
+\label{fig:sicepsilon}
 \end{figure}
@@
 \includegraphics[width=8.5cm]{hBN_comparison_thy_vs_expt.pdf}
-\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk hexagonal boron nitride.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk hexagonal boron nitride from using Equation (12).
+\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk hexagonal boron nitride.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk hexagonal boron nitride from using Equation \ref{eq:pitoepsilon}.
 The phonon properties and $\epsilon_{\infty}$ calculated from the ABINIT package.
@@
 The component of the real part of the permittivity parallel to the c-axis is shown in (a) component parallel is shown in (b).}
-\label{fig:phonons}
+\label{fig:hbnepsilon}
 \end{figure}
@@
 For $z \neq 0$, the Maxwell equation for the magnetic field is $(\nabla\times\nabla\times - \epsilon\frac{\omega^2}{c^2})\mathbf{H} = 0$. This equation, for the TM mode described above, becomes 
-\begin{equation}
+\begin{equation}\label{eq:2dmaxwell}
 \left(-\frac{d^2}{dz^2}+\mathbf{q}^2-\epsilon\frac{\omega^2}{c^2} \right)H = 0,
@@
 Taking the case where $x$ coincides with a principal axis of the system, we have that
-\begin{equation}
+\begin{equation}\label{eq:2ddispersion}
 q\Pi_{xx}(q,\omega)  = -2\epsilon.
@@
 To connect this relation to the discussion of the three-dimensional case, we write a microscopic form for the polarization response function of the ionic lattice. It is simply
-\begin{equation}
+\begin{equation}\label{eq:2dsusceptibility}
 \boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 A}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i0^+}\left(e^{\beta E_m}-e^{\beta E_n} \right),
@@
 Note that although the lattice is two dimensional, in general, the ions may be displaced in any three directions and thus $\boldsymbol{\Pi}$ is still a $3\times3$ matrix.
-Following the same kind of reasoning that lead to Equation (13), if we parameterize the effect of the different masses and Born charges in the unit cell by $M^*$ and $Q^*$, we will find a form for the polarization response function which is given by the second term of Equation (13) but with $V$ replaced by $A$.
-Defining $N/A \equiv n_s$, Equation (14) for the dispersion relation of 2D phonon polaritons tells us that
-\begin{equation}
+Following the same kind of reasoning that lead to Equation \ref{eq:lorentzoscillator}, if we parameterize the effect of the different masses and Born charges in the unit cell by $M^*$ and $Q^*$, we will find a form for the polarization response function which is given by the second term of Equation \ref{eq:lorentzoscillator} but with $V$ replaced by $A$.
+Defining $N/A \equiv n_s$, Equation \ref{eq:2dmaxwell} for the dispersion relation of 2D phonon polaritons tells us that
+\begin{equation}\label{eq:2dphononpolaritondispersion}
 \omega = \sqrt{\omega_{TO}^2+\frac{n_sQ^{*2}}{2M^*\epsilon_0\epsilon}q}.
@@
 
-We note that the dispersion relation of Equation (8) can in fact be obtained by taking the infinitely thin-limit of a bulk polar dielectric, and finding the dispersion of the resulting phonon polaritons in the quasi-electrostatic limit.
+We note that the dispersion relation of Equation \ref{eq:2dphononpolaritondispersion} can in fact be obtained by taking the infinitely thin-limit of a bulk polar dielectric, and finding the dispersion of the resulting phonon polaritons in the quasi-electrostatic limit.
 In particular, for an isotropic polar dielectric of thickness $2d$ with boundaries at $z=\pm d$, solving Laplace's equations yields as the dispersion of the even parity mode:
-\begin{equation}
+\begin{equation}\label{eq:thinfilmphononpolaritondispersion}
 \tanh(qd) = -\frac{\epsilon}{\epsilon(\omega)} \rightarrow qd\epsilon(\omega) = -\epsilon,
@@
 The right-hand side is obtained in the limit of $d\rightarrow 0$.
-Taking Equation (13) for the dielectric function in bulk, we have that (in the absence of losses) $\epsilon + qd\epsilon_{\infty} + q\frac{NQ^{*2}}{2\epsilon_0 M^*A}\frac{1}{\omega_{TO}^2-\omega^2} = 0$, where we have identified $V = 2Ad$.
-In the limit of small $d$, i.e., $qd \ll 1$, we recover Equations (16) and (18) with $\Pi = \frac{NQ^{*2}}{\epsilon_0 MA}\frac{1}{\omega^2_{TO}-\omega^2}$.
+Taking Equation \ref{eq:lorentzoscillator} for the dielectric function in bulk, we have that (in the absence of losses) $\epsilon + qd\epsilon_{\infty} + q\frac{NQ^{*2}}{2\epsilon_0 M^*A}\frac{1}{\omega_{TO}^2-\omega^2} = 0$, where we have identified $V = 2Ad$.
+In the limit of small $d$, i.e., $qd \ll 1$, we recover Equations \ref{eq:2ddispersion} and \ref{eq:2dphononpolaritondispersion} with $\Pi = \frac{NQ^{*2}}{\epsilon_0 MA}\frac{1}{\omega^2_{TO}-\omega^2}$.
 A useful result which follows is that we may also derive another important figure of merit, the $p-$polarized reflectivity of a 2D polar dielectric, which is related to the local density of states of the polar dielectric, which quantifies the strength of coupling to external probes such as a nano-antenna or a quantum emitter.
 By considering the infinitely thin limit of a polar dielectric with dielectric function $\epsilon(\omega)$ and defining $\epsilon(\omega) = \frac{\Pi}{2d}$, we immediately find that the p-polarized reflectivity is given by
-\begin{equation}
+\begin{equation}\label{eq:2dtmreflectivity}
 r_p = \frac{1}{1+\frac{2\epsilon}{q\Pi}}.
@@
 
-We provided this thin-film derivation as a simple way to understand the content of Equation (16).
-In general Equation (17) should be used to calculate $\Pi$ and the corresponding dispersion.
-We note that a similar situation appears in the treatment of 2D plasmons: namely that a relation like (16) appears, except with $\Pi$ the polarization-polarization response of a 2D electron gas.
+We provided this thin-film derivation as a simple way to understand the content of Equations \ref{eq:2ddispersion} and \ref{eq:2dphononpolaritondispersion}.
+In general Equation \ref{eq:2dsusceptibility} should be used to calculate $\Pi$ and the corresponding dispersion.
+We note that a similar situation appears in the treatment of 2D plasmons: namely that a relation like \ref{eq:2ddispersion} appears, except with $\Pi$ the polarization-polarization response of a 2D electron gas.
 And furthermore that one can also obtain the same equation by considering the ultra-thin limit of a 3D plasmonic material \cite{jablan2009plasmonics,jablan2013plasmons}.
-
-% We now move to generalize the derivation of the previous section to describe the optical phonon contribution to Maxwell's equations coming from a two-dimensional material.
-Because of the change of symmetry in the system, we will not be able to perform the analysis by Fourier transforming Maxwell's equations in three dimensions.
-Nevertheless, we may proceed as follows.
-Consider Maxwell's equations with a source density given by
-% \begin{equation}
-% \mathbf{P}(\mathbf{r}) = \mathbf{P}(\mathbf{r}_{||})\delta(z) = \sum\limits_{\mathbf{R}}\mathbf{d}(\mathbf{R})\delta(\mathbf{R}-\mathbf{r})\delta(z).
-% \end{equation}
-% Taking the fourier transform in two dimensions gives:
-% \begin{equation}
-% \mathbf{P}(\mathbf{q},\omega) = \delta(z)\sum\limits_{\mathbf{R}}\mathbf{d}(\mathbf{R})e^{-i\mathbf{q}\cdot\mathbf{R}}.
-% \end{equation}
-% Fourier transforming the Maxwell equations in two dimensions, leaving the third unchanged connects the fourier transformed polarization density to the  
-% \begin{equation}
-% \left(-\frac{d^2}{dz^2} + \mathbf{q}^2 - \frac{\omega^2}{c^2} \right)\mathbf{E}(\mathbf{q},\omega) = \mu_0\omega^2\mathbf{P}(\mathbf{q},\omega)\delta(z)
-% \end{equation}
-% To connect to response functions, we note that: the Hamiltonian is given by Equation (1), integrated over two dimensions, the linear response relation is $\langle \mathbf{P}(\mathbf{q},\omega) = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, and that the susceptibility is still defined by Equation (6), except that the units of the polarization density are different, and the volume factor becomes an area factor $A$.
-Given this 2D polarizability of the ionic lattice, we may solve Maxwell's equations for the resultant fields, as well as the dispersion relation of the phonon-polariton modes.
-We consider for simplicity an isotropic case
-
-%While we could work in Fourier space in two dimensions and real-space in the third, we employ a simpler argument for the phonon-polariton modes of a 2D structure and the response function that captures such phonon-polariton response.
-
-%In particular, what we will do is, in the spirit of the self-consistent mean-field approximation, find an expression for the expectation value of the dipole moment at each site in the two-dimensional lattice.
-Then, using Maxwell's equations, write the electromagnetic field as a superposition of fields from the various dipoles.
-In particular, suppose we have a two-dimensional lattice with sites $\mathbf{R}$ in the xy-plane (i.e., the two-dimensional layer is located at $z=0$).
-The polarization density can then be expressed as
-%\begin{equation}
-%\mathbf{P}(\mathbf{r}) = \sum\limits_{\mathbf{R}}\mathbf{d}(\mathbf{R})\delta(\mathbf{r}-\mathbf{R})\delta(z)
-%\end{equation}
 
@@
 \section{Acknowledgements}
-This research used resources of the National Energy Research Scientific Computing Center, a DOE Office of Science User Facility supported by the Office of Science
-of the U.S. Department of Energy under Contract No. DE-AC02-05CH11231.
-NR and JC recognize the support of the DOE Computational Science Graduate Fellowship (CSGF).
-PN acknowledges start-up funding from the Harvard John A.
+This research used resources of the National Energy Research Scientific Computing Center, a DOE Office of Science User Facility supported by the Office of Science of the U.S.
+Department of Energy under Contract No.
+DE-AC02-05CH11231.
+N.
+R.
+and J.
+C.
+recognize the support of the DOE Computational Science Graduate Fellowship (CSGF) fellowship no.
+DE-FG02-97ER25308.
+P.
+N.
+acknowledges start-up funding from the Harvard John A.
 Paulson School of Engineering and Applied Sciences.
 %The authors thank XXXX for helpful discussions.
+Tentatively Josh Caldwell, Jeremy Baumberg, Dirk England, etc.
 

### 2018-08-06 paper.tex
added some references

@@
 In recent years, it has been shown that phonon polaritons offer the ability to confine light to volumes over $10^6$ times smaller than that of a photon in free-space at the diffraction limit \cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra}.
-Due to this remarkable confinement and their relatively high lifetimes of around picoseconds, phonon polaritons open new opportunities for vibrational spectroscopy, for radiative heat transfer \cite{hillenbrand2002phonon}, and for control of dynamics in quantum emitters \cite{kumar2015tunable,rivera2017making}.
+Due to this remarkable confinement and their relatively high lifetimes of around picoseconds, phonon polaritons open new opportunities for vibrational spectroscopy, for radiative heat transfer \cite{hillenbrand2002phonon}, and for control of dynamics in quantum emitters \cite{kumar2015tunable,rivera2017making,kurman2018controlling}.
 Typically, the core features of phonon polaritons, such as mode shape, confinement, and propagation characteristics, can be understood through the properties of the Lorentz oscillator dielectric function, which has been successfully applied to model experiments involving the launching and propagation of these polaritons.
@@
 Here, we develop a unified
-theoretical framework to analyze phonon-polaritons in materials, which can address effects of spatial dispersion as well as effects that emerge from arbitrary dimensionality.
+theoretical framework to analyze phonon-polaritons in materials which can address effects of spatial dispersion as well as effects that emerge from arbitrary dimensionality.
 Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in polar materials of arbitrary dimensionality.
@@
 \end{equation}
-%where $\Gamma$ is a positive infinitesimal, and it is to be understood that rigorously, the limit $\Gamma \rightarrow 0$ is to be taken.
-In a situation where there is phonon dissipation, typically due to electron-phonon and phonon-phonon scattering, we make take a relaxation-time approximation and interpret $\Gamma$ as the finite relaxation time.
 where $m$ and $n$ are taken to be eigenstates of the lattice, $V$ is a normalization volume, and we have written the polarization-polarization susceptibility in Fourier space.
@@
 \end{equation}
-From this equation, let us consider an extremely simplified limit.
+From this equation, let us consider an highly simplified limiting case.
 Assume that the permittivity is isotropic.
@@
 We calculated the phonon modes, their frequencies (shown in Figure \ref{fig:phonons}(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation \ref{eq:pitoepsilon}, using experimental values for the dissipation rates $\Gamma$.
-We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering and three-phonon decay processes.
+We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering \cite{sundararaman2014theoretical,brown2015nonradiative,brown2016ab,narang2017effects} and three-phonon decay processes \cite{srivastava1990physics}.
 

### 2018-08-06 paper.tex
fixed ref compile issue

@@
 In recent years, it has been shown that phonon polaritons offer the ability to confine light to volumes over $10^6$ times smaller than that of a photon in free-space at the diffraction limit \cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra}.
-Due to this remarkable confinement and their relatively high lifetimes of around picoseconds, phonon polaritons open new opportunities for vibrational spectroscopy, for radiative heat transfer \cite{hillenbrand2002phonon}, and for control of dynamics in quantum emitters \cite{kumar2015tunable,rivera2017making,kurman2018controlling}.
+Due to this remarkable confinement and their relatively high lifetimes of around picoseconds, phonon polaritons open new opportunities for vibrational spectroscopy, for radiative heat transfer \cite{hillenbrand2002phonon}, and for control of dynamics in quantum emitters \cite{kumar2015tunable,rivera2017making,kurman2018control}.
 Typically, the core features of phonon polaritons, such as mode shape, confinement, and propagation characteristics, can be understood through the properties of the Lorentz oscillator dielectric function, which has been successfully applied to model experiments involving the launching and propagation of these polaritons.

### 2018-08-06 paper.tex
update
filling in the manuscript

@@
 
-\title{Theory of Phonon Polaritons in Polar Dielectrics}
+\title{Variational theory of the ground state of cavity quantum electrodynamics}
 
 \author{Nicholas Rivera$^\perp$}\email{nrivera@seas.harvard.edu}\affiliation{\HarvardSEAS}\affiliation{\MITPhy}
-\author{Jennifer Coulter$^\perp$}\email{jcoulter@g.harvard.edu}\affiliation{\HarvardSEAS}
+\author{Johannes Flick}\email{flick@g.harvard.edu}\affiliation{\HarvardSEAS}
 \author{Prineha Narang}\email{prineha@seas.harvard.edu}\affiliation{\HarvardSEAS}
@@
 \begin{abstract}
-The ability to use photonic quasiparticles to control electromagnetic energy far below the diffraction limit is a defining paradigm in nanophotonics.
-An important recent development in this field is the measurement and manipulation of extremely confined phonon-polariton modes in polar dielectrics such as silicon carbide and hexagonal boron nitride, which pave the way for nanophotonics and extreme light-matter interactions in the mid-IR to THz frequency band.
-To further advance this promising field, it is of great interest to predict the optical response of polar dielectrics from first principles.
-Here we develop a linear response theory for an ionic lattice coupled to the electromagnetic field in arbitrary dimensions and derive expressions for the dielectric function.
-For a 3D bulk in the spatially local regime, our results qualitatively match the famous Lorentz oscillator model.
-However, unlike previous empirical descriptions, we are able to specify Lorentz oscillator parameters in terms of \emph{ab initio} material properties namely the masses of the atoms in the unit cell, the tensor of Born effective charges of each atom in the unit cell, and the phonon-dispersion (bandstructure).
-Moreover, our results allow us to identify a regime where quantum nonlocal effects may correct the phonon polariton dispersion, extremely relevant in atomic-scale experiments which confine electromagnetic fields to the scale of 1 nm.
-Our results also allow us, for the first time, to derive the dielectric function of phonon polaritons in a monolayer, which may also be specified in terms of the same \emph{ab initio}-derived parameters of the underlying lattice.
-To demonstrate the framework derived, we provide \emph{ab initio} calculations for bulk silicon carbide and bulk hexagonal boron nitride, and show that they agree well with the measured values of their dielectric functions.
-Finally, to show the generality of the framework developed here, we derive the electrodynamics of 2D phonon polaritons, finding a universal form for their electromagnetic modes and dispersion relation, which can be specified in terms of \emph{ab initio} parameters.
-Our results point the way towards first principles descriptions of the effect of interface phonons, phonon strong coupling, and chiral phonons on the properties of phonon polaritons.
+
 \end{abstract}
@@
 
-Phonon polaritons, hybrid quasiparticles of photons and optical phonons, offer great promise for deeply sub-diffractional control of electromagnetic fields at mid-IR and THz frequencies.
-These polaritons share many features in common with plasmon-polaritons in conductors, a heavily well-studied topic.
-In recent years, it has been shown that phonon polaritons offer the ability to confine light to volumes over $10^6$ times smaller than that of a photon in free-space at the diffraction limit \cite{caldwell2013low,xu2014mid,caldwell2014sub,dai2014tunable,tomadin2015accessing,yoxall2015direct,li2015hyperbolic,dai2015subdiffractional,dai2015graphene,caldwell2015low,li2016reversible,Basov:2016,basov2017towards,low2017polaritons,giles2017ultra}.
-Due to this remarkable confinement and their relatively high lifetimes of around picoseconds, phonon polaritons open new opportunities for vibrational spectroscopy, for radiative heat transfer \cite{hillenbrand2002phonon}, and for control of dynamics in quantum emitters \cite{kumar2015tunable,rivera2017making,kurman2018control}.
-Typically, the core features of phonon polaritons, such as mode shape, confinement, and propagation characteristics, can be understood through the properties of the Lorentz oscillator dielectric function, which has been successfully applied to model experiments involving the launching and propagation of these polaritons.
+\section{Variational ansatz}
 
-There are nevertheless some critical questions regarding the behavior of phonon polaritons.
-Here, we present a few.
-First, would the very successful Lorentz oscillator model ever break down and how?
-One particular mechanism of breakdown, which could be present in the case of extremely confined polaritons, concerns effects coming from spatial dispersion.
-Such short phonon polariton wavelengths are indeed predicted by classical electrodynamics of a polar dielectric, especially in the case of a very thin film (nm-thick).
-Second, what are the properties of phonon polaritons in reduced dimensions, such as 2D polar insulators and can they support extremely confined polaritonic modes which allow for extreme light-matter interactions, in analogy to extremely confined plasmonic modes in 2D plasmonic materials such as graphene \cite{jablan2009plasmonics, koppens2011graphene, fei2012gate, chen2012optical,rivera2016shrinking,lundeberg2017tuning}?
-In general, how would the properties of optical phonons inherited by the phonon polariton?
-Finally, what is the influence of non-trivial spatial shaping of phonons on the resulting polariton?
-Examples of non-trivial spatial shaping include but are not limited to chiral phonons \cite{zhu2018observation} and mechanical heterostructures \cite{ratchford2018controlling}.
-The ability to design new phonon polariton materials by designing the optical phonon band structure allows a predictive way to generate polaritonic nano-materials.
-
-To address these questions an \emph{ab initio} framework to predict the phonon contribution to the dielectric tensor in arbitrary settings is essential.
-Methodologies exist to calculate the local permittivity contribution from phonons in a 3D bulk, as in \cite{Gonze1997dynamical}, and are implemented in \emph{ab initio} materials physics software packages\cite{abinit1,abinit2,abinit3}.
-Here, we develop a unified
-theoretical framework to analyze phonon-polaritons in materials which can address effects of spatial dispersion as well as effects that emerge from arbitrary dimensionality.
-Specifically, we develop a means to arrive at the dielectric function associated with optical phonons in polar materials of arbitrary dimensionality.
-We pay close attention to develop a description of the optical response that shows clearly the parallels to the description of plasmonic response from first principles.
-Importantly, from this theoretical framework, we not only accurately predict surface phonon polaritons in thick materials where a bulk permittivity is valid but also provide tools to predict phonon polariton dispersion in monolayers and anticipate when spatially nonlocal corrections to the phonon contribution to the dielectric function become relevant.
-We apply this formalism to calculate the dielectric function of silicon carbide and hexagonal boron nitride from first principles, and show that the results agree well with measurements.
-Finally, we develop the theory of phonon polaritons in a 2D polar material, specifying the dielectric function in terms of \emph{ab initio} parameters, and then finding a universal form for the dispersion relation of 2D phonon polaritons.
-
-\section{Linear response of phonon polaritons in three dimensions}
-
-We start by developing the theory of phonon polaritons in a 3D bulk in a way that is parallel to the linear response theory of electronic contributions to the permittivity and plasmonic properties.
-Here, the matter subsystem which couples to light is the ionic lattice of the crystal.
-Thus, we develop the linear response theory of an ionic lattice coupled to an electromagnetic field.
-The interaction Hamiltonian of the lattice with the electromagnetic field is taken as:
-\begin{equation}\label{eq:interaction}
-H_{\text{int}} = -\int d^3x ~\mathbf{E}(\mathbf{x},t)\cdot\mathbf{P}(\mathbf{x},t),
-\end{equation}
-where $\mathbf{E}$ is a driving electric field, and $\mathbf{P}$ is the polarization associated with the lattice.
-We write the polarization as a sum of contributions from all unit cells in the system.
-Namely
-\begin{equation}\label{eq:dipolesum}
-\mathbf{P}(\mathbf{r},t) = \sum\limits_{\text{sites, }\mathbf{R}}\mathbf{d}(\mathbf{R},t)\delta(\mathbf{r}-\mathbf{R}),
-\end{equation}
-where $\mathbf{d}(\mathbf{R},t)$ is the time-dependent dipole moment associated with lattice site $\mathbf{R}$.
-In expressing the polarization in this fashion, we are assuming that the displacement of the ions is far smaller than the size of the unit cell, which is, even in a deeply anharmonic regime, a good approximation.
-This said, we operate in the harmonic regime.
-The dipole moment is given by $\mathbf{d}(\mathbf{R},t) = \sum\limits_{\text{basis sites,}\kappa}  \mathbf{Z}_{\kappa}\mathbf{u}(\mathbf{R}_{\kappa} \equiv \mathbf{R}+\mathbf{b}_{\kappa},t)$, where $\mathbf{Z}_{\kappa}$ is the tensor of Born effective charges for atom $\kappa$ in the unit cell at basis vector $\mathbf{b}_{\kappa}$ and $\mathbf{u}(\mathbf{R}_{\kappa},t)$ is the displacement of atom $\kappa$ at basis site $\mathbf{R}_{\kappa}$.
-Note that the Born effective charges are both tensorial and site-dependent.
-This displacement can be expanded in phonon modes as
-\begin{equation}\label{eq:phononfield}
-\mathbf{u}(\mathbf{R}_{\kappa},t) = \sum\limits_{\mathbf{q}\sigma}\sqrt{\frac{\hbar}{2M_{\kappa}N\omega_{\mathbf{q}\sigma}}}\left( e^{i\mathbf{q}\cdot\mathbf{R}_{\kappa}-i\omega_{\mathbf{q}\sigma}t}\hat{e}_{\mathbf{q}\sigma}(\mathbf{b}_{\kappa})a_{\mathbf{q}\sigma} + \text{h.c.} \right),
-\end{equation}
-where $M_{\kappa}$ is the mass of atom $\kappa$, $\omega_{qs}$ is the phonon frequency at wavevector $\mathbf{q}$ and branch $s$, and $N$ is the number of unit cells.
-Assuming that the field is weak, the expectation value of the polarization in the lattice will depend linearly on the total field acting on the ions: this field is the sum of the external field with the field resulting from induced electronic and ionic polarization, and it is this field which appears in the interaction Hamiltonian of Equation \ref{eq:interaction}.
-We define the linear-response relation for a bulk crystal in Fourier space as $\langle\mathbf{P}(\mathbf{q},\omega)\rangle = \epsilon_0\boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$, where we refer to $\boldsymbol{\Pi}(\mathbf{q},\omega)$ as the polarization-polarization response function.
-
-To connect this polarization-polarization response function to the observed dielectric function, we write down Maxwell's equations for the electric field sourced by a polarization density.
-We have that
-\begin{equation}\label{eq:maxwellsourced}
-\left(-\mathbf{q}\times\mathbf{q}\times - \epsilon_{\infty}\frac{\omega^2}{c^2} \right)\mathbf{E} = \omega^2\mu_0\mathbf{P}(\mathbf{q},\omega),
-\end{equation}
-where $\omega$ is the frequency of light under consideration $\mathbf{P}$ is the polarization at that frequency, and $\epsilon_{\infty}$ is the high-frequency dielectric function associated with the electrons.
-Now, we identify the average polarization $\langle\mathbf{P}\rangle$ with the classical polarization $\mathbf{P}$ in Maxwell's equations.
-This is essentially the random-phase approximation (RPA).
-We note that an analogous procedure is also employed in the linear-response theory of the longitudinal response of electrons, where instead of the polarization density being averaged, the charge density is averaged.
-And instead of the solving Maxwell's equations for the field, we solve Laplace's equation for the potential.
-Using the linear response relation $\mathbf{P} = \epsilon_0\boldsymbol{\Pi}\mathbf{E}$, we have simply that
-\begin{equation}\label{eq:maxwellepsilon}
-\left(-\mathbf{q}\times\mathbf{q}\times - \left(\epsilon_{\infty} + \boldsymbol{\Pi} \right)\frac{\omega^2}{c^2}\right)\mathbf{E} = 0,
-\end{equation}
-leading to the identification of of the dielectric tensor $\boldsymbol{\epsilon} = \epsilon_{\infty} + \boldsymbol{\Pi}$.
-
-Now, we determine the polarization-polarization response function. It is given by the Kubo formula at finite temperature via
-\begin{equation}\label{eq:polarizationkubo}
-\boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 V}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i\frac{\Gamma}{2}}\left(e^{\beta E_m}-e^{\beta E_n} \right),
-\end{equation}
-where $m$ and $n$ are taken to be eigenstates of the lattice, $V$ is a normalization volume, and we have written the polarization-polarization susceptibility in Fourier space.
-It is understood that the polarization operator is now the Schrodinger operator.
-$\Gamma$ is a positive infinitesimal, and it is to be understood that rigorously, the limit $\Gamma \rightarrow 0$ is to be taken.
-In cases where there is phonon dissipation, typically due to electron-phonon and phonon-phonon scattering, we make take a relaxation-time approximation and interpret $\Gamma$ as the finite relaxation time.
-The eigenstates span the phononic Fock space.
-For simplicity, we will work at zero temperature from here onwards.
-In this case, in the first term, the states $m$ must be zero-phonon states, while in the second term, the states $n$ must be zero-phonon states.
-Due to the displacement being linear in creation and annihilation operators, the states $n$ and $m$ in the first and second terms (resp.) must be one-phonon states.
-The Fourier transformed polarization operators are found to be
-\begin{equation}\label{eq:fourierpolarization}
-\mathbf{P}(\mathbf{q}) = \sum\limits_{\mathbf{R}}\mathbf{d}(\mathbf{R})e^{-i\mathbf{q}\cdot\mathbf{R}} \equiv \mathbf{d}(\mathbf{q}),
-\end{equation}
-where it is to be understood that $\mathbf{P}(\mathbf{q})$ refers to a continuous Fourier transform while $\mathbf{d}(\mathbf{q})$ refers to a discrete Fourier transform.
-Substituting in the displacement operator of Equation \ref{eq:phononfield}, we find that $\mathbf{d}(\mathbf{q})$ is given by
-\begin{equation}\label{eq:phononstodipole}
-\mathbf{d}(\mathbf{q}) = \sum\limits_{\sigma}\sqrt{\frac{\hbar N}{2\omega_{\mathbf{q}\sigma}}} \left(\mathbf{S}_{\mathbf{q}\sigma}a_{\mathbf{q}\sigma} + \text{h.c.}\right),
-\end{equation}
-where 
-\begin{equation}\label{eq:oscillatordefinition}
-\mathbf{S}_{\mathbf{q}\sigma} = \sum\limits_{\kappa} \frac{1}{\sqrt{M_{\kappa}}}\mathbf{Z}_{\kappa}\hat{e}_{\mathbf{q}\sigma}(\mathbf{R}_{\kappa}).
-\end{equation}
-This can be expressed in terms of more standard outputs of \emph{ab initio} materials physics methods that calculate phonon properties.
-In particular, we can take $\hat{e}_{\mathbf{q}\sigma}$, the eigenvectors of the dynamical matrix, and define eigendisplacements via $\boldsymbol{\eta}_{\mathbf{q}\sigma\kappa}=\frac{\hat{e}_{\mathbf{q}\sigma}}{\sqrt{M_{\kappa}}}$.
-In which case, we may cast Equation \ref{eq:oscillatordefinition} into the form:
-\begin{equation}\label{eq:oscillatorstrength}
-\mathbf{S}_{\mathbf{q}\sigma} = \sum\limits_{\kappa} \mathbf{Z}_{\kappa}\boldsymbol{\eta}_{\mathbf{q}\sigma\kappa}.
-\end{equation}
-We may now use this in concurrence with Equation \cite{eq:polarizationkubo} to evaluate the dielectric function.
-Consider $\boldsymbol{\Pi}(\mathbf{q},\omega)$.
-From the form of Equation \ref{eq:phononstodipole}, we see that at zero temperature, the first term only sees contributions from $m=|0_{\mathbf{q}\sigma}\rangle$ and $n=|1_{\mathbf{q}\sigma}\rangle$, where $0_{\mathbf{q}\sigma}$ and $1_{\mathbf{q}\sigma}$ refer to the number of phonons in mode $\mathbf{q}\sigma$.
-Here $\sigma$ can be any branch of the phonon band structure.
-In the second term, $m$ and $n$ switch.
-In the limit where the phonons are very weakly damped, $\Gamma_{\mathbf{q}\sigma} \\ \omega_{\mathbf{q}\sigma}$.
-this leads to:
-\begin{equation}\label{eq:fullsusceptibility}
-\boldsymbol{\Pi}(\mathbf{q},\omega) = \frac{1}{\Omega}\sum\limits_{\sigma}\frac{\mathbf{S}_{\mathbf{q}\sigma}\otimes \mathbf{S}_{\mathbf{q}\sigma}^*}{\omega^2_{\mathbf{q}\sigma}-\omega^2-i\Gamma_{\mathbf{q}\sigma}\omega},
-\end{equation}
-where $\Omega=N/V$ is the volume of the unit cell.
-Rigorously, the phonon frequencies in this equations are real.
-However, in the realistic situation where there is phonon dissipation, typically due to electron-phonon and phonon-phonon scattering, we may make take a "relaxation-time approximation" which effectively results in expressing $\omega_{\mathbf{q}\sigma} \rightarrow \omega_{\mathbf{q}\sigma} - i\frac{\Gamma_{\mathbf{q}\sigma}}{2}$, where $\Gamma_{\mathbf{q}\sigma}$ is the dissipation rate of the phonon.
-Finally, the dielectric function takes the form
-\begin{equation}\label{eq:pitoepsilon}
-\epsilon(\mathbf{q},\omega) = \epsilon_{\infty} + \frac{1}{\Omega}\sum\limits_{\sigma}\frac{\mathbf{S}_{\mathbf{q}\sigma}\otimes \mathbf{S}_{\mathbf{q}\sigma}^*}{\omega^2_{\mathbf{q}\sigma}-\omega^2-i\Gamma_{\mathbf{q}\sigma}\omega}.
-\end{equation}
-From this equation, let us consider an highly simplified limiting case.
-Assume that the permittivity is isotropic.
-Further assume that the site-specific masses collapse to a single effective mass $M^*$, and that the site-specific and tensorial Born effective charges collapse to a single effective charge $Q^*$.
-Finally, assume that the only phonon with oscillator strength (non-zero $\mathbf{S}$) is the transverse optical phonon at frequency $\omega_{TO}$.
-In this case, one arises at a dielectric function
-\begin{equation}\label{eq:lorentzoscillator}
-\epsilon(\omega) = \epsilon_{\infty} + \frac{NQ^{*2}}{\epsilon_0 M^*V}\frac{1}{\omega^2_{TO}-\omega^2-i\Gamma_{TO}\omega},
-\end{equation}
-which coincides precisely with the phenomenological Lorentz oscillator model, showing the consistency, of our approach.
-That said, our approach takes into account complications arising from the tensorial and site-dependent nature of the Born charges, the effects of a complex unit-cell, as well as a detailed understanding of the phonon band-structure and how it affects the features of the predicted dielectric function.
-
-Before applying this formalism to a real material and discussion of the two-dimensional case, we briefly consider and comment on `nonlocality' in the phonon contribution to the dielectric function.
-Note that Equation \ref{eq:pitoepsilon} has a wavevector dependence which is implicit through the phonon dispersion.
-Therefore, it is possible that the phonon polariton resonances red-shift or blue-shift, depending on how the relative position of the phonons at finite-wavevector versus zero wavevector.
-For example, in 4H-SiC, upon moving along the $\Gamma-L$ direction of the Brillouin zone, the optical phonon which determines the phonon polariton, calculated to be at 764 cm$^{-1}$ at $\Gamma$, red-shifts by 2  cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, by 5 cm$^{-1}$ at 1/10 of the way along the $\Gamma-L$ direction, and 13 cm$^{-1}$ at 1/2 of the way along the $\Gamma-L$ direction.
-This will lead to a red-shift of the Reststrahlen bands by these amounts.
+\subsection{Concept: Effectively non-interacting matter and photon systems}
 \begin{figure*}[t]
-\includegraphics[width=16cm]{combined_structure.pdf}
-\caption{\textbf{Crystal structure and phonon dispersion of bulk 4H silicon carbide and hexagonal boron nitride.} (a) The hexagonal unit cell structure of silicon carbide allows different optical response in the a-b plane and along the c axis.
-(b) The full phonon bandstructure of silicon carbide, from which the phonon eigendisplacements associated with the gamma point were used to calculate the frequency-dependent dielectric function.
-Structure (looking from stacking plane) (c) and phonon bandstructure (d) as before but for bulk hexagonal boron nitride.}
-\label{fig:phonons}
+\includegraphics[width=18cm]{conceptfigure.pdf}
+\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} .}
+\label{fig:ansatz}
 \end{figure*}
 
-%$\epsilon(\mathbf{q},\omega)$ measures the response to an electromagnetic perturbation of frequency $\omega$ and wavevector $\mathbf{q}$.  I
-In the vast majority of experiments that have been performed on phonon polaritons, the wavevector used to probe the polariton is on the order of $\frac{\omega}{c}$ which is much less than the wavevector scale at which the optical phonon dispersion varies, which is of the order of $\frac{\pi}{a}$ where $a$ is a lattice constant of the polar dielectric.
-Even in scattering near-field optical microscope experiments, where one uses an AFM tip of radius of about $20$ nm to probe the polaritons, the wavevectors accessed would still be extremely small compared to the extent of the Brillouin zone.
-It would then seem extremely challenging to observe nonlocal behavior of phonon-polaritons.
-However, a recent experiment \cite{iranzo2018probing} has shown that plasmons in graphene, by means of a gold mirror, can be confined to dimensions of about 1 nm in the dimension transverse to the graphene sheet.
-Additionally, another recent experiment \cite{benz2016single} has leveraged so-called `picocavities', sub-nanometer gaps between a metal nanoparticle and metal film, to access very strong light-matter couplings and extreme variations in fields.
-These two experiments show a way to get large optically-accessed wavevectors to probe nonlocal behavior in polar dielectrics.
+\subsection{Correlation corrections coming from virtual matter and photon quasiparticles}
 
-\section{Application to bulk semiconductor systems}
-% calculation basics
-To demonstrate the practical use of this theoretical framework, we present an example calculation of the dielectric function for both hBN and the 4H polytype of silicon carbide (SiC), as is shown in \ref{fig:phonons}(a).
-Phonon bandstructures for SiC and hBN were calculated within density functional perturbation
-theory (DFPT) as implemented in the ABINIT package\cite{abinit1,abinit2,abinit3,Gonze1997dynamical,Hamann2005metric}, with electric field response included to appropriately capture LO-TO splitting.
-Initial ground state density functional theory (DFT) properties were predicted using norm-conserving Vanderbilt pseudopotentials of the PBEsol parameterization for SiC, and PBE with a D2 (Grimme) van der Waals correction for hBN \cite{PBEsol,PBE,ONCV_PPs,pseudodojo,grimme}.
-We calculated the phonon modes, their frequencies (shown in Figure \ref{fig:phonons}(b)), and their eigendisplacements at the gamma point to construct the dielectric function from Equation \ref{eq:pitoepsilon}, using experimental values for the dissipation rates $\Gamma$.
-We note that while here, we have taken both the SiC bandgap and the SiC optical phonon lifetime from experiments, these quantities may also be obtained \emph{ab initio} by calculating lifetimes associated with electron-phonon scattering \cite{sundararaman2014theoretical,brown2015nonradiative,brown2016ab,narang2017effects} and three-phonon decay processes \cite{srivastava1990physics}.
+\section{Application to a multi-level system coupled to a multi-mode cavity}
 
-% about the scissor shift -- and a defense sentence in case someone hits us with "semi-empirical" later on
-Additionally, for SiC a scissor shift of 1.01 eV was applied to shift the calculated band gap of 2.26 eV to match the experimental gap of $\sim$3.27 eV \cite{band_gap_expt1,band_gap_expt2}.
-This is necessary to correct for the well-known underestimation of the band gap in ground state density functional theory calculations and the corresponding effects to the dielectric response.
-In materials like hBN where excitonic effects play a significant role, the optical properties of the system are not determined solely by the energy associated with band to band recombination, and instead must also account for the exciton binding energy\cite{robert2016excitonic}.
-Therefore, for the calculation of the dielectric function of hBN, we consider the optical gap, E$_{opt}$ = E$_g$ - E$_{b}$, with Eg = 5.2 eV from experiment\cite{levinshtein2001properties} and E$_{b}$ = 0.67 eV from recent \emph{ab initio} calculations\cite{attaccalite2018two}, so that E$_{opt}$ is approximately 4.5 eV.
-As the bandgap we compute from DFT is 4.48 eV, we did not require a correction to the gap in hBN.
-From these calculations, the high-frequency dielectric tensor for SiC was found to be $\epsilon^{\infty}_{\parallel}$ = 6.986, $\epsilon^{\infty}_{\perp}$ = 7.307 and for hBN, $\epsilon^{\infty}_{\parallel}$ = 4.634, $\epsilon^{\infty}_{\perp}$ = 2.593.
-This also leads to a modification of the phonon spectra.
-Note that for a materials where the optical band gap is not as well known as in SiC, a more computationally intensive GW calculation could be used while retaining the \emph{ab initio} implementation of our theory.
-Further, if excitonic effects contribute significantly to the optical properties of a chosen material, a Bethe-Salpeter Equation prediction might be used to further improve predictive accuracy within the same framework presented here.
-The resulting frequency-dependent dielectric tensor, compared to experimental values from \cite{tiwald1999carrier,caldwell2014sub}, is shown in Figures \ref{fig:sicepsilon} and \ref{fig:hbnepsilon}.
-As can be seen, the agreement between the theoretical calculation and the experimental results is quite good for silicon carbide, and good for the upper Reststrahlen band of hBN.
-For the lower Reststrahlen band, the major discrepancy results from a roughly 5\% deviation between calculated and measured transverse optical phonon frequencies.
-
-% can also put this table just by listing eps^inf_perp and parallel in the main text.  
-\begin{figure}[t]
-\includegraphics[width=8.5cm]{SiC_comparison_thy_vs_expt.pdf}
-\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk 4H silicon carbide.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk 4H silicon carbide from using Equation \ref{eq:pitoepsilon}.
-The phonon properties and $\epsilon_{\infty}$ calculated from the ABINIT package.
-Overlaid in blue dots are the experimental values for the dielectric function, taken from \cite{tiwald1999carrier}.
-The real part is shown in (a) while the imaginary part is shown in (b).
-}
-\label{fig:sicepsilon}
-\end{figure}
-
-\begin{figure}[t]
-\includegraphics[width=8.5cm]{hBN_comparison_thy_vs_expt.pdf}
-\caption{\textbf{Comparison of theoretical predictions and experimental values of the infrared dielectric function of bulk hexagonal boron nitride.} The frequency-depended dielectric tensor in the direction perpendicular to the c-axis (orange), calculated for bulk hexagonal boron nitride from using Equation \ref{eq:pitoepsilon}.
-The phonon properties and $\epsilon_{\infty}$ calculated from the ABINIT package.
-Overlaid in blue dots are the experimental values for the dielectric function, taken from \cite{caldwell2014sub}.
-The component of the real part of the permittivity parallel to the c-axis is shown in (a) component parallel is shown in (b).}
-\label{fig:hbnepsilon}
-\end{figure}
-
-%Thus, to evaluate it, we must know the microscopic states of the lattice, in addition to matrix elements of displacement.
-Working in the harmonic approximation, the states are simply phonon-occupation states $|\{n_{qs}\}\rangle$, labeled by their wavevector $\mathbf{q}$ in the Brillouin zone and their band $s$ which encodes transverse/longitudinal polarization of acoustic/optical branches.
-The phonons are also labeled by their frequency $\omega_{qs}$ and the polarization vector at each basis site $b$, $\hat{e}_{\mathbf{q}s}(\mathbf{b})$.
-The displacement of each ion in the lattice can be expressed as an expansion over phonon modes as:
-% \begin{equation}
-% \mathbf{u}(\mathbf{R},\mathbf{b},t) = .
-% \end{equation}
-% From the displacement, we then must construct the dipole moment associated with each unit cell in the lattice. This is given by:
-% \begin{equation}
-% \mathbf{d}(\mathbf{R}) = e\sum\limits_{\mathbf{b}} Z_{ij}u_{j},
-% \end{equation}
-% where $\mathbf{Z}$ is a tensor of Born effective charges.
-These charges formally are defined as the derivative of the lattice polarization with respect to charge displacement.
-They act as effective charges for the ions, taking into account the motion of electrons and other ions in the system.
-Once we have the dipole moment in each unit cell, we can now complete the formal evaluation of the polarization-polarization response function.
-Assuming that we are operating at zero temperature, it is given by:
-
-\section{Linear response of phonon polaritons in two dimensions}
-We now move to generalize the derivation of the previous section to describe the optical phonon contribution to Maxwell's equations coming from a two-dimensional material.
-Unlike the previous section, we will not demonstrate the results of the calculation on a particular material.
-Here, we are only interested in finding an expression for the optical response of the surface in terms of \emph{ab initio}-derivable parameters, as well as the general form of the dispersion relation of phonon-polaritons in 2D.
-Because of the change of symmetry in the system, we will not be able to perform the analysis by Fourier transforming Maxwell's equations in three dimensions.
-However, we will apply a technique which is often used to analyze electromagnetic modes in 2D systems such as plasmons in 2D conductors.
-An example of the technique is shown in detail in \cite{jablan2009plasmonics}, but we summarize the essential details here.
-
-Consider a two-dimensional ionic lattice located in the $xy$ plane ($z=0$) surrounded by a homogeneous environment of permittivity $\epsilon$.
-We consider the simplified case where the lattice is free-standing, to show the essential concept.
-We search for a mode whose magnetic field $\mathbf{H}$ is of the form $\mathbf{H}(z)e^{i\mathbf{q}\cdot\mathbf{r}_{||}-i\omega  t}$, where $\mathbf{r}_{||}=(x,y,0)$ is the position in the plane, $\mathbf{q}$ is the wavevector and $\omega_{\mathbf{q}}$ is the frequency.
-To simplify the discussion even further and furnish analytical forms for the phonon-polariton modes in 2D, we assume this material has mirror symmetry along the axis of $\mathbf{q}$.
-In this case, the electromagnetic modes can be decomposed into TE and TM-polarized modes.
-As is typical in the study of highly-confined polaritons, whether they are plasmon- or phonon- polaritons, the TM mode is the highly confined one, and occurs when $\epsilon < 0$.
-In our convention, the TM mode is such that $\mathbf{H}$ is in the plane of the lattice and perpendicular to $\mathbf{q}$.
-Without loss of generality, define the $\mathbf{q}$ direction to be the $x$ direction, making the magnetic field in the $y$ direction.
-Moreover, the magnetic field in this mode flips sign under $z \rightarrow -z$.
-
-For $z \neq 0$, the Maxwell equation for the magnetic field is $(\nabla\times\nabla\times - \epsilon\frac{\omega^2}{c^2})\mathbf{H} = 0$. This equation, for the TM mode described above, becomes 
-\begin{equation}\label{eq:2dmaxwell}
-\left(-\frac{d^2}{dz^2}+\mathbf{q}^2-\epsilon\frac{\omega^2}{c^2} \right)H = 0,
-\end{equation}
-where we have written $\mathbf{H} = H\hat{y}$. The solution in the region $z > 0$ is simply \begin{align}
-H &= e^{i\mathbf{q}\cdot\mathbf{r}_{||}-\kappa z}, ~z>0 \nonumber \\
-H &= -e^{i\mathbf{q}\cdot\mathbf{r}_{||}-\kappa z}, ~z>0 \nonumber \\
-\end{align}
-where $\kappa = \sqrt{\mathbf{q}^2-\epsilon\frac{\omega^2}{c^2}}$ and we have implicitly assumed $q > \epsilon\frac{\omega}{c}$ as we are interested in confined evanescent modes.
-To join these two solutions consistently, we apply the interface conditions for the magnetic field.
-They are $\hat{z}\times(\mathbf{H}^{(+)}-\mathbf{H}^{(-)}) = \mathbf{K} = -i\omega\mathbf{P}_s$ where $\mathbf{K}$ is the surface current density, expressed through the surface polarization density $\mathbf{P}_s$.
-Now we use the linear response relation $\mathbf{P}(\mathbf{q},\omega) = \boldsymbol{\Pi}(\mathbf{q},\omega)\mathbf{E}(\mathbf{q},\omega)$ in addition to Ampere's law $\nabla\times\mathbf{H} = -i\omega\epsilon\mathbf{E}$.
-Plugging these relations in yields an implicit equation connecting the polarization response function, the frequency of the 2D phonon polariton, and its frequency.
-Taking the case where $x$ coincides with a principal axis of the system, we have that
-\begin{equation}\label{eq:2ddispersion}
-q\Pi_{xx}(q,\omega)  = -2\epsilon.
-\end{equation}
-
-To connect this relation to the discussion of the three-dimensional case, we write a microscopic form for the polarization response function of the ionic lattice. It is simply
-\begin{equation}\label{eq:2dsusceptibility}
-\boldsymbol{\Pi}(\mathbf{q},\omega) =  \frac{1}{\epsilon_0 A}\sum\limits_{m,n}\frac{\mathbf{P}_{mn}(\mathbf{q})\otimes\mathbf{P}_{nm}(\mathbf{q})}{\hbar\omega + E_{nm}+i0^+}\left(e^{\beta E_m}-e^{\beta E_n} \right),
-\end{equation}
-where the only change is to replace a normalization volume with a normalization area $A$.
-Here, $\mathbf{P}(\mathbf{q})$ is the 2D Fourier transform of the polarization density.
-As the form of the displacement of the ions is still given by Equation (3), $\mathbf{P}(\mathbf{q})$ is still given by Equations (7-10), but with the Born charges and eigendisplacements appropriate to the 2D system of interest.
-Note that although the lattice is two dimensional, in general, the ions may be displaced in any three directions and thus $\boldsymbol{\Pi}$ is still a $3\times3$ matrix.
-Following the same kind of reasoning that lead to Equation \ref{eq:lorentzoscillator}, if we parameterize the effect of the different masses and Born charges in the unit cell by $M^*$ and $Q^*$, we will find a form for the polarization response function which is given by the second term of Equation \ref{eq:lorentzoscillator} but with $V$ replaced by $A$.
-Defining $N/A \equiv n_s$, Equation \ref{eq:2dmaxwell} for the dispersion relation of 2D phonon polaritons tells us that
-\begin{equation}\label{eq:2dphononpolaritondispersion}
-\omega = \sqrt{\omega_{TO}^2+\frac{n_sQ^{*2}}{2M^*\epsilon_0\epsilon}q}.
-\end{equation}
-Note that this dispersion has a strong resemblance to the plasmon dispersion relation of two-dimensional electron gases, which is normally of the form\cite{stern1967polarizability} $\omega = \sqrt{\frac{n_s e^2  }{2m\epsilon_0\epsilon}q}$ .
-The only difference is the replacement of electron parameters (density, charge, mass) by the corresponding lattice parameters in the phononic case, and a shift in the minimum frequency by the transverse optical phonon frequency.
-
-We note that the dispersion relation of Equation \ref{eq:2dphononpolaritondispersion} can in fact be obtained by taking the infinitely thin-limit of a bulk polar dielectric, and finding the dispersion of the resulting phonon polaritons in the quasi-electrostatic limit.
-In particular, for an isotropic polar dielectric of thickness $2d$ with boundaries at $z=\pm d$, solving Laplace's equations yields as the dispersion of the even parity mode:
-\begin{equation}\label{eq:thinfilmphononpolaritondispersion}
-\tanh(qd) = -\frac{\epsilon}{\epsilon(\omega)} \rightarrow qd\epsilon(\omega) = -\epsilon,
-\end{equation}
-with $\epsilon$ the dielectric constant of the surrounding background and $\epsilon(\omega)$ that of the polar dielectric.
-The right-hand side is obtained in the limit of $d\rightarrow 0$.
-Taking Equation \ref{eq:lorentzoscillator} for the dielectric function in bulk, we have that (in the absence of losses) $\epsilon + qd\epsilon_{\infty} + q\frac{NQ^{*2}}{2\epsilon_0 M^*A}\frac{1}{\omega_{TO}^2-\omega^2} = 0$, where we have identified $V = 2Ad$.
-In the limit of small $d$, i.e., $qd \ll 1$, we recover Equations \ref{eq:2ddispersion} and \ref{eq:2dphononpolaritondispersion} with $\Pi = \frac{NQ^{*2}}{\epsilon_0 MA}\frac{1}{\omega^2_{TO}-\omega^2}$.
-A useful result which follows is that we may also derive another important figure of merit, the $p-$polarized reflectivity of a 2D polar dielectric, which is related to the local density of states of the polar dielectric, which quantifies the strength of coupling to external probes such as a nano-antenna or a quantum emitter.
-By considering the infinitely thin limit of a polar dielectric with dielectric function $\epsilon(\omega)$ and defining $\epsilon(\omega) = \frac{\Pi}{2d}$, we immediately find that the p-polarized reflectivity is given by
-\begin{equation}\label{eq:2dtmreflectivity}
-r_p = \frac{1}{1+\frac{2\epsilon}{q\Pi}}.
-\end{equation}
-
-We note that, in general, when considering bulk and 2D versions of a material, the effective parameters $Q^*$ and $M$ will change as a result of changes to the phonon band structure.
-The losses may also change drastically in the reduction of dimensions.
-For example, it was found in several experiments that the damping rate of the TO phonon in hexagonal boron nitride nearly doubles upon going from bulk to 2D\cite{gorbachev2011hunting,tran2016quantum}, which may strongly suppress the ability to image the polariton with near-field probes.
-This is due to the small thickness where the dispersion of the phonon polariton is quite flat, and very high wavevector modes, corresponding to the maximum wavevector detectable by a near-field tip, occur at frequencies very close to the TO phonon frequency.
-Near the TO phonon-frequency however, the ratio $\frac{\text{Im }\Pi}{\text{Re }\Pi}$ is the highest, corresponding to very low quality-factor modes.
-In addition to these considerations, the values of the parameters may depend on the background material, which may change the phonon band structure.
-
-We provided this thin-film derivation as a simple way to understand the content of Equations \ref{eq:2ddispersion} and \ref{eq:2dphononpolaritondispersion}.
-In general Equation \ref{eq:2dsusceptibility} should be used to calculate $\Pi$ and the corresponding dispersion.
-We note that a similar situation appears in the treatment of 2D plasmons: namely that a relation like \ref{eq:2ddispersion} appears, except with $\Pi$ the polarization-polarization response of a 2D electron gas.
-And furthermore that one can also obtain the same equation by considering the ultra-thin limit of a 3D plasmonic material \cite{jablan2009plasmonics,jablan2013plasmons}.
-
-\section{Summary and Outlook}
-
-In summary, we have provided a theoretical framework based on linear response theory to calculate the phonon contribution to the dielectric function from first principles.
-Our framework is rather versatile, allowing us to: use first principles calculations to get the dielectric function, predict how nonlocality enters the dielectric function, and predict the effect of reduced dimensionality.
-A particularly interesting case to consider in future work would be to find a system where the optical phonons are drastically different from 3D to 2D.
-Perhaps it is possible that there are some materials in which the 2D optical phonons experience lower losses due to a reduced scattering phase space.
-The formalism we provide here may also be extended to understand phonon-polaritons in other more atypical reduced-dimensional settings, such as zero-dimensional settings in single emitters, i.e., `molecular phonon polaritons', in analogy to recent work on `molecular plasmons' \cite{manjavacas2013tunable,lauchner2015molecular}.
-Notably, we go beyond oversimplifications of the Lorentz oscillator model in which the Born charges are treated as a single scalar quantity and can treat the influence and interplay of many phonon modes that contribute to the dielectric function.
-We corroborated our approach through density functional theoretic calculations.
-
-In future work, besides furnishing an \emph{ab initio} description of nonlocality and reduced-dimension phonon polaritons, there are a number of interesting questions that can be addressed by the framework provided here.
-For example, we may consider situations such as the phonon polaritons associated with optical phonons at the interface of two materials.
-Another system of interest would be to consider two nearby layers of material in which the optical phonons of each material strongly couple to each other.
-In that case, it would be relevant to evaluate how this strong coupling manifests itself in the infrared dielectric function, and ultimately the confinement and propagation of the phonon polaritons.
+\section{Outlook}
 
 \section{Acknowledgements}
-This research used resources of the National Energy Research Scientific Computing Center, a DOE Office of Science User Facility supported by the Office of Science of the U.S.
-Department of Energy under Contract No.
-DE-AC02-05CH11231.
+% Pri, Johannes: Please insert any other relevant acknowledgement information here!
 N.
@@
 %The authors thank XXXX for helpful discussions.
-Tentatively Josh Caldwell, Jeremy Baumberg, Dirk England, etc.
+Tentatively Joel Yuen Zhou and Javier Aizpurua.
 

### 2018-08-19 paper.tex
intro claims

@@
 
+In what follows, we develop a variational ansatz for the ground state of quantum electrodynamical Hamiltonians.
+In particular, we develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states.
+This ansatz, analogous to the Hartree-Fock ansatz, leads to coupled eigen-equations describing the ground-state of the light-matter system.
+We benchmark our ansatz against a multi-level and multi-mode Rabi model parameterized by the strength of the characteristic momentum matrix elements describing the coupling of the matter and the electromagnetic fields.
+We find that when the momentum matrix elements are sufficiently weak, our theory finds the ground state energy to a remarkable accuracy of less than 1\% error, even in the non-perturbative strong coupling regime.
+Even in regimes where the correlation energy is 200\% of the total energy, far outside of the validity domain of our ansatz, our ansatz predicts trends in the evolution of the ground state energy which are not predicted by perturbative analysis.
+In regimes where our results are accurate, we claim we have found the effective quasiparticle description of the ground state.
+Moreover, in this regime, the ground state interaction energy manifests itself as the difference in zero-point energies between the quasiparticles and the bare particles.
+This generalizes the notion of the Casimir energy to a single atom in a cavity.
+
+Our results accomplish several aims.
+First, they provide a starting point for the description of light-matter systems beyond simple two-level and/or single-mode descriptions.
+Secondly, they prove a general framework for understanding light-matter decoupling effects in the regime of strong light-matter interactions.
+Third, they prove a non-perturbative theory of Casimir-Polder forces.
+Finally, they provide a rigorous concept of correlation energy in quantum electrodynamics in a way analogous to electronic structure theory.
+
+The outline of this work is as follows: in section I, we develop the variational ansatz for the ground state of the Hamiltonian of macroscopic quantum electrodynamics.
+In section II, we calculate the ground state energy according to this ansatz for the multi-level and multi-mode Rabi model and compare it to numerical diagonalization of the same problem.
+In section III, we study the accuracy of our ansatz as a function of the correlation energy in the system, and test several approaches to recover the correlation energy.
+
 \section{Variational ansatz}
@@
 \section{Application to a multi-level system coupled to a multi-mode cavity}
+
+\section{Strategies for recovering the correlation energy}
 

### 2018-08-29 paper.tex
filling out more

@@
 We benchmark our ansatz against a multi-level and multi-mode Rabi model parameterized by the strength of the characteristic momentum matrix elements describing the coupling of the matter and the electromagnetic fields.
-We find that when the momentum matrix elements are sufficiently weak, our theory finds the ground state energy to a remarkable accuracy of less than 1\% error, even in the non-perturbative strong coupling regime.
-Even in regimes where the correlation energy is 200\% of the total energy, far outside of the validity domain of our ansatz, our ansatz predicts trends in the evolution of the ground state energy which are not predicted by perturbative analysis.
+We find that when the momentum matrix elements are sufficiently weak, our theory finds the ground state energy to a remarkable accuracy of less than >99.9\%, even in the non-perturbative coupling regime.
+Even in regimes where the correlation energy is 50\% of the total energy, far outside of the validity domain of our ansatz, our ansatz predicts the energy up to a few percent.
 In regimes where our results are accurate, we claim we have found the effective quasiparticle description of the ground state.
@@
 
-Our results accomplish several aims.
-First, they provide a starting point for the description of light-matter systems beyond simple two-level and/or single-mode descriptions.
-Secondly, they prove a general framework for understanding light-matter decoupling effects in the regime of strong light-matter interactions.
+Our results service several aims.
+First, they suggest a novel variational approach for understanding ground states in ultra-strongly coupled light-matter systems.
+Second, they provide a general framework for understanding light-matter decoupling effects in the regime of strong light-matter interactions.
 Third, they prove a non-perturbative theory of Casimir-Polder forces.
@@
 In section III, we study the accuracy of our ansatz as a function of the correlation energy in the system, and test several approaches to recover the correlation energy.
+We consider both perturbation theory formulated in the interacting modes as well as a self-consistent perturbation theory which allows the perturbative energy shift to influence the modes.
 
 \section{Variational ansatz}
+
+In this work, we consider the light-matter Hamiltonian corresponding to an emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction.
+As the cavity is considered for simplicity to be one-dimensional, the electric field is oriented along a single direction, denoted $x$, while the magnetic field is oriented along a direction transverse to both the electric field and the cavity length, denoted $y$.
+We consider the interaction in the velocity gauge.
+The Hamiltonian can then be written as:
+\begin{equation}
+H = H_{\text{matter}}+\frac{\epsilon_0}{2}\int dz~(E^2+c^2B^2)+\frac{e}{m}A(d)p + \frac{q^2}{2m}A^2(d).
+\end{equation}
+The fields can be expressed as a mode expansion via:
+\begin{equation}
+A(z,t) = \sum\limits_{n=1}^{\infty} \sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L}\right)(a_ne^{-i\omega_n t}+a_n^{\dagger}e^{i\omega_n t}),
+\end{equation}
+with $E(z,t) = -\partial_t A(z,t)$ and $B(z,t)=\partial_z A(z,t)$.
+The mode frequencies of the cavity are $\omega_n = \frac{n\pi c}{L}$.
+The matter Hamiltonian we take to be a multilevel system with an arbitrary number of levels, $N_a$.
+The matter system we describe as an $N_a$ site system, perhaps describing a molecule in an LCAO framework.
+Thus we parameterize the general family of matter Hamiltonians as:
+\begin{equation}
+H_{\text{matter}} = \sum\limits_{i=1}^{{N_a-1}} V_i|i\rangle\langle i|+t(|i\rangle\langle i+1|+|i+1\rangle\langle i|) 
+\end{equation}
 
@@
 \includegraphics[width=18cm]{conceptfigure.pdf}
-\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} .}
+\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons.
+(Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
 \label{fig:ansatz}
 \end{figure*}
-
-\subsection{Correlation corrections coming from virtual matter and photon quasiparticles}
 
@@
 % Pri, Johannes: Please insert any other relevant acknowledgement information here!
+We thank Joel Yuen-Zhou for useful discussions.
 N.
 R.
-and J.
-C.
-recognize the support of the DOE Computational Science Graduate Fellowship (CSGF) fellowship no.
+recognizes the support of the DOE Computational Science Graduate Fellowship (CSGF) fellowship no.
 DE-FG02-97ER25308.

### 2018-08-29 paper.tex
filling out more theory section

@@
 
-\section{Variational ansatz}
-
 In this work, we consider the light-matter Hamiltonian corresponding to an emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction.
@@
 \begin{equation}
-A(z,t) = \sum\limits_{n=1}^{\infty} \sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L}\right)(a_ne^{-i\omega_n t}+a_n^{\dagger}e^{i\omega_n t}),
+A(z,t) = \sum\limits_{n=1}^{\infty} \sqrt{\frac{\hbar}{2\epsilon_0\omega_n }}(F_n(z)a_ne^{-i\omega_n t}+F^*_n(z)a_n^{\dagger}e^{i\omega_n t}),
 \end{equation}
 with $E(z,t) = -\partial_t A(z,t)$ and $B(z,t)=\partial_z A(z,t)$.
-The mode frequencies of the cavity are $\omega_n = \frac{n\pi c}{L}$.
+The $F_n(z)$ are the mode functions of the cavity, normalized such that $\int dz ~|F_n(z)|^2 = 1$.
+For a cavity of length $L$ the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$.
 The matter Hamiltonian we take to be a multilevel system with an arbitrary number of levels, $N_a$.
@@
 \end{equation}
+
+\section{Variational ansatz}
+
+The concept behind the variational ansatz is shown in Figure (1).
+We seek a description of the ground state of the system as $\Psi\rangle \approx |\tilde{g}\rangle\otimes|\tilde{0}\rangle + |\delta\rangle,$ where a $\sim$ denotes an effective quantity.
+In other words, we seek a description in which the state is a factorizable state of matter and photon \textit{quasi}particles, up to some correlations quantified by the state $|\delta\rangle$.
+These correlations essentially hybridize $|\tilde{g}\rangle\otimes|\tilde{0}\rangle$ with excitations of the effective emitter and virtual effective photons.
+
+We start our analysis by considering a family of ansatze in which $\delta = 0$ (uncorrelated ansatz).  The expectation value of the Hamiltonian is 
+\begin{align}
+\langle \Psi | H | \Psi \rangle &= \langle \tilde{g} |H_{\text{matter}} | \tilde{g}\rangle \nonumber \\ &+ \frac{\hbar}{4}\int dz ~\sum\limits_{n=1}^{\infty}\left(\omega_n|F_n|^2 - \frac{c^2}{\omega_n}F_n^*\partial_z^2F_n\right) \nonumber \\ &+ \frac{\hbar q^2}{4m\epsilon_0}\sum\limits_{n=1}^{\infty} \int dz~\delta(z-d)|F_n|^2
+\end{align}
+Notably, in this ansatz, the term in the Hamiltonian coupling the momentum of the matter to the photon makes no contribution.
+We impose constraints of matter normalization and photon mode normalization through Lagrange multipliers $\epsilon_a(\langle \tilde{g}|\tilde{g}\rangle-1)$ and $\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right)$.
 

### 2018-08-30 paper.tex
more theory

@@
 
-\title{Variational theory of the ground state of cavity quantum electrodynamics}
+\title{Variational theory ground states in quantum electrodynamics}
 
@@
 The matter Hamiltonian we take to be a multilevel system with an arbitrary number of levels, $N_a$.
-The matter system we describe as an $N_a$ site system, perhaps describing a molecule in an LCAO framework.
+The matter system we describe as an $N_a$ site system.
+This could be considered as a simple model of a molecule within a tight-binding description.
 Thus we parameterize the general family of matter Hamiltonians as:
@@
 \section{Variational ansatz}
+\begin{figure*}[t]
+\includegraphics[width=16cm]{conceptfigure.pdf}
+\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons.
+(Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
+\label{fig:ansatz}
+\end{figure*}
 
@@
 \begin{align}
-\langle \Psi | H | \Psi \rangle &= \langle \tilde{g} |H_{\text{matter}} | \tilde{g}\rangle \nonumber \\ &+ \frac{\hbar}{4}\int dz ~\sum\limits_{n=1}^{\infty}\left(\omega_n|F_n|^2 - \frac{c^2}{\omega_n}F_n^*\partial_z^2F_n\right) \nonumber \\ &+ \frac{\hbar q^2}{4m\epsilon_0}\sum\limits_{n=1}^{\infty} \int dz~\delta(z-d)|F_n|^2
+\langle \Psi | H | \Psi \rangle &= \langle \tilde{g} |H_{\text{matter}} | \tilde{g}\rangle \nonumber \\ &+ \frac{\hbar}{4}\int dz ~\sum\limits_{n=1}^{\infty}\left(\omega_n|F_n|^2 - \frac{c^2}{\omega_n}F_n^*\partial_z^2F_n\right) \nonumber \\ &+ \frac{\hbar q^2}{4m\epsilon_0\omega_n}\sum\limits_{n=1}^{\infty} \int dz~\delta(z-d)|F_n|^2
 \end{align}
 Notably, in this ansatz, the term in the Hamiltonian coupling the momentum of the matter to the photon makes no contribution.
-We impose constraints of matter normalization and photon mode normalization through Lagrange multipliers $\epsilon_a(\langle \tilde{g}|\tilde{g}\rangle-1)$ and $\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right)$.
+We impose constraints of matter normalization and photon mode normalization by defining a Lagrange function $\mathcal{L} \equiv \langle \Psi | H | \Psi \rangle - \epsilon(\langle \tilde{g}|\tilde{g}\rangle-1)-\sum\limits_{n=1}^{\infty}\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right)$.
+To find the ground state, we minimize the Lagrange function with respect to the matter orbital $|\tilde{g}\rangle$ and with respect to the mode functions $F_n$.
+The minimization with respect to the matter leads to the trivial equation $H_{\text{matter}} |\tilde{g}\rangle = \epsilon|\tilde{g}\rangle$ which leaves the effective matter ground state as simply the ground state of $H_{\text{matter}}$.
+On the other hand, the minimization with respect to the photon mode functions leads to  the equation
+\begin{equation}
+\left(\partial_z^2-\frac{\omega^2_n}{c^2}+2\frac{\omega_n\lambda_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
+\end{equation}
+We may constrain $\lambda$ by differentiating the Lagrange function also with respect to the $\omega_n$. The equation which follows is:
+\begin{equation}
+\int dz ~\left(|F_n|^2 + \frac{c^2}{\omega^2_n}F_n^*\partial_z^2F_n\right) - \frac{ q^2}{m\epsilon_0\omega^2_n} \int dz~\delta(z-d)|F_n|^2 = 0
+\end{equation}
+Performing $\frac{\omega_n^2}{c^2}\int dz~F_n^*$ on both sides of Equation (5), and adding this equation to Equation (6), one immediately finds that $\lambda_n = \omega_n$ and that
+\begin{equation}
+\left(\partial_z^2+\frac{\omega^2_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
+\end{equation}
 
-\subsection{Concept: Effectively non-interacting matter and photon systems}
-\begin{figure*}[t]
-\includegraphics[width=18cm]{conceptfigure.pdf}
-\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons.
-(Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
-\label{fig:ansatz}
-\end{figure*}
+This is an ordinary second-order differential equation with the conditions that $F_n$ is continuous at $d$ and that its derivative is discontinuous according to 
+\begin{equation}
+\partial_zF_n\Big|_{z=d^+}-\partial_zF_n\Big|_{z=d^-} = \frac{q^2}{m\epsilon_0 c^2}F_n(d),
+\end{equation}
+in addition to the usual condition of the modes vanishing at the cavity walls $z=0$ and $z=L$. It can be shown that the solution to Equation (7) satisfying such boundary conditions is:
+
+provided that the auxiliary condition
+\begin{equation}
+\cot\left(\frac{\omega_n}{c}d \right)+\cot\left(\frac{\omega_n}{c}(L-d) \right) = \frac{q^2}{4m\epsilon_0\omega_nc}
+\end{equation}
+is met.
+The condition of Equation (9) determines the resonance frequencies of the photon quasiparticle modes.
+After calculating the modes and the frequencies, we may move to an evaluation of the energy of the ground state.
+We now show that the energy of the ground state in the interacting light-matter system ($q \neq 0$) is different from the energy of the ground state in the non-interacting system by an amount equal to the difference in zero-point energy between the photon quasiparticle modes and the bare photon modes.
+
+The result just derived establishes that the interaction energy within this ansatz is nothing more than the Casimir energy of the system.
 

### 2018-08-30 paper.tex
even more theory wow

@@
 in addition to the usual condition of the modes vanishing at the cavity walls $z=0$ and $z=L$. It can be shown that the solution to Equation (7) satisfying such boundary conditions is:
-
+\begin{align}
+&\theta (z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}-\right) \nonumber \\ 
+-&\theta (z-d) \left(\frac{\cos\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\sin\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\ 
++&\theta (d-z) \sin\left(\frac{\omega_n z}{c} \right)
+\end{align}
 provided that the auxiliary condition
@@
 \end{equation}
-is met.
+is met. To ensure that the modes are normalized according to the constraint, we have that the solutions of Equation (9) must be multiplied by a normalization factor $N_n$ given by
+\begin{equation}
+N_n = 2\sqrt{\frac{c}{\omega_n\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
+\end{equation}
 The condition of Equation (9) determines the resonance frequencies of the photon quasiparticle modes.
 After calculating the modes and the frequencies, we may move to an evaluation of the energy of the ground state.
+We proceed by taking the solutions of Equation (9), normalized with the normalization factor of Equation (11), and plugging it back into the expectation value of the Hamiltonian of the ground state (Equation (4))
+
 We now show that the energy of the ground state in the interacting light-matter system ($q \neq 0$) is different from the energy of the ground state in the non-interacting system by an amount equal to the difference in zero-point energy between the photon quasiparticle modes and the bare photon modes.
-
-The result just derived establishes that the interaction energy within this ansatz is nothing more than the Casimir energy of the system.
+The result just derived establishes that the interaction energy within this ansatz is the Casimir energy of the system.
+This is quite interesting as typically when one calculates a Casimir energy, what is calculated is the zero-point energy associated with the electromagnetic modes of \textit{macroscopic polarizable objects}.
+Here on the other hand, we consider the \textit{electromagnetic modes of a single atom in a cavity}.
 

### 2018-08-30 paper.tex
about 50% of text written now

@@
 
-\title{Variational theory ground states in quantum electrodynamics}
+\title{Variational theory of the ground state of quantum electrodynamics}
 
@@
 \begin{abstract}
-
+In this work, we develop a variational theory of the ground state of a quantum electrodynamical system of coupled light and matter.
+Essential to our ansatz is the notion an effective photonic vacuum.
+As a first step towards an eventually \textit{ab initio} approach to ground states in QED, we apply our ansatz to a Rabi model involving a multi-level emitter with many optical modes, a model which cannot be analytically solved.
+We find a compact semi-analytic formula which describes the ground state energy very accurately in all regimes of coupling parameters.
 \end{abstract}
@@
 \maketitle
+
+The past decade has brought an explosion of effort and experimental progress in realizing the coupling of matter and quantized electromagnetic fields beyond a regime of weak coupling, in which perturbative treatments can accurately describe the dynamics.
 
@@
 \end{equation}
+The momentum operator, we write in a finite-different representation as
+\begin{equation}
+p = \frac{i}{R}\sum\limits_{i=1}^{N_a-1} \left(|i\rangle\langle i+1|-|i+1\rangle\langle i| \right),
+\end{equation}
+where $R$ is a constant with units of length representing roughly the difference in positions between sites.
+This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the following sum rule is enforced:
+\begin{equation}
+\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1,
+\end{equation}
+which is the Thomas-Reiche-Kuhn sum rule.
+Of course, this sum rule does not rigorously apply to a discrete-level system.
+However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot.
+As we shall show later in the text however, it does place a bound on how strong the effect of the $Ap$ term can be on the total energy.
+The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.
 
@@
 The concept behind the variational ansatz is shown in Figure (1).
-We seek a description of the ground state of the system as $\Psi\rangle \approx |\tilde{g}\rangle\otimes|\tilde{0}\rangle + |\delta\rangle,$ where a $\sim$ denotes an effective quantity.
+We seek a description of the ground state of the system as $|\Psi\rangle \approx |\tilde{g}\rangle\otimes|\tilde{0}\rangle + |\delta\rangle,$ where a $\sim$ denotes an effective quantity.
 In other words, we seek a description in which the state is a factorizable state of matter and photon \textit{quasi}particles, up to some correlations quantified by the state $|\delta\rangle$.
@@
 After calculating the modes and the frequencies, we may move to an evaluation of the energy of the ground state.
-We proceed by taking the solutions of Equation (9), normalized with the normalization factor of Equation (11), and plugging it back into the expectation value of the Hamiltonian of the ground state (Equation (4))
-
-We now show that the energy of the ground state in the interacting light-matter system ($q \neq 0$) is different from the energy of the ground state in the non-interacting system by an amount equal to the difference in zero-point energy between the photon quasiparticle modes and the bare photon modes.
+We proceed by taking the solutions of Equation (9), normalized with the normalization factor of Equation (11), and plugging it back into the expectation value of the Hamiltonian of the ground state (Equation (4)).
+After plugging in the quasiparticle photon modes into Equation (4), we find that
+\begin{equation}
+\langle \Psi | H|\Psi\rangle = E_{g}+\frac{1}{2}\sum\limits_{n=1}^{\infty}\hbar\omega_n.
+\end{equation}
+The interaction energy, which is the difference between the ground state energy with $q\neq 0$ and the non-interacting ground state energy, is $\frac{1}{2}\sum\limits_{n=1}^{\infty}\hbar(\omega_n-\omega_n^0)$, where $\omega_n^0 = \frac{n\pi c}{L}$ are the frequencies of the bare cavity modes.
 The result just derived establishes that the interaction energy within this ansatz is the Casimir energy of the system.
@@
 
-\section{Strategies for recovering the correlation energy}
+%\section{Strategies for recovering the correlation energy}
 

### 2018-09-27 paper.tex
paper draft

@@
 
-\author{Nicholas Rivera$^\perp$}\email{nrivera@seas.harvard.edu}\affiliation{\HarvardSEAS}\affiliation{\MITPhy}
+\author{Nicholas Rivera}\email{nrivera@seas.harvard.edu}\affiliation{\HarvardSEAS}\affiliation{\MITPhy}
 \author{Johannes Flick}\email{flick@g.harvard.edu}\affiliation{\HarvardSEAS}
@@
 \begin{abstract}
-In this work, we develop a variational theory of the ground state of a quantum electrodynamical system of coupled light and matter.
-Essential to our ansatz is the notion an effective photonic vacuum.
+The past decade has brought rapid theoretical and experimental advances in the field of ultra-strong coupling of light and matter from microwave to optical frequencies.
+Many of the theoretical analyses of such systems proceed by considering simple two-level systems coupled to a single cavity mode.
+There remains a need to understand light-matter coupling  in contexts of complicated electronic systems coupling to many-mode or continuum photonic systems.
+A promising recent development along these lines is quantum electrodynamical density functional theory, which describes the ground state of a QED system in terms of reduced quantities such as the electronic density and electromagnetic displacement field.
+In this work, complement the set of theoretical tools for describing ground states of light-matter systems by developing a variational theory of the ground state of a quantum electrodynamical system of coupled light and matter.
+Essential to our ansatz is the notion an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling.
 As a first step towards an eventually \textit{ab initio} approach to ground states in QED, we apply our ansatz to a Rabi model involving a multi-level emitter with many optical modes, a model which cannot be analytically solved.
-We find a compact semi-analytic formula which describes the ground state energy very accurately in all regimes of coupling parameters.
+We find a compact semi-analytic formula which describes the ground state energy very accurately (to less than 1\% error) in all regimes of coupling parameters allowed by the Thomas-Reiche-Kuhn sum rule.
 \end{abstract}
@@
 The past decade has brought an explosion of effort and experimental progress in realizing the coupling of matter and quantized electromagnetic fields beyond a regime of weak coupling, in which perturbative treatments can accurately describe the dynamics.
+Ultra-strong, or even deep-strong coupling is now regularly observed in superconducting qubits at microwave frequencies, ensembles of molecules in optical cavities, 2D electron gas cyclotron resonances coupled to cavities, intersubband transitions in quantum wells coupled to cavities, and single-emitters in ultra-confined plasmonic geometries.
+Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators, highly ionized atoms coupled to optical media via the Cerenkov effect, and many more.
+The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems, altering chemical reactivity and electronic transport properties, realizing analogues of nonlinear optical processes with vacuum fluctuations, as new sources of photons through the dynamical Casimir effect, and even as new operating principles for Cerenkov detectors.
+
+In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved.
+In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model, which describes the coupling of a two-level atom to a single mode, the Hopfield model, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators.
+Any realistic system which deviates from these simpler models require numerical diagonalization approaches, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space.
+It also becomes unwieldy when one needs to keep track of many cavity modes, which can be populated with many photons.
+Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling.
+With such density-functional descriptions, workers in this field have been able to compute ground-state and time-dependent properties of realistic molecules in optical cavities.
+\textcolor{blue}{I still need to figure out a segue from here to the next paragraph.}
 
@@
 Finally, they provide a rigorous concept of correlation energy in quantum electrodynamics in a way analogous to electronic structure theory.
-
 The outline of this work is as follows: in section I, we develop the variational ansatz for the ground state of the Hamiltonian of macroscopic quantum electrodynamics.
 In section II, we calculate the ground state energy according to this ansatz for the multi-level and multi-mode Rabi model and compare it to numerical diagonalization of the same problem.
-In section III, we study the accuracy of our ansatz as a function of the correlation energy in the system, and test several approaches to recover the correlation energy.
-We consider both perturbation theory formulated in the interacting modes as well as a self-consistent perturbation theory which allows the perturbative energy shift to influence the modes.
+In Section III, we propose a self-consistent extension of our ansatz.
 
@@
 \begin{equation}
-H = H_{\text{matter}}+\frac{\epsilon_0}{2}\int dz~(E^2+c^2B^2)+\frac{e}{m}A(d)p + \frac{q^2}{2m}A^2(d).
+H = H_{\text{matter}}+\frac{\epsilon_0}{2}\int dz~(E^2+c^2B^2)+\frac{q}{m}A(d)p + \frac{q^2}{2m}A^2(d).
 \end{equation}
@@
 \end{equation}
-which is the Thomas-Reiche-Kuhn sum rule.
+which is the Thomas-Reiche-Kuhn (TRK) sum rule.
 Of course, this sum rule does not rigorously apply to a discrete-level system.
@@
 \end{equation}
-
+\begin{figure*}[t]
+\includegraphics[width=16cm]{fig2.pdf}
+\caption{\textbf{Comparison of the accuracy of the variational ansatz to exact diagonalization of the multi-level and multi-mode Rabi model.} Ratio of approximate ground state energy (calculated via our ansatz and perturbatively) to the energy obtained by numerical diagonalization of the Hamiltonian, truncating the Hilbert space to four photons for (left) a two-level system, (middle) a three-level system, and (right) a four-level system.
+The length parameter which sets the strength of the momentum matrix elements is chosen to saturate the TRK sum rule.}
+\label{fig:decoupling}
+\end{figure*}
 This is an ordinary second-order differential equation with the conditions that $F_n$ is continuous at $d$ and that its derivative is discontinuous according to 
@@
 
+We now look to understand the effect of correlations on our ansatz.
+Interestingly, at lowest-order, correlations are entirely a result of the coupling between the momentum and the vector potential.
+Here, we capture the effect of correlations perturbatively.
+Treating such correlations perturbatively, it follows that the change in the ground state wavefunction $|\delta\rangle$ is given by
+\begin{equation}
+|\delta\rangle = \sum\limits_{b = 2}^{N_a}\sum\limits_{n=1}^{\infty} \frac{q}{m}\sqrt{\frac{1}{2\epsilon_0\omega_n}}\frac{F_n(d)p_{ba}}{\omega_{ab}-\omega_n}|b\rangle|n\rangle.
+\end{equation}
+The corresponding change in the ground state energy, defined as the correlation energy $E_c$, is
+\begin{equation}
+E_{c} = \frac{q^2}{2m^2\epsilon_0\omega_n}\frac{|F_n(d)p_{ba}|^2}{\omega_{ab}-\omega_n}.
+\end{equation}
+Summarizing the result of this section, the quantum electrodynamics interaction energy in the ground state, according to our variational ansatz, should be well-approximated by
+\begin{equation}
+E_{int} = \frac{1}{2}\sum\limits_{n=1}^{\infty}\left(\hbar\omega_n - \hbar\omega_n^{(0)} \right) + \frac{q^2}{2m^2\epsilon_0\omega_n}\frac{|F_n(d)p_{ba}|^2}{\omega_{ab}-\omega_n},
+\end{equation}
+where the eigenmodes $F_n$ and their eigenfrequencies $\omega_n$ are given by Equations (11-13).
+
 \section{Application to a multi-level system coupled to a multi-mode cavity}
 
-%\section{Strategies for recovering the correlation energy}
+In this section, we test the validity of the ansatz proposed and developed in the previous section.
+In particular, we consider an atomic system of varying number of levels (two through four), and a cavity with fifty modes: this system cannot be solved analytically.
+In particular, the Rabi model can only be solved analytically for a two level system interacting with one cavity mode.
 
-\section{Outlook}
+In Figure 2, we consider the ground state energy of the Hamiltonian of Equation (1) calculated by Equation (17) and also calculated by perturbation theory.
+In Figure 2 we take the ratio of these ground state energies to the "exact" ground state energy, computed by numerical diagonalization of the Hamiltonian.
+To do so, we must truncate the Fock space for the photons.
+For the range of couplings considered in Figure 2, up to four photons are needed in the Hilbert space, rendering the dimension of the Hilbert space in the problem to be $6.25\times10^6 \times N_a$ with $N_a$ the number of atomic levels.
+For two, three, and four-level systems, we find that our ansatz captures the ground state energy derived by numerical diagonalization nearly exactly, within an error of less than 1\%.
+On the other hand, perturbation theory can feature an error of over 20\% for the largest coupling considered on this plot.
+The level of accuracy appears to be independent of the number of levels, suggesting that our accuracy is not an artifact of a two-level approximation.
+It is also important that our ansatz does well in a multi-level system, because as pointed out recently, there are severe discrepancies between the Coulomb gauge and length gauges for few level systems, and the slow decay of matrix elements as a function of energy suggests that for few level systems, the length gauge is the appropriate gauge.
+
+Before moving to further analyze the results, we note that we vary the coupling by varying the charge $q$.
+In principle, we could vary the coupling by varying the strength of the momentum matrix elements, which are controlled by $R$.
+However, there is a minimum $R=R_{m}$ that is allowed by the TRK sum rule.
+When $R < R_m$, there is no real-space matter system that our TRK-violating  few-level model can be an approximation of.
+Thus, in all cases, we chose $R=R_m$ as the value, thus taking the effect of correlations to be as large as they can possibly be.
+
+\begin{figure}[t]
+\includegraphics[width=9cm]{fig3.pdf}
+\caption{\textbf{Sources of light-matter decoupling in the QED ground state.} (a) Plots of the mode profile of Equations (11-13) for the lowest mode as a function of position in the cavity for different emitter positions.
+The most striking effect is the screening of the cavity fields in the emitter region, which is most pronounced when the emitter is in the middle of the cavity and interacts strongest with the bare cavity mode.
+(b) Frequencies of the first ten modes as a function of coupling when the emitter is placed in the center of the cavity.
+(c) Normalized change in the coupling constant between matter and electromagnetic fields as a function of position in the cavity and coupling constant, assuming that the emitter is placed in the center of the cavity.
+}
+\label{fig:decoupling}
+\end{figure}
+
+In Figure 3, we explore an interesting feature of Figure 2, which is that the ground-state energy predicted by our ansatz is always substantially lower than predicted by perturbation theory for the coupling parameters considered here.
+To elaborate on this, we refer to Figure 3(a), which shows the normalized lowest frequency electromagnetic mode $F_1(z)$ given by Equations (11-13) for a coupling strength of $0.33\omega_{21}$ and for different positions of the emitter in the cavity.
+The most obvious feature one sees is a strong dip (screening) in the electromagnetic fields where the emitter is.
+An analogous effect, known as a light-matter decoupling effect, was noted in a recent work analytically in the context of expectation values of photodetection signals in the Hopfield model in cavity QED in the limit of deep-strong coupling.
+This dip can be understood physically as a result of the fact that the $A^2$ term in Equation (1) is positive-definite, increasing the energy when the emitter overlaps with strong intensity of the vacuum fluctuations of the electromagnetic field.
+Thus the ground state should screen the electromagnetic field fluctuations out of the region where the emitter is.
+This dip leads to a self-consistent decrease in the magnitude of the interaction energy between the matter and the electromagnetic field, both for the $A^2$ and $Ap$ terms.
+
+Another feature in Figure 3(b) which also leads to a decrease in the magnitude of the interaction relative to perturbation theory is a blue shift of all of the modes relative to their unperturbed values.
+This is consistent with the fields being pushed out of the emitter region, which increases the curvature of the mode functions.
+The effect of this blue-shift to reduce the coupling can be seen from the form of the vector potential of a single photon in mode $n$, which depends on the mode via $\frac{F_n}{\sqrt{\omega_n}}$.
+Thus a blue-shift in the frequencies self-consistently decreases the magnitude of the coupling constant.
+Additionally, blue shifts lead to a further detuning in the energy denominators of Equations (15) and (16), serving to also self-consistently reduce the effect of correlations.
+The combined effects of the field screening and blue shifts are summarized in Figure 3(c), where we plot the normalized change in the self-consistent coupling constant
+\begin{equation}
+\frac{g_n-g_n^{(0)}}{g_n^{(0)}} = \frac{q}{mR}\sqrt{\frac{\hbar}{2\epsilon_0}}\left(\frac{F_n(d)}{F^{(0)}_n(d)}\sqrt{\frac{\omega^{(0)}_n}{\omega_n}}-1\right)
+\end{equation}
+for $n=1$ and $b=2$.
+As can be seen, for bare coupling constants approaching $0.5$, well into the ultra-strong coupling regime, the self-consistent coupling can be reduced by over 80\% as a result of the field screening and blue-shifts.
+
+\section{Extensions}
+Moving forward, there are a few interesting ways to proceed with this ansatz.
+Of course, in the spirit of recent work on ab-initio treatments of quantum electrodynamics, it will be of great interest to apply this ansatz on problems involving realistic molecules ultra strongly coupled to cavities, and other photonic and plasmonic platforms which facilitate the ultra-strong coupling of light and matter.
+To do this, one must take into account the effect of electronic interactions.
+Another interesting direction is to apply this ansatz on a system in which a perturbative treatment of correlations fails to capture their effect accurately.
+One way to remedy this in the framework of our ansatz is to treat the terms in the correlation energy of Equation (16) as variational parameters: this leads to a revised set of equations for the field modes, as well as the matter wavefunctions.
+In particular, one will find for the fields:
+\begin{widetext}
+\begin{equation*}
+\left(\frac{d^2}{dz^2} + \frac{\omega_n^2}{c^2} \right)F_n(z) = \left(\frac{q^2}{4m\epsilon_0 c^2 A} - \sum\limits_{b=2}^{N_a}\frac{1}{2\epsilon_0 A}\left( \frac{q\hbar}{mc}\right)^2\frac{|p_{ba}|^2}{\omega_{ba}+\omega_n}\right)F_n(z)\delta(z-d),
+\end{equation*}
+\end{widetext}
+For the matter, one similarly finds,
+\begin{widetext}
+\begin{equation*}
+\left(H_{\mathrm{matter}} - \frac{1}{2\epsilon_0 A}\left(\frac{q\hbar}{m}\right)^2\sum\limits_{n=1}^{\infty}\sum\limits_{b=2}^{N_a}\frac{F^2_n(L/2)}{\omega_n(\omega_{ba}+\omega_n)}p|\psi_b\rangle\langle \psi_b|p\right)|\psi_a\rangle = \hbar\omega_a|\psi_a\rangle.
+\end{equation*}
+\end{widetext}
+In the case of the model studied in this work, this self-consistent approach is just as accurate as the perturbative approach.
+However, in this case, new fundamental insights are obtained.
+In particular, one sees that the effect of the correlations on the field modes is to counteract the screening.
+This is because for the ground state, the $Ap$ term lowers the energy and this is achieved by pulling the fields into the regions where the matter is.
+
+\section{Summary and Outlook}
+In summary, we have developed a variational approximation to the ground state of quantum electrodynamical problems and have benchmarked it against the multi-mode and multi-level Rabi model, which has no analytical solutions, and whose accurate numerical diagonalization requires a very high-dimensional Hilbert space.
+Despite our ansatz being centered on the assumption of weak correlations, with correlations from the $Ap$ term being treated perturbatively, we find that our approximation does very well in predicting the ground state energy, even when the strength of the correlations saturates the TRK sum rule.
+
+From a more fundamental perspective, the ansatz studied in this work is interesting because they give a rigorous meaning to the notion of correlation energy in quantum electrodynamics which is parallel to correlation energy in electronic structure theory.
+In particular, in the context of electronic structure, the correlation energy is the difference between the true ground state energy and that computed by Hartree-Fock, which is an independent electron ansatz.
+Similarly, our ansatz here is an independent electron-independent photon ansatz whose structure is parallel to the Hartree-Fock ansatz.
+
+Another fundamentally interesting aspect of our ansatz is that unlike current formulations of quantum-electrodynamical density functional theory, it considers the real-space properties of photon modes as they are affected by matter.
+With the advent of ultra-strong coupling and deep-strong coupling being realizable in quantum electrodynamical systems, one may use this real-space knowledge of how matter affect photons to design a photonic mode atom-by-atom.
 
@@
 % Pri, Johannes: Please insert any other relevant acknowledgement information here!
-We thank Joel Yuen-Zhou for useful discussions.
+We thank Prof.
+Joel Yuen-Zhou (University of California San Diego), Prof.
+Ido Kaminer (Technion Israel Institute of Technology), Prof.
+Marin Solja\v{c}i\'{c} (Massachusetts Institute of Technology) and Prof.
+John D.
+Joannopoulos (Massachusetts Institute of Technology), for useful discussions.
 N.
@@
 Paulson School of Engineering and Applied Sciences.
-%The authors thank XXXX for helpful discussions.
-Tentatively Joel Yuen Zhou and Javier Aizpurua.
+We also acknowledge support from the STC Center for Integrated Quantum Materials NSF Grant No.
+DMR-1231319 and from the Harvard John A.
+Paulson School of Engineering and Applied Sciences.
+J.
+F.
+acknowledges financial support from the Deutsche Forschungsgemeinschaft under Contract No.
+FL 997/1-1.
 

### 2018-09-28 paper.tex
first set of comments

@@
 
+\usepackage[normalem]{ulem}
+\newcommand{\Jadd}[1]{\textcolor{blue}{#1}}
+\newcommand{\Jrem}[1]{\textcolor{blue}{\sout{#1}}}
+
 %\usepackage[usenames]{color}
@@
 \author{Nicholas Rivera}\email{nrivera@seas.harvard.edu}\affiliation{\HarvardSEAS}\affiliation{\MITPhy}
-\author{Johannes Flick}\email{flick@g.harvard.edu}\affiliation{\HarvardSEAS}
+\author{Johannes Flick}\email{flick@seas.harvard.edu}\affiliation{\HarvardSEAS}
 \author{Prineha Narang}\email{prineha@seas.harvard.edu}\affiliation{\HarvardSEAS}
@@
 A promising recent development along these lines is quantum electrodynamical density functional theory, which describes the ground state of a QED system in terms of reduced quantities such as the electronic density and electromagnetic displacement field.
-In this work, complement the set of theoretical tools for describing ground states of light-matter systems by developing a variational theory of the ground state of a quantum electrodynamical system of coupled light and matter.
+In this work, \Jadd{we} complement the set of theoretical tools for describing ground states of light-matter systems by developing a variational theory of the ground state of a quantum electrodynamical system of coupled light and matter.
 Essential to our ansatz is the notion an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling.
@@
 
+\Jadd{\section{Introduction}}
+\label{sec:Introduction}
+
 The past decade has brought an explosion of effort and experimental progress in realizing the coupling of matter and quantized electromagnetic fields beyond a regime of weak coupling, in which perturbative treatments can accurately describe the dynamics.
@@
 Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators, highly ionized atoms coupled to optical media via the Cerenkov effect, and many more.
-The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems, altering chemical reactivity and electronic transport properties, realizing analogues of nonlinear optical processes with vacuum fluctuations, as new sources of photons through the dynamical Casimir effect, and even as new operating principles for Cerenkov detectors.
+The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems, altering chemical reactivity~\cite{hutchison2012, thomas2016} and electronic transport properties, realizing analogues of nonlinear optical processes with vacuum fluctuations, as new sources of photons through the dynamical Casimir effect, and even as new operating principles for Cerenkov detectors.
 
 In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved.
-In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model, which describes the coupling of a two-level atom to a single mode, the Hopfield model, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators.
+In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators.
 Any realistic system which deviates from these simpler models require numerical diagonalization approaches, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space.
@@
 In particular, we develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states.
-This ansatz, analogous to the Hartree-Fock ansatz, leads to coupled eigen-equations describing the ground-state of the light-matter system.
+This ansatz, analogous to the Hartree-Fock ansatz~\Jadd{of electronic structure theory}, leads to coupled eigen-equations describing the ground-state of the light-matter system.
 We benchmark our ansatz against a multi-level and multi-mode Rabi model parameterized by the strength of the characteristic momentum matrix elements describing the coupling of the matter and the electromagnetic fields.
@@
 Finally, they provide a rigorous concept of correlation energy in quantum electrodynamics in a way analogous to electronic structure theory.
-The outline of this work is as follows: in section I, we develop the variational ansatz for the ground state of the Hamiltonian of macroscopic quantum electrodynamics.
-In section II, we calculate the ground state energy according to this ansatz for the multi-level and multi-mode Rabi model and compare it to numerical diagonalization of the same problem.
-In Section III, we propose a self-consistent extension of our ansatz.
+The outline of this work is as follows: in section~\ref{sec:variational_ansatz}, we develop the variational ansatz for the ground state of the Hamiltonian of macroscopic quantum electrodynamics.
+In section~\ref{sec:multi-level}, we calculate the ground state energy according to this ansatz for the multi-level and multi-mode Rabi model and compare it to numerical diagonalization of the same problem.
+In Section~\ref{sec:extensions}, we propose a self-consistent extension of our ansatz.
 
+\Jadd{\section{Theory}}
+\label{sec:theory}
+\Jadd{(Nick: Can we write the first part more general in 3d?)}
 In this work, we consider the light-matter Hamiltonian corresponding to an emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction.
@@
 We consider the interaction in the velocity gauge.
-The Hamiltonian can then be written as:
+The Hamiltonian can then be written as: \Jadd{(no spin-magnetic field interaction)}
 \begin{equation}
@@
 The momentum operator, we write in a finite-different representation as
+\Jadd{($\hbar$ missing?)}
 \begin{equation}
@@
 where $R$ is a constant with units of length representing roughly the difference in positions between sites.
-This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the following sum rule is enforced:
+This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the following sum rule is enforced: \Jadd{(Nick: define $p_{ig}$?)}
+
 \begin{equation}
@@
 \section{Variational ansatz}
+\label{sec:variational_ansatz}
 \begin{figure*}[t]
@@
 
-The concept behind the variational ansatz is shown in Figure (1).
+The concept behind the variational ansatz is shown in Figure~(\ref{fig:ansatz}).
 We seek a description of the ground state of the system as $|\Psi\rangle \approx |\tilde{g}\rangle\otimes|\tilde{0}\rangle + |\delta\rangle,$ where a $\sim$ denotes an effective quantity.
@@
 
-We start our analysis by considering a family of ansatze in which $\delta = 0$ (uncorrelated ansatz).  The expectation value of the Hamiltonian is 
+We start our analysis by considering a family of ansatze in which $\delta = 0$ (uncorrelated ansatz).
+The expectation value of the Hamiltonian is \Jadd{(the matter part is assumed to be a single atom, we could use the density here.)}
 \begin{align}
@@
 We impose constraints of matter normalization and photon mode normalization by defining a Lagrange function $\mathcal{L} \equiv \langle \Psi | H | \Psi \rangle - \epsilon(\langle \tilde{g}|\tilde{g}\rangle-1)-\sum\limits_{n=1}^{\infty}\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right)$.
-To find the ground state, we minimize the Lagrange function with respect to the matter orbital $|\tilde{g}\rangle$ and with respect to the mode functions $F_n$.
+\Jadd{(Nick: can we write $\mathcal{L}$ as $\mathcal{L}(\Psi, \Psi^*, F, F^*, \omega_n, \lambda_n)$)}To find the ground state, we minimize the Lagrange function with respect to the matter orbital $|\tilde{g}\rangle$ and with respect to the mode functions $F_n$.
 The minimization with respect to the matter leads to the trivial equation $H_{\text{matter}} |\tilde{g}\rangle = \epsilon|\tilde{g}\rangle$ which leaves the effective matter ground state as simply the ground state of $H_{\text{matter}}$.
@@
 Here, we capture the effect of correlations perturbatively.
-Treating such correlations perturbatively, it follows that the change in the ground state wavefunction $|\delta\rangle$ is given by
+Treating such correlations perturbatively, it follows that the change in the ground state wavefunction $|\delta\rangle$ is given by \Jadd{(add that this is a first-order correction, maybe also add in the spirit of MP2)}
 \begin{equation}
@@
 \section{Application to a multi-level system coupled to a multi-mode cavity}
+\label{sec:multi-level}
 
@@
 
-In Figure 2, we consider the ground state energy of the Hamiltonian of Equation (1) calculated by Equation (17) and also calculated by perturbation theory.
-In Figure 2 we take the ratio of these ground state energies to the "exact" ground state energy, computed by numerical diagonalization of the Hamiltonian.
+In Figure~\ref{fig:decoupling}, we consider the ground state energy of the Hamiltonian of Equation (1) calculated by Equation (17) and also calculated by perturbation theory \Jadd{(briefly define what is meant by this, to distinguish with our approach.
+maybe non SCF-perturbation theory)}.
+In Figure~(\ref{fig:decoupling}) we take the ratio of these ground state energies to the "exact" ground state energy, computed by numerical diagonalization of the Hamiltonian.
 To do so, we must truncate the Fock space for the photons.
-For the range of couplings considered in Figure 2, up to four photons are needed in the Hilbert space, rendering the dimension of the Hilbert space in the problem to be $6.25\times10^6 \times N_a$ with $N_a$ the number of atomic levels.
+For the range of couplings considered in Figure~\ref{fig:decoupling}, up to four photons are needed in the Hilbert space, rendering the dimension of the Hilbert space in the problem to be $6.25\times10^6 \times N_a$ with $N_a$ the number of atomic levels.
 For two, three, and four-level systems, we find that our ansatz captures the ground state energy derived by numerical diagonalization nearly exactly, within an error of less than 1\%.
@@
 The level of accuracy appears to be independent of the number of levels, suggesting that our accuracy is not an artifact of a two-level approximation.
-It is also important that our ansatz does well in a multi-level system, because as pointed out recently, there are severe discrepancies between the Coulomb gauge and length gauges for few level systems, and the slow decay of matrix elements as a function of energy suggests that for few level systems, the length gauge is the appropriate gauge.
+It is also important that our ansatz does well in a multi-level system, because as pointed out recently, there are severe discrepancies between the Coulomb gauge and length gauges for few level systems~\cite{schaefer2018, bernardis2018}, and the slow decay of matrix elements as a function of energy suggests that for few level systems, the length gauge is the appropriate gauge.
 
@@
 }
-\label{fig:decoupling}
+\label{fig:sources_decoupling}
 \end{figure}
 
-In Figure 3, we explore an interesting feature of Figure 2, which is that the ground-state energy predicted by our ansatz is always substantially lower than predicted by perturbation theory for the coupling parameters considered here.
-To elaborate on this, we refer to Figure 3(a), which shows the normalized lowest frequency electromagnetic mode $F_1(z)$ given by Equations (11-13) for a coupling strength of $0.33\omega_{21}$ and for different positions of the emitter in the cavity.
+In Figure~\ref{fig:sources_decoupling}, we explore an interesting feature of Figure ~\ref{fig:decoupling}, which is that the ground-state energy predicted by our ansatz is always substantially lower than predicted by perturbation theory for the coupling parameters considered here.
+To elaborate on this, we refer to Figure ~\ref{fig:sources_decoupling}(a), which shows the normalized lowest frequency electromagnetic mode $F_1(z)$ given by Equations (11-13) for a coupling strength of $0.33\omega_{21}$ and for different positions of the emitter in the cavity.
 The most obvious feature one sees is a strong dip (screening) in the electromagnetic fields where the emitter is.
-An analogous effect, known as a light-matter decoupling effect, was noted in a recent work analytically in the context of expectation values of photodetection signals in the Hopfield model in cavity QED in the limit of deep-strong coupling.
+An analogous effect, known as a light-matter decoupling effect~\cite{liberato2014}, was noted in a recent work analytically in the context of expectation values of photodetection signals in the Hopfield model in cavity QED in the limit of deep-strong coupling.
 This dip can be understood physically as a result of the fact that the $A^2$ term in Equation (1) is positive-definite, increasing the energy when the emitter overlaps with strong intensity of the vacuum fluctuations of the electromagnetic field.
@@
 
-Another feature in Figure 3(b) which also leads to a decrease in the magnitude of the interaction relative to perturbation theory is a blue shift of all of the modes relative to their unperturbed values.
+Another feature in Figure~\ref{fig:sources_decoupling}(b) which also leads to a decrease in the magnitude of the interaction relative to perturbation theory is a blue shift of all of the modes relative to their unperturbed values.
 This is consistent with the fields being pushed out of the emitter region, which increases the curvature of the mode functions.
@@
 Additionally, blue shifts lead to a further detuning in the energy denominators of Equations (15) and (16), serving to also self-consistently reduce the effect of correlations.
-The combined effects of the field screening and blue shifts are summarized in Figure 3(c), where we plot the normalized change in the self-consistent coupling constant
+The combined effects of the field screening and blue shifts are summarized in Figure ~\ref{fig:sources_decoupling}(c), where we plot the normalized change in the self-consistent coupling constant
 \begin{equation}
@@
 
+\Jadd{(Nick, do you solve it actually in a scf cycle? if so, we could also show how the energy converges per iteration)}
+
 \section{Extensions}
+\label{sec:extensions}
 Moving forward, there are a few interesting ways to proceed with this ansatz.

### 2018-10-02 supp.tex
pri, edit away. also notice there's a supp file now

\documentclass[aps,prb,onecolumn,
	groupedaddress,superscriptaddress,
	amsfonts,amssymb,amsmath,floatfix,
	citeautoscript]{revtex4-1}

\usepackage{graphicx}
\usepackage[centering,hmargin=20mm,tmargin=30mm,bmargin=25mm]{geometry}
\usepackage{multirow}
\usepackage{newtxtext}
\usepackage[cmintegrals]{newtxmath}

%----- References -----
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks,
	linkcolor={blue!75!black!80!yellow},
	citecolor={blue!75!black!80!yellow},
	urlcolor={blue!75!black!80!yellow}
}

%----- Captions in sans font -----
\makeatletter
\renewcommand\@make@capt@title[2]{%
	\@ifx@empty\float@link{\@firstofone}{\expandafter\href\expandafter{\float@link}}%
	\sffamily{\textbf{#1}}\@caption@fignum@sep#2
}%
\renewcommand\figurename{Fig.}
\makeatother

\thickmuskip=5mu plus 2mu minus 1mu  %binary relations (default, 5mu plus 5mu)
\medmuskip=4mu plus 2mu minus 2mu    %binary operations (default, 4mu plus 2mu minus 4mu)

\frenchspacing %Ensure that revTeX does not do "double spaces" after punctuation

\renewcommand{\Im}{\operatorname{Im}}
\renewcommand{\Re}{\operatorname{Re}}
\newcommand{\sub}[1]{\ensuremath{_{\textrm{#1}}}} %Upright multi-character subscript
\newcommand{\super}[1]{\ensuremath{^{\textrm{#1}}}} %Upright multi-character superscript

\newcommand{\HarvardSEAS}{John A. Paulson School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, USA}
\newcommand{\MITPhy}{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA, USA}

\usepackage[normalem]{ulem}
\newcommand{\Jadd}[1]{\textcolor{blue}{#1}}
\newcommand{\Jrem}[1]{\textcolor{blue}{\sout{#1}}}

%\usepackage[usenames]{color}
%\newcommand{\edited}[1]{{\color{red} #1}}

\begin{document}

\title{Supplementary information for: Variational theory of the ground state of quantum electrodynamics}

\author{Nicholas Rivera}\email{nrivera@seas.harvard.edu}\affiliation{\HarvardSEAS}\affiliation{\MITPhy}
\author{Johannes Flick}\email{flick@seas.harvard.edu}\affiliation{\HarvardSEAS}
\author{Prineha Narang}\email{prineha@seas.harvard.edu}\affiliation{\HarvardSEAS}

\date{\today}

\maketitle
In this supplement, we derive a self-consistent set of equations describing the ground state of a general quantum electrodynamical system with the effect of correlations included. We derive them in the case of a multi-electron system with electron-electron interactions in three spatial dimensions with an arbitrary number of photonic modes. The resulting equations are very similar in spirit to the equations of quantum electrodynamical density functional theory in an optimized-effective potential scheme, except that the orbitals used in that scheme are Kohn-Sham orbitals, while ours essentially are mean-field theory orbitals. The one-dimensional problem studied for concreteness in the main text is a special case of these equations for a single-electron emitter in a finite-dimensional space and in a one-dimensional cavity.
\section{Equations for the ground state of the coupled light-matter system}

The QED Hamiltonian takes the form:
\begin{equation}
H = H_{mat}+H_{em}+H_{int},
\end{equation}
where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter. The matter Hamiltonian takes the form:

%\begin{align}
%H_{el} = &\int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext} \right)\psi(x) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
%\end{align}
\begin{equation}
H_{el} = \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext} \right)\psi(x) + \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
\end{equation}
where $v_{ext}$ is the one-body potential, $V(x-x')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.

Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$. This renders the free electromagnetic Hamiltonian as
\begin{equation}
H_{em} = \int d^3x~ \epsilon (\partial_t \mathbf{A}(x))^2 + \mathbf{A}(x)\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(x)),
\end{equation}
where $\epsilon$ and $\mu$ represent a non-dispersive and positive dielectric and magnetic background that the matter and photon occupy. For cases we consider in this work, these will be taken to be unity, but we leave them here for the sake of generality.

The interaction Hamiltonian takes the form:
%\begin{align}
%H_{int} = &\frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(x) + \nonumber \\ &\frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(x).
%\end{align}
\begin{equation}
H_{int} = \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(x) + \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(x).
\end{equation}

This Hamiltonian, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions $\{\psi_n\}$, and in terms of a set of photonic mode functions $\{\mathbf{F}_i\}$ via mode expansions for the electronic and photonic field operators. The electron field operator takes the form:
\begin{equation}
\psi(x) = \sum_n \psi_n(x)c_n.
\end{equation}
In this second-quantized description, the $c_n$ is an annihilation operator for an electron corresponding to state $n$. The electromagnetic field operator takes the form
\begin{equation}
\mathbf{A}(x) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(x)a_i+\mathbf{F}^*_i(x)a^{\dagger}_i\right).
\end{equation}
In the electromagnetic field operator, we parameterize not only by the mode functions but also by modal frequencies. We have chosen a form for the vector potential operator which mirrors the choice generally adopted in QED. The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}.$

With the Hamiltonian set up, we now move to describe a variational theory of the ground state. In the variational theorem, we choose some ansatz $|\Omega\rangle$ for the ground state of $H$. The variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy. Parameterizing the ground state to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy for the chosen family of ground states.

We choose as our ground state
\begin{equation}
|\Omega\rangle = |\psi\rangle \otimes |\phi\rangle,
\end{equation}
where
\begin{equation}
|\psi\rangle = \prod\limits_n c_n^{\dagger}|0_n\rangle,
\end{equation}
and where
\begin{equation}
|\phi\rangle = \bigotimes_i|0_i\rangle.
\end{equation}
In such an ansatz $|\psi\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons. Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions. Thus, we shall minimize the expectational value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$. Note that the electromagnetic field gets two sets of (distinct) variational parameters, $\omega_i$ and $\mathbf{F}_i$. Intuitively, this happens because the electromagnetic field is described by an electric and magnetic field, which are related to $\mathbf{F}_i$ and $\partial_t \mathbf{F}_i \rightarrow \omega_i\mathbf{F}_i$.

The last thing to note before taking the expectation values and minimization is that we should focus on the submanifold of variations that keep the complete set of electron and photon functions normalized. This shall be enforced by constructing the Lagrange function:
\begin{widetext}
\begin{equation}
\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle - \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right).
\end{equation}
\end{widetext}

\subsection{System of equations for the ground state}
We now proceed with the program of evaluating the expectation value and carrying out the functional minimization with respect to all of the free parameters.

The minimization with respect to the electron orbital $\psi_i^*$ yields the equation:
% \begin{widetext}
% \begin{equation}
% \left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) + \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)  = E_i\psi_i(x).
% \end{equation}
% \end{widetext}
\begin{align}
\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) &+  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)  = E_i\psi_i(x).
\end{align}
This differs from the usual Hartree-Fock equation of electronic theory by the last term on the left-hand-side, representing a polarization induced by the $\mathbf{A}^2$ term.

The minimization with respect to the photonic mode function $\mathbf{F}_n^*$ gives:
\begin{equation}
\left(\epsilon\omega_n+\frac{e^2\sum_i|\psi_i(x)|^2}{m\epsilon_0\omega_n} \right)\mathbf{F}_n + \frac{c^2}{\omega_n}\nabla\times\mu^{-1}\nabla\times\mathbf{F}_n = 2\lambda_n\omega_n\mathbf{F}_n.
\end{equation}
Meanwhile, the minimization with respect to frequency $\omega_n$ gives:
\begin{equation}
\int d^3x\left(\epsilon - \frac{e^2\sum_i|\psi_i(x)|^2}{m\epsilon_0\omega^2_n}\right)|\mathbf{F}_n|^2 - \frac{c^2}{\omega_n^2}\mathbf{F}_n^*\cdot\nabla\times\mu^{-1}\nabla\times\mathbf{F}_n=0.
\end{equation}
Equations (12) and (13), to be mutually consistent, force $\lambda_n = \omega_n$, reducing Equation (12) to the equivalent form:
\begin{equation}
\nabla\times\mu^{-1}\nabla\times\mathbf{F}_n = \left(\epsilon - \frac{\omega_p^2(x)}{\omega_n^2} \right)\mathbf{F}_n,
\end{equation}
where $\omega_p \equiv \frac{e^2}{2m\epsilon_0}\sum_n|\psi_n(x)|^2$ is the position-dependent plasma frequency. Notice that taking the divergence of both sides of Equation (14) requires that
\begin{equation}
\nabla\cdot\left(\epsilon - \frac{\omega_p^2(x)}{\omega_n^2} \right)\mathbf{F}_n = 0,
\end{equation}
which is a statement of Gauss' law for the electric field.

\section{Lamb shift correction to the equations for the quantum electrodynamical ground state}

In the derivation of Equation (11), it is notable that the term linear in the vector potential vanishes. In many quantum electrodynamical problems, this linear term is important. To first order, it leads to spontaneous emission of a photon. To second order, it leads to van der Waals / Casimir-Polder forces on emitters, as well as effective interactions betweeen distinct emitters. Thus, we seek to capture the effect of this term. Physically, this term will mix the factorizable ground state of Equation (7) with terms that have virtual excitations of the matter, as well as virtual excitations of the electromagnetic field. The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.

We capture the effect of correlations perturbatively. In other words, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential. That correction is given by
\begin{equation}
\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{q=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_q^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_q(\omega_{mn} -\lambda_q)},
\end{equation}
where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$,$\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.  In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (11) and (14), and then all the information necessary to calculate the correlation energy is present. In what follows, we add this energy correction $\delta E$ to the expectation value of the energy in the ground state, with the orbitals and eigenvalues as variational parameters. As a result, the orbitals and eigenvalues obtained will be different from Equation (11) and (14), and this difference will be small in the limit of weak correlations. Strictly speaking, this approach is only justified for weak correlations, but can be applied to systems with strong-correlations as is often done with self-consistent methods.

\subsection{Physics contained in these equations}

As the equations which will result from this new expectation value of the energy will be quite complicated, it is useful to discuss what physics should contained by introducing $\delta E$. From the standpoint of the electrons, the equation for them should differ from Equation (11) by a term which is proportional to the Casimir-Polder, or van der Waals, potential that the electrons feel.  These potentials can also be thought of as Lamb shifts due to the QED coupling. It is well-known that this potential depends on all occupied and unoccupied electron orbitals and the photon orbitals, as well as their respective eigenvalues. In the equation we will derive for the electron orbitals and their energies, there should be a new term relative to Equation (11) which is simply this potential but with self-consistent electron and photon orbitals and energies.

For the photon, the equation will differ from Equation (14) by the introduction of a new term which has the appearance of a source term proportional to a sum over the transition current densities of the electronic system, $\mathbf{j}_{nm}$. The weight of these transition current densities will be proportional to the coupling between the current densities and the photonic orbital. In other words, this new term expresses a coupling of fields to fluctuating currents associated with the ground state of the QED system.

In the next section, we derive the equations implied by the correction to the energy of Equation (16) and confirm the physical understanding presented in the previous two paragraphs.

\subsection{Equations for the ground state of quantum electrodynamics}

The derivative of $\delta E$ with respect to an occupied electron orbital $k$ is given by
\begin{equation}
\frac{\partial\delta E}{\partial \psi_k^*} = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{nk}(\mathbf{y})}{\omega_q(\omega_{kn}-\lambda_q)}\int d^3x~\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)
\end{equation}
A similar equation arises for the derivative with respect to an unoccupied electron orbital, except that the summation should now be over unoccupied electron orbitals.

The derivative of $\delta E$ with respect to a photonic orbital $q$ is given by
\begin{equation}
\frac{\partial\delta E}{\partial \mathbf{F_q^*}}=\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot \mathbf{j}_{mn}(\mathbf{y})}{\omega_q(\omega_{mn}-\lambda_q)}\mathbf{j}_{nm}.
\end{equation}

The derivative of $\delta E$ with respect to a photonic frequency $\omega_q$ is given by
\begin{equation}
-\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\Big|\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})\Big|^2}{\omega_q^2(\omega_{mn}-\lambda_q)}
\end{equation}

Using these derivatives, Equation (11) now becomes:
\begin{align}
&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(x).
\end{align}

Meanwhile, the equation derived by setting to zero the derivative with respect to the photonic orbital now becomes:
\begin{equation}
\frac{\hbar}{4}\left(\omega_q\mathbf{F}_q + \frac{c^2}{\omega_q}\nabla\times\nabla\times\mathbf{F}_q\right) + \frac{\hbar\sum\limits_{m=1}^{N_{\sigma}}|\psi_m|^2}{4m\epsilon_0\omega_q}+\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{\infty}\frac{\int d^3y ~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_q(\omega_{mn}-\lambda_q)}\mathbf{j}_{nm} = \frac{\hbar\lambda_q}{2}\mathbf{F}_q
\end{equation}

Setting the derivative with respect to $\omega_q$ to zero gives the equation:
\begin{equation}
\frac{\hbar}{4}\left(\int d^3x ~|\mathbf{F}_q|^2 - \frac{c^2}{\omega_q^2}\mathbf{F}^*_q\cdot\nabla\times\nabla\times\mathbf{F}_q \right) - \frac{e^2\hbar}{4m\epsilon_0\omega_q^2}\int d^3x~\sum\limits_{m=1}^{N_{\sigma}} |\psi_m|^2 -\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\Big|\int d^3x~\mathbf{F}_q\cdot\mathbf{j}_{mn}\Big|^2}{\omega_q^2(\omega_{mn}-\lambda_q)}=  0 
\end{equation}
Performing the operation $\frac{1}{\omega_q}\int d^3x \mathbf{F}_q^* \cdot$ on Equation (21), and adding the result to Equation (22), gives the result $\lambda_q = \omega_q$, just as before $\delta E$ was introduced. Using this key simplification, along with the definition of the spatially-dependent plasma frequency as $\omega_p^2 = \frac{e^2}{m\epsilon_0}\sum\limits_{m=1}^{N_p}$, we have that
\begin{equation}
\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_q^2} \right)\right)\mathbf{F}_q = -\frac{e^2\hbar}{2m^2\epsilon_0c^2}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_{mn}-\omega_{q}}\mathbf{j}_{nm}.
\end{equation}
Equations (20) and (23), represent main results of this work.

\bibliographystyle{apsrev4-1}
\bibliography{references}

\end{document}

### 2018-10-04 paper.tex
some more comments

@@
 A promising recent development along these lines is quantum electrodynamical density functional theory, which describes the ground state of a QED system in terms of reduced quantities such as the electronic density and electromagnetic displacement field.
-In this work, \Jadd{we} complement the set of theoretical tools for describing ground states of light-matter systems by developing a variational theory of the ground state of a quantum electrodynamical system of coupled light and matter.
+In this work, we complement the set of theoretical tools for describing ground states of light-matter systems by developing a variational theory of the ground state of a quantum electrodynamical system of coupled light and matter.
 Essential to our ansatz is the notion an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling.
@@
 
-\Jadd{\section{Introduction}}
+\section{Introduction}
 \label{sec:Introduction}
 
-The past decade has brought an explosion of effort and experimental progress in realizing the coupling of matter and quantized electromagnetic fields beyond a regime of weak coupling, in which perturbative treatments can accurately describe the dynamics.
-Ultra-strong, or even deep-strong coupling is now regularly observed in superconducting qubits at microwave frequencies, ensembles of molecules in optical cavities, 2D electron gas cyclotron resonances coupled to cavities, intersubband transitions in quantum wells coupled to cavities, and single-emitters in ultra-confined plasmonic geometries.
-Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators, highly ionized atoms coupled to optical media via the Cerenkov effect, and many more.
-The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems, altering chemical reactivity~\cite{hutchison2012, thomas2016} and electronic transport properties, realizing analogues of nonlinear optical processes with vacuum fluctuations, as new sources of photons through the dynamical Casimir effect, and even as new operating principles for Cerenkov detectors.
+The past decade has brought an explosion of effort and experimental progress in realizing the coupling of matter and quantized electromagnetic fields beyond a regime of weak coupling, in which perturbative treatments of quantum electrodynamical couplings can no longer accurately describe the dynamics \cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,kockum2018ultrastrong}.
+Ultra-strong, or even deep-strong coupling is now regularly observed in superconducting qubits at microwave frequencies \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, ensembles of molecules in optical cavities \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, 2D electron gas cyclotron resonances coupled to cavities \cite{hagenmuller2010ultrastrong,scalari2012ultrastrong,zhang2016collective}, intersubband transitions in quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and single-emitters in ultra-confined plasmonic geometries \cite{benz2016,chikkaraddy2016}.
+Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, highly ionized atoms coupled to optical media via the Cerenkov effect, and many more.
+The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems \cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties \cite{orgiu2015}, realizing analogues of nonlinear optical processes with vacuum fluctuations \cite{kockum2017deterministic}, and even as new sources of photons through the dynamical Casimir effect \cite{ciuti2005quantum}.
 
 In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved.
-In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators.
+In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model \cite{hopfield1958theory}, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators \cite{dicke1954coherence}.
 Any realistic system which deviates from these simpler models require numerical diagonalization approaches, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space.
-It also becomes unwieldy when one needs to keep track of many cavity modes, which can be populated with many photons.
-Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling.
-With such density-functional descriptions, workers in this field have been able to compute ground-state and time-dependent properties of realistic molecules in optical cavities.
-\textcolor{blue}{I still need to figure out a segue from here to the next paragraph.}
+It also becomes unwieldy when one needs to keep track of \Jrem{many}\Jadd{multi} cavity modes, which can be populated with many photons.
+Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}.
+With such density-functional descriptions, workers in this field have been able to compute ground-state and time-dependent properties of realistic molecules in optical cavities \cite{flick2017c}.
 
-In what follows, we develop a variational ansatz for the ground state of quantum electrodynamical Hamiltonians.
+In what follows, we propose the use \Jadd{of} variational methods for the solution of \emph{ab initio} QED problems.
+In particular, here we develop a particular class of mean-field variational ansatzes for the ground state of quantum electrodynamical Hamiltonians.
 In particular, we develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states.
-This ansatz, analogous to the Hartree-Fock ansatz~\Jadd{of electronic structure theory}, leads to coupled eigen-equations describing the ground-state of the light-matter system.
+This ansatz, \Jrem{analogous} \Jadd{reminiscent} to the Hartree-Fock ansatz of electronic structure theory, leads to coupled eigen-equations describing the ground-state of the light-matter system.
 We benchmark our ansatz against a multi-level and multi-mode Rabi model parameterized by the strength of the characteristic momentum matrix elements describing the coupling of the matter and the electromagnetic fields.
-We find that when the momentum matrix elements are sufficiently weak, our theory finds the ground state energy to a remarkable accuracy of less than >99.9\%, even in the non-perturbative coupling regime.
-Even in regimes where the correlation energy is 50\% of the total energy, far outside of the validity domain of our ansatz, our ansatz predicts the energy up to a few percent.
+We find that when the momentum matrix elements are sufficiently weak, our theory finds the ground state energy to a remarkable accuracy of \Jrem{less}\Jadd{more?} than >99.9\%, even in the non-perturbative coupling regime.
+Even in regimes where the correlation energy is 50\% of the total energy, far outside of the validity domain of our ansatz, our ansatz predicts the energy up to a few percent \Jadd{due to self-consistency}.
 In regimes where our results are accurate, we claim we have found the effective quasiparticle description of the ground state.
 Moreover, in this regime, the ground state interaction energy manifests itself as the difference in zero-point energies between the quasiparticles and the bare particles.
-This generalizes the notion of the Casimir energy to a single atom in a cavity.
+This generalizes the notion of the Casimir energy \Jadd{(cite)} to a single atom in a cavity.
 
@@
 Second, they provide a general framework for understanding light-matter decoupling effects in the regime of strong light-matter interactions.
-Third, they prove a non-perturbative theory of Casimir-Polder forces.
+Third, they prove a non-perturbative theory of Casimir-Polder forces \Jadd{(cite)}.
 Finally, they provide a rigorous concept of correlation energy in quantum electrodynamics in a way analogous to electronic structure theory.
+
 The outline of this work is as follows: in section~\ref{sec:variational_ansatz}, we develop the variational ansatz for the ground state of the Hamiltonian of macroscopic quantum electrodynamics.
@@
 
-\Jadd{\section{Theory}}
+\section{Theory}
 \label{sec:theory}
-\Jadd{(Nick: Can we write the first part more general in 3d?)}
-In this work, we consider the light-matter Hamiltonian corresponding to an emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction.
+In the Supplementary Information (SI), we derive a set of nonlinear coupled eigenequations which can be applied to describe the ground state of any \Jadd{nonrelativistic} macroscopic quantum electrodynamical system, in three dimensions, with electron-electron interactions, and with arbitrarily many photon modes.
+These are real-space equations that give single-particle electron and photon orbitals \Jadd{(should we call them orbitals?)} in the presence of QED coupling, thus giving real-space information about the electron density and the electromagnetic field fluctuations in the ground state.
+The equations derived are in general fairly complicated, and thus in the main text, we analyze a much simpler case of the equations in order to extract the basic physics of the equations.
+
+In the main text, we consider the light-matter Hamiltonian corresponding to a single emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction.
 As the cavity is considered for simplicity to be one-dimensional, the electric field is oriented along a single direction, denoted $x$, while the magnetic field is oriented along a direction transverse to both the electric field and the cavity length, denoted $y$.
-We consider the interaction in the velocity gauge.
-The Hamiltonian can then be written as: \Jadd{(no spin-magnetic field interaction)}
-\begin{equation}
+We consider the interaction in the velocity gauge, and neglect the weak interaction between spins and magnetic fields, thus neglecting spin altogether.
+The Hamiltonian can then be written as:
+\begin{equation}\label{eq:hamiltonian}
 H = H_{\text{matter}}+\frac{\epsilon_0}{2}\int dz~(E^2+c^2B^2)+\frac{q}{m}A(d)p + \frac{q^2}{2m}A^2(d).
 \end{equation}
+\Jadd{(this is dipole approximation, or dipole approximation is exact for point-like single atom? )}
 The fields can be expressed as a mode expansion via:
-\begin{equation}
+\begin{equation}\label{eq:mode_expansion}
 A(z,t) = \sum\limits_{n=1}^{\infty} \sqrt{\frac{\hbar}{2\epsilon_0\omega_n }}(F_n(z)a_ne^{-i\omega_n t}+F^*_n(z)a_n^{\dagger}e^{i\omega_n t}),
@@
 Thus we parameterize the general family of matter Hamiltonians as:
-\begin{equation}
+\begin{equation}\label{eq:matter_hamiltonian}
 H_{\text{matter}} = \sum\limits_{i=1}^{{N_a-1}} V_i|i\rangle\langle i|+t(|i\rangle\langle i+1|+|i+1\rangle\langle i|) 
@@
 The momentum operator, we write in a finite-different representation as
-\Jadd{($\hbar$ missing?)}
-\begin{equation}
-p = \frac{i}{R}\sum\limits_{i=1}^{N_a-1} \left(|i\rangle\langle i+1|-|i+1\rangle\langle i| \right),
+\begin{equation}\label{eq:momentum_operator}
+p = \frac{-i\hbar}{R}\sum\limits_{i=1}^{N_a-1} \left(|i\rangle\langle i+1|-|i+1\rangle\langle i| \right),
 \end{equation}
 where $R$ is a constant with units of length representing roughly the difference in positions between sites.
-This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the following sum rule is enforced: \Jadd{(Nick: define $p_{ig}$?)}
-
-\begin{equation}
+This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the following sum rule is enforced:
+\begin{equation}\label{eq:TRK_sum_rule}
 \frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1,
 \end{equation}
-which is the Thomas-Reiche-Kuhn (TRK) sum rule.
-Of course, this sum rule does not rigorously apply to a discrete-level system.
+where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.
+Equation \ref{eq:TRK_sum_rule} is the Thomas-Reiche-Kuhn (TRK) sum rule \Jadd{(cite)}.
+\Jrem{Of course} \Jadd{Since the TRK is based on a full electronic real-space description}, this sum rule does not rigorously apply to a discrete-level system.
 However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot.
-As we shall show later in the text however, it does place a bound on how strong the effect of the $Ap$ term can be on the total energy.
+As we shall show later in the text however, it does place a bound on how strong the effect of the $A\Jadd{\cdot}p$ term can be on the total energy.
 The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.
@@
 We start our analysis by considering a family of ansatze in which $\delta = 0$ (uncorrelated ansatz).
-The expectation value of the Hamiltonian is \Jadd{(the matter part is assumed to be a single atom, we could use the density here.)}
-\begin{align}
+The expectation value of the Hamiltonian is %\Jadd{(the matter part is assumed to be a single atom, we could use the density here.)}
+\begin{align}\label{eq:expectation_val1}
 \langle \Psi | H | \Psi \rangle &= \langle \tilde{g} |H_{\text{matter}} | \tilde{g}\rangle \nonumber \\ &+ \frac{\hbar}{4}\int dz ~\sum\limits_{n=1}^{\infty}\left(\omega_n|F_n|^2 - \frac{c^2}{\omega_n}F_n^*\partial_z^2F_n\right) \nonumber \\ &+ \frac{\hbar q^2}{4m\epsilon_0\omega_n}\sum\limits_{n=1}^{\infty} \int dz~\delta(z-d)|F_n|^2
@@
 Notably, in this ansatz, the term in the Hamiltonian coupling the momentum of the matter to the photon makes no contribution.
-We impose constraints of matter normalization and photon mode normalization by defining a Lagrange function $\mathcal{L} \equiv \langle \Psi | H | \Psi \rangle - \epsilon(\langle \tilde{g}|\tilde{g}\rangle-1)-\sum\limits_{n=1}^{\infty}\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right)$.
-\Jadd{(Nick: can we write $\mathcal{L}$ as $\mathcal{L}(\Psi, \Psi^*, F, F^*, \omega_n, \lambda_n)$)}To find the ground state, we minimize the Lagrange function with respect to the matter orbital $|\tilde{g}\rangle$ and with respect to the mode functions $F_n$.
+We impose constraints of matter normalization and photon mode normalization by defining a Lagrange function
+\begin{align}\label{eq:lagrange}
+&\mathcal{L}(\Psi, \Psi^*, \{F_n, F_n^*, \omega_n, \lambda_n\}) \equiv \langle \Psi | H | \Psi \rangle \nonumber \\ &- \epsilon(\langle \tilde{g}|\tilde{g}\rangle-1)-\sum\limits_{n=1}^{\infty}\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right).
+\end{align} 
+To find the ground state, we minimize the Lagrange function with respect to the matter orbital $|\tilde{g}\rangle$ and with respect to the mode functions $F_n$.
 The minimization with respect to the matter leads to the trivial equation $H_{\text{matter}} |\tilde{g}\rangle = \epsilon|\tilde{g}\rangle$ which leaves the effective matter ground state as simply the ground state of $H_{\text{matter}}$.
 On the other hand, the minimization with respect to the photon mode functions leads to  the equation
-\begin{equation}
+\begin{equation}\label{eq:maxwell}
 \left(\partial_z^2-\frac{\omega^2_n}{c^2}+2\frac{\omega_n\lambda_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
 \end{equation}
-We may constrain $\lambda$ by differentiating the Lagrange function also with respect to the $\omega_n$. The equation which follows is:
-\begin{equation}
+We may constrain the $\lambda_n$ by differentiating the Lagrange function also with respect to the $\omega_n$. The equation which follows is:
+\begin{equation}\label{eq:other_maxwell}
 \int dz ~\left(|F_n|^2 + \frac{c^2}{\omega^2_n}F_n^*\partial_z^2F_n\right) - \frac{ q^2}{m\epsilon_0\omega^2_n} \int dz~\delta(z-d)|F_n|^2 = 0
 \end{equation}
-Performing $\frac{\omega_n^2}{c^2}\int dz~F_n^*$ on both sides of Equation (5), and adding this equation to Equation (6), one immediately finds that $\lambda_n = \omega_n$ and that
-\begin{equation}
+Performing $\frac{\omega_n^2}{c^2}\int dz~F_n^*$ on both sides of Equation \ref{eq:maxwell}, and adding this equation to Equation \ref{eq:other_maxwell}, one immediately finds that $\lambda_n = \omega_n$ and that
+\begin{equation}\label{eq:final_maxwell}
 \left(\partial_z^2+\frac{\omega^2_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
@@
 This is an ordinary second-order differential equation with the conditions that $F_n$ is continuous at $d$ and that its derivative is discontinuous according to 
-\begin{equation}
+\begin{equation}\label{eq:boundary_condition}
 \partial_zF_n\Big|_{z=d^+}-\partial_zF_n\Big|_{z=d^-} = \frac{q^2}{m\epsilon_0 c^2}F_n(d),
 \end{equation}
-in addition to the usual condition of the modes vanishing at the cavity walls $z=0$ and $z=L$. It can be shown that the solution to Equation (7) satisfying such boundary conditions is:
-\begin{align}
+in addition to the usual condition of the modes vanishing at the cavity walls $z=0$ and $z=L$.
+It can be shown that the solution to Equation \ref{eq:final_maxwell} satisfying such boundary conditions is:
+\begin{align}\label{eq:field_mode}
 &\theta (z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}-\right) \nonumber \\ 
@@
 provided that the auxiliary condition
-\begin{equation}
+\begin{equation}\label{eq:mode_transcendental}
 \cot\left(\frac{\omega_n}{c}d \right)+\cot\left(\frac{\omega_n}{c}(L-d) \right) = \frac{q^2}{4m\epsilon_0\omega_nc}
 \end{equation}
-is met. To ensure that the modes are normalized according to the constraint, we have that the solutions of Equation (9) must be multiplied by a normalization factor $N_n$ given by
-\begin{equation}
+is met. To ensure that the modes are normalized according to the constraint, we have that the solutions of Equation \ref{eq:field_mode} must be multiplied by a normalization factor $N_n$ given by
+\begin{equation}\label{eq:mode_normalization}
 N_n = 2\sqrt{\frac{c}{\omega_n\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
 \end{equation}
-The condition of Equation (9) determines the resonance frequencies of the photon quasiparticle modes.
-After calculating the modes and the frequencies, we may move to an evaluation of the energy of the ground state.
-We proceed by taking the solutions of Equation (9), normalized with the normalization factor of Equation (11), and plugging it back into the expectation value of the Hamiltonian of the ground state (Equation (4)).
-After plugging in the quasiparticle photon modes into Equation (4), we find that
-\begin{equation}
+The condition of Equation \ref{eq:mode_transcendental} determines the resonance frequencies of the photon quasiparticle modes.
+\Jadd{(put eq.12-14 to SI?)} After calculating the modes and the frequencies, we may move to an evaluation of the energy of the ground state.
+We proceed by taking the solutions of Equation \ref{eq:field_mode}, normalized with the normalization factor of Equation \ref{eq:mode_normalization}, and plugging it back into the expectation value of the Hamiltonian of the ground state (Equation \ref{eq:expectation_val1}).
+We find that
+\begin{equation}\label{eq:casimir}
 \langle \Psi | H|\Psi\rangle = E_{g}+\frac{1}{2}\sum\limits_{n=1}^{\infty}\hbar\omega_n.
@@
 Here on the other hand, we consider the \textit{electromagnetic modes of a single atom in a cavity}.
+\Jadd{( Can we expand the discussion of the connection to the casimir energy/casimir-polder effect?
+Maybe even put a new subsection to give some focus.)}
 
 We now look to understand the effect of correlations on our ansatz.
-Interestingly, at lowest-order, correlations are entirely a result of the coupling between the momentum and the vector potential.
-Here, we capture the effect of correlations perturbatively.
-Treating such correlations perturbatively, it follows that the change in the ground state wavefunction $|\delta\rangle$ is given by \Jadd{(add that this is a first-order correction, maybe also add in the spirit of MP2)}
-\begin{equation}
+\Jadd{(Briefly define what we call here correlations, one definition could be, energy contributions beyond a single (slater) determinant)} Interestingly, at lowest-order, correlations are entirely a result of the coupling between the momentum and the vector potential.
+\Jadd{(Why only lowest order?
+Maybe briefly discuss here the perturbation serie)} Here, we capture the effect of correlations perturbatively.
+We briefly note that in the context of quantum chemistry, the perturbative approach we take now is analogous to Moller-Plesset perturbation theory (MP2).
+\Jadd{Connecting to quantum chemistry: They use the terms static correlation and dynamical correlations, the first occurs in the dissociation limit of streched molecules, where one needs two slater determinants.
+In contrast, dynamical correlations appear also close to the equilibrium distance.
+What would be the analogy here?
+Is there a case, where a superposition of two MHF wave functions become exact?}
+Treating such correlations perturbatively, it follows that the change in the ground state wavefunction $|\delta\rangle$ is given to first order as:
+\begin{equation}\label{eq:MP2wavefunction}
 |\delta\rangle = \sum\limits_{b = 2}^{N_a}\sum\limits_{n=1}^{\infty} \frac{q}{m}\sqrt{\frac{1}{2\epsilon_0\omega_n}}\frac{F_n(d)p_{ba}}{\omega_{ab}-\omega_n}|b\rangle|n\rangle.
@@
 The corresponding change in the ground state energy, defined as the correlation energy $E_c$, is
-\begin{equation}
+\begin{equation}\label{eq:MP2energy}
 E_{c} = \frac{q^2}{2m^2\epsilon_0\omega_n}\frac{|F_n(d)p_{ba}|^2}{\omega_{ab}-\omega_n}.
@@
 Summarizing the result of this section, the quantum electrodynamics interaction energy in the ground state, according to our variational ansatz, should be well-approximated by
-\begin{equation}
+\begin{equation}\label{eq:total_interaction_energy}
 E_{int} = \frac{1}{2}\sum\limits_{n=1}^{\infty}\left(\hbar\omega_n - \hbar\omega_n^{(0)} \right) + \frac{q^2}{2m^2\epsilon_0\omega_n}\frac{|F_n(d)p_{ba}|^2}{\omega_{ab}-\omega_n},
 \end{equation}
-where the eigenmodes $F_n$ and their eigenfrequencies $\omega_n$ are given by Equations (11-13).
+where the eigenmodes $F_n$ and their eigenfrequencies $\omega_n$ are given by Equations (12-14).
 
@@
 
-In Figure~\ref{fig:decoupling}, we consider the ground state energy of the Hamiltonian of Equation (1) calculated by Equation (17) and also calculated by perturbation theory \Jadd{(briefly define what is meant by this, to distinguish with our approach.
-maybe non SCF-perturbation theory)}.
+In Figure~\ref{fig:decoupling}, we consider the ground state energy of the Hamiltonian of Equation \ref{eq:hamiltonian}  calculated by Equation \ref{eq:total_interaction_energy} and also calculated by perturbation theory.
+By "calculated by perturbation theory", we mean that we apply perturbation theory in which the photonic modes used are the bare sine modes, rather than those of Equation \ref{eq:field_mode}.
 In Figure~(\ref{fig:decoupling}) we take the ratio of these ground state energies to the "exact" ground state energy, computed by numerical diagonalization of the Hamiltonian.
@@
 It is also important that our ansatz does well in a multi-level system, because as pointed out recently, there are severe discrepancies between the Coulomb gauge and length gauges for few level systems~\cite{schaefer2018, bernardis2018}, and the slow decay of matrix elements as a function of energy suggests that for few level systems, the length gauge is the appropriate gauge.
+\Jadd{(Can we discuss or show that this is related to how we treat the $A^2$ term?)}
 
@@
 \includegraphics[width=9cm]{fig3.pdf}
-\caption{\textbf{Sources of light-matter decoupling in the QED ground state.} (a) Plots of the mode profile of Equations (11-13) for the lowest mode as a function of position in the cavity for different emitter positions.
+\caption{\textbf{Sources of light-matter decoupling in the QED ground state.} (a) Plots of the mode profile of Equations (12-14) for the lowest mode as a function of position in the cavity for different emitter positions.
 The most striking effect is the screening of the cavity fields in the emitter region, which is most pronounced when the emitter is in the middle of the cavity and interacts strongest with the bare cavity mode.
@@
 In Figure~\ref{fig:sources_decoupling}, we explore an interesting feature of Figure ~\ref{fig:decoupling}, which is that the ground-state energy predicted by our ansatz is always substantially lower than predicted by perturbation theory for the coupling parameters considered here.
-To elaborate on this, we refer to Figure ~\ref{fig:sources_decoupling}(a), which shows the normalized lowest frequency electromagnetic mode $F_1(z)$ given by Equations (11-13) for a coupling strength of $0.33\omega_{21}$ and for different positions of the emitter in the cavity.
-The most obvious feature one sees is a strong dip (screening) in the electromagnetic fields where the emitter is.
-An analogous effect, known as a light-matter decoupling effect~\cite{liberato2014}, was noted in a recent work analytically in the context of expectation values of photodetection signals in the Hopfield model in cavity QED in the limit of deep-strong coupling.
-This dip can be understood physically as a result of the fact that the $A^2$ term in Equation (1) is positive-definite, increasing the energy when the emitter overlaps with strong intensity of the vacuum fluctuations of the electromagnetic field.
+To elaborate on this, we refer to Figure ~\ref{fig:sources_decoupling}(a), which shows the normalized lowest frequency electromagnetic mode $F_1(z)$ given by Equations (12-14) for a coupling strength of $0.33\omega_{21}$ and for different positions of the emitter in the cavity.
+The most obvious feature one sees is a strong dip (screening) in the electromagnetic fields where the emitter is \Jadd{located}.
+An analogous effect, known as a light-matter decoupling effect~\cite{liberato2014,garcia2015light,bayer2017terahertz}, was noted in a recent work analytically in the context of expectation values of photodetection signals in the Hopfield model in cavity QED in the limit of deep-strong coupling.
+This dip can be understood physically as a result of the fact that the $A^2$ term in the Hamiltonian is positive-definite, increasing the energy when the emitter overlaps with strong intensity of the vacuum fluctuations of the electromagnetic field.
 Thus the ground state should screen the electromagnetic field fluctuations out of the region where the emitter is.
-This dip leads to a self-consistent decrease in the magnitude of the interaction energy between the matter and the electromagnetic field, both for the $A^2$ and $Ap$ terms.
+This dip leads to a self-consistent decrease in the magnitude of the interaction energy between the matter and the electromagnetic field, both for the $A^2$ and $A\Jadd{\cdot}p$ terms.
 
@@
 Thus a blue-shift in the frequencies self-consistently decreases the magnitude of the coupling constant.
-Additionally, blue shifts lead to a further detuning in the energy denominators of Equations (15) and (16), serving to also self-consistently reduce the effect of correlations.
+Additionally, blue shifts lead to a further detuning in the energy denominators of Equations \ref{eq:MP2wavefunction} and \ref{eq:MP2energy}, serving to also self-consistently reduce the effect of correlations.
 The combined effects of the field screening and blue shifts are summarized in Figure ~\ref{fig:sources_decoupling}(c), where we plot the normalized change in the self-consistent coupling constant
@@
 
-\Jadd{(Nick, do you solve it actually in a scf cycle? if so, we could also show how the energy converges per iteration)}
+%\Jadd{(Nick, do you solve it actually in a scf cycle? if so, we could also show how the energy converges per iteration)}
 
@@
 Another interesting direction is to apply this ansatz on a system in which a perturbative treatment of correlations fails to capture their effect accurately.
-One way to remedy this in the framework of our ansatz is to treat the terms in the correlation energy of Equation (16) as variational parameters: this leads to a revised set of equations for the field modes, as well as the matter wavefunctions.
+One way to remedy this in the framework of our ansatz is to treat the terms in the correlation energy of Equation \ref{eq:MP2energy} as variational parameters: this leads to a revised set of equations for the field modes, as well as the matter wavefunctions.
 In particular, one will find for the fields:
 \begin{widetext}
-\begin{equation*}
+\begin{equation}\label{eq:self_consistent_maxwell}
 \left(\frac{d^2}{dz^2} + \frac{\omega_n^2}{c^2} \right)F_n(z) = \left(\frac{q^2}{4m\epsilon_0 c^2 A} - \sum\limits_{b=2}^{N_a}\frac{1}{2\epsilon_0 A}\left( \frac{q\hbar}{mc}\right)^2\frac{|p_{ba}|^2}{\omega_{ba}+\omega_n}\right)F_n(z)\delta(z-d),
-\end{equation*}
+\end{equation}
 \end{widetext}
@@
 \begin{widetext}
-\begin{equation*}
+\begin{equation}\label{eq:self_consistent_matter}
 \left(H_{\mathrm{matter}} - \frac{1}{2\epsilon_0 A}\left(\frac{q\hbar}{m}\right)^2\sum\limits_{n=1}^{\infty}\sum\limits_{b=2}^{N_a}\frac{F^2_n(L/2)}{\omega_n(\omega_{ba}+\omega_n)}p|\psi_b\rangle\langle \psi_b|p\right)|\psi_a\rangle = \hbar\omega_a|\psi_a\rangle.
-\end{equation*}
+\end{equation}
 \end{widetext}

### 2018-10-12 paper4pg.tex
new version of manuscript

\documentclass[aps,prl,twocolumn,
	groupedaddress,superscriptaddress,
	amsfonts,amssymb,amsmath,floatfix,
	citeautoscript]{revtex4-1}

\usepackage{graphicx}
\usepackage[centering,hmargin=20mm,tmargin=30mm,bmargin=25mm]{geometry}
\usepackage{multirow}
\usepackage{newtxtext}
\usepackage[cmintegrals]{newtxmath}

%----- References -----
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{colorlinks,
	linkcolor={blue!75!black!80!yellow},
	citecolor={blue!75!black!80!yellow},
	urlcolor={blue!75!black!80!yellow}
}

%----- Captions in sans font -----
\makeatletter
\renewcommand\@make@capt@title[2]{%
	\@ifx@empty\float@link{\@firstofone}{\expandafter\href\expandafter{\float@link}}%
	\sffamily{\textbf{#1}}\@caption@fignum@sep#2
}%
\renewcommand\figurename{Fig.}
\makeatother

\thickmuskip=5mu plus 2mu minus 1mu  %binary relations (default, 5mu plus 5mu)
\medmuskip=4mu plus 2mu minus 2mu    %binary operations (default, 4mu plus 2mu minus 4mu)

\frenchspacing %Ensure that revTeX does not do "double spaces" after punctuation

\renewcommand{\Im}{\operatorname{Im}}
\renewcommand{\Re}{\operatorname{Re}}
\newcommand{\sub}[1]{\ensuremath{_{\textrm{#1}}}} %Upright multi-character subscript
\newcommand{\super}[1]{\ensuremath{^{\textrm{#1}}}} %Upright multi-character superscript

\newcommand{\HarvardSEAS}{John A. Paulson School of Engineering and Applied Sciences, Harvard University, Cambridge, MA, USA}
\newcommand{\MITPhy}{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA, USA}

\usepackage[normalem]{ulem}
\newcommand{\Jadd}[1]{\textcolor{blue}{#1}}
\newcommand{\Jrem}[1]{\textcolor{blue}{\sout{#1}}}

%\usepackage[usenames]{color}
%\newcommand{\edited}[1]{{\color{red} #1}}

\begin{document}

\title{Variational theory of the ground state of quantum electrodynamics}

\author{Nicholas Rivera}\email{nrivera@seas.harvard.edu}\affiliation{\HarvardSEAS}\affiliation{\MITPhy}
\author{Johannes Flick}\email{flick@seas.harvard.edu}\affiliation{\HarvardSEAS}
\author{Prineha Narang}\email{prineha@seas.harvard.edu}\affiliation{\HarvardSEAS}

\date{\today}

\begin{abstract}
The past decade has brought rapid theoretical and experimental advances in the field of ultra-strong coupling of light and matter from microwave to optical frequencies. Many of the theoretical analyses of such systems proceed by considering simple two-level systems coupled to a single cavity mode. There remains a need to understand light-matter coupling  in contexts of complicated electronic systems coupling to many-mode or continuum photonic systems, as well as a need to understand how vacuum fluctuations of electromagnetic fields are altered in real space by strong light-matter coupling in such complex systems.  In this work, we develop a variational theory of the ground state of a quantum electrodynamical system of coupled light and matter. Essential to our ansatz is the notion an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling. Our variational formulation leads to a set of general equations that can describe the ground state of multi-electron systems coupled to many photonic modes in real space. As a first step towards an  \textit{ab initio} approach to ground states in QED, we apply our ansatz to a Rabi model involving a multi-level emitter with many optical modes, a model which cannot be analytically solved. We find a compact semi-analytic formula which describes the ground state energy very accurately (to less than 1\% error) in all regimes of coupling parameters allowed by the Thomas-Reiche-Kuhn sum rule. We expect our method to give rise to highly accurate descriptions of ground-state phenomena in quantum electrodynamics in complicated systems, such as Casimir forces, Lamb shifts, and fluctuational electrodynamics phenomena. 
\end{abstract}

\maketitle

%\section{Introduction}
%\label{sec:Introduction}

The past decade has brought an explosion of effort and experimental progress in realizing the coupling of matter and quantized electromagnetic fields beyond a regime of weak coupling, in which perturbative treatments of quantum electrodynamical couplings can no longer accurately describe the dynamics \cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,kockum2018ultrastrong}. Ultra-strong, or even deep-strong coupling is now regularly observed in superconducting qubits at microwave frequencies \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, ensembles of molecules in optical cavities \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, 2D electron gas cyclotron resonances coupled to cavities \cite{hagenmuller2010ultrastrong,scalari2012ultrastrong,zhang2016collective}, intersubband transitions in quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and single-emitters in ultra-confined plasmonic geometries \cite{benz2016,chikkaraddy2016}. Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, highly ionized atoms coupled to optical media via the Cerenkov effect, and many more. The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems \cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties \cite{orgiu2015}, realizing analogues of nonlinear optical processes with vacuum fluctuations \cite{kockum2017deterministic}, and even as new sources of photons through the dynamical Casimir effect \cite{ciuti2005quantum}.

In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved. In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model \cite{hopfield1958theory}, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators \cite{dicke1954coherence}. Any realistic system which deviates from these simpler models require numerical diagonalization approaches for an exact solution, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space. It also becomes unwieldy when one needs to keep track of many cavity modes which can also be populated with many photons. Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}. With such density-functional descriptions, workers in this field have been able to compute ground-state and time-dependent properties of realistic molecules in optical cavities \cite{flick2017c}.

In what follows, we propose the use of variational methods for the solution of \emph{ab initio} QED problems.  In particular, we develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states. This ansatz, reminiscent to the Hartree-Fock ansatz of electronic structure theory, leads to coupled eigen-equations describing the ground-state of the light-matter system. We benchmark our ansatz against a multi-level and multi-mode Rabi model parameterized by the strength of the characteristic momentum matrix elements describing the coupling of the matter and the electromagnetic fields. We find that for light-matter couplings that respect the Thomas-Reiche-Kuhn sum rule, our theory finds the ground state energy to a remarkable accuracy of more than >99\%, even in non-perturbative coupling regimes.  In regimes where our results are accurate, we claim we have found the effective quasiparticle description of the ground state. Moreover, in this regime, the ground state interaction energy manifests itself as the difference in zero-point energies between the quasiparticles and the bare particles.  Our findings furnish a non-perturbative theory of the Lamb shift, allowing us to non-perturbatively calculate position-dependent Lamb shifts of ground and excited states that give rise to Casimir-Polder forces.

%Our results service several aims. First, they suggest a novel variational approach for understanding ground states in ultra-strongly coupled light-matter systems. Second, they provide a general framework for understanding light-matter decoupling effects in the regime of strong light-matter interactions. Third, they prove a non-perturbative theory of Casimir-Polder forces \Jadd{(cite)}. Finally, they provide a rigorous concept of correlation energy in quantum electrodynamics in a way analogous to electronic structure theory.

%The outline of this work is as follows: in section~\ref{sec:variational_ansatz}, we develop the variational ansatz for the ground state of the Hamiltonian of macroscopic quantum electrodynamics. In section~\ref{sec:multi-level}, we calculate the ground state energy according to this ansatz for the multi-level and multi-mode Rabi model and compare it to numerical diagonalization of the same problem. In Section~\ref{sec:extensions}, we propose a self-consistent extension of our ansatz.

The QED Hamiltonian takes the form $H = H_{mat}+H_{em}+H_{int}$ where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter. The matter Hamiltonian takes the form:
\begin{align}
H_{el} &= \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext}(x) \right)\psi(x) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
\end{align}
where $v_{ext}$ is the one-body potential, $V(x-x')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field. %\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}

Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$. This renders the free electromagnetic Hamiltonian as
\begin{equation}
H_{em} = \int d^3x~ \epsilon (\partial_t \mathbf{A}(x))^2 + \mathbf{A}(x)\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(x)),
\end{equation}
where $\epsilon$ and $\mu$ represent a non-dispersive and positive dielectric and magnetic background that the matter and photon occupy. For cases we consider in this work, these will be taken to be unity, but we leave them here for the sake of generality.

The interaction Hamiltonian takes the form:
%\begin{align}
%H_{int} = &\frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(x) + \nonumber \\ &\frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(x).
%\end{align}
\begin{align}
H_{int} &= \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}(x)\cdot\nabla +  \nabla \cdot \mathbf{A}(x))\psi(x) \nonumber \\ &+ \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}(x)\psi(x)\mathbf{A}^2(x).
\end{align}

This Hamiltonian, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions $\{\psi_n\}$, and in terms of a set of photonic mode functions $\{\mathbf{F}_i\}$ via mode expansions for the electronic and photonic field operators. The electron field operator takes the form $\psi(x) = \sum_n \psi_n(x)c_n$.
In this second-quantized description, the $c_n$ is an annihilation operator for an electron corresponding to state $n$. The electromagnetic field operator takes the form $\mathbf{A}(x) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(x)a_i+\mathbf{F}^*_i(x)a^{\dagger}_i\right)$. In the electromagnetic field operator, we parameterize not only by the mode functions but also by modal frequencies. We have chosen a form for the vector potential operator which mirrors the choice generally adopted in QED. The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}.$

With the Hamiltonian set up, we now move to describe a variational theory of the ground state. In the variational theorem, we choose some ansatz $|\Omega\rangle$ for the ground state of $H$. The variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy. Parameterizing the ground state to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy for the chosen family of ground states.  We choose as our ansatz
\begin{equation}
|\Omega\rangle = \left( \prod\limits_n c_n^{\dagger}|0_n\rangle,\right) \otimes \left(\bigotimes_i|0_i\rangle\right),
\end{equation}
In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons. Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions. Thus, we shall minimize the expectational value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$. Note that the electromagnetic field gets two sets of (distinct) variational parameters, $\omega_i$ and $\mathbf{F}_i$. The last thing to note before taking the expectation values and minimization is that we should focus on the submanifold of variations that keep the complete set of electron and photon functions normalized. This shall be enforced by constructing the Lagrange function:
\begin{widetext}
\begin{equation}
\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle - \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right),
\end{equation}
\end{widetext}
with the $E_n$ and $\lambda_i$ being the Lagrange multipliers that enforce the normalization condition. Evaluating the expectation value of the Hamiltonian, and minimizing the Lagrange function  with respect to the orbitals, photon modes, and photon frequencies, immediately yields:
\begin{align}
&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi^*_j(x')\psi_j(x')\psi_i(x) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi_j(x')\psi_j(x)\psi_i(x')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)   = E_i\psi_i(x).
\end{align}
for the electron orbitals and energies. One sees that in addition to the one-body and Hartree-Fock terms for the electrons, the effect of the QED coupling is to add a one-body ponderomotive potential. We note that in the often used dipole approximation, the ponderomotive potential has no effect as it becomes independent of position.
\begin{figure}[t]
\includegraphics[width=8.5cm]{conceptfigure.pdf}
\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons. (Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
\label{fig:ansatz}
\end{figure}
For the photon orbitals and energies, the minimization yields:
\begin{equation}
\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_q^2} \right)\right)\mathbf{F}_q = 0,
\end{equation}
where $\omega_p^2 = \frac{e^2}{m\epsilon_0}\sum\limits_{i=1}^N |\psi_i(x)|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located.

Immediately, one notices that the term in the interaction Hamiltonian linear in the vector potential makes no contribution to the expectation value of the ground state of the energy in this ansatz. In many quantum electrodynamical problems, this linear term, which has zero expectation value in our ansatz, is important. To first order, it leads to spontaneous emission of a single photon. To second order, it leads to two (virtual) photon processes, such as van der Waals / Casimir-Polder forces on emitters, real photon processes such as two-photon emission, and effective interactions betweeen distinct emitters. Thus, we seek to capture the effect of this term. Physically, this term will mix the factorizable ground state of Equation (7) with terms that have virtual excitations of the matter, as well as virtual excitations of the electromagnetic field. The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.
\begin{figure}[t]
\includegraphics[width=5cm]{figure2new.pdf}
\caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} Lowest two eigenvalues as a function of coupling for the multi-mode Rabi model describing the coupling of an emitter to a one-dimensional cavity. The variational results (blue) agree to within roughly a percent with the result derived by numerical diagonalization (orange) retaining up to four photons in the Hilbert space. The result from perturbation theory (green) disagrees strongly with the other methods of calculation for the first excited state due to a near-resonance between the bare transition energy and the bare lowest cavity mode.}
\label{fig:ansatz}
\end{figure}
We capture the effect of correlations perturbatively. In other words, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential. That correction is given by 
\begin{equation}
\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{q=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_q^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_q(\omega_{mn} -\lambda_q)}
\end{equation}
where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.  In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (6) and (7), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Equation (8).
\begin{figure}[t]
\includegraphics[width=8.4cm]{figure3new.pdf}
\caption{\textbf{Variational theory of ground state energy: accuracy as a function of number of emitter levels.} (Left) Ground-state energy of the one-dimensional multi-mode Rabi model as a function of the dimensionless coupling for two (top), three (middle), and four (bottom) level emitter systems with momentum matrix elements chosen to saturate the Thomas-Reiche-Kuhn sum rule. (Right) Ratio of the energy calculated with the variational theory and with numerical diagonalization. These results show that our accuracy is robust to the number of levels in the system. }
\label{fig:ansatz}
\end{figure}
In what follows, we provide a proof-of-concept demonstration of the efficacy of the variational theory derived here. We consider the QED Hamiltonian corresponding to a single emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction. As the cavity is considered for simplicity to be one-dimensional, the electric field is oriented along a single direction, denoted $x$, while the magnetic field is oriented along a direction transverse to both the electric field and the cavity length, denoted $y$.

Working under the long-wavelength or dipole approximation, the Hamiltonian can then be written as:
\begin{equation}\label{eq:hamiltonian}
H = H_{\text{matter}}+\frac{\epsilon_0}{2}\int dz~(E^2+c^2B^2)+\frac{q}{m}A(d)p + \frac{q^2}{2m}A^2(d).
\end{equation}
The fields can be expressed as a mode expansion via:
\begin{equation}\label{eq:mode_expansion}
A(z,t) = \sum\limits_{n=1}^{\infty} \sqrt{\frac{\hbar}{2\epsilon_0\omega_n }}(F_n(z)a_ne^{-i\omega_n t}+F^*_n(z)a_n^{\dagger}e^{i\omega_n t}),
\end{equation}
with $E(z,t) = -\partial_t A(z,t)$ and $B(z,t)=\partial_z A(z,t)$.  The $F_n(z)$ are the mode functions of the cavity, normalized such that $\int dz ~|F_n(z)|^2 = 1$. For a cavity of length $L$ the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$. The matter Hamiltonian we take to be a multilevel system with an arbitrary number of levels, $N_a$. The matter system we describe as an $N_a$ site system. This could be considered as a simple model of a molecule within a tight-binding description. Thus we parameterize the general family of matter Hamiltonians as:
\begin{equation}\label{eq:matter_hamiltonian}
H_{\text{matter}} = \sum\limits_{i=1}^{{N_a-1}} V_i|i\rangle\langle i|+t(|i\rangle\langle i+1|+|i+1\rangle\langle i|) 
\end{equation}
The momentum operator, we write in a finite-different representation as
\begin{equation}\label{eq:momentum_operator}
p = \frac{-i\hbar}{R}\sum\limits_{i=1}^{N_a-1} \left(|i\rangle\langle i+1|-|i+1\rangle\langle i| \right),
\end{equation}
where $R$ is a constant with units of length representing roughly the difference in positions between sites. This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the Thomas-Reiche-Kuhn (TRK) sum rule is enforced. In other words: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states. Equation \ref{eq:TRK_sum_rule} is the Thomas-Reiche-Kuhn (TRK) sum rule \cite{ciuti2005quantum}. Since the TRK is based on a full electronic real-space description, this sum rule does not rigorously apply to a discrete-level system. However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot. As we shall show later in the text however, it does place a bound on how strong the effect of the $A\Jadd{\cdot}p$ term can be on the total energy. The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.

The detailed calculations of the energies of states via the formalism introduced here will be shown in a longer paper. Here, we state the main results. Using an essentially one-dimensional version of Equation (7), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations. In the absence of correlations, we found that the ground state energy satisfies
\begin{equation}
E = E_g + \frac{1}{2}\sum\limits_{n=1}^{\infty}(\omega_n - \omega_n^0),
\end{equation}
where the $\omega_n$ are found in our framework and $\omega_n^0 = \frac{n\pi c}{L}$. This is a remarkable formula, as it says that in the absence of correlations, the energy of the system is the Casimir energy of the system. In particular, it has long been known that when two conducting plates are placed near each other, there is a Casimir energy associated with the fact that the zero-point energy of the nearby plates is different than the zero-point energy of plates infinitely apart. This is simply because the electromagnetic mode structure of two nearby plates is very different from that of two infinitely separated plates. This Casimir energy is simply the difference between the interacting and non-interacting zero-point energies. This logic can be applied to any arrangement of macroscopic polarizable objects. What is remarkable about the result of Equation (13) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.

In the presence of correlations induced by the $A \cdot p$ term, in addition to Equation (13), we must add a contribution of the form of Equation (8), specialized to the case of an emitter in a one-dimensional cavity. We can actually apply the correlation correction to excited states as well, by using the second-order perturbation theory formula for the energy shift of excited states due to the $A \cdot p$ term, using the same electron and photon orbitals and frequencies as derived in Equations (6) and (7). It is worthwhile to note that for this one-dimensional case, as we show in a longer paper, we arrive at semi-analytic results for all variational energies that we calculate.

In Figure 2, we show the result of this procedure when applied to calculate the two lowest energies of a two-level system coupled to a one-dimensional cavity. For the two-level system considered here, the energy gap between the bare emitter levels is 0.5 eV, while the lowest bare cavity frequency is 0.61 eV. Thus, the two lowest eigenvalues are similar to the bare ground and excited states of the emitter with zero photons in the cavity. In blue is our estimate for these two energies, in orange is the exact result by numerical diagonalization of the Hilbert space, and in green is the result obtained by perturbation theory in terms of the bare atomic and cavity parameters. One immediately sees that the variational result is quite accurate when compared to numerical diagonalization, much more so than perturbation theory, which completely fails for the first excited state.

The reason it fails for the first excited state more than for the ground state is that the first excited state is nearly resonant with the ground state, leading to a very large negative contribution from the $A\cdot p$ of nearly 2 eV, which is far larger than the spacing of the bare emitter levels. On the other hand, the variational estimate finds no such large negative energy shift, and leads to an energy gap between the first two levels which is similar to the bare gap, and in agreement with numerical diagonalization. The reason for this behavior is that the effect of the plasma term in Equation (7) is to blue-shift all of the photon frequencies. In particular, for the largest coupling considered in Figure 2, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition. This result very clearly demonstrates not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime. Moreover, our results demonstrate a successful non-perturbative theory of the Lamb shift and consequently Casimir-Polder forces. In particular, it is long known that energy levels of emitters can shift as a result of virtual photon emission and re-absorption. These energy shifts, called Lamb shifts, depend on the particular position of the emitter in the photonic structure it is embedded in. These shifts not only lead to changes in the transition frequencies of the emitter, but the position dependence of these energy shifts also implies forces on the emitter, typically called Casimir-Polder forces. Such Casimir-Polder forces are typically calculated using the celebrated Lifshitz theory, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Equation (8) using bare atomic and photonic properties. Thus, our calculations, which find the energy shifts non-perturbatively, lead to a consequently non-perturbative theory of Casimir-Polder forces.

In Figure 3, we consider the ground state of the QED Hamiltonian, calculated through the procedure above, but now extended to three and four level systems. In all cases, the parameter $R$ which controls the strength of the momentum matrix elements and the correlation energy, is chosen to saturate the TRK sum rule. We find that the accuracy of our approach is not compromised by having additional energy levels in the Hamiltonian, indicating the potential generality of our approach.

Finally, we note that Equation (8), which incorporates correlations, can also be employed self-consistently, by adding it to the Lagrange function of Equation (5), and minimizing the result with respect to electronic and photonic properties. The set of equations that arise are:

\begin{widetext}
\begin{align}
&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(x).
\end{align}
\end{widetext}

\begin{widetext}
\begin{equation}
\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_q^2} \right)\right)\mathbf{F}_q = -\frac{e^2\hbar}{2m^2\epsilon_0c^2}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_{mn}-\omega_{q}}\mathbf{j}_{nm}.
\end{equation}
\end{widetext}
These two equations are highly general, providing real-space information about the electronic and photonic properties of the QED ground state for a many-electron system in a many-mode system in three dimensions.

In summary, we have developed a variational approximation to the ground state of quantum electrodynamical problems and have benchmarked it against the multi-mode and multi-level Rabi model, which has no analytical solutions, and whose accurate numerical diagonalization requires a very high-dimensional Hilbert space. Despite our ansatz being centered on the assumption of weak correlations, with correlations from the $Ap$ term being treated perturbatively, we find that our approximation does very well in predicting the ground state energy, even when the strength of the correlations saturates the TRK sum rule.  With the advent of ultra-strong coupling and deep-strong coupling being realizable in quantum electrodynamical systems, one may use this real-space knowledge of how matter affect photons to design a photonic mode atom-by-atom.

%In what follows, we add this energy correction $\delta E$ to the expectation value of the energy in the ground state, with the orbitals and eigenvalues as variational parameters. As a result, the orbitals and eigenvalues obtained will be different from Equation (11) and (14), and this difference will be small in the limit of weak correlations. Strictly speaking, this approach is only justified for weak correlations, but can be applied to systems with strong-correlations as is often done with self-consistent methods. Minimizing with respect to the orbitals, photon modes, and photon frequencies, immediately yields:
%\begin{widetext}
%\begin{align}
%&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(x).
%\end{align}
%\end{widetext}
%
%\begin{widetext}
%\begin{equation}
%\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_q^2} \right)\right)\mathbf{F}_q = -\frac{e^2\hbar}{2m^2\epsilon_0c^2}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_{mn}-\omega_{q}}\mathbf{j}_{nm}.
%\end{equation}
%\end{widetext}
%These two equations are highly general, providing real-space information about the electronic and photonic properties of the QED ground state for a many-electron system in a many-mode system in three dimensions. Commensurately, these equations are complicated, and to understand them, we now consider a simple case of them.

%\section{Summary and Outlook}
%In summary, we have developed a variational approximation to the ground state of quantum electrodynamical problems and have benchmarked it against the multi-mode and multi-level Rabi model, which has no analytical solutions, and whose accurate numerical diagonalization requires a very high-dimensional Hilbert space. Despite our ansatz being centered on the assumption of weak correlations, with correlations from the $Ap$ term being treated perturbatively, we find that our approximation does very well in predicting the ground state energy, even when the strength of the correlations saturates the TRK sum rule.

%From a more fundamental perspective, the ansatz studied in this work is interesting because they give a rigorous meaning to the notion of correlation energy in quantum electrodynamics which is parallel to correlation energy in electronic structure theory. In particular, in the context of electronic structure, the correlation energy is the difference between the true ground state energy and that computed by Hartree-Fock, which is an independent electron ansatz. Similarly, our ansatz here is an independent electron-independent photon ansatz whose structure is parallel to the Hartree-Fock ansatz.

%Another fundamentally interesting aspect of our ansatz is that unlike current formulations of quantum-electrodynamical density functional theory, it considers the real-space properties of photon modes as they are affected by matter. With the advent of ultra-strong coupling and deep-strong coupling being realizable in quantum electrodynamical systems, one may use this real-space knowledge of how matter affect photons to design a photonic mode atom-by-atom.

\section{Acknowledgements}
% Pri, Johannes: Please insert any other relevant acknowledgement information here!
We thank Prof. Joel Yuen-Zhou (University of California San Diego), Prof. Ido Kaminer (Technion Israel Institute of Technology), Prof. Marin Solja\v{c}i\'{c} (Massachusetts Institute of Technology) and Prof. John D. Joannopoulos (Massachusetts Institute of Technology), for useful discussions. N. R. recognizes the support of the DOE Computational Science Graduate Fellowship (CSGF) fellowship no.  DE-FG02-97ER25308. P. N. acknowledges start-up funding from the Harvard John A. Paulson School of Engineering and Applied Sciences.  We also acknowledge support from the STC Center for Integrated Quantum Materials NSF Grant No. DMR-1231319 and from the Harvard John A. Paulson School of Engineering and Applied Sciences. J. F. acknowledges financial support from the Deutsche Forschungsgemeinschaft under Contract No. FL 997/1-1.

\bibliographystyle{apsrev4-1}
\bibliography{references}

\end{document}

### 2018-10-12 paper4pg.tex
small change

@@
 The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}.$
-
+\begin{figure}[t]
+\includegraphics[width=8.5cm]{conceptfigure.pdf}
+\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons.
+(Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
+\label{fig:ansatz}
+\end{figure}
 With the Hamiltonian set up, we now move to describe a variational theory of the ground state.
@@
 \end{equation}
-In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons.
+In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons (see Figure 1).
 Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions.
@@
 We note that in the often used dipole approximation, the ponderomotive potential has no effect as it becomes independent of position.
-\begin{figure}[t]
-\includegraphics[width=8.5cm]{conceptfigure.pdf}
-\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons.
-(Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
-\label{fig:ansatz}
-\end{figure}
+
 For the photon orbitals and energies, the minimization yields:

### 2018-10-16 paper.tex

@@
 Any realistic system which deviates from these simpler models require numerical diagonalization approaches, which become unwieldy when the dimension of the electronic Hilbert space becomes large, as is the case when describing an electronic system in real space.
-It also becomes unwieldy when one needs to keep track of \Jrem{many}\Jadd{multi} cavity modes, which can be populated with many photons.
+It also becomes unwieldy when one needs to keep track of many cavity modes, which can be populated with many photons.
 Exciting new developments towards ab-initio studies of QED have proceeded by proposing reduced-quantity descriptions such as density-functional descriptions of the light-matter coupling \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}.
@@
 
-In what follows, we propose the use \Jadd{of} variational methods for the solution of \emph{ab initio} QED problems.
+In what follows, we propose the use variational methods for the solution of \emph{ab initio} QED problems.
 In particular, here we develop a particular class of mean-field variational ansatzes for the ground state of quantum electrodynamical Hamiltonians.
 In particular, we develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states.
-This ansatz, \Jrem{analogous} \Jadd{reminiscent} to the Hartree-Fock ansatz of electronic structure theory, leads to coupled eigen-equations describing the ground-state of the light-matter system.
+This ansatz, analogous to the Hartree-Fock ansatz of electronic structure theory, leads to coupled eigen-equations describing the ground-state of the light-matter system.
 We benchmark our ansatz against a multi-level and multi-mode Rabi model parameterized by the strength of the characteristic momentum matrix elements describing the coupling of the matter and the electromagnetic fields.
-We find that when the momentum matrix elements are sufficiently weak, our theory finds the ground state energy to a remarkable accuracy of \Jrem{less}\Jadd{more?} than >99.9\%, even in the non-perturbative coupling regime.
-Even in regimes where the correlation energy is 50\% of the total energy, far outside of the validity domain of our ansatz, our ansatz predicts the energy up to a few percent \Jadd{due to self-consistency}.
+We find that when the momentum matrix elements are sufficiently weak, our theory finds the ground state energy to a remarkable accuracy of less than >99.9\%, even in the non-perturbative coupling regime.
+Even in regimes where the correlation energy is 50\% of the total energy, far outside of the validity domain of our ansatz, our ansatz predicts the energy up to a few percent.
 In regimes where our results are accurate, we claim we have found the effective quasiparticle description of the ground state.
 Moreover, in this regime, the ground state interaction energy manifests itself as the difference in zero-point energies between the quasiparticles and the bare particles.
-This generalizes the notion of the Casimir energy \Jadd{(cite)} to a single atom in a cavity.
+This generalizes the notion of the Casimir energy to a single atom in a cavity.
 
@@
 Second, they provide a general framework for understanding light-matter decoupling effects in the regime of strong light-matter interactions.
-Third, they prove a non-perturbative theory of Casimir-Polder forces \Jadd{(cite)}.
+Third, they prove a non-perturbative theory of Casimir-Polder forces.
 Finally, they provide a rigorous concept of correlation energy in quantum electrodynamics in a way analogous to electronic structure theory.
@@
 \label{sec:theory}
-In the Supplementary Information (SI), we derive a set of nonlinear coupled eigenequations which can be applied to describe the ground state of any \Jadd{nonrelativistic} macroscopic quantum electrodynamical system, in three dimensions, with electron-electron interactions, and with arbitrarily many photon modes.
-These are real-space equations that give single-particle electron and photon orbitals \Jadd{(should we call them orbitals?)} in the presence of QED coupling, thus giving real-space information about the electron density and the electromagnetic field fluctuations in the ground state.
+In the Supplementary Information (SI), we derive a set of nonlinear coupled eigenequations which can be applied to describe the ground state of any macroscopic quantum electrodynamical system, in three dimensions, with electron-electron interactions, and with arbitrarily many photon modes.
+These are real-space equations that give single-particle electron and photon orbitals in the presence of QED coupling, thus giving real-space information about the electron density and the electromagnetic field fluctuations in the ground state.
 The equations derived are in general fairly complicated, and thus in the main text, we analyze a much simpler case of the equations in order to extract the basic physics of the equations.
@@
 \end{equation}
-\Jadd{(this is dipole approximation, or dipole approximation is exact for point-like single atom? )}
 The fields can be expressed as a mode expansion via:
@@
 where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.
-Equation \ref{eq:TRK_sum_rule} is the Thomas-Reiche-Kuhn (TRK) sum rule \Jadd{(cite)}.
-\Jrem{Of course} \Jadd{Since the TRK is based on a full electronic real-space description}, this sum rule does not rigorously apply to a discrete-level system.
+Equation \ref{eq:TRK_sum_rule} is the Thomas-Reiche-Kuhn (TRK) sum rule.
+Of course, this sum rule does not rigorously apply to a discrete-level system.
 However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot.
-As we shall show later in the text however, it does place a bound on how strong the effect of the $A\Jadd{\cdot}p$ term can be on the total energy.
+As we shall show later in the text however, it does place a bound on how strong the effect of the $Ap$ term can be on the total energy.
 The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.
@@
 The condition of Equation \ref{eq:mode_transcendental} determines the resonance frequencies of the photon quasiparticle modes.
-\Jadd{(put eq.12-14 to SI?)} After calculating the modes and the frequencies, we may move to an evaluation of the energy of the ground state.
+After calculating the modes and the frequencies, we may move to an evaluation of the energy of the ground state.
 We proceed by taking the solutions of Equation \ref{eq:field_mode}, normalized with the normalization factor of Equation \ref{eq:mode_normalization}, and plugging it back into the expectation value of the Hamiltonian of the ground state (Equation \ref{eq:expectation_val1}).
@@
 Here on the other hand, we consider the \textit{electromagnetic modes of a single atom in a cavity}.
-\Jadd{( Can we expand the discussion of the connection to the casimir energy/casimir-polder effect?
-Maybe even put a new subsection to give some focus.)}
 
 We now look to understand the effect of correlations on our ansatz.
-\Jadd{(Briefly define what we call here correlations, one definition could be, energy contributions beyond a single (slater) determinant)} Interestingly, at lowest-order, correlations are entirely a result of the coupling between the momentum and the vector potential.
-\Jadd{(Why only lowest order?
-Maybe briefly discuss here the perturbation serie)} Here, we capture the effect of correlations perturbatively.
+Interestingly, at lowest-order, correlations are entirely a result of the coupling between the momentum and the vector potential.
+Here, we capture the effect of correlations perturbatively.
 We briefly note that in the context of quantum chemistry, the perturbative approach we take now is analogous to Moller-Plesset perturbation theory (MP2).
-\Jadd{Connecting to quantum chemistry: They use the terms static correlation and dynamical correlations, the first occurs in the dissociation limit of streched molecules, where one needs two slater determinants.
-In contrast, dynamical correlations appear also close to the equilibrium distance.
-What would be the analogy here?
-Is there a case, where a superposition of two MHF wave functions become exact?}
 Treating such correlations perturbatively, it follows that the change in the ground state wavefunction $|\delta\rangle$ is given to first order as:
@@
 It is also important that our ansatz does well in a multi-level system, because as pointed out recently, there are severe discrepancies between the Coulomb gauge and length gauges for few level systems~\cite{schaefer2018, bernardis2018}, and the slow decay of matrix elements as a function of energy suggests that for few level systems, the length gauge is the appropriate gauge.
-\Jadd{(Can we discuss or show that this is related to how we treat the $A^2$ term?)}
 
@@
 To elaborate on this, we refer to Figure ~\ref{fig:sources_decoupling}(a), which shows the normalized lowest frequency electromagnetic mode $F_1(z)$ given by Equations (12-14) for a coupling strength of $0.33\omega_{21}$ and for different positions of the emitter in the cavity.
-The most obvious feature one sees is a strong dip (screening) in the electromagnetic fields where the emitter is \Jadd{located}.
+The most obvious feature one sees is a strong dip (screening) in the electromagnetic fields where the emitter is.
 An analogous effect, known as a light-matter decoupling effect~\cite{liberato2014,garcia2015light,bayer2017terahertz}, was noted in a recent work analytically in the context of expectation values of photodetection signals in the Hopfield model in cavity QED in the limit of deep-strong coupling.
@@
 Thus the ground state should screen the electromagnetic field fluctuations out of the region where the emitter is.
-This dip leads to a self-consistent decrease in the magnitude of the interaction energy between the matter and the electromagnetic field, both for the $A^2$ and $A\Jadd{\cdot}p$ terms.
+This dip leads to a self-consistent decrease in the magnitude of the interaction energy between the matter and the electromagnetic field, both for the $A^2$ and $Ap$ terms.
 

### 2018-10-16 paper.tex

--- 2018-10-04 paper.tex
@@
 We consider the interaction in the velocity gauge, and neglect the weak interaction between spins and magnetic fields, thus neglecting spin altogether.
+We also work under the long-wavelength or dipole approximation.
 The Hamiltonian can then be written as:
@@
 with $E(z,t) = -\partial_t A(z,t)$ and $B(z,t)=\partial_z A(z,t)$.
-The $F_n(z)$ are the mode functions of the cavity, normalized such that $\int dz ~|F_n(z)|^2 = 1$.
 For a cavity of length $L$ the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$.
@@
 where $R$ is a constant with units of length representing roughly the difference in positions between sites.
-This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the following sum rule is enforced:
-\begin{equation}\label{eq:TRK_sum_rule}
-\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1,
-\end{equation}
-where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.
+This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the Thomas-Reiche-Kuhn (TRK) sum rule is enforced.
+In other words: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.
 Equation \ref{eq:TRK_sum_rule} is the Thomas-Reiche-Kuhn (TRK) sum rule \Jadd{(cite)}.
-\Jrem{Of course} \Jadd{Since the TRK is based on a full electronic real-space description}, this sum rule does not rigorously apply to a discrete-level system.
+Since the TRK is based on a full electronic real-space description, this sum rule does not rigorously apply to a discrete-level system.
 However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot.

### 2018-10-16 paper4pg.tex

--- 2018-10-12 paper4pg.tex
@@
 
-\title{Variational theory of the ground state of quantum electrodynamics}
+\title{Variational theory of non-relativistic quantum electrodynamics}
 
@@
 \begin{abstract}
-The past decade has brought rapid theoretical and experimental advances in the field of ultra-strong coupling of light and matter from microwave to optical frequencies.
-Many of the theoretical analyses of such systems proceed by considering simple two-level systems coupled to a single cavity mode.
-There remains a need to understand light-matter coupling  in contexts of complicated electronic systems coupling to many-mode or continuum photonic systems, as well as a need to understand how vacuum fluctuations of electromagnetic fields are altered in real space by strong light-matter coupling in such complex systems.
-In this work, we develop a variational theory of the ground state of a quantum electrodynamical system of coupled light and matter.
+The almost routine ability to achieve ultra-strong coupling between light and matter promises to bring about new means to control  material properties, new concepts for manipulating light at the atomic scale, and fundamentally new insights into quantum electrodynamics (QED).
+Thus, there is a need to develop quantitative theories of QED phenomena in complex electronic and photonic systems.
+In this letter, we develop a variational theory of general non-relativistic QED systems of coupled light and matter.
 Essential to our ansatz is the notion an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling.
-Our variational formulation leads to a set of general equations that can describe the ground state of multi-electron systems coupled to many photonic modes in real space.
-As a first step towards an  \textit{ab initio} approach to ground states in QED, we apply our ansatz to a Rabi model involving a multi-level emitter with many optical modes, a model which cannot be analytically solved.
-We find a compact semi-analytic formula which describes the ground state energy very accurately (to less than 1\% error) in all regimes of coupling parameters allowed by the Thomas-Reiche-Kuhn sum rule.
-We expect our method to give rise to highly accurate descriptions of ground-state phenomena in quantum electrodynamics in complicated systems, such as Casimir forces, Lamb shifts, and fluctuational electrodynamics phenomena.
+This variational formulation leads to a set of general equations that can describe the ground state of multi-electron systems coupled to many photonic modes in real space.
+As a first step towards an  \textit{ab initio} approach to ground states in QED, we apply our ansatz to a QED system involving a multi-level emitter coupled to many optical modes.
+We find a compact semi-analytic formula which describes ground and excited state energies to less than 1\% error in all regimes of coupling parameters allowed by sum rules.
+Moreover, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as "the Casimir energy of a single atom in a cavity".
+Our method should give rise to highly accurate descriptions of ground-state phenomena in general QED systems, such as Casimir forces, Lamb shifts, and other fluctational electrodynamical phenomena.
 \end{abstract}
@@
 
-The past decade has brought an explosion of effort and experimental progress in realizing the coupling of matter and quantized electromagnetic fields beyond a regime of weak coupling, in which perturbative treatments of quantum electrodynamical couplings can no longer accurately describe the dynamics \cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,kockum2018ultrastrong}.
-Ultra-strong, or even deep-strong coupling is now regularly observed in superconducting qubits at microwave frequencies \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, ensembles of molecules in optical cavities \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, 2D electron gas cyclotron resonances coupled to cavities \cite{hagenmuller2010ultrastrong,scalari2012ultrastrong,zhang2016collective}, intersubband transitions in quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and single-emitters in ultra-confined plasmonic geometries \cite{benz2016,chikkaraddy2016}.
-Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, highly ionized atoms coupled to optical media via the Cerenkov effect, and many more.
-The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems \cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties \cite{orgiu2015}, realizing analogues of nonlinear optical processes with vacuum fluctuations \cite{kockum2017deterministic}, and even as new sources of photons through the dynamical Casimir effect \cite{ciuti2005quantum}.
+Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED) \cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,kockum2018ultrastrong}.
+Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems \cite{hagenmuller2010ultrastrong,scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems \cite{benz2016,chikkaraddy2016}.
+Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect, and many more.
+The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems \cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties \cite{orgiu2015}, realizing analogues of nonlinear optical processes with vacuum fluctuations \cite{kockum2017deterministic}, and as probes of exotic phenomena such as the dynamical Casimir effect \cite{ciuti2005quantum}.
+Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
+Through "reduced quantity theories" such as quantum electrodynamical density functional theory (QEDFT) \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}, one is now able to calculate observables in large molecules coupled to realistic optical cavities  \cite{flick2017c}.
 
-In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved.
+In this letter, we propose a variational framework to analyze complex light-matter systems from first principles.
+Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of functionals to describe the ground state, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the use of the long-wavelength (dipole) approximation.
+A variational framework, as we shall show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
+Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom.
+A variational framework also allows compact semi-analytic formulas to describe complex systems which may allow the development of functionals for use in QEDFT.
+%In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved.
 In particularly, the only exactly solvable models in all regimes of coupling parameters include the Rabi model~\cite{braak2011}, which describes the coupling of a two-level atom to a single mode, the Hopfield model \cite{hopfield1958theory}, which describes the coupling of a harmonic oscillator to any number of cavity modes, and Dicke models based on two-level systems and harmonic oscillators \cite{dicke1954coherence}.
@@
 
-In what follows, we propose the use of variational methods for the solution of \emph{ab initio} QED problems.
-In particular, we develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states.
-This ansatz, reminiscent to the Hartree-Fock ansatz of electronic structure theory, leads to coupled eigen-equations describing the ground-state of the light-matter system.
-We benchmark our ansatz against a multi-level and multi-mode Rabi model parameterized by the strength of the characteristic momentum matrix elements describing the coupling of the matter and the electromagnetic fields.
-We find that for light-matter couplings that respect the Thomas-Reiche-Kuhn sum rule, our theory finds the ground state energy to a remarkable accuracy of more than >99\%, even in non-perturbative coupling regimes.
-In regimes where our results are accurate, we claim we have found the effective quasiparticle description of the ground state.
-Moreover, in this regime, the ground state interaction energy manifests itself as the difference in zero-point energies between the quasiparticles and the bare particles.
-Our findings furnish a non-perturbative theory of the Lamb shift, allowing us to non-perturbatively calculate position-dependent Lamb shifts of ground and excited states that give rise to Casimir-Polder forces.
+Motivated by all of these potential advantages, we now develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states.
+This ansatz $-$ reminiscent to, but qualitatively distinct from, the Hartree-Fock ansatz of electronic structure theory $-$ leads to coupled eigen-equations describing the ground-state of the light-matter system.
+We apply our ansatz to describe ground and excited states in a multi-level emitter coupled to many photonic modes.
+We find that for light-matter couplings that respect sum rules, our theory finds the ground state energy to a remarkable accuracy of about 99\%, even in deeply non-perturbative coupling regimes.
+In regimes where our results are accurate, we have found the effective quasiparticle description of the ground state of QED.
+Our findings also furnish a non-perturbative theory of the position-dependent Lamb shifts of ground and excited states that give rise to Casimir-Polder forces.
 
@@
 
-The QED Hamiltonian takes the form $H = H_{mat}+H_{em}+H_{int}$ where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter.
+The QED Hamiltonian  $H = H_{mat}+H_{em}+H_{int}$ where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter.
 The matter Hamiltonian takes the form:
@@
 %\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}
-
+\begin{figure}[t]
+\includegraphics[width=8.5cm]{conceptfigure.pdf}
+\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons.
+(Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
+\label{fig:ansatz}
+\end{figure}
 Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$.
@@
 where $\epsilon$ and $\mu$ represent a non-dispersive and positive dielectric and magnetic background that the matter and photon occupy.
-For cases we consider in this work, these will be taken to be unity, but we leave them here for the sake of generality.
+For cases we consider in this work, these will be taken to be unity, but we leave them here for generality.
 
@@
 
-This Hamiltonian, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions $\{\psi_n\}$, and in terms of a set of photonic mode functions $\{\mathbf{F}_i\}$ via mode expansions for the electronic and photonic field operators.
+This Hamiltonian, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$.
 The electron field operator takes the form $\psi(x) = \sum_n \psi_n(x)c_n$.
-In this second-quantized description, the $c_n$ is an annihilation operator for an electron corresponding to state $n$.
-The electromagnetic field operator takes the form $\mathbf{A}(x) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(x)a_i+\mathbf{F}^*_i(x)a^{\dagger}_i\right)$.
+The $c_n$ is an annihilation operator for an electron corresponding to state $n$.
+The electromagnetic field operator takes the form $\mathbf{A}(x) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(x)a_i+\mathbf{F}^*_i(x)a^{\dagger}_i\right)$, where the $a_i^{(\dagger)}$ annihilate (create) a photon in mode $i$.
 In the electromagnetic field operator, we parameterize not only by the mode functions but also by modal frequencies.
-We have chosen a form for the vector potential operator which mirrors the choice generally adopted in QED.
 The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}.$
 
-With the Hamiltonian set up, we now move to describe a variational theory of the ground state.
+With the Hamiltonian described, we now move to develop a variational theory of the ground state.
 In the variational theorem, we choose some ansatz $|\Omega\rangle$ for the ground state of $H$.
@@
 Note that the electromagnetic field gets two sets of (distinct) variational parameters, $\omega_i$ and $\mathbf{F}_i$.
-The last thing to note before taking the expectation values and minimization is that we should focus on the submanifold of variations that keep the complete set of electron and photon functions normalized.
-This shall be enforced by constructing the Lagrange function:
+We enforce that the matter and photon remain normalized by constructing the Lagrange function:
 \begin{widetext}
@@
 \end{widetext}
-with the $E_n$ and $\lambda_i$ being the Lagrange multipliers that enforce the normalization condition.
-Evaluating the expectation value of the Hamiltonian, and minimizing the Lagrange function  with respect to the orbitals, photon modes, and photon frequencies, immediately yields:
+with the $E_n$ and $\lambda_i$ being the Lagrange multipliers that enforce the normalization conditions.
+Evaluating the expectation value of the Hamiltonian, and minimizing the Lagrange function  immediately yields:
 \begin{align}
@@
 for the electron orbitals and energies.
-One sees that in addition to the one-body and Hartree-Fock terms for the electrons, the effect of the QED coupling is to add a one-body ponderomotive potential.
+We see that in addition to the one-body and Hartree-Fock terms for the electrons, the effect of the QED coupling is to add a one-body ponderomotive potential.
 We note that in the often used dipole approximation, the ponderomotive potential has no effect as it becomes independent of position.
-\begin{figure}[t]
-\includegraphics[width=8.5cm]{conceptfigure.pdf}
-\caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons.
-(Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
-\label{fig:ansatz}
-\end{figure}
+
 For the photon orbitals and energies, the minimization yields:
 \begin{equation}
-\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_q^2} \right)\right)\mathbf{F}_q = 0,
+\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_i^2} \right)\right)\mathbf{F}_i = 0,
 \end{equation}
-where $\omega_p^2 = \frac{e^2}{m\epsilon_0}\sum\limits_{i=1}^N |\psi_i(x)|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located.
+where $\omega_p^2 = \frac{e^2}{m\epsilon_0}\sum\limits_{n=1}^N |\psi_n(x)|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located.
 
 Immediately, one notices that the term in the interaction Hamiltonian linear in the vector potential makes no contribution to the expectation value of the ground state of the energy in this ansatz.
-In many quantum electrodynamical problems, this linear term, which has zero expectation value in our ansatz, is important.
-To first order, it leads to spontaneous emission of a single photon.
-To second order, it leads to two (virtual) photon processes, such as van der Waals / Casimir-Polder forces on emitters, real photon processes such as two-photon emission, and effective interactions betweeen distinct emitters.
+In many quantum electrodynamical problems, this linear term.
+To second order, it leads to two (virtual) photon processes, such as van der Waals / Casimir-Polder forces on emitters, two-photon spontaneous emission, and effective interactions betweeen distinct emitters.
 Thus, we seek to capture the effect of this term.
-Physically, this term will mix the factorizable ground state of Equation (7) with terms that have virtual excitations of the matter, as well as virtual excitations of the electromagnetic field.
+Physically, this term will mix the factorizable ground state of Eq.
+(5) with states that simultaneously have virtual excitations of the matter and the electromagnetic field.
 The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.
@@
 \begin{equation}
-\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{q=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_q^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_q(\omega_{mn} -\lambda_q)}
+\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{n=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)}
 \end{equation}
 where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.
-In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (6) and (7), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Equation (8).
+In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs.
+(6) and (7), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Eq.
+(8).
 \begin{figure}[t]
@@
 \end{figure}
+
 In what follows, we provide a proof-of-concept demonstration of the efficacy of the variational theory derived here.
@@
 As the cavity is considered for simplicity to be one-dimensional, the electric field is oriented along a single direction, denoted $x$, while the magnetic field is oriented along a direction transverse to both the electric field and the cavity length, denoted $y$.
-
 Working under the long-wavelength or dipole approximation, the Hamiltonian can then be written as:
@@
 \end{equation}
-The fields can be expressed as a mode expansion via:
-\begin{equation}\label{eq:mode_expansion}
-A(z,t) = \sum\limits_{n=1}^{\infty} \sqrt{\frac{\hbar}{2\epsilon_0\omega_n }}(F_n(z)a_ne^{-i\omega_n t}+F^*_n(z)a_n^{\dagger}e^{i\omega_n t}),
-\end{equation}
-with $E(z,t) = -\partial_t A(z,t)$ and $B(z,t)=\partial_z A(z,t)$.
-The $F_n(z)$ are the mode functions of the cavity, normalized such that $\int dz ~|F_n(z)|^2 = 1$.
-For a cavity of length $L$ the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$.
+The fields can be expressed as a mode expansion, where for a cavity of length $L$, the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$.
 The matter Hamiltonian we take to be a multilevel system with an arbitrary number of levels, $N_a$.
@@
 \end{equation}
-The momentum operator, we write in a finite-different representation as
+The momentum operator, we write as 
 \begin{equation}\label{eq:momentum_operator}
@@
 In other words: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.
-Equation \ref{eq:TRK_sum_rule} is the Thomas-Reiche-Kuhn (TRK) sum rule \cite{ciuti2005quantum}.
 Since the TRK is based on a full electronic real-space description, this sum rule does not rigorously apply to a discrete-level system.
 However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot.
-As we shall show later in the text however, it does place a bound on how strong the effect of the $A\Jadd{\cdot}p$ term can be on the total energy.
+It thus places a bound on how strong the effect of the $A\Jadd{\cdot}p$ term can be.
 The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.
 
-The detailed calculations of the energies of states via the formalism introduced here will be shown in a longer paper.
+The detailed calculations of the energies of states via the formalism introduced here will be shown elsewhere.
 Here, we state the main results.
-Using an essentially one-dimensional version of Equation (7), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations.
+Using an essentially one-dimensional version of Eq.
+(7), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations.
 In the absence of correlations, we found that the ground state energy satisfies
 \begin{equation}
-E = E_g + \frac{1}{2}\sum\limits_{n=1}^{\infty}(\omega_n - \omega_n^0),
+E = E_g + \frac{1}{2}\sum\limits_{n=1}^{\infty}(\hbar\omega_n - \hbar\omega_n^0),
 \end{equation}
 where the $\omega_n$ are found in our framework and $\omega_n^0 = \frac{n\pi c}{L}$.
-This is a remarkable formula, as it says that in the absence of correlations, the energy of the system is the Casimir energy of the system.
+This is a remarkable result, as it says that in the absence of correlations, the energy of the system is the Casimir energy of the system.
 In particular, it has long been known that when two conducting plates are placed near each other, there is a Casimir energy associated with the fact that the zero-point energy of the nearby plates is different than the zero-point energy of plates infinitely apart.
@@
 This logic can be applied to any arrangement of macroscopic polarizable objects.
-What is remarkable about the result of Equation (13) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
+What is remarkable about the result of Eq.
+(13) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
 
-In the presence of correlations induced by the $A \cdot p$ term, in addition to Equation (13), we must add a contribution of the form of Equation (8), specialized to the case of an emitter in a one-dimensional cavity.
-We can actually apply the correlation correction to excited states as well, by using the second-order perturbation theory formula for the energy shift of excited states due to the $A \cdot p$ term, using the same electron and photon orbitals and frequencies as derived in Equations (6) and (7).
-It is worthwhile to note that for this one-dimensional case, as we show in a longer paper, we arrive at semi-analytic results for all variational energies that we calculate.
+In the presence of correlations  we must add to the energy a contribution of the form of Eq.
+(8), specialized to the case of an emitter in a one-dimensional cavity.
+We can actually apply the correlation correction to excited states as well, by using the second-order perturbation theory formula for the energy shift of excited states due to the $A \cdot p$ term, using the same electron and photon orbitals and frequencies as derived in Eqs.
+(6) and (7).
+It is worthwhile to note that for this one-dimensional case, as we show elsewhere, we arrive at semi-analytic results for all variational energies that we calculate.
 
@@
 Thus, the two lowest eigenvalues are similar to the bare ground and excited states of the emitter with zero photons in the cavity.
-In blue is our estimate for these two energies, in orange is the exact result by numerical diagonalization of the Hilbert space, and in green is the result obtained by perturbation theory in terms of the bare atomic and cavity parameters.
-One immediately sees that the variational result is quite accurate when compared to numerical diagonalization, much more so than perturbation theory, which completely fails for the first excited state.
+We immediately see that the variational result is quite accurate when compared to numerical diagonalization, much more so than perturbation theory, which completely fails for the first excited state.
 
-The reason it fails for the first excited state more than for the ground state is that the first excited state is nearly resonant with the ground state, leading to a very large negative contribution from the $A\cdot p$ of nearly 2 eV, which is far larger than the spacing of the bare emitter levels.
-On the other hand, the variational estimate finds no such large negative energy shift, and leads to an energy gap between the first two levels which is similar to the bare gap, and in agreement with numerical diagonalization.
-The reason for this behavior is that the effect of the plasma term in Equation (7) is to blue-shift all of the photon frequencies.
+Importantly, the reason perturbation theory fails much more for the first excited state more than for the ground state is that the first excited state is nearly resonant with the ground state, leading to a very large negative contribution from the $A\cdot p$ of nearly 2 eV, which is far larger than the spacing of the bare emitter levels.
+On the other hand, the variational estimate from our formalism finds no such large negative energy shift, and leads to an energy gap between the first two levels which is similar to the bare gap, and in agreement with numerical diagonalization.
+The reason for this behavior is that the effect of the plasma term in Eq.
+(7) is to blue-shift all of the photon frequencies.
 In particular, for the largest coupling considered in Figure 2, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition.
+
 This result very clearly demonstrates not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime.
@@
 These shifts not only lead to changes in the transition frequencies of the emitter, but the position dependence of these energy shifts also implies forces on the emitter, typically called Casimir-Polder forces.
-Such Casimir-Polder forces are typically calculated using the celebrated Lifshitz theory, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Equation (8) using bare atomic and photonic properties.
+Such Casimir-Polder forces are typically calculated using the celebrated Lifshitz theory, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq.
+(8) using bare atomic and photonic properties.
 Thus, our calculations, which find the energy shifts non-perturbatively, lead to a consequently non-perturbative theory of Casimir-Polder forces.
@@
 In Figure 3, we consider the ground state of the QED Hamiltonian, calculated through the procedure above, but now extended to three and four level systems.
-In all cases, the parameter $R$ which controls the strength of the momentum matrix elements and the correlation energy, is chosen to saturate the TRK sum rule.
-We find that the accuracy of our approach is not compromised by having additional energy levels in the Hamiltonian, indicating the potential generality of our approach.
+In all cases, the parameter $R$ is chosen to saturate the TRK sum rule.
+We find that the accuracy of our approach is not compromised by having additional energy levels in the Hamiltonian, indicating the potential of our approach.
 
-Finally, we note that Equation (8), which incorporates correlations, can also be employed self-consistently, by adding it to the Lagrange function of Equation (5), and minimizing the result with respect to electronic and photonic properties.
+Finally, we note that Eq.
+(8), which incorporates correlations, can also be employed self-consistently, by adding it to the Lagrange function of Eq.
+(5), and minimizing the result with respect to electronic and photonic properties.
 The set of equations that arise are:
-
+%\vspace{-1.5cm}
 \begin{widetext}
@@
 \end{widetext}
-These two equations are highly general, providing real-space information about the electronic and photonic properties of the QED ground state for a many-electron system in a many-mode system in three dimensions.
+Eqs.
+(13) and (14) are quite general, providing real-space information about the electronic and photonic properties of the QED ground state for a many-electron system in a many-mode system in three dimensions.
+With the advent of ultra-strong coupling and deep-strong coupling in QED systems, we could in the future use this non-perturbative real-space knowledge of how matter affect photons to design a photonic mode atom-by-atom.
 
-In summary, we have developed a variational approximation to the ground state of quantum electrodynamical problems and have benchmarked it against the multi-mode and multi-level Rabi model, which has no analytical solutions, and whose accurate numerical diagonalization requires a very high-dimensional Hilbert space.
+%In summary, we have developed a variational approximation to the ground state of quantum electrodynamical problems and have benchmarked it against the multi-mode and multi-level Rabi model, which has no analytical solutions, and whose accurate numerical diagonalization requires a very high-dimensional Hilbert space.
 Despite our ansatz being centered on the assumption of weak correlations, with correlations from the $Ap$ term being treated perturbatively, we find that our approximation does very well in predicting the ground state energy, even when the strength of the correlations saturates the TRK sum rule.
-With the advent of ultra-strong coupling and deep-strong coupling being realizable in quantum electrodynamical systems, one may use this real-space knowledge of how matter affect photons to design a photonic mode atom-by-atom.
 
 %In what follows, we add this energy correction $\delta E$ to the expectation value of the energy in the ground state, with the orbitals and eigenvalues as variational parameters.
-As a result, the orbitals and eigenvalues obtained will be different from Equation (11) and (14), and this difference will be small in the limit of weak correlations.
+As a result, the orbitals and eigenvalues obtained will be different from Eq.
+(11) and (14), and this difference will be small in the limit of weak correlations.
 Strictly speaking, this approach is only justified for weak correlations, but can be applied to systems with strong-correlations as is often done with self-consistent methods.
@@
 
-\section{Acknowledgements}
+%\section{Acknowledgements}
 % Pri, Johannes: Please insert any other relevant acknowledgement information here!
-We thank Prof.
+%We thank Prof.
 Joel Yuen-Zhou (University of California San Diego), Prof.

### 2018-10-16 supp.tex

@@
 
-\title{Supplementary information for: Variational theory of the ground state of quantum electrodynamics}
+\title{Supplementary information: Variational theory of the ground state of quantum electrodynamics}
 
@@
 \begin{equation}
-H_{el} = \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext} \right)\psi(x) + \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
+H_{el} = \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext}\Jadd{(x)} \right)\psi(x) + \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
 \end{equation}
 where $v_{ext}$ is the one-body potential, $V(x-x')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.
+\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}
 
@@
 \begin{equation}
-H_{int} = \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(x) + \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(x).
+H_{int} = \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\Jadd{(x)}\cdot\nabla +  \nabla \cdot \mathbf{A}\Jadd{(x)})\psi(x) + \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\Jadd{(x)}\psi\Jadd{(x)}\mathbf{A}^2(x).
 \end{equation}
@@
 \end{widetext}
+\Jadd{(say where the $\lambda$ is coming from)}
 
@@
 
-The minimization with respect to the electron orbital $\psi_i^*$ yields the equation:
+The minimization with respect to the electron orbital $\psi_i^*$ \Jadd{($\delta\mathcal{L}/\delta\psi_i^*=0$) }yields the equation:
 % \begin{widetext}
@@
 \begin{align}
-\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) &+  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)  = E_i\psi_i(x).
+\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) &+  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n\Jadd{(x)}|^2\psi_i(x)  = E_i\psi_i(x).
 \end{align}
@@
 In many quantum electrodynamical problems, this linear term is important.
-To first order, it leads to spontaneous emission of a photon.
-To second order, it leads to van der Waals / Casimir-Polder forces on emitters, as well as effective interactions betweeen distinct emitters.
+To first order, it leads to spontaneous emission of a \Jadd{single} photon.
+To second order, it leads to \Jadd{two-photon processes, such as} van der Waals / Casimir-Polder forces on emitters, as well as effective interactions betweeen distinct emitters.
 Thus, we seek to capture the effect of this term.
@@
 \end{equation}
-where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$,$\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.
+where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.
 In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (11) and (14), and then all the information necessary to calculate the correlation energy is present.
@@
 
-The derivative of $\delta E$ with respect to an occupied electron orbital $k$ is given by
+The derivative of $\delta E$ with respect to an occupied electron orbital $k$ is given by \Jadd{(here you have the y, and x bold, best if also do so above)}
 \begin{equation}

### 2018-10-17 paper4pg.tex
small corrections

@@
 Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED) \cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,kockum2018ultrastrong}.
-Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems \cite{hagenmuller2010ultrastrong,scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems \cite{benz2016,chikkaraddy2016}.
+Ultra-strong, or even deep-strong coupling has been observed in systems involving electromagnetic cavities coupled to superconducting qubits \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems \cite{hagenmuller2010ultrastrong,scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems \cite{benz2016,chikkaraddy2016}.
 Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect, and many more.
@@
 Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
-Through "reduced quantity theories" such as quantum electrodynamical density functional theory (QEDFT) \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}, one is now able to calculate observables in large molecules coupled to realistic optical cavities  \cite{flick2017c}.
+Through reduced quantity theories such as quantum electrodynamical density functional theory (QEDFT) \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}, one is now able to calculate observables in large molecules coupled to realistic optical cavities  \cite{flick2017c}.
 
 In this letter, we propose a variational framework to analyze complex light-matter systems from first principles.
-Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of functionals to describe the ground state, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the use of the long-wavelength (dipole) approximation.
-A variational framework, as we shall show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
-Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom.
-A variational framework also allows compact semi-analytic formulas to describe complex systems which may allow the development of functionals for use in QEDFT.
+Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise, such as the shortage of exchange-correlation functionals to describe the ground state, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the use of the long-wavelength (dipole) approximation.
+A variational framework flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
+Beyond these advantages, a variational framework also allows conceptual insights into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom.
+A variational framework also allows compact semi-analytic formulas to describe complex systems which can drive the development of functionals for use in QEDFT.
 %In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved.
@@
 Motivated by all of these potential advantages, we now develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states.
-This ansatz $-$ reminiscent to, but qualitatively distinct from, the Hartree-Fock ansatz of electronic structure theory $-$ leads to coupled eigen-equations describing the ground-state of the light-matter system.
+This ansatz $-$ reminiscent to, but qualitatively distinct from, the Hartree-Fock ansatz~\cite{szabo1989} of electronic structure theory $-$ leads to coupled eigen-equations describing the ground-state of the light-matter system.
 We apply our ansatz to describe ground and excited states in a multi-level emitter coupled to many photonic modes.
@@
 
-The QED Hamiltonian  $H = H_{mat}+H_{em}+H_{int}$ where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter.
+The QED Hamiltonian is given by $H = H_{mat}+H_{em}+H_{int}$ where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter.
 The matter Hamiltonian takes the form:
@@
 \end{figure}
-Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$.
-This renders the free electromagnetic Hamiltonian as
+Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$ renders the free electromagnetic Hamiltonian as
 \begin{equation}
@@
 where $\epsilon$ and $\mu$ represent a non-dispersive and positive dielectric and magnetic background that the matter and photon occupy.
-For cases we consider in this work, these will be taken to be unity, but we leave them here for generality.
+For cases we consider in this work, these will be taken to be unity.
 
@@
 With the Hamiltonian described, we now move to develop a variational theory of the ground state.
-In the variational theorem, we choose some ansatz $|\Omega\rangle$ for the ground state of $H$.
+In the variational theorem, we choose an ansatz $|\Omega\rangle$ for the ground state of $H$.
 The variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy.
@@
 \begin{equation}
-|\Omega\rangle = \left( \prod\limits_n c_n^{\dagger}|0_n\rangle,\right) \otimes \left(\bigotimes_i|0_i\rangle\right),
+|\Omega\rangle = \left( \prod\limits_n c_n^{\dagger}|0_n\rangle\right) \otimes \left(\bigotimes_i|0_i\rangle\right),
 \end{equation}

### 2018-10-17 paper4pg.tex
some more corrections, I think we need some citations for the casimir related discussion

@@
 We enforce that the matter and photon remain normalized by constructing the Lagrange function:
-\begin{widetext}
-\begin{equation}
-\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle - \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right),
-\end{equation}
-\end{widetext}
+\begin{align}
+&\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle \\
+&- \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right),\nonumber
+\end{align}
 with the $E_n$ and $\lambda_i$ being the Lagrange multipliers that enforce the normalization conditions.
@@
 Immediately, one notices that the term in the interaction Hamiltonian linear in the vector potential makes no contribution to the expectation value of the ground state of the energy in this ansatz.
-In many quantum electrodynamical problems, this linear term.
-To second order, it leads to two (virtual) photon processes, such as van der Waals / Casimir-Polder forces on emitters, two-photon spontaneous emission, and effective interactions betweeen distinct emitters.
+To second order, it leads to two (virtual) photon processes, such as van der Waals / Casimir-Polder forces on emitters, two-photon spontaneous emission, and effective interactions betweeen distinct emitters \Jadd{(Nick: here we need citations)}.
 Thus, we seek to capture the effect of this term.
@@
 
-In what follows, we provide a proof-of-concept demonstration of the efficacy of the variational theory derived here.
+In what follows, we provide a proof-of-concept demonstration of the efficiency of the variational theory derived here.
 We consider the QED Hamiltonian corresponding to a single emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction.
@@
 (13) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
-
+\Jadd{(Nick, the upper section would benefit from 1 or 2 citations)}
 In the presence of correlations  we must add to the energy a contribution of the form of Eq.
@@
 These shifts not only lead to changes in the transition frequencies of the emitter, but the position dependence of these energy shifts also implies forces on the emitter, typically called Casimir-Polder forces.
-Such Casimir-Polder forces are typically calculated using the celebrated Lifshitz theory, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq.
+Such Casimir-Polder forces are typically calculated using the celebrated Lifshitz theory\Jadd{(Nick, cite)}, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq.
 (8) using bare atomic and photonic properties.
-Thus, our calculations, which find the energy shifts non-perturbatively, lead to a consequently non-perturbative theory of Casimir-Polder forces.
+Thus, our calculations, which find the energy shifts non-perturbatively, lead to a consequently non-perturbative theory of Casimir-Polder forces \Jadd{(Nick, cite)}.
 

### 2018-10-17 paper4pg.tex
...

--- 2018-10-16 paper4pg.tex
@@
 We find a compact semi-analytic formula which describes ground and excited state energies to less than 1\% error in all regimes of coupling parameters allowed by sum rules.
-Moreover, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as "the Casimir energy of a single atom in a cavity".
+Moreover, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
 Our method should give rise to highly accurate descriptions of ground-state phenomena in general QED systems, such as Casimir forces, Lamb shifts, and other fluctational electrodynamical phenomena.
@@
 Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
-Through "reduced quantity theories" such as quantum electrodynamical density functional theory (QEDFT) \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}, one is now able to calculate observables in large molecules coupled to realistic optical cavities  \cite{flick2017c}.
+Through `reduced quantity theories' such as quantum electrodynamical density functional theory (QEDFT) \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}, one is now able to calculate observables in large molecules coupled to realistic optical cavities  \cite{flick2017c}.
 
-In this letter, we propose a variational framework to analyze complex light-matter systems from first principles.
+In this Letter, we establish a variational framework to analyze complex light-matter systems from first principles.
 Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of functionals to describe the ground state, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the use of the long-wavelength (dipole) approximation.

### 2018-10-19 paper4pg.tex
following most of chris recommendations

@@
 \begin{abstract}
- The almost routine ability to achieve ultra-strong coupling between light and matter promises to bring about new means to control  material properties, new concepts for manipulating light at the atomic scale, and fundamentally new insights into quantum electrodynamics (QED).
+ The ability to achieve ultra-strong coupling between light and matter promises to bring about new means to control  material properties, new concepts for manipulating light at the atomic scale, and fundamentally new insights into quantum electrodynamics (QED).
 Thus, there is a need to develop quantitative theories of QED phenomena in complex electronic and photonic systems.
-In this letter, we develop a variational theory of general non-relativistic QED systems of coupled light and matter.
+In this Letter, we develop a variational theory of general non-relativistic QED systems of coupled light and matter.
 Essential to our ansatz is the notion an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling.
@@
 We find a compact semi-analytic formula which describes ground and excited state energies to less than 1\% error in all regimes of coupling parameters allowed by sum rules.
-Moreover, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
+Additionally, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
 Our method should give rise to highly accurate descriptions of ground-state phenomena in general QED systems, such as Casimir forces, Lamb shifts, and other fluctational electrodynamical phenomena.
@@
 \begin{equation}
-|\Omega\rangle = \left( \prod\limits_n c_n^{\dagger}|0_n\rangle\right) \otimes \left(\bigotimes_i|0_i\rangle\right),
+|\Omega\rangle = \left( \prod\limits_n c_n^{\dagger}|0_n\rangle\right) \otimes \left(\bigotimes_i|0_i\rangle\right).
 \end{equation}
@@
 \begin{align}
-&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi^*_j(x')\psi_j(x')\psi_i(x) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi_j(x')\psi_j(x)\psi_i(x')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)   = E_i\psi_i(x).
+&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi^*_j(x')\psi_j(x')\psi_i(x) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi_j(x')\psi_j(x)\psi_i(x')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)   = E_i\psi_i(x),
 \end{align}
@@
 \begin{equation}
-\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{n=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)}
+\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{n=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)},
 \end{equation}
@@
 Since the TRK is based on a full electronic real-space description, this sum rule does not rigorously apply to a discrete-level system.
-However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot.
+However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot \Jadd{exist physically}.
 It thus places a bound on how strong the effect of the $A\Jadd{\cdot}p$ term can be.
@@
 These shifts not only lead to changes in the transition frequencies of the emitter, but the position dependence of these energy shifts also implies forces on the emitter, typically called Casimir-Polder forces.
-Such Casimir-Polder forces are typically calculated using the celebrated Lifshitz theory\Jadd{(Nick, cite)}, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq.
+Such Casimir-Polder forces are often calculated using the celebrated Lifshitz theory\Jadd{(Nick, cite)}, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq.
 (8) using bare atomic and photonic properties.
@@
 (5), and minimizing the result with respect to electronic and photonic properties.
-The set of equations that arise are:
+The set of equations that arise are \Jadd{(Nick, should we remove the argument in F(x) in the following equation?)}:
 %\vspace{-1.5cm}

### 2018-10-21 paper4pg.tex
restructuring of figures and end, shortened paper to 4.25 pg

@@
 \begin{figure}[t]
-\includegraphics[width=5cm]{figure2new.pdf}
+\includegraphics[width=9cm]{Figure2and3combined.pdf}
 \caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} Lowest two eigenvalues as a function of coupling for the multi-mode Rabi model describing the coupling of an emitter to a one-dimensional cavity.
@@
 (8).
-\begin{figure}[t]
-\includegraphics[width=8.4cm]{figure3new.pdf}
-\caption{\textbf{Variational theory of ground state energy: accuracy as a function of number of emitter levels.} (Left) Ground-state energy of the one-dimensional multi-mode Rabi model as a function of the dimensionless coupling for two (top), three (middle), and four (bottom) level emitter systems with momentum matrix elements chosen to saturate the Thomas-Reiche-Kuhn sum rule.
+%\begin{figure}[t]
+%\includegraphics[width=8.4cm]{figure3new.pdf}
+%\caption{\textbf{Variational theory of ground state energy: accuracy as a function of number of emitter levels.} (Left) Ground-state energy of the one-dimensional multi-mode Rabi model as a function of the dimensionless coupling for two (top), three (middle), and four (bottom) level emitter systems with momentum matrix elements chosen to saturate the Thomas-Reiche-Kuhn sum rule.
 (Right) Ratio of the energy calculated with the variational theory and with numerical diagonalization.
@@
 }
-\label{fig:ansatz}
-\end{figure}
+%\label{fig:ansatz}
+%\end{figure}
 
@@
 
-In Figure 3, we consider the ground state of the QED Hamiltonian, calculated through the procedure above, but now extended to three and four level systems.
+%In Figure 3, we consider the ground state of the QED Hamiltonian, calculated through the procedure above, but now extended to three and four level systems.
 In all cases, the parameter $R$ is chosen to saturate the TRK sum rule.
@@
 
-Finally, we note that Eq.
+%Finally, we note that Eq.
 (8), which incorporates correlations, can also be employed self-consistently, by adding it to the Lagrange function of Eq.
@@
 The set of equations that arise are \Jadd{(Nick, should we remove the argument in F(x) in the following equation?)}:
-%\vspace{-1.5cm}
-\begin{widetext}
-\begin{align}
-&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(x).
-\end{align}
-\end{widetext}
+%%\vspace{-1.5cm}
+%\begin{widetext}
+%\begin{align}
+%&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(x).
+%\end{align}
+%\end{widetext}
+%
+%\begin{widetext}
+%\begin{equation}
+%\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_q^2} \right)\right)\mathbf{F}_q = -\frac{e^2\hbar}{2m^2\epsilon_0c^2}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_{mn}-\omega_{q}}\mathbf{j}_{nm}.
+%\end{equation}
+%\end{widetext}
+%Eqs.
+(13) and (14) are quite general, providing real-space information about the electronic and photonic properties of the QED ground state for a many-electron system in a many-mode system in three dimensions.
 
-\begin{widetext}
-\begin{equation}
-\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_q^2} \right)\right)\mathbf{F}_q = -\frac{e^2\hbar}{2m^2\epsilon_0c^2}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_{mn}-\omega_{q}}\mathbf{j}_{nm}.
-\end{equation}
-\end{widetext}
-Eqs.
-(13) and (14) are quite general, providing real-space information about the electronic and photonic properties of the QED ground state for a many-electron system in a many-mode system in three dimensions.
 With the advent of ultra-strong coupling and deep-strong coupling in QED systems, we could in the future use this non-perturbative real-space knowledge of how matter affect photons to design a photonic mode atom-by-atom.

### 2018-10-21 paper4pg.tex
some more edits. only a small few remaining; working on it now

@@
 Additionally, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
-Our method should give rise to highly accurate descriptions of ground-state phenomena in general QED systems, such as Casimir forces, Lamb shifts, and other fluctational electrodynamical phenomena.
+Our method should give rise to highly accurate descriptions of ground-state phenomena in general QED systems, such as Casimir forces, Lamb shifts, spontaneous emission, and other fluctational electrodynamical phenomena.
  \end{abstract}
@@
 Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom.
-A variational framework also allows compact semi-analytic formulas to describe complex systems which may allow the development of functionals for use in QEDFT.
+A variational framework also allows compact semi-analytic formulas to describe complex systems which may assist the development of functionals for use in QEDFT.
  %In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved.
@@
 We apply our ansatz to describe ground and excited states in a multi-level emitter coupled to many photonic modes.
-We find that for light-matter couplings that respect sum rules, our theory finds the ground state energy to a remarkable accuracy of about 99\%, even in deeply non-perturbative coupling regimes.
+We find that for light-matter couplings that respect sum rules, our method yields ground and excited state energies to a remarkable accuracy of up to 99\%, even in deeply non-perturbative coupling regimes.
 In regimes where our results are accurate, we have found the effective quasiparticle description of the ground state of QED.
@@
 \begin{align}
-H_{el} &= \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext}(x) \right)\psi(x) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
+H_{el} &= \int d^3x ~\psi^{\dagger}(\mathbf{x})\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext}(\mathbf{x}) \right)\psi(\mathbf{x}) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(\mathbf{x})\psi^{\dagger}(\mathbf{x}')V(\mathbf{x}-\mathbf{x}')\psi(\mathbf{x}')\psi(\mathbf{x}),
 \end{align}
-where $v_{ext}$ is the one-body potential, $V(x-x')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.
+where $v_{ext}$ is the one-body potential, $V(\mathbf{x}-\mathbf{x}')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.
 %\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}
@@
 \begin{equation}
-H_{em} = \int d^3x~ \epsilon (\partial_t \mathbf{A}(x))^2 + \mathbf{A}(x)\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(x)),
+H_{em} = \int d^3x~ \epsilon (\partial_t \mathbf{A}(\mathbf{x}))^2 + \mathbf{A}(\mathbf{x})\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(\mathbf{x})),
 \end{equation}
@@
 %\begin{align}
-%H_{int} = &\frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(x) + \nonumber \\ &\frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(x).
+%H_{int} = &\frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(\mathbf{x})(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(\mathbf{x}) + \nonumber \\ &\frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(\mathbf{x}).
 %\end{align}
 \begin{align}
-H_{int} &= \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}(x)\cdot\nabla +  \nabla \cdot \mathbf{A}(x))\psi(x) \nonumber \\ &+ \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}(x)\psi(x)\mathbf{A}^2(x).
+H_{int} &= \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(\mathbf{x})(\mathbf{A}(\mathbf{x})\cdot\nabla +  \nabla \cdot \mathbf{A}(\mathbf{x}))\psi(\mathbf{x}) \nonumber \\ &+ \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}(\mathbf{x})\psi(\mathbf{x})\mathbf{A}^2(\mathbf{x}).
 \end{align}
@@
 This Hamiltonian, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$.
-The electron field operator takes the form $\psi(x) = \sum_n \psi_n(x)c_n$.
+The electron field operator takes the form $\psi(\mathbf{x}) = \sum_n \psi_n(\mathbf{x})c_n$.
 The $c_n$ is an annihilation operator for an electron corresponding to state $n$.
-The electromagnetic field operator takes the form $\mathbf{A}(x) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(x)a_i+\mathbf{F}^*_i(x)a^{\dagger}_i\right)$, where the $a_i^{(\dagger)}$ annihilate (create) a photon in mode $i$.
+The electromagnetic field operator takes the form $\mathbf{A}(\mathbf{x}) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(\mathbf{x})a_i+\mathbf{F}^*_i(\mathbf{x})a^{\dagger}_i\right)$, where the $a_i^{(\dagger)}$ annihilate (create) a photon in mode $i$.
 In the electromagnetic field operator, we parameterize not only by the mode functions but also by modal frequencies.
-The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}.$
+The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}$ \cite{joannopoulos2011photonic}.
 
@@
 \begin{align}
-&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi^*_j(x')\psi_j(x')\psi_i(x) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\psi_j(x')\psi_j(x)\psi_i(x')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)   = E_i\psi_i(x),
+&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
 \end{align}
@@
 \end{equation}
-where $\omega_p^2 = \frac{e^2}{m\epsilon_0}\sum\limits_{n=1}^N |\psi_n(x)|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located.
+where $\omega_p^2 = \frac{e^2}{m\epsilon_0}\sum\limits_{n=1}^N |\psi_n(\mathbf{x})|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located.
+Equations (6) and (7) are main results and can be used to describe ultra-strongly coupled systems in three dimensions, in an arbitrary photonic system, and with multi-electron matter.
+These equations can also be generalized to find excited states by replacing Equation (4) with a suitable ansatz for an excited state and modifying the derivations that lead to Equations (6) and (7) accordingly.
 
-Immediately, one notices that the term in the interaction Hamiltonian linear in the vector potential makes no contribution to the expectation value of the ground state of the energy in this ansatz.
-To second order, it leads to two (virtual) photon processes, such as van der Waals / Casimir-Polder forces on emitters, two-photon spontaneous emission, and effective interactions betweeen distinct emitters \Jadd{(Nick: here we need citations)}.
+Immediately, we notice that the term in the interaction Hamiltonian linear in the vector potential makes no contribution to the expectation value of the ground state of the energy in this ansatz.
+To second order, it leads to virtual photon processes, such as Lamb shifts, whose position dependence gives rise to van der Waals and Casimir-Polder forces on emitters \cite{scheel2009macroscopic}.
 Thus, we seek to capture the effect of this term.
@@
 The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.
+We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations.
 \begin{figure}[t]
-\includegraphics[width=9cm]{Figure2and3combined.pdf}
-\caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} Lowest two eigenvalues as a function of coupling for the multi-mode Rabi model describing the coupling of an emitter to a one-dimensional cavity.
-The variational results (blue) agree to within roughly a percent with the result derived by numerical diagonalization (orange) retaining up to four photons in the Hilbert space.
-The result from perturbation theory (green) disagrees strongly with the other methods of calculation for the first excited state due to a near-resonance between the bare transition energy and the bare lowest cavity mode.}
+\includegraphics[width=8cm]{Figure2and3combined.pdf}
+\caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} (a) Lowest few energy levels of a two (top), three (middle), and four (bottom) level system embedded in the middle of a one-dimensional cavity.
+The results of our variational method (blue) are compared to perturbation theory (orange), as well as numerical diagonalization (red) with the Fock space truncated to fifty cavity modes with no more than four photons.
+(b) Mechanism of overestimation of couplings and resonances in perturbation theory: the modes derived from the variational theorem are always suppressed in the vicinity of the emitter, and this self-consistently decreases the coupling between the emitter and the field.
+}
 \label{fig:ansatz}
 \end{figure}
+
 We capture the effect of correlations perturbatively.
-In other words, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential.
+For the example of the ground state, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential.
 That correction is given by
@@
 (8).
+By taking $m$ as an ansatz for an excited state, correlation corrections to excited states can also be found.
 %\begin{figure}[t]
@@
 
-In what follows, we provide a proof-of-concept demonstration of the efficiency of the variational theory derived here.
+In what follows, we provide a proof-of-concept demonstration of the accuracy and content of the variational theory derived here.
 We consider the QED Hamiltonian corresponding to a single emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction.
@@
 The matter system we describe as an $N_a$ site system.
-This could be considered as a simple model of a molecule within a tight-binding description.
+This could be considered as a simplified model of a molecule within a tight-binding description.
 Thus we parameterize the general family of matter Hamiltonians as:
@@
 Since the TRK is based on a full electronic real-space description, this sum rule does not rigorously apply to a discrete-level system.
-However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot \Jadd{exist physically}.
-It thus places a bound on how strong the effect of the $A\Jadd{\cdot}p$ term can be.
+However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot exist physically.
+It thus places a bound on how strong the effect of the $A\cdot p$ term can be.
 The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.
@@
 In particular, it has long been known that when two conducting plates are placed near each other, there is a Casimir energy associated with the fact that the zero-point energy of the nearby plates is different than the zero-point energy of plates infinitely apart.
-This is simply because the electromagnetic mode structure of two nearby plates is very different from that of two infinitely separated plates.
+This is simply because the electromagnetic mode structure of two nearby plates is different from that of two infinitely separated plates.
 This Casimir energy is simply the difference between the interacting and non-interacting zero-point energies.
@@
 (5), and minimizing the result with respect to electronic and photonic properties.
-The set of equations that arise are \Jadd{(Nick, should we remove the argument in F(x) in the following equation?)}:
+The set of equations that arise are \Jadd{(Nick, should we remove the argument in F(\mathbf{x}) in the following equation?)}:
 %%\vspace{-1.5cm}
@@
 %\begin{align}
-%&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(x).
+%&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) +  \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\left(\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) - \psi_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(\mathbf{x}) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(\mathbf{x}).
 %\end{align}

### 2018-10-22 paper4pg.tex
finished another round of edits. 4.5 pages. will give it a once over tomorrow

@@
 
- Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED) \cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,kockum2018ultrastrong}.
-Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems \cite{hagenmuller2010ultrastrong,scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems \cite{benz2016,chikkaraddy2016}.
+ Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED) \cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong}.
+Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems \cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems \cite{benz2016,chikkaraddy2016}.
 Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect, and many more.
-The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems \cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties \cite{orgiu2015}, realizing analogues of nonlinear optical processes with vacuum fluctuations \cite{kockum2017deterministic}, and as probes of exotic phenomena such as the dynamical Casimir effect \cite{ciuti2005quantum}.
+The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems \cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties \cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations \cite{kockum2017deterministic}.
 Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
@@
 \end{equation}
-In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons (see Figure 1).
+In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons (see Fig.
+1).
 Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions.
@@
 \begin{figure}[t]
-\includegraphics[width=8cm]{Figure2and3combined.pdf}
+\includegraphics[width=8.5cm]{Figure2and3combined.pdf}
 \caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} (a) Lowest few energy levels of a two (top), three (middle), and four (bottom) level system embedded in the middle of a one-dimensional cavity.
 The results of our variational method (blue) are compared to perturbation theory (orange), as well as numerical diagonalization (red) with the Fock space truncated to fifty cavity modes with no more than four photons.
+(Inset) Fourth and fifth energy levels over the full range of energies shows a weak anti-crossing behavior.
+Blue denotes variational while red denotes numerical.
 (b) Mechanism of overestimation of couplings and resonances in perturbation theory: the modes derived from the variational theorem are always suppressed in the vicinity of the emitter, and this self-consistently decreases the coupling between the emitter and the field.
@@
 
-The detailed calculations of the energies of states via the formalism introduced here will be shown elsewhere.
+The detailed derivations of the energies of states via the formalism introduced here will be shown elsewhere.
 Here, we state the main results.
@@
 \end{equation}
-where the $\omega_n$ are found in our framework and $\omega_n^0 = \frac{n\pi c}{L}$.
-This is a remarkable result, as it says that in the absence of correlations, the energy of the system is the Casimir energy of the system.
+where the $\omega_n$ are found in our framework and $\omega_n^0 = \frac{n\pi c}{L}$. The modes found in our framework have their frequencies given by
+\begin{equation}
+\cot\left(\frac{\omega_n}{c}d \right)+\cot\left(\frac{\omega_n}{c}(L-d) \right) = \frac{q^2}{4m\epsilon_0\omega_nc}.
+\end{equation}
+The "interacting" field modes are given by
+\begin{align}\label{eq:field_mode}
+N_n^{-1}F_n(z) =~ &\theta(z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}-\right) \nonumber \\
+-&\theta (z-d) \left(\frac{\cos\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\sin\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\ 
++&\theta (d-z) \sin\left(\frac{\omega_n z}{c} \right),
+\end{align}
+with the normalization constant
+\begin{equation}\label{eq:mode_normalization}
+N_n = 2\sqrt{\frac{c}{\omega_n\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
+\end{equation}
+The result of Eq.
+(12) is remarkable, as it says that in the absence of correlations, the energy of the system is the Casimir energy of the system.
 In particular, it has long been known that when two conducting plates are placed near each other, there is a Casimir energy associated with the fact that the zero-point energy of the nearby plates is different than the zero-point energy of plates infinitely apart.
 This is simply because the electromagnetic mode structure of two nearby plates is different from that of two infinitely separated plates.
-This Casimir energy is simply the difference between the interacting and non-interacting zero-point energies.
+This Casimir energy is simply the difference between the interacting and non-interacting zero-point energies \cite{casimir1948attraction,lifshitz1956theory}.
 This logic can be applied to any arrangement of macroscopic polarizable objects.
@@
 (13) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
-\Jadd{(Nick, the upper section would benefit from 1 or 2 citations)}
+
 In the presence of correlations  we must add to the energy a contribution of the form of Eq.
 (8), specialized to the case of an emitter in a one-dimensional cavity.
-We can actually apply the correlation correction to excited states as well, by using the second-order perturbation theory formula for the energy shift of excited states due to the $A \cdot p$ term, using the same electron and photon orbitals and frequencies as derived in Eqs.
+Th interaction energy, given by Eqs.
+(8) and (12) is semi-analytic once the bare emitter states are known, as it is fully specified by Eqs.
+(13-15) once the transcendental equation of Eq.
+(13) is solved.
+We also apply the correlation correction to excited states as well, by using the second-order perturbation theory formula for the energy shift of excited states due to the $A \cdot p$ term, using the same electron and photon orbitals and frequencies as derived in Eqs.
 (6) and (7).
 It is worthwhile to note that for this one-dimensional case, as we show elsewhere, we arrive at semi-analytic results for all variational energies that we calculate.
+In Fig.
+2(a), we show the result of this procedure when applied to calculate ground- and excited- state energies for two-, three-, and four-level systems coupled to a one-dimensional cavity.
+In all cases, the agreement between our variational approach and numerical diagonalization is excellent, suggesting that our variational method is sufficiently flexible to capture ground states and excited states both with and without photonic excitations.
+This is to be contrasted with perturbation theory in the bare matter and photon states, which can both strongly over- and underestimate the energies.
+The most interesting case of disagreement arises in the case of the two-level system (top panel).
+For the two-level system considered here, the variational result agrees very well with numerical diagonalization, while perturbation theory predicts an energy which evolves with coupling in the wrong direction and is off from the true energy by over 100\%.
 
-In Figure 2, we show the result of this procedure when applied to calculate the two lowest energies of a two-level system coupled to a one-dimensional cavity.
-For the two-level system considered here, the energy gap between the bare emitter levels is 0.5 eV, while the lowest bare cavity frequency is 0.61 eV.
-Thus, the two lowest eigenvalues are similar to the bare ground and excited states of the emitter with zero photons in the cavity.
-We immediately see that the variational result is quite accurate when compared to numerical diagonalization, much more so than perturbation theory, which completely fails for the first excited state.
-
-Importantly, the reason perturbation theory fails much more for the first excited state more than for the ground state is that the first excited state is nearly resonant with the ground state, leading to a very large negative contribution from the $A\cdot p$ of nearly 2 eV, which is far larger than the spacing of the bare emitter levels.
+Importantly, the reason perturbation theory fails for first excited state, much more so than for the ground state, is that the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ of nearly 2 eV, which is far larger than the spacing of the bare emitter levels.
 On the other hand, the variational estimate from our formalism finds no such large negative energy shift, and leads to an energy gap between the first two levels which is similar to the bare gap, and in agreement with numerical diagonalization.
@@
 In particular, for the largest coupling considered in Figure 2, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition.
+The plasma term, as shown in Fig.
+2b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (7) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (8).
+This is a so-called light-matter decoupling effect \cite{de2014light}.
+The results of Fig.
+2 very clearly demonstrates not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime.
 
-This result very clearly demonstrates not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime.
-Moreover, our results demonstrate a successful non-perturbative theory of the Lamb shift and consequently Casimir-Polder forces.
+Our results also demonstrate a non-perturbative theory of the Lamb shift and consequently Casimir-Polder forces.
 In particular, it is long known that energy levels of emitters can shift as a result of virtual photon emission and re-absorption.
@@
 These shifts not only lead to changes in the transition frequencies of the emitter, but the position dependence of these energy shifts also implies forces on the emitter, typically called Casimir-Polder forces.
-Such Casimir-Polder forces are often calculated using the celebrated Lifshitz theory\Jadd{(Nick, cite)}, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq.
-(8) using bare atomic and photonic properties.
-Thus, our calculations, which find the energy shifts non-perturbatively, lead to a consequently non-perturbative theory of Casimir-Polder forces \Jadd{(Nick, cite)}.
+Such Casimir-Polder forces are often calculated using the celebrated Lifshitz theory \cite{lifshitz1956theory}, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq.
+(8) using bare atomic and photonic properties \cite{scheel2009macroscopic}.
+Thus our calculation of the energy shifts via Eq.
+(8), which uses the interacting photon modes and frequencies (Eqs.
+(13-15)), which differ greatly from the bare modes and frequencies in the non-perturbative regime, provide a compact and simple extension of the theory of Lamb shifts and Casimir-Polder forces to the non-perturbative regime.
+
+With the advent of ultra-strong coupling and deep-strong coupling in QED systems, the theory posed here, when applied to more complex systems, could form the basis for understanding Casimir phenomena in the ultra-strong coupling regime.
+Additionally, the results developed here could be extended to matter  or photon systems with infinitely many degrees of freedom, also allowing one to capture effects like spontaneous emission in the non-perturbative regime.
+Finally, one could use the non-perturbative real-space knowledge provided by the variational theory of how matter affects photons in order to design a photonic mode atom-by-atom.
 
@@
 (13) and (14) are quite general, providing real-space information about the electronic and photonic properties of the QED ground state for a many-electron system in a many-mode system in three dimensions.
-
-With the advent of ultra-strong coupling and deep-strong coupling in QED systems, we could in the future use this non-perturbative real-space knowledge of how matter affect photons to design a photonic mode atom-by-atom.
 

### 2018-10-22 paper4pg.tex
Another round of edits

@@
 This variational formulation leads to a set of general equations that can describe the ground state of multi-electron systems coupled to many photonic modes in real space.
-As a first step towards an  \textit{ab initio} approach to ground states in QED, we apply our ansatz to a QED system involving a multi-level emitter coupled to many optical modes.
-We find a compact semi-analytic formula which describes ground and excited state energies to less than 1\% error in all regimes of coupling parameters allowed by sum rules.
+As a first step towards a new  \textit{ab initio} approach to ground and excited state energies in QED, we apply our ansatz to describe a multi-level emitter coupled to many optical modes, a system with no analytical solution.
+We find a compact semi-analytical formula which describes ground and excited state energies to less than 1\% error in all regimes of coupling parameters allowed by sum rules.
 Additionally, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
-Our method should give rise to highly accurate descriptions of ground-state phenomena in general QED systems, such as Casimir forces, Lamb shifts, spontaneous emission, and other fluctational electrodynamical phenomena.
+Our method should give rise to highly accurate descriptions of phenomena in general QED systems, such as Casimir forces, Lamb shifts, spontaneous emission, and other fluctational electrodynamical effects.
  \end{abstract}
@@
 Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom.
-A variational framework also allows compact semi-analytic formulas to describe complex systems which may assist the development of functionals for use in QEDFT.
+A variational framework also allows compact semi-analytical formulae to describe complex systems which may assist the development of functionals for use in QEDFT.
  %In the nonperturbative regime of quantum electrodynamics, there are very few models which can be analytically solved.
@@
 
-Motivated by all of these potential advantages, we now develop an ansatz in which the ground state can be considered as a factorizable state of an effective matter and effective photon quasiparticle both in their respective ground states.
-This ansatz $-$ reminiscent to, but qualitatively distinct from, the Hartree-Fock ansatz~\cite{szabo1989} of electronic structure theory $-$ leads to coupled eigen-equations describing the ground-state of the light-matter system.
+Motivated by all of these potential advantages, we now develop an ansatz in which the ground state can be considered as a factorizable state of effective matter and effective photon quasiparticles, both in their respective vacuum states.
+This ansatz $-$ reminiscent to, but qualitatively distinct from, the Hartree-Fock ansatz~\cite{szabo1989} of electronic structure theory $-$ leads to coupled eigen-equations describing ground and excited states of the light-matter system.
 We apply our ansatz to describe ground and excited states in a multi-level emitter coupled to many photonic modes.
@@
 In regimes where our results are accurate, we have found the effective quasiparticle description of the ground state of QED.
-Our findings also furnish a non-perturbative theory of the position-dependent Lamb shifts of ground and excited states that give rise to Casimir-Polder forces.
+Our findings also furnish a non-perturbative theory of the position-dependent energy (Lamb) shifts of ground and excited states that give rise to Casimir-Polder forces.
 
@@
 The electromagnetic field operator takes the form $\mathbf{A}(\mathbf{x}) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(\mathbf{x})a_i+\mathbf{F}^*_i(\mathbf{x})a^{\dagger}_i\right)$, where the $a_i^{(\dagger)}$ annihilate (create) a photon in mode $i$.
-In the electromagnetic field operator, we parameterize not only by the mode functions but also by modal frequencies.
+In the electromagnetic field operator, we parameterize not only by the mode functions but also by mode frequencies.
 The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}$ \cite{joannopoulos2011photonic}.
@@
 \end{equation}
-In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons (see Fig.
+In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $\left(\bigotimes_i|0_i\rangle\right)$ represents a "photonic vacuum" for effectively non-interacting photons (see Fig.
 1).
-Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions.
+Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions,  photon mode functions, and photon mode frequencies.
 Thus, we shall minimize the expectational value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$.
-Note that the electromagnetic field gets two sets of (distinct) variational parameters, $\omega_i$ and $\mathbf{F}_i$.
 We enforce that the matter and photon remain normalized by constructing the Lagrange function:
@@
 \end{align}
-with the $E_n$ and $\lambda_i$ being the Lagrange multipliers that enforce the normalization conditions.
+with the $E_n$ and $\frac{\hbar\lambda_i}{2}$ being the Lagrange multipliers that enforce the normalization conditions.
 Evaluating the expectation value of the Hamiltonian, and minimizing the Lagrange function  immediately yields:
 \begin{align}
-&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
+&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\left(\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\right)\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
 \end{align}
@@
 We see that in addition to the one-body and Hartree-Fock terms for the electrons, the effect of the QED coupling is to add a one-body ponderomotive potential.
-We note that in the often used dipole approximation, the ponderomotive potential has no effect as it becomes independent of position.
+%We note that in the often used dipole approximation, the ponderomotive potential has no effect as it becomes independent of position.
 
@@
 \begin{equation}
-\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_i^2} \right)\right)\mathbf{F}_i = 0,
+\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2(\mathbf{x})}{\omega_i^2} \right)\right)\mathbf{F}_i = 0,
 \end{equation}
-where $\omega_p^2 = \frac{e^2}{m\epsilon_0}\sum\limits_{n=1}^N |\psi_n(\mathbf{x})|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located.
+where $\omega_p^2(\mathbf{x}) = \frac{e^2}{m\epsilon_0}\sum\limits_{n=1}^N |\psi_n(\mathbf{x})|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located.
 Equations (6) and (7) are main results and can be used to describe ultra-strongly coupled systems in three dimensions, in an arbitrary photonic system, and with multi-electron matter.
-These equations can also be generalized to find excited states by replacing Equation (4) with a suitable ansatz for an excited state and modifying the derivations that lead to Equations (6) and (7) accordingly.
+Excited states in this framework can be identified with matter and photon quasiparticle excitations.
 
-Immediately, we notice that the term in the interaction Hamiltonian linear in the vector potential makes no contribution to the expectation value of the ground state of the energy in this ansatz.
-To second order, it leads to virtual photon processes, such as Lamb shifts, whose position dependence gives rise to van der Waals and Casimir-Polder forces on emitters \cite{scheel2009macroscopic}.
+Immediately, we notice that the term in the interaction Hamiltonian linear in the vector potential (the "$A\cdot p$ term") makes no contribution to the expectation value of the ground state of the energy in this ansatz.
+At second order in the $A\cdot p$ term, virtual photon processes arise, such as Lamb shifts, whose emitter-position-dependence gives rise to van der Waals and Casimir-Polder forces  \cite{scheel2009macroscopic}.
 Thus, we seek to capture the effect of this term.
 Physically, this term will mix the factorizable ground state of Eq.
-(5) with states that simultaneously have virtual excitations of the matter and the electromagnetic field.
-The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.
+(4) with states that simultaneously have virtual excitations of the matter and the electromagnetic field.
+The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of QED ground and excited states.
 We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations.
@@
 \begin{equation}
-\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{n=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)},
+\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{\infty}\sum_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)},
 \end{equation}
-where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.
+where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained).
 In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs.
@@
 In what follows, we provide a proof-of-concept demonstration of the accuracy and content of the variational theory derived here.
-We consider the QED Hamiltonian corresponding to a single emitter placed at position $z=d$ in a one-dimensional cavity oriented along the $z$-direction.
+We consider the QED Hamiltonian corresponding to a single emitter placed at position $z=d$ in a one-dimensional cavity whose axis is along the $z$-direction.
 As the cavity is considered for simplicity to be one-dimensional, the electric field is oriented along a single direction, denoted $x$, while the magnetic field is oriented along a direction transverse to both the electric field and the cavity length, denoted $y$.
-Working under the long-wavelength or dipole approximation, the Hamiltonian can then be written as:
+Working under the long-wavelength (dipole) approximation, the Hamiltonian can then be written as:
 \begin{equation}\label{eq:hamiltonian}
@@
 The fields can be expressed as a mode expansion, where for a cavity of length $L$, the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$.
-The matter Hamiltonian we take to be a multilevel system with an arbitrary number of levels, $N_a$.
-The matter system we describe as an $N_a$ site system.
-This could be considered as a simplified model of a molecule within a tight-binding description.
+The matter Hamiltonian we take to be a multilevel system with $N_a$ levels.
+The matter system we describe can thus be mapped $N_a$ site system, which be considered as a simplified model of a molecule within a tight-binding description.
 Thus we parameterize the general family of matter Hamiltonians as:
@@
 Since the TRK is based on a full electronic real-space description, this sum rule does not rigorously apply to a discrete-level system.
-However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot exist physically.
+However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian.
+Thus, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot exist physically.
 It thus places a bound on how strong the effect of the $A\cdot p$ term can be.
 The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.
+These considerations also imply that when we plot observables as a function of parameter, for fixed $R$, we vary the coupling by varying some external sum-rule independent measure such as the charge of the emitter, or the number of emitters collectively coupled to the mode.
+We choose the former.
 
-The detailed derivations of the energies of states via the formalism introduced here will be shown elsewhere.
+The detailed derivations of the energies of states via the formalism introduced here are shown in the Supplementary Materials (SM).
 Here, we state the main results.
@@
 (7), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations.
-In the absence of correlations, we found that the ground state energy satisfies
+In the absence of correlations, we found for example that the energy of any state with no photonic quasiparticles is given by:
 \begin{equation}
-E = E_g + \frac{1}{2}\sum\limits_{n=1}^{\infty}(\hbar\omega_n - \hbar\omega_n^0),
+E_a = E_a + \frac{1}{2}\sum\limits_{n=1}^{\infty}(\hbar\omega_n - \hbar\omega_n^0),
 \end{equation}
-where the $\omega_n$ are found in our framework and $\omega_n^0 = \frac{n\pi c}{L}$. The modes found in our framework have their frequencies given by
+where the $\omega_n$ are found in our framework and $\omega_n^0 = \frac{n\pi c}{L}$.
+Imposing continuity of the modes and discontinuity of their derivatives at $z=d$, the modes found in our framework have their frequencies given by
 \begin{equation}
@@
 \end{equation}
-The "interacting" field modes are given by
+The corresponding "interacting" field mode profiles are given by
 \begin{align}\label{eq:field_mode}
@@
 In particular, it has long been known that when two conducting plates are placed near each other, there is a Casimir energy associated with the fact that the zero-point energy of the nearby plates is different than the zero-point energy of plates infinitely apart.
-This is simply because the electromagnetic mode structure of two nearby plates is different from that of two infinitely separated plates.
+This is because the electromagnetic mode structure of two nearby plates is different from that of two infinitely separated plates.
 This Casimir energy is simply the difference between the interacting and non-interacting zero-point energies \cite{casimir1948attraction,lifshitz1956theory}.
@@
 What is remarkable about the result of Eq.
-(13) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
+(12) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
 
@@
 Th interaction energy, given by Eqs.
-(8) and (12) is semi-analytic once the bare emitter states are known, as it is fully specified by Eqs.
+(8) and (12) is semi-analytical once the bare emitter states are known, as it is fully specified by Eqs.
 (13-15) once the transcendental equation of Eq.
@@
 (6) and (7).
-It is worthwhile to note that for this one-dimensional case, as we show elsewhere, we arrive at semi-analytic results for all variational energies that we calculate.
 In Fig.
@@
 (8), which uses the interacting photon modes and frequencies (Eqs.
-(13-15)), which differ greatly from the bare modes and frequencies in the non-perturbative regime, provide a compact and simple extension of the theory of Lamb shifts and Casimir-Polder forces to the non-perturbative regime.
+(13-15)), which differ greatly from the bare modes and frequencies in the non-perturbative regime, provide a compact, relatively simple, and semi-analytical extension of the theory of Lamb shifts and Casimir-Polder forces to the non-perturbative regime.
 
@@
 
-%\section{Acknowledgements}
-% Pri, Johannes: Please insert any other relevant acknowledgement information here!
-%We thank Prof.
+\section{Acknowledgements}
+We thank Prof.
 Joel Yuen-Zhou (University of California San Diego), Prof.
@@
 Paulson School of Engineering and Applied Sciences.
-We also acknowledge support from the STC Center for Integrated Quantum Materials NSF Grant No.
-DMR-1231319 and from the Harvard John A.
-Paulson School of Engineering and Applied Sciences.
 J.
@@
 FL 997/1-1.
+This work was supported by the DOE Photonics at Thermodynamic Limits Energy Frontier Research Center under grant no.
+DE-SC0019140.
+
+%We also acknowledge support from the STC Center for Integrated Quantum Materials NSF Grant No. DMR-1231319 and from the Harvard John A. Paulson School of Engineering and Applied Sciences.
 

### 2018-10-22 paper4pg.tex
1. adding labels for equations and the figures, this way the hyperlink in the document is working 2. otherwise some small changes and changed order in acknowledgements

@@
 In this Letter, we develop a variational theory of general non-relativistic QED systems of coupled light and matter.
-Essential to our ansatz is the notion an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling.
+Essential to our ansatz is the notion \Jadd{of} an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling.
 This variational formulation leads to a set of general equations that can describe the ground state of multi-electron systems coupled to many photonic modes in real space.
@@
 
- Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED) \cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong}.
-Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits \cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules \cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems \cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities \cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems \cite{benz2016,chikkaraddy2016}.
-Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect, and many more.
-The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems \cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties \cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations \cite{kockum2017deterministic}.
+ Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED)~\cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong}.
+Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
+Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect\Jadd{(cite)}, and many more.
+The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
 Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
-Through `reduced quantity theories' such as quantum electrodynamical density functional theory (QEDFT) \cite{ruggenthaler2014,pellegrini2015,flick2015,dimitrov2017,flick2018,flick2018b,schaefer2018}, one is now able to calculate observables in large molecules coupled to realistic optical cavities  \cite{flick2017c}.
+Through `reduced quantity theories' such as quantum electrodynamical density functional theory (QEDFT)~\cite{tokatly2013,ruggenthaler2014,flick2015,dimitrov2017,flick2018,flick2018b}, one is now able to calculate observables in large molecules coupled to realistic optical cavities~\cite{pellegrini2015,flick2017c}.
 
  In this Letter, we establish a variational framework to analyze complex light-matter systems from first principles.
-Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of functionals to describe the ground state, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the use of the long-wavelength (dipole) approximation.
+Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of \Jadd{simple exchange-correlation} functionals to describe the ground state \Jadd{energy, as well as other more involved observables}, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the \Jadd{often} use of the long-wavelength (dipole) approximation.
 A variational framework, as we shall show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
@@
 \end{align}
-where $v_{ext}$ is the one-body potential, $V(\mathbf{x}-\mathbf{x}')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.
+where $v_{ext}$ is the one-body \Jadd{external} potential, $V(\mathbf{x}-\mathbf{x}')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.
 %\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}
@@
 \caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons.
-(Right) Quasiparticle description of the coupled system as a factorable state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
+(Right) Quasiparticle description of the coupled system as a factorizablee state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
 \label{fig:ansatz}
@@
 
-This Hamiltonian, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$.
+\Jadd{The full} Hamiltonian \Jadd{${H}$}, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$.
 The electron field operator takes the form $\psi(\mathbf{x}) = \sum_n \psi_n(\mathbf{x})c_n$.
@@
 |\Omega\rangle = \left( \prod\limits_n c_n^{\dagger}|0_n\rangle\right) \otimes \left(\bigotimes_i|0_i\rangle\right).
+\label{eq:ansatz}
 \end{equation}
 In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $\left(\bigotimes_i|0_i\rangle\right)$ represents a "photonic vacuum" for effectively non-interacting photons (see Fig.
-1).
+~\ref{fig:ansatz}).
 Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions,  photon mode functions, and photon mode frequencies.
-Thus, we shall minimize the expectational value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$.
+Thus, we shall minimize the expectation value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$.
 We enforce that the matter and photon remain normalized by constructing the Lagrange function:
@@
 &\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\left(\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\right)\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
+\label{eq:mhf-electron}
 \end{align}
@@
 \left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2(\mathbf{x})}{\omega_i^2} \right)\right)\mathbf{F}_i = 0,
+\label{eq:mhf-photon}
 \end{equation}
 where $\omega_p^2(\mathbf{x}) = \frac{e^2}{m\epsilon_0}\sum\limits_{n=1}^N |\psi_n(\mathbf{x})|^2$ is a position-dependent squared-plasma frequency which will push the photon orbitals out of the region where the emitter is located.
-Equations (6) and (7) are main results and can be used to describe ultra-strongly coupled systems in three dimensions, in an arbitrary photonic system, and with multi-electron matter.
+Equations (\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}) are main results and can be used to describe ultra-strongly coupled systems in three dimensions, in an arbitrary photonic system, and with multi-electron matter.
 Excited states in this framework can be identified with matter and photon quasiparticle excitations.
@@
 Thus, we seek to capture the effect of this term.
-Physically, this term will mix the factorizable ground state of Eq.
-(4) with states that simultaneously have virtual excitations of the matter and the electromagnetic field.
+Physically, this term will mix the factorizable ground state of Eq.~(\ref{eq:ansatz}) with states that simultaneously have virtual excitations of the matter and the electromagnetic field.
 The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of QED ground and excited states.
@@
 (b) Mechanism of overestimation of couplings and resonances in perturbation theory: the modes derived from the variational theorem are always suppressed in the vicinity of the emitter, and this self-consistently decreases the coupling between the emitter and the field.
-}
-\label{fig:ansatz}
+\Jadd{($R_m\rightarrow R$, zoom more in the inset)}  }
+\label{fig:results}
 \end{figure}
@@
 \delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{i=1}^{\infty}\sum_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_i^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_i(\omega_{mn} -\omega_i)},
+\label{eq:e-shift}
 \end{equation}
@@
 In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs.
-(6) and (7), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Eq.
-(8).
+(\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}), and then the electron energies and orbitals as well as the photon frequencies and orbitals, are plugged into Eq.
+(\ref{eq:e-shift}).
 By taking $m$ as an ansatz for an excited state, correlation corrections to excited states can also be found.
@@
 \begin{equation}\label{eq:matter_hamiltonian}
-H_{\text{matter}} = \sum\limits_{i=1}^{{N_a-1}} V_i|i\rangle\langle i|+t(|i\rangle\langle i+1|+|i+1\rangle\langle i|) 
+H_{\text{matter}} = \sum\limits_{i=1}^{{N_a-1}} V_i|i\rangle\langle i|+t(|i\rangle\langle i+1|+|i+1\rangle\langle i|).
 \end{equation}
@@
 Using an essentially one-dimensional version of Eq.
-(7), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations.
+\Jadd{(\ref{eq:mhf-electron}) and }(\ref{eq:mhf-photon}), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations.
 In the absence of correlations, we found for example that the energy of any state with no photonic quasiparticles is given by:
@@
 E_a = E_a + \frac{1}{2}\sum\limits_{n=1}^{\infty}(\hbar\omega_n - \hbar\omega_n^0),
+\label{eq:casimir}
 \end{equation}
@@
 \cot\left(\frac{\omega_n}{c}d \right)+\cot\left(\frac{\omega_n}{c}(L-d) \right) = \frac{q^2}{4m\epsilon_0\omega_nc}.
+\label{eq:cot}
 \end{equation}
@@
 The result of Eq.
-(12) is remarkable, as it says that in the absence of correlations, the energy of the system is the Casimir energy of the system.
+(\ref{eq:casimir}) is remarkable \Jadd{(maybe use a different wording)}, as it says that in the absence of correlations, the energy of the system is the Casimir energy of the system.
 In particular, it has long been known that when two conducting plates are placed near each other, there is a Casimir energy associated with the fact that the zero-point energy of the nearby plates is different than the zero-point energy of plates infinitely apart.
@@
 What is remarkable about the result of Eq.
-(12) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
+(\ref{eq:casimir}) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
 
 In the presence of correlations  we must add to the energy a contribution of the form of Eq.
-(8), specialized to the case of an emitter in a one-dimensional cavity.
-Th interaction energy, given by Eqs.
-(8) and (12) is semi-analytical once the bare emitter states are known, as it is fully specified by Eqs.
-(13-15) once the transcendental equation of Eq.
-(13) is solved.
+(\ref{eq:e-shift}), specialized to the case of an emitter in a one-dimensional cavity.
+The interaction energy, given by Eqs.
+(\ref{eq:e-shift}) and (\ref{eq:casimir}) is semi-analytical once the bare emitter states are known, as it is fully specified by Eqs.
+(\ref{eq:cot}-\ref{eq:mode_normalization}) once the transcendental equation of Eq.
+(\ref{eq:cot}) is solved.
 We also apply the correlation correction to excited states as well, by using the second-order perturbation theory formula for the energy shift of excited states due to the $A \cdot p$ term, using the same electron and photon orbitals and frequencies as derived in Eqs.
-(6) and (7).
+(\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}).
 In Fig.
-2(a), we show the result of this procedure when applied to calculate ground- and excited- state energies for two-, three-, and four-level systems coupled to a one-dimensional cavity.
+~\ref{fig:results}(a), we show the result of this procedure when applied to calculate ground- and excited- state energies for two-, three-, and four-level systems coupled to a one-dimensional cavity.
 In all cases, the agreement between our variational approach and numerical diagonalization is excellent, suggesting that our variational method is sufficiently flexible to capture ground states and excited states both with and without photonic excitations.
@@
 
-Importantly, the reason perturbation theory fails for first excited state, much more so than for the ground state, is that the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ of nearly 2 eV, which is far larger than the spacing of the bare emitter levels.
+Importantly, the reason perturbation theory fails for first excited state, much more so than for the ground state, is that the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ of nearly $2$ eV, which is far larger than the spacing of the bare emitter levels.
 On the other hand, the variational estimate from our formalism finds no such large negative energy shift, and leads to an energy gap between the first two levels which is similar to the bare gap, and in agreement with numerical diagonalization.
 The reason for this behavior is that the effect of the plasma term in Eq.
-(7) is to blue-shift all of the photon frequencies.
-In particular, for the largest coupling considered in Figure 2, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition.
-The plasma term, as shown in Fig.
-2b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (7) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (8).
+(\ref{eq:mhf-electron}) is to blue-shift all of the photon frequencies.
+In particular, for the largest coupling considered in Figure~\ref{fig:results}, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition.
+The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-electron}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}).
 This is a so-called light-matter decoupling effect \cite{de2014light}.
 The results of Fig.
-2 very clearly demonstrates not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime.
+~\ref{fig:results} very clearly demonstrates not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime.
 
@@
 Such Casimir-Polder forces are often calculated using the celebrated Lifshitz theory \cite{lifshitz1956theory}, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq.
-(8) using bare atomic and photonic properties \cite{scheel2009macroscopic}.
+(\ref{eq:e-shift}) using bare atomic and photonic properties \cite{scheel2009macroscopic}.
 Thus our calculation of the energy shifts via Eq.
-(8), which uses the interacting photon modes and frequencies (Eqs.
-(13-15)), which differ greatly from the bare modes and frequencies in the non-perturbative regime, provide a compact, relatively simple, and semi-analytical extension of the theory of Lamb shifts and Casimir-Polder forces to the non-perturbative regime.
+(\ref{eq:e-shift}), which uses the interacting photon modes and frequencies (Eqs.
+(\ref{eq:cot}-\ref{eq:mode_normalization})), which differ greatly from the bare modes and frequencies in the non-perturbative regime, provide a compact, relatively simple, and semi-analytical extension of the theory of Lamb shifts and Casimir-Polder forces to the non-perturbative regime.
 
 With the advent of ultra-strong coupling and deep-strong coupling in QED systems, the theory posed here, when applied to more complex systems, could form the basis for understanding Casimir phenomena in the ultra-strong coupling regime.
-Additionally, the results developed here could be extended to matter  or photon systems with infinitely many degrees of freedom, also allowing one to capture effects like spontaneous emission in the non-perturbative regime.
+Additionally, the results developed here could be extended to matter  or photon systems with infinitely many degrees of freedom, also allowing to capture effects like spontaneous emission in the non-perturbative regime.
 Finally, one could use the non-perturbative real-space knowledge provided by the variational theory of how matter affects photons in order to design a photonic mode atom-by-atom.
@@
 DE-FG02-97ER25308.
+J.
+F.
+acknowledges financial support from the Deutsche Forschungsgemeinschaft under Contract No.
+FL 997/1-1.
 P.
@@
 Paulson School of Engineering and Applied Sciences.
-J.
-F.
-acknowledges financial support from the Deutsche Forschungsgemeinschaft under Contract No.
-FL 997/1-1.
 This work was supported by the DOE Photonics at Thermodynamic Limits Energy Frontier Research Center under grant no.

### 2018-10-22 paper4pg.tex
adding a sentence regarding oep
changes in acknowledgments and two equations were not correctly labeled

@@
 We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations.
+\Jadd{We briefly note that this behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes.
+(this could also be a footnote) }
 \begin{figure}[t]
@@
 The reason for this behavior is that the effect of the plasma term in Eq.
-(\ref{eq:mhf-electron}) is to blue-shift all of the photon frequencies.
+(\ref{eq:mhf-photon}) is to blue-shift all of the photon frequencies.
 In particular, for the largest coupling considered in Figure~\ref{fig:results}, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition.
-The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-electron}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}).
+The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}).
 This is a so-called light-matter decoupling effect \cite{de2014light}.
@@
 F.
-acknowledges financial support from the Deutsche Forschungsgemeinschaft under Contract No.
-FL 997/1-1.
-P.
-N.
-acknowledges start-up funding from the Harvard John A.
-Paulson School of Engineering and Applied Sciences.
+acknowledges financial support from the Deutsche Forschungsgemeinschaft (DFG Forschungsstipendium FL 997/1-1).
 This work was supported by the DOE Photonics at Thermodynamic Limits Energy Frontier Research Center under grant no.

### 2018-10-22 paper4pg.tex
adding some citations to seminal work of feist, herrera, and joel

@@
 Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect\Jadd{(cite)}, and many more.
-The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
+The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017, herrera2016,feist2017,martinez2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
 Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.

### 2018-10-22 paper4pg.tex
replacing one of the ebbesen refs by the most recent one

--- 2018-10-22 paper4pg.tex
@@
  Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED)~\cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong}.
-Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
+Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016,thomas2018}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
 Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect\Jadd{(cite)}, and many more.
-The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
+The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,thomas2018, flick2017, herrera2016,feist2017,martinez2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
 Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
@@
 The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}).
-This is a so-called light-matter decoupling effect \cite{de2014light}.
+This is a so-called light-matter decoupling effect \cite{liberato2014}.
 The results of Fig.

### 2018-10-22 paper4pg.tex
another round, another dollar

--- 2018-10-22 paper4pg.tex
@@
 In this Letter, we develop a variational theory of general non-relativistic QED systems of coupled light and matter.
-Essential to our ansatz is the notion \Jadd{of} an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling.
+Essential to our ansatz is the notion of an effective photonic vacuum whose modes are different than the modes in the absence of light-matter coupling.
 This variational formulation leads to a set of general equations that can describe the ground state of multi-electron systems coupled to many photonic modes in real space.
@@
 Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
-Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect\Jadd{(cite)}, and many more.
+Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect \cite{carmes2018non}, and many more.
 The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
@@
  In this Letter, we establish a variational framework to analyze complex light-matter systems from first principles.
-Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of \Jadd{simple exchange-correlation} functionals to describe the ground state \Jadd{energy, as well as other more involved observables}, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the \Jadd{often} use of the long-wavelength (dipole) approximation.
+Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of simple exchange-correlation functionals to describe the ground state energy, as well as other more involved observables, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the often use of the long-wavelength (dipole) approximation.
 A variational framework, as we shall show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
@@
 \end{align}
-where $v_{ext}$ is the one-body \Jadd{external} potential, $V(\mathbf{x}-\mathbf{x}')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.
+where $v_{ext}$ is the one-body external potential, $V(\mathbf{x}-\mathbf{x}')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.
 %\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}
@@
 \begin{equation}
-H_{em} = \int d^3x~ \epsilon (\partial_t \mathbf{A}(\mathbf{x}))^2 + \mathbf{A}(\mathbf{x})\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(\mathbf{x})),
+H_{em} = \frac{\epsilon_0}{2}\int d^3x~ \epsilon (\partial_t \mathbf{A}(\mathbf{x}))^2 + \mathbf{A}(\mathbf{x})\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(\mathbf{x})),
 \end{equation}
@@
 
-\Jadd{The full} Hamiltonian \Jadd{${H}$}, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$.
+The full Hamiltonian $H$, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$.
 The electron field operator takes the form $\psi(\mathbf{x}) = \sum_n \psi_n(\mathbf{x})c_n$.
@@
 \begin{align}
-&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\left(\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\right)\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
+&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\left(\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\right)\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
 \label{eq:mhf-electron}
@@
 We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations.
-\Jadd{We briefly note that this behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes.
-(this could also be a footnote) }
+\footnote{We briefly note that this behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes.
+}
 \begin{figure}[t]
@@
 (b) Mechanism of overestimation of couplings and resonances in perturbation theory: the modes derived from the variational theorem are always suppressed in the vicinity of the emitter, and this self-consistently decreases the coupling between the emitter and the field.
-\Jadd{($R_m\rightarrow R$, zoom more in the inset)}  }
+}
 \label{fig:results}
@@
 Using an essentially one-dimensional version of Eq.
-\Jadd{(\ref{eq:mhf-electron}) and }(\ref{eq:mhf-photon}), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations.
-In the absence of correlations, we found for example that the energy of any state with no photonic quasiparticles is given by:
+(\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations.
+In the absence of correlations, we found for example that the energy of any matter state $a$ with no photonic quasiparticles is given by:
 \begin{equation}
-E_a = E_a + \frac{1}{2}\sum\limits_{n=1}^{\infty}(\hbar\omega_n - \hbar\omega_n^0),
+E_{a} = E^{0}_a + \frac{1}{2}\sum\limits_{n=1}^{\infty}(\hbar\omega_n - \hbar\omega_n^0),
 \label{eq:casimir}
 \end{equation}
-where the $\omega_n$ are found in our framework and $\omega_n^0 = \frac{n\pi c}{L}$.
+where $E_a^{(0)}$ is the energy of the matter state in the absence of coupling, $\omega_n$ are found in our framework, $\omega_n^0 = \frac{n\pi c}{L}$.
 Imposing continuity of the modes and discontinuity of their derivatives at $z=d$, the modes found in our framework have their frequencies given by
 \begin{equation}
-\cot\left(\frac{\omega_n}{c}d \right)+\cot\left(\frac{\omega_n}{c}(L-d) \right) = \frac{q^2}{4m\epsilon_0\omega_nc}.
+\cot\left(\frac{\omega_n}{c}d \right)+\cot\left(\frac{\omega_n}{c}(L-d) \right) = -\frac{q^2}{m\epsilon_0\omega_nc}.
 \label{eq:cot}
@@
 \begin{align}\label{eq:field_mode}
-N_n^{-1}F_n(z) =~ &\theta(z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}-\right) \nonumber \\
+N_n^{-1}F_n(z) =~ &\theta(z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\
 -&\theta (z-d) \left(\frac{\cos\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\sin\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\ 
@@
 \begin{equation}\label{eq:mode_normalization}
-N_n = 2\sqrt{\frac{c}{\omega_n\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
+N_n = 2\sqrt{\frac{1}{\frac{c}{\omega_n}\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
 \end{equation}
 The result of Eq.
-(\ref{eq:casimir}) is remarkable \Jadd{(maybe use a different wording)}, as it says that in the absence of correlations, the energy of the system is the Casimir energy of the system.
+(\ref{eq:casimir})  says that in the absence of correlations, the energy of the system is the Casimir energy of the system.
 In particular, it has long been known that when two conducting plates are placed near each other, there is a Casimir energy associated with the fact that the zero-point energy of the nearby plates is different than the zero-point energy of plates infinitely apart.
@@
 This logic can be applied to any arrangement of macroscopic polarizable objects.
-What is remarkable about the result of Eq.
+What is notable about the result of Eq.
 (\ref{eq:casimir}) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
@@
 The reason for this behavior is that the effect of the plasma term in Eq.
-(\ref{eq:mhf-photon}) is to blue-shift all of the photon frequencies.
+(\ref{eq:mhf-electron}) is to blue-shift all of the photon frequencies.
 In particular, for the largest coupling considered in Figure~\ref{fig:results}, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition.
-The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}).
+The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-electron}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}).
 This is a so-called light-matter decoupling effect \cite{de2014light}.
@@
 F.
-acknowledges financial support from the Deutsche Forschungsgemeinschaft (DFG Forschungsstipendium FL 997/1-1).
+acknowledges financial support from the Deutsche Forschungsgemeinschaft under Contract No.
+FL 997/1-1.
+P.
+N.
+acknowledges start-up funding from the Harvard John A.
+Paulson School of Engineering and Applied Sciences.
 This work was supported by the DOE Photonics at Thermodynamic Limits Energy Frontier Research Center under grant no.

### 2018-10-22 paper4pg.tex

@@
 The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
+Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
+Through `reduced quantity theories' such as quantum electrodynamical density functional theory (QEDFT)~\cite{tokatly2013,ruggenthaler2014,flick2015,dimitrov2017,flick2018,flick2018b}, one is now able to calculate observables in large molecules coupled to realistic optical cavities~\cite{pellegrini2015,flick2017c}.
+
+Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED)~\cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong}.
+Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016,thomas2018}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
+Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect \cite{carmes2018non}, and many more.
+The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,thomas2018, flick2017, herrera2016,feist2017,martinez2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
 Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
@@
 The reason for this behavior is that the effect of the plasma term in Eq.
-(\ref{eq:mhf-electron}) is to blue-shift all of the photon frequencies.
+(\ref{eq:mhf-photon}) is to blue-shift all of the photon frequencies.
 In particular, for the largest coupling considered in Figure~\ref{fig:results}, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition.
-The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-electron}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}).
-This is a so-called light-matter decoupling effect \cite{de2014light}.
+The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}).
+This is a so-called light-matter decoupling effect \cite{liberato2014}.
 The results of Fig.

### 2018-10-22 paper4pg.tex
changes

@@
 %\label{sec:Introduction}
-
-Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED)~\cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong}.
-Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,hutchison2013,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
-Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect \cite{carmes2018non}, and many more.
-The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,flick2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
-Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
-Through `reduced quantity theories' such as quantum electrodynamical density functional theory (QEDFT)~\cite{tokatly2013,ruggenthaler2014,flick2015,dimitrov2017,flick2018,flick2018b}, one is now able to calculate observables in large molecules coupled to realistic optical cavities~\cite{pellegrini2015,flick2017c}.
 

### 2018-10-22 paper4pg.tex
citing t.shegai and boerjesson

@@
 
-Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED)~\cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong}.
-Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016,thomas2018}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
+Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED)~\cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,baranov2018}.
+Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016,stranius2018,thomas2018}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
 Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect \cite{carmes2018non}, and many more.
@@
 Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
-Through `reduced quantity theories' such as quantum electrodynamical density functional theory (QEDFT)~\cite{tokatly2013,ruggenthaler2014,flick2015,dimitrov2017,flick2018,flick2018b}, one is now able to calculate observables in large molecules coupled to realistic optical cavities~\cite{pellegrini2015,flick2017c}.
+Through `reduced quantity theories' such as quantum electrodynamical density functional theory (QEDFT)~\cite{tokatly2013,ruggenthaler2014,flick2015,dimitrov2017,flick2018,flick2018b}, one is now able to calculate observables in large molecules coupled to realistic optical cavities~\cite{flick2017c, flick2018b,flick2018}.
 

### 2018-10-22 paper4pg.tex
I think we're ready.

@@
  In this Letter, we establish a variational framework to analyze complex light-matter systems from first principles.
-Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of simple exchange-correlation functionals to describe the ground state energy, as well as other more involved observables, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the often use of the long-wavelength (dipole) approximation.
+Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of simple exchange-correlation functionals to describe the ground state energy, as well as other more involved observables, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the common use of the long-wavelength (dipole) approximation.
 A variational framework, as we shall show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
@@
 \caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons.
-(Right) Quasiparticle description of the coupled system as a factorizablee state an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
+(Right) Quasiparticle description of the coupled system as a factorizable state of an effective emitter in its ground state and the vacuum of an effective photonic degree of freedom.}
 \label{fig:ansatz}
@@
 \begin{equation}
-\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2(\mathbf{x})}{\omega_i^2} \right)\right)\mathbf{F}_i = 0,
+\left( \nabla\times\nabla\times - \frac{\omega_i^2}{c^2}\left(1-\frac{\omega_p^2(\mathbf{x})}{\omega_i^2} \right)\right)\mathbf{F}_i = 0,
 \label{eq:mhf-photon}
@@
 The results of our variational method (blue) are compared to perturbation theory (orange), as well as numerical diagonalization (red) with the Fock space truncated to fifty cavity modes with no more than four photons.
-(Inset) Fourth and fifth energy levels over the full range of energies shows a weak anti-crossing behavior.
+(Inset) Fourth and fifth energy levels shows a weak anti-crossing behavior which is well-reproduced by the variational theory.
 Blue denotes variational while red denotes numerical.
@@
 \begin{equation}\label{eq:hamiltonian}
-H = H_{\text{matter}}+\frac{\epsilon_0}{2}\int dz~(E^2+c^2B^2)+\frac{q}{m}A(d)p + \frac{q^2}{2m}A^2(d).
+H = H_{\text{matter}}+\frac{\epsilon_0S}{2}\int dz~(E^2+c^2B^2)+\frac{q}{m}A(d)p + \frac{q^2}{2m}A^2(d),
 \end{equation}
+with the emitter charge now expressed as $q$, $E, B, A$ being the electric field, magnetic field, and vector potential, and $S$ being a normalization area of the cavity in the $xy$ plane.
 The fields can be expressed as a mode expansion, where for a cavity of length $L$, the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$.
 The matter Hamiltonian we take to be a multilevel system with $N_a$ levels.
-The matter system we describe can thus be mapped $N_a$ site system, which be considered as a simplified model of a molecule within a tight-binding description.
+The matter system we describe can thus be mapped to an $N_a$ site system, which be considered as a simplified model of a molecule within a tight-binding description.
 Thus we parameterize the general family of matter Hamiltonians as:
@@
 ~\ref{fig:results}(a), we show the result of this procedure when applied to calculate ground- and excited- state energies for two-, three-, and four-level systems coupled to a one-dimensional cavity.
+The relevant parameters for Fig.
+2(a) are listed in the SM.
 In all cases, the agreement between our variational approach and numerical diagonalization is excellent, suggesting that our variational method is sufficiently flexible to capture ground states and excited states both with and without photonic excitations.

### 2018-10-22 paper4pg.tex
fluctational

@@
 Additionally, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
-Our method should give rise to highly accurate descriptions of phenomena in general QED systems, such as Casimir forces, Lamb shifts, spontaneous emission, and other fluctational electrodynamical effects.
+Our method should give rise to highly accurate descriptions of phenomena in general QED systems, such as Casimir forces, Lamb shifts, spontaneous emission, and other fluctuational electrodynamical effects.
 \end{abstract}

### 2018-10-22 paper4pg.tex
punctutation

@@
 The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of QED ground and excited states.
-We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations.
-\footnote{We briefly note that this behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes.
-}
+We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations \footnote{We briefly note that this behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes.
+}.
 \begin{figure}[t]

### 2018-10-31 supp2.tex
about 80% there...

--- 2018-10-16 supp.tex
@@
-\documentclass[aps,prb,onecolumn,
+\documentclass[aps,prb,onecolumn,preprint,
 	groupedaddress,superscriptaddress,
@@
 
-\title{Supplementary information: Variational theory of the ground state of quantum electrodynamics}
+\title{Supplementary information: Variational theory of non-relativistic quantum electrodynamics}
 
@@
 \maketitle
-In this supplement, we derive a self-consistent set of equations describing the ground state of a general quantum electrodynamical system with the effect of correlations included.
+In this supplement, we derive a self-consistent extension of the equations describing the ground state of a general quantum electrodynamical system with the effect of correlations included.
 We derive them in the case of a multi-electron system with electron-electron interactions in three spatial dimensions with an arbitrary number of photonic modes.
 The resulting equations are very similar in spirit to the equations of quantum electrodynamical density functional theory in an optimized-effective potential scheme, except that the orbitals used in that scheme are Kohn-Sham orbitals, while ours essentially are mean-field theory orbitals.
-The one-dimensional problem studied for concreteness in the main text is a special case of these equations for a single-electron emitter in a finite-dimensional space and in a one-dimensional cavity.
-\section{Equations for the ground state of the coupled light-matter system}
+We then derive the one-dimensional model used in the main text, as well as describe the parameters used in generating the data of Figure (2) of the main test.
 
-The QED Hamiltonian takes the form:
-\begin{equation}
-H = H_{mat}+H_{em}+H_{int},
-\end{equation}
-where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter.
-The matter Hamiltonian takes the form:
-
-%\begin{align}
-%H_{el} = &\int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext} \right)\psi(x) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
-%\end{align}
-\begin{equation}
-H_{el} = \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext}\Jadd{(x)} \right)\psi(x) + \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
-\end{equation}
-where $v_{ext}$ is the one-body potential, $V(x-x')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.
-\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}
-
-Parameterizing the electromagnetic field purely in terms of a vector potential: $\mathbf{E} = -\partial_t\mathbf{A}$ and $\mathbf{B} = \nabla\times\mathbf{A}$.
-This renders the free electromagnetic Hamiltonian as
-\begin{equation}
-H_{em} = \int d^3x~ \epsilon (\partial_t \mathbf{A}(x))^2 + \mathbf{A}(x)\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(x)),
-\end{equation}
-where $\epsilon$ and $\mu$ represent a non-dispersive and positive dielectric and magnetic background that the matter and photon occupy.
-For cases we consider in this work, these will be taken to be unity, but we leave them here for the sake of generality.
-
-The interaction Hamiltonian takes the form:
-%\begin{align}
-%H_{int} = &\frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(x) + \nonumber \\ &\frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(x).
-%\end{align}
-\begin{equation}
-H_{int} = \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\Jadd{(x)}\cdot\nabla +  \nabla \cdot \mathbf{A}\Jadd{(x)})\psi(x) + \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\Jadd{(x)}\psi\Jadd{(x)}\mathbf{A}^2(x).
-\end{equation}
-
-This Hamiltonian, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions $\{\psi_n\}$, and in terms of a set of photonic mode functions $\{\mathbf{F}_i\}$ via mode expansions for the electronic and photonic field operators.
-The electron field operator takes the form:
-\begin{equation}
-\psi(x) = \sum_n \psi_n(x)c_n.
-\end{equation}
-In this second-quantized description, the $c_n$ is an annihilation operator for an electron corresponding to state $n$. The electromagnetic field operator takes the form
-\begin{equation}
-\mathbf{A}(x) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(x)a_i+\mathbf{F}^*_i(x)a^{\dagger}_i\right).
-\end{equation}
-In the electromagnetic field operator, we parameterize not only by the mode functions but also by modal frequencies.
-We have chosen a form for the vector potential operator which mirrors the choice generally adopted in QED.
-The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}.$
-
-With the Hamiltonian set up, we now move to describe a variational theory of the ground state.
-In the variational theorem, we choose some ansatz $|\Omega\rangle$ for the ground state of $H$.
-The variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy.
-Parameterizing the ground state to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy for the chosen family of ground states.
-
-We choose as our ground state
-\begin{equation}
-|\Omega\rangle = |\psi\rangle \otimes |\phi\rangle,
-\end{equation}
-where
-\begin{equation}
-|\psi\rangle = \prod\limits_n c_n^{\dagger}|0_n\rangle,
-\end{equation}
-and where
-\begin{equation}
-|\phi\rangle = \bigotimes_i|0_i\rangle.
-\end{equation}
-In such an ansatz $|\psi\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $|\phi\rangle$ represents a "photonic vacuum" for effectively non-interacting photons.
-Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions and photon mode functions.
-Thus, we shall minimize the expectational value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$.
-Note that the electromagnetic field gets two sets of (distinct) variational parameters, $\omega_i$ and $\mathbf{F}_i$.
-Intuitively, this happens because the electromagnetic field is described by an electric and magnetic field, which are related to $\mathbf{F}_i$ and $\partial_t \mathbf{F}_i \rightarrow \omega_i\mathbf{F}_i$.
-
-The last thing to note before taking the expectation values and minimization is that we should focus on the submanifold of variations that keep the complete set of electron and photon functions normalized.
-This shall be enforced by constructing the Lagrange function:
-\begin{widetext}
-\begin{equation}
-\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle - \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right).
-\end{equation}
-\end{widetext}
-\Jadd{(say where the $\lambda$ is coming from)}
-
-\subsection{System of equations for the ground state}
-We now proceed with the program of evaluating the expectation value and carrying out the functional minimization with respect to all of the free parameters.
-
-The minimization with respect to the electron orbital $\psi_i^*$ \Jadd{($\delta\mathcal{L}/\delta\psi_i^*=0$) }yields the equation:
-% \begin{widetext}
-% \begin{equation}
-% \left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) + \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)  = E_i\psi_i(x).
-% \end{equation}
-% \end{widetext}
-\begin{align}
-\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) &+  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n\Jadd{(x)}|^2\psi_i(x)  = E_i\psi_i(x).
-\end{align}
-This differs from the usual Hartree-Fock equation of electronic theory by the last term on the left-hand-side, representing a polarization induced by the $\mathbf{A}^2$ term.
-
-The minimization with respect to the photonic mode function $\mathbf{F}_n^*$ gives:
-\begin{equation}
-\left(\epsilon\omega_n+\frac{e^2\sum_i|\psi_i(x)|^2}{m\epsilon_0\omega_n} \right)\mathbf{F}_n + \frac{c^2}{\omega_n}\nabla\times\mu^{-1}\nabla\times\mathbf{F}_n = 2\lambda_n\omega_n\mathbf{F}_n.
-\end{equation}
-Meanwhile, the minimization with respect to frequency $\omega_n$ gives:
-\begin{equation}
-\int d^3x\left(\epsilon - \frac{e^2\sum_i|\psi_i(x)|^2}{m\epsilon_0\omega^2_n}\right)|\mathbf{F}_n|^2 - \frac{c^2}{\omega_n^2}\mathbf{F}_n^*\cdot\nabla\times\mu^{-1}\nabla\times\mathbf{F}_n=0.
-\end{equation}
-Equations (12) and (13), to be mutually consistent, force $\lambda_n = \omega_n$, reducing Equation (12) to the equivalent form:
-\begin{equation}
-\nabla\times\mu^{-1}\nabla\times\mathbf{F}_n = \left(\epsilon - \frac{\omega_p^2(x)}{\omega_n^2} \right)\mathbf{F}_n,
-\end{equation}
-where $\omega_p \equiv \frac{e^2}{2m\epsilon_0}\sum_n|\psi_n(x)|^2$ is the position-dependent plasma frequency. Notice that taking the divergence of both sides of Equation (14) requires that
-\begin{equation}
-\nabla\cdot\left(\epsilon - \frac{\omega_p^2(x)}{\omega_n^2} \right)\mathbf{F}_n = 0,
-\end{equation}
-which is a statement of Gauss' law for the electric field.
-
-\section{Lamb shift correction to the equations for the quantum electrodynamical ground state}
+\section{Self-consistent lamb shift correction to the equations for the quantum electrodynamical ground state}
 

### 2018-10-31 supp2.tex
added 1d to supp

@@
 
+\section{Derivation of results for one-dimensional cavity model in the main text}
+
+We seek a description of the ground state of the system as $|\Psi\rangle \approx |\tilde{g}\rangle\otimes|\tilde{0}\rangle + |\delta\rangle,$ where a $\sim$ denotes an effective quantity.
+In other words, we seek a description in which the state is a factorizable state of matter and photon \textit{quasi}particles, up to some correlations quantified by the state $|\delta\rangle$.
+These correlations essentially hybridize $|\tilde{g}\rangle\otimes|\tilde{0}\rangle$ with excitations of the effective emitter and virtual effective photons.
+
+We start our analysis by considering a family of ansatze in which $\delta = 0$ (uncorrelated ansatz).
+The expectation value of the Hamiltonian is %\Jadd{(the matter part is assumed to be a single atom, we could use the density here.)}
+\begin{align}\label{eq:expectation_val1}
+\langle \Psi | H | \Psi \rangle &= \langle \tilde{g} |H_{\text{matter}} | \tilde{g}\rangle \nonumber \\ &+ \frac{\hbar}{4}\int dz ~\sum\limits_{n=1}^{\infty}\left(\omega_n|F_n|^2 - \frac{c^2}{\omega_n}F_n^*\partial_z^2F_n\right) \nonumber \\ &+ \frac{\hbar q^2}{4m\epsilon_0\omega_n}\sum\limits_{n=1}^{\infty} \int dz~\delta(z-d)|F_n|^2
+\end{align}
+Notably, in this ansatz, the term in the Hamiltonian coupling the momentum of the matter to the photon makes no contribution.
+We impose constraints of matter normalization and photon mode normalization by defining a Lagrange function
+\begin{align}\label{eq:lagrange}
+&\mathcal{L}(\Psi, \Psi^*, \{F_n, F_n^*, \omega_n, \lambda_n\}) \equiv \langle \Psi | H | \Psi \rangle \nonumber \\ &- \epsilon(\langle \tilde{g}|\tilde{g}\rangle-1)-\sum\limits_{n=1}^{\infty}\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right).
+\end{align} 
+To find the ground state, we minimize the Lagrange function with respect to the matter orbital $|\tilde{g}\rangle$ and with respect to the mode functions $F_n$.
+The minimization with respect to the matter leads to the trivial equation $H_{\text{matter}} |\tilde{g}\rangle = \epsilon|\tilde{g}\rangle$ which leaves the effective matter ground state as simply the ground state of $H_{\text{matter}}$.
+On the other hand, the minimization with respect to the photon mode functions leads to  the equation
+\begin{equation}\label{eq:maxwell}
+\left(\partial_z^2-\frac{\omega^2_n}{c^2}+2\frac{\omega_n\lambda_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
+\end{equation}
+We may constrain the $\lambda_n$ by differentiating the Lagrange function also with respect to the $\omega_n$. The equation which follows is:
+\begin{equation}\label{eq:other_maxwell}
+\int dz ~\left(|F_n|^2 + \frac{c^2}{\omega^2_n}F_n^*\partial_z^2F_n\right) - \frac{ q^2}{m\epsilon_0\omega^2_n} \int dz~\delta(z-d)|F_n|^2 = 0
+\end{equation}
+Performing $\frac{\omega_n^2}{c^2}\int dz~F_n^*$ on both sides of Equation \ref{eq:maxwell}, and adding this equation to Equation \ref{eq:other_maxwell}, one immediately finds that $\lambda_n = \omega_n$ and that
+\begin{equation}\label{eq:final_maxwell}
+\left(\partial_z^2+\frac{\omega^2_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
+\end{equation}
+
+This is an ordinary second-order differential equation with the conditions that $F_n$ is continuous at $d$ and that its derivative is discontinuous according to 
+\begin{equation}\label{eq:boundary_condition}
+\partial_zF_n\Big|_{z=d^+}-\partial_zF_n\Big|_{z=d^-} = \frac{q^2}{m\epsilon_0 c^2}F_n(d),
+\end{equation}
+in addition to the usual condition of the modes vanishing at the cavity walls $z=0$ and $z=L$.
+It can be shown that the solution to Equation \ref{eq:final_maxwell} satisfying such boundary conditions is:
+\begin{align}\label{eq:field_mode}
+&\theta (z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}-\right) \nonumber \\ 
+-&\theta (z-d) \left(\frac{\cos\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\sin\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\ 
++&\theta (d-z) \sin\left(\frac{\omega_n z}{c} \right)
+\end{align}
+provided that the auxiliary condition
+\begin{equation}\label{eq:mode_transcendental}
+\cot\left(\frac{\omega_n}{c}d \right)+\cot\left(\frac{\omega_n}{c}(L-d) \right) = \frac{q^2}{4m\epsilon_0\omega_nc}
+\end{equation}
+is met. To ensure that the modes are normalized according to the constraint, we have that the solutions of Equation \ref{eq:field_mode} must be multiplied by a normalization factor $N_n$ given by
+\begin{equation}\label{eq:mode_normalization}
+N_n = 2\sqrt{\frac{c}{\omega_n\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
+\end{equation}
+The condition of Equation \ref{eq:mode_transcendental} determines the resonance frequencies of the photon quasiparticle modes. \Jadd{(put eq.12-14 to SI?)}
+
+\subsection{Model parameters for Figure 2}
+
 \bibliographystyle{apsrev4-1}

### 2018-11-06 supp2.tex
re-vamped supp

--- 2018-10-31 supp2.tex
@@
 \maketitle
-In this supplement, we derive a self-consistent extension of the equations describing the ground state of a general quantum electrodynamical system with the effect of correlations included.
-We derive them in the case of a multi-electron system with electron-electron interactions in three spatial dimensions with an arbitrary number of photonic modes.
-The resulting equations are very similar in spirit to the equations of quantum electrodynamical density functional theory in an optimized-effective potential scheme, except that the orbitals used in that scheme are Kohn-Sham orbitals, while ours essentially are mean-field theory orbitals.
-We then derive the one-dimensional model used in the main text, as well as describe the parameters used in generating the data of Figure (2) of the main test.
+In this Supplement, we derive a self-consistent extension of the equations describing the ground state of a general quantum electrodynamical system with the effect of correlations included.
+We derive the equations in the case of a multi-electron system in three spatial dimensions interacting quantum electrodynamically with an arbitrary number of photonic modes.
+The resulting equations are similar in spirit to the equations of quantum electrodynamical density functional theory in an optimized-effective potential scheme, except that the orbitals used in that scheme are Kohn-Sham orbitals, while ours essentially are mean-field orbitals.
+We then derive the one-dimensional model used in the main text, as well as describe the parameters used in generating the data of Fig.
+2 of the main text.
 
@@
 
-In the derivation of Equation (11), it is notable that the term linear in the vector potential vanishes.
+In the derivation of Equations (6) and (7) of the main text, it is notable that the term linear in the vector potential makes no contribution to the expectation value of the Hamiltonian in the ground state.
 In many quantum electrodynamical problems, this linear term is important.
-To first order, it leads to spontaneous emission of a \Jadd{single} photon.
-To second order, it leads to \Jadd{two-photon processes, such as} van der Waals / Casimir-Polder forces on emitters, as well as effective interactions betweeen distinct emitters.
-Thus, we seek to capture the effect of this term.
-Physically, this term will mix the factorizable ground state of Equation (7) with terms that have virtual excitations of the matter, as well as virtual excitations of the electromagnetic field.
-The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.
+To first order, it leads to spontaneous emission of a single photon.
+To second order, it leads to Casimir-Polder forces on emitters, which arise from virtual emission and re-absorption of photons.
+To second order, the term linear in the vector potential also leads to effective interactions betweeen distinct emitters.
+Thus, we must capture the effect of this term.
+Physically, this term mixes the factorizable ground state of Equation (4) of the main text with states that have virtual excitations of the matter, as well as virtual excitations of the electromagnetic field.
+The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the \textit{correlation energy} of the quantum electrodynamical ground state.
 
@@
 \begin{equation}
-\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{q=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_q^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_q(\omega_{mn} -\lambda_q)},
+\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{q=1}^{\infty}\sum_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_q^*(\mathbf{x})\cdot\mathbf{j}_{nm}(\mathbf{x})\Big|^2}{\omega_q(\omega_{mn} -\lambda_q)},
 \end{equation}
-where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.
-In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (11) and (14), and then all the information necessary to calculate the correlation energy is present.
-In what follows, we add this energy correction $\delta E$ to the expectation value of the energy in the ground state, with the orbitals and eigenvalues as variational parameters.
-As a result, the orbitals and eigenvalues obtained will be different from Equation (11) and (14), and this difference will be small in the limit of weak correlations.
+where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals.
+In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (6) and (7) of the main text.
+This non-self-consistent procedure was applied in Fig.
+2 of the main text.
+In what follows, we add this energy correction $\delta E$ to the expectation value of the energy in the ground state self-consistently, with the orbitals and eigenvalues as variational parameters.
+As a result, the orbitals and eigenvalues obtained will be different from Equations (6) and (7) of the main text, this difference being small in the case of weak correlations.
 Strictly speaking, this approach is only justified for weak correlations, but can be applied to systems with strong-correlations as is often done with self-consistent methods.
@@
 
-As the equations which will result from this new expectation value of the energy will be quite complicated, it is useful to discuss what physics should contained by introducing $\delta E$.
-From the standpoint of the electrons, the equation for them should differ from Equation (11) by a term which is proportional to the Casimir-Polder, or van der Waals, potential that the electrons feel.
-These potentials can also be thought of as Lamb shifts due to the QED coupling.
-It is well-known that this potential depends on all occupied and unoccupied electron orbitals and the photon orbitals, as well as their respective eigenvalues.
-In the equation we will derive for the electron orbitals and their energies, there should be a new term relative to Equation (11) which is simply this potential but with self-consistent electron and photon orbitals and energies.
+As the equations which result from self-consistence are complicated, it is useful to discuss what physics should contained by introducing $\delta E$ to the Lagrange function to be minimized.
+From the standpoint of the electrons, the equation for the electron orbitals should differ from Equation (6) of the main text by a term which is proportional to the spatially dependent Lamb shift potential that the electrons feel.
+This Lamb shift is due to virtual emission and re-absorption of photons.
+We note in passing that the gradient of this potential with respect to position gives rise to a force called the Casimir-Polder force.
+It is known that this potential depends on all occupied and unoccupied electron orbitals and the photon orbitals, as well as their respective eigenvalues.
 
-For the photon, the equation will differ from Equation (14) by the introduction of a new term which has the appearance of a source term proportional to a sum over the transition current densities of the electronic system, $\mathbf{j}_{nm}$.
-The weight of these transition current densities will be proportional to the coupling between the current densities and the photonic orbital.
-In other words, this new term expresses a coupling of fields to fluctuating currents associated with the ground state of the QED system.
-
-In the next section, we derive the equations implied by the correction to the energy of Equation (16) and confirm the physical understanding presented in the previous two paragraphs.
+For the photon, the equation will differ from Equation (7) of the main text by the introduction of a term which has the appearance of a source term proportional to a sum over the transition current densities of the electronic system, $\mathbf{j}_{nm}$.
+These transition currents connect occupied and unoccupied electronic orbitals.
+The weight of these transition current densities will be proportional to the coupling between the current densities and the photonic modes.
+In other words, this new term expresses a coupling of the electromagnetic field to fluctuating currents associated with the matter part of the ground state of the QED system.
 
@@
 
-The derivative of $\delta E$ with respect to an occupied electron orbital $k$ is given by \Jadd{(here you have the y, and x bold, best if also do so above)}
+Here, we derive the equations implied by the correction to the energy of Equation (1) and confirm the physical understanding presented in the previous two paragraphs.
+
+The derivative of $\delta E$ with respect to an occupied electron orbital $k$ is given by
 \begin{equation}
@@
 \begin{equation}
-\frac{\partial\delta E}{\partial \mathbf{F_q^*}}=\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot \mathbf{j}_{mn}(\mathbf{y})}{\omega_q(\omega_{mn}-\lambda_q)}\mathbf{j}_{nm}.
+\frac{\partial\delta E}{\partial \mathbf{F}_q^*}=\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot \mathbf{j}_{mn}(\mathbf{y})}{\omega_q(\omega_{mn}-\lambda_q)}\mathbf{j}_{nm}(\mathbf{x}).
 \end{equation}
@@
 \begin{equation}
--\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\Big|\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})\Big|^2}{\omega_q^2(\omega_{mn}-\lambda_q)}
+\frac{\partial\delta E}{\partial\omega_q}=-\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\Big|\int d^3x~\mathbf{F}_q(\mathbf{x})\cdot\mathbf{j}_{mn}(\mathbf{x})\Big|^2}{\omega_q^2(\omega_{mn}-\lambda_q)}
 \end{equation}
 
-Using these derivatives, Equation (11) now becomes:
+Using these derivatives, Equation (6) of the main text is generalized to:
 \begin{align}
-&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(x).
+&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) +  \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\left(\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) - \psi_j^*(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n(\mathbf{x})|^2\psi_i(\mathbf{x}) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(\mathbf{x}).
 \end{align}
 
-Meanwhile, the equation derived by setting to zero the derivative with respect to the photonic orbital now becomes:
+Setting the derivative with respect to photonic modes zero gives the equation:
 \begin{equation}
-\frac{\hbar}{4}\left(\omega_q\mathbf{F}_q + \frac{c^2}{\omega_q}\nabla\times\nabla\times\mathbf{F}_q\right) + \frac{\hbar\sum\limits_{m=1}^{N_{\sigma}}|\psi_m|^2}{4m\epsilon_0\omega_q}+\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{\infty}\frac{\int d^3y ~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_q(\omega_{mn}-\lambda_q)}\mathbf{j}_{nm} = \frac{\hbar\lambda_q}{2}\mathbf{F}_q
+\frac{\hbar}{4}\left(\omega_q\mathbf{F}_q(\mathbf{x}) + \frac{c^2}{\omega_q}\nabla\times\nabla\times\mathbf{F}_q(\mathbf{x})\right) + \frac{\hbar\sum\limits_{m=1}^{N_{\sigma}}|\psi_m(\mathbf{x})|^2}{4m\epsilon_0\omega_q}+\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\int d^3y ~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_q(\omega_{mn}-\lambda_q)}\mathbf{j}_{nm}(\mathbf{x}) = \frac{\hbar\lambda_q}{2}\mathbf{F}_q(\mathbf{x})
 \end{equation}
 
-Setting the derivative with respect to $\omega_q$ to zero gives the equation:
+And setting the derivative with respect to the photonic frequencies to zero gives the equation:
 \begin{equation}
-\frac{\hbar}{4}\left(\int d^3x ~|\mathbf{F}_q|^2 - \frac{c^2}{\omega_q^2}\mathbf{F}^*_q\cdot\nabla\times\nabla\times\mathbf{F}_q \right) - \frac{e^2\hbar}{4m\epsilon_0\omega_q^2}\int d^3x~\sum\limits_{m=1}^{N_{\sigma}} |\psi_m|^2 -\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\Big|\int d^3x~\mathbf{F}_q\cdot\mathbf{j}_{mn}\Big|^2}{\omega_q^2(\omega_{mn}-\lambda_q)}=  0
+\frac{\hbar}{4}\left(\int d^3x ~|\mathbf{F}_q(\mathbf{x})|^2 - \frac{c^2}{\omega_q^2}\mathbf{F}^*_q(\mathbf{x})\cdot\nabla\times\nabla\times\mathbf{F}_q(\mathbf{x}) \right) - \frac{e^2\hbar}{4m\epsilon_0\omega_q^2}\int d^3x~\sum\limits_{m=1}^{N_{\sigma}} |\psi_m(\mathbf{x})|^2 -\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\Big|\int d^3x~\mathbf{F}_q(\mathbf{x})\cdot\mathbf{j}_{mn}(\mathbf{x})\Big|^2}{\omega_q^2(\omega_{mn}-\lambda_q)}=  0
 \end{equation}
-Performing the operation $\frac{1}{\omega_q}\int d^3x \mathbf{F}_q^* \cdot$ on Equation (21), and adding the result to Equation (22), gives the result $\lambda_q = \omega_q$, just as before $\delta E$ was introduced.
-Using this key simplification, along with the definition of the spatially-dependent plasma frequency as $\omega_p^2 = \frac{e^2}{m\epsilon_0}\sum\limits_{m=1}^{N_p}$, we have that
+Performing the operation $\frac{1}{\omega_q}\int d^3x \mathbf{F}_q^*(\mathbf{x})  \cdot$ on Equation (6), and adding the result to Equation (7), gives the result $\lambda_q = \omega_q$, just as before $\delta E$ was introduced.
+Using this key simplification, along with the definition of the spatially-dependent plasma frequency as $\omega_p^2(\mathbf{x}) = \frac{e^2}{m\epsilon_0}\sum\limits_{m=1}^{N_{\sigma}}|\psi_m(\mathbf{x})|^2$, we have that Equation (7) of the main text is generalized to:
 \begin{equation}
-\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2}{\omega_q^2} \right)\right)\mathbf{F}_q = -\frac{e^2\hbar}{2m^2\epsilon_0c^2}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_{mn}-\omega_{q}}\mathbf{j}_{nm}.
+\left( \nabla\times\nabla\times - \left(1-\frac{\omega_p^2(\mathbf{x})}{\omega_q^2} \right)\right)\mathbf{F}_q(\mathbf{x}) = -\frac{e^2\hbar}{2m^2\epsilon_0c^2}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}} \frac{\int d^3y~\mathbf{F}_q(\mathbf{y})\cdot\mathbf{j}_{mn}(\mathbf{y})}{\omega_{mn}-\omega_{q}}\mathbf{j}_{nm}(\mathbf{x}).
 \end{equation}
-Equations (20) and (23), represent main results of this work.
+\textit{Equations (5) and (8) represent main results of this work and provide a general starting point for first-principles analysis of ground states of QED systems in the non-perturbative regime.}
+
+\section{Derivation of results for one-dimensional cavity model in the main text}
+
+In this section, we provide some additional details on the one-dimensional cavity QED model considered in the main text.
+Given the Hamiltonian of Equation (9) in the main text describing the coupling of an emitter to a one-dimensional cavity, with the matter being described by the site model of Equations (10) and (11) of the main text, the expectation value of the Hamiltonian according to the ansatz of Equation (4) in the main text is given by
+\begin{equation}
+\langle \Psi | H | \Psi \rangle &= \langle \tilde{g} |H_{\text{matter}} | \tilde{g}\rangle + \frac{\hbar}{4}\int dz ~\sum\limits_{n=1}^{\infty}\left(\omega_n|F_n|^2 - \frac{c^2}{\omega_n}F_n^*\partial_z^2F_n\right) + \frac{\hbar q^2}{4m\epsilon_0\omega_n}\sum\limits_{n=1}^{\infty} \int dz~\delta(z-d)|F_n|^2
+\end{equation}
+In this equation, $|\tilde{g}\rangle$ is the ground state of the effective matter part of the Hamiltonian, and $H_{\mathrm{matter}}$ is the Hamiltonian of Equation (10) of the main text.
+We impose constraints of matter normalization and photon mode normalization by defining a Lagrange function
+\begin{align}\label{eq:lagrange}
+\mathcal{L}(|\tilde{g}\rangle, \langle \tilde{g}|,\epsilon, \{F_n, F_n^*, \omega_n, \lambda_n\}) \equiv \langle \Psi | H | \Psi \rangle - \epsilon(\langle \tilde{g}|\tilde{g}\rangle-1)-\sum\limits_{n=1}^{\infty}\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right).
+\end{equation} 
+To find the ground state, we minimize the Lagrange function with respect to the matter orbital $|\tilde{g}\rangle$ and with respect to the mode functions $F_n$.
+The minimization with respect to the matter leads to the trivial equation $H_{\text{matter}} |\tilde{g}\rangle = \epsilon|\tilde{g}\rangle$ which leaves the effective matter ground state as simply the ground state of $H_{\text{matter}}$.
+On the other hand, the minimization with respect to the photon mode functions leads to  the equation
+\begin{equation}\label{eq:maxwell}
+\left(\partial_z^2-\frac{\omega^2_n}{c^2}+2\frac{\omega_n\lambda_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
+\end{equation}
+We may constrain the $\lambda_n$ by differentiating the Lagrange function with respect to the $\omega_n$. The equation which follows is:
+\begin{equation}\label{eq:other_maxwell}
+\int dz ~\left(|F_n|^2 + \frac{c^2}{\omega^2_n}F_n^*\partial_z^2F_n\right) - \frac{ q^2}{m\epsilon_0\omega^2_n} \int dz~\delta(z-d)|F_n|^2 = 0
+\end{equation}
+Performing $\frac{\omega_n^2}{c^2}\int dz~F_n^*$ on both sides of Equation (11), and adding this equation to Equation (12), one immediately finds that $\lambda_n = \omega_n$ and that
+\begin{equation}\label{eq:final_maxwell}
+\left(\partial_z^2+\frac{\omega^2_n}{c^2}-\frac{q^2 }{m\epsilon_0 c^2}\delta(z-d)\right)F_n  = 0.
+\end{equation}
+
+This is an ordinary second-order differential equation with the conditions that $F_n$ is continuous at $d$ and that its derivative is discontinuous according to 
+\begin{equation}\label{eq:boundary_condition}
+\partial_zF_n\Big|_{z=d^+}-\partial_zF_n\Big|_{z=d^-} = \frac{q^2}{m\epsilon_0 c^2}F_n(d),
+\end{equation}
+in addition to the usual condition of the modes vanishing at the cavity walls $z=0$ and $z=L$. It can be shown that the solution to Equation (13) satisfying such boundary conditions is:
+\begin{align}\label{eq:field_mode}
+&\theta (z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}-\right) \nonumber \\ 
+-&\theta (z-d) \left(\frac{\cos\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\sin\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\ 
++&\theta (d-z) \sin\left(\frac{\omega_n z}{c} \right)
+\end{align}
+provided that the auxiliary condition
+\begin{equation}\label{eq:mode_transcendental}
+\cot\left(\frac{\omega_n}{c}d \right)+\cot\left(\frac{\omega_n}{c}(L-d) \right) = -\frac{q^2}{m\epsilon_0\omega_nc}
+\end{equation}
+is met. To ensure that the modes are normalized according to the constraint, we have that the solutions of Equation (15) must be multiplied by a normalization factor $N_n$ given by
+\begin{equation}\label{eq:mode_normalization}
+N_n = 2\sqrt{\frac{1}{\frac{c}{\omega_n}\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
+\end{equation}
+The condition of Equation \ref{eq:mode_transcendental} determines the resonance frequencies of the photon quasiparticle modes. \Jadd{(put eq.12-14 to SI?)}
+
+\subsection{Model parameters for Figure 2}
+
+Here, we note the parameters used in Fig. 2.
+\begin{enumerate}
+\item{The hopping matrix elements $t$ were taken to be 0.25 eV for the two-, three-, and four-level systems.
+Meanwhile, the on-site energies were taken to be equal on all sites in the two-, three-, and four-level systems.}
+\item{The cavity length was taken to be 1 micron.}
+\item{The area of the cavity in the transverse direction was taken to be 100 nm$^2$.}
+\item{The maximum number of cavity modes retained in the calculations was 100. Our results were converged with respect to the number of cavity modes.}
+\item{In the numerical diagonalization results (red lines of Fig.
+2 of the main text), the Fock space was truncated such that the number of photons retained was no more than four.
+For the largest couplings plotted in Fig.
+2, this was sufficient.
+But for higher couplings, more photons in the numerical diagonalization are needed.
+For four photons and 100 cavity modes coupled to a four-level system, the dimension of the Hilbert space is $4\times 10^8$.
+}
+\item{The largest couplings plotted in Fig.
+2 correspond to either a single emitter with a charge of $200e$, or an ensemble of emitters (as is the case in many ultra-strong coupling experiments) in which there are 40,000 emitters in the cavity.
+}
+\end{enumerate}
 

### 2018-11-06 supp2.tex
fixed compilation

@@
 \begin{align}
-&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) +  \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\left(\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) - \psi_j^*(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n(\mathbf{x})|^2\psi_i(\mathbf{x}) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(\mathbf{x}).
+&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) +  \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\left(\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) - \psi_j^*(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n(\mathbf{x})|^2\psi_i(\mathbf{x}) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  \nonumber \\ &= E_i\psi_i(\mathbf{x}).
 \end{align}
@@
 And setting the derivative with respect to the photonic frequencies to zero gives the equation:
-\begin{equation}
-\frac{\hbar}{4}\left(\int d^3x ~|\mathbf{F}_q(\mathbf{x})|^2 - \frac{c^2}{\omega_q^2}\mathbf{F}^*_q(\mathbf{x})\cdot\nabla\times\nabla\times\mathbf{F}_q(\mathbf{x}) \right) - \frac{e^2\hbar}{4m\epsilon_0\omega_q^2}\int d^3x~\sum\limits_{m=1}^{N_{\sigma}} |\psi_m(\mathbf{x})|^2 -\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\Big|\int d^3x~\mathbf{F}_q(\mathbf{x})\cdot\mathbf{j}_{mn}(\mathbf{x})\Big|^2}{\omega_q^2(\omega_{mn}-\lambda_q)}=  0
-\end{equation}
+\begin{align}
+&\frac{\hbar}{4}\left(\int d^3x ~|\mathbf{F}_q(\mathbf{x})|^2 - \frac{c^2}{\omega_q^2}\mathbf{F}^*_q(\mathbf{x})\cdot\nabla\times\nabla\times\mathbf{F}_q(\mathbf{x}) \right) - \frac{e^2\hbar}{4m\epsilon_0\omega_q^2}\int d^3x~\sum\limits_{m=1}^{N_{\sigma}} |\psi_m(\mathbf{x})|^2 \nonumber \\ &-\frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{m=1}^{N_{\sigma}}\frac{\Big|\int d^3x~\mathbf{F}_q(\mathbf{x})\cdot\mathbf{j}_{mn}(\mathbf{x})\Big|^2}{\omega_q^2(\omega_{mn}-\lambda_q)}=  0
+\end{align}
 Performing the operation $\frac{1}{\omega_q}\int d^3x \mathbf{F}_q^*(\mathbf{x})  \cdot$ on Equation (6), and adding the result to Equation (7), gives the result $\lambda_q = \omega_q$, just as before $\delta E$ was introduced.
@@
 \begin{equation}
-\langle \Psi | H | \Psi \rangle &= \langle \tilde{g} |H_{\text{matter}} | \tilde{g}\rangle + \frac{\hbar}{4}\int dz ~\sum\limits_{n=1}^{\infty}\left(\omega_n|F_n|^2 - \frac{c^2}{\omega_n}F_n^*\partial_z^2F_n\right) + \frac{\hbar q^2}{4m\epsilon_0\omega_n}\sum\limits_{n=1}^{\infty} \int dz~\delta(z-d)|F_n|^2
+\langle \Psi | H | \Psi \rangle = \langle \tilde{g} |H_{\text{matter}} | \tilde{g}\rangle + \frac{\hbar}{4}\int dz ~\sum\limits_{n=1}^{\infty}\left(\omega_n|F_n|^2 - \frac{c^2}{\omega_n}F_n^*\partial_z^2F_n\right) + \frac{\hbar q^2}{4m\epsilon_0\omega_n}\sum\limits_{n=1}^{\infty} \int dz~\delta(z-d)|F_n|^2
 \end{equation}
@@
 We impose constraints of matter normalization and photon mode normalization by defining a Lagrange function
-\begin{align}\label{eq:lagrange}
+\begin{equation}\label{eq:lagrange}
 \mathcal{L}(|\tilde{g}\rangle, \langle \tilde{g}|,\epsilon, \{F_n, F_n^*, \omega_n, \lambda_n\}) \equiv \langle \Psi | H | \Psi \rangle - \epsilon(\langle \tilde{g}|\tilde{g}\rangle-1)-\sum\limits_{n=1}^{\infty}\frac{\hbar\lambda_n}{2}\left( \int dz~|F_n|^2-1\right).
@@
 \end{equation}
-The condition of Equation \ref{eq:mode_transcendental} determines the resonance frequencies of the photon quasiparticle modes. \Jadd{(put eq.12-14 to SI?)}
+The condition of Equation \ref{eq:mode_transcendental} determines the resonance frequencies of the photon quasiparticle modes.
 
@@
 2 correspond to either a single emitter with a charge of $200e$, or an ensemble of emitters (as is the case in many ultra-strong coupling experiments) in which there are 40,000 emitters in the cavity.
-}
+The largest couplings correspond to rather extreme coupling parameters and are shown mostly to demonstrate that our ansatz is quite accurate even in regimes of extremely high coupling.}
 \end{enumerate}

### 2018-11-06 supp2.tex
first set of corrections of the supp2, need to check equations

@@
 We derive the equations in the case of a multi-electron system in three spatial dimensions interacting quantum electrodynamically with an arbitrary number of photonic modes.
-The resulting equations are similar in spirit to the equations of quantum electrodynamical density functional theory in an optimized-effective potential scheme, except that the orbitals used in that scheme are Kohn-Sham orbitals, while ours essentially are mean-field orbitals.
+The resulting equations are similar in spirit to the equations of quantum electrodynamical density-functional theory~\cite{tokatly2013,ruggenthaler2014} in an optimized-effective potential scheme~\cite{pellegrini2015,flick2017c}, except that the orbitals used in that scheme are Kohn-Sham orbitals, while ours essentially are mean-field orbitals.
 We then derive the one-dimensional model used in the main text, as well as describe the parameters used in generating the data of Fig.
@@
 To second order, it leads to Casimir-Polder forces on emitters, which arise from virtual emission and re-absorption of photons.
-To second order, the term linear in the vector potential also leads to effective interactions betweeen distinct emitters.
+\Jrem{To second order} \Jadd{In the same order}, the term linear in the vector potential also leads to effective interactions betweeen distinct emitters.
 Thus, we must capture the effect of this term.
@@
 \end{equation}
-where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals.
+where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$ \Jadd{are current-transition moments}, $\omega_{mn} = \omega_m - \omega_n$ \Jadd{the transition frequencies}, \Jadd{and} $N_{\sigma}$ is the number of occupied orbitals.
 In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (6) and (7) of the main text.
@@
 As the equations which result from self-consistence are complicated, it is useful to discuss what physics should contained by introducing $\delta E$ to the Lagrange function to be minimized.
-From the standpoint of the electrons, the equation for the electron orbitals should differ from Equation (6) of the main text by a term which is proportional to the spatially dependent Lamb shift potential that the electrons feel.
+From the standpoint of the electrons, the equation for the electron orbitals should differ from Equation (6) of the main text by a term which is proportional to the spatially dependent Lamb shift potential \Jadd{(Nick: not clear what you refer to here, we should define the spatially dependent Lamb shift potential)} that the electrons feel.
 This Lamb shift is due to virtual emission and re-absorption of photons.
@@
 But for higher couplings, more photons in the numerical diagonalization are needed.
-For four photons and 100 cavity modes coupled to a four-level system, the dimension of the Hilbert space is $4\times 10^8$.
-}
+For four photons and 100 cavity modes coupled to a four-level system, the dimension of the Hilbert space is \Jadd{ $4\times 1 \times 100 \times 5050 \times 171.700 \times 4.421.275 = 18.392.504$.} \Jadd{Also see Ref.~\cite{flick2017} for more details on the exact numerical diagonalization methods for light-matter coupled problems.}}
 \item{The largest couplings plotted in Fig.

### 2018-11-06 supp2.tex
johannes' comments. he still needs to go through the equations

@@
 To second order, it leads to Casimir-Polder forces on emitters, which arise from virtual emission and re-absorption of photons.
-\Jrem{To second order} \Jadd{In the same order}, the term linear in the vector potential also leads to effective interactions betweeen distinct emitters.
+At the same order, the term linear in the vector potential also leads to effective interactions betweeen distinct emitters.
 Thus, we must capture the effect of this term.
@@
 \end{equation}
-where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$ \Jadd{are current-transition moments}, $\omega_{mn} = \omega_m - \omega_n$ \Jadd{the transition frequencies}, \Jadd{and} $N_{\sigma}$ is the number of occupied orbitals.
+where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$ are transition current densities, $\omega_{mn} = \omega_m - \omega_n$ are transition frequencies, and $N_{\sigma}$ is the number of occupied orbitals.
 In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (6) and (7) of the main text.
@@
 As the equations which result from self-consistence are complicated, it is useful to discuss what physics should contained by introducing $\delta E$ to the Lagrange function to be minimized.
-From the standpoint of the electrons, the equation for the electron orbitals should differ from Equation (6) of the main text by a term which is proportional to the spatially dependent Lamb shift potential \Jadd{(Nick: not clear what you refer to here, we should define the spatially dependent Lamb shift potential)} that the electrons feel.
+From the standpoint of the electrons, the equation for the electron orbitals should differ from Equation (6) of the main text by a potential energy term which corresponds to the spatially dependent Lamb shift  that the electrons feel.
 This Lamb shift is due to virtual emission and re-absorption of photons.
-We note in passing that the gradient of this potential with respect to position gives rise to a force called the Casimir-Polder force.
+We note in passing that the gradient of the Lamb shift with respect to position gives rise to a force called the Casimir-Polder force.
 It is known that this potential depends on all occupied and unoccupied electron orbitals and the photon orbitals, as well as their respective eigenvalues.
@@
 But for higher couplings, more photons in the numerical diagonalization are needed.
-For four photons and 100 cavity modes coupled to a four-level system, the dimension of the Hilbert space is \Jadd{ $4\times 1 \times 100 \times 5050 \times 171.700 \times 4.421.275 = 18.392.504$.} \Jadd{Also see Ref.~\cite{flick2017} for more details on the exact numerical diagonalization methods for light-matter coupled problems.}}
+For four photons and 100 cavity modes coupled to a four-level system, the dimension of the Hilbert space is  $4\times \left( 1 + 100 + 5050 + 171700 + 4421275 \right) = 18.392.504$.
+The four terms in the parentheses correspond to the dimension of the properly symmetrized zero-, one-, two-, three-, and four-photon Hilbert spaces respectively.
+Also see Ref.~\cite{flick2017} for more details on the exact numerical diagonalization methods for light-matter coupled problems.}
 \item{The largest couplings plotted in Fig.

### 2018-11-08 paper4pg.tex
merge.
two small changes

@@
 \end{equation}
-where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained).
+where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons \Jrem{(divided by 2 if spin is retained)}.
 In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs.

### 2018-11-08 supp2.tex
two small changes

@@
 \end{equation}
-where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$ are transition current densities, $\omega_{mn} = \omega_m - \omega_n$ are transition frequencies, and $N_{\sigma}$ is the number of occupied orbitals.
+where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$ are transition current densities, $\omega_{mn} = \omega_m - \omega_n$ are transition frequencies, $N_{\sigma}$ is the number of occupied orbitals, \Jadd{$\textbf{F}_q$ is the cavity mode function, $\omega_q$ is the cavity frequency, and $\lambda_q$ a lagrange multiplier}.
 In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (6) and (7) of the main text.
@@
 \begin{equation}
-\frac{\partial\delta E}{\partial \psi_k^*} = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{nk}(\mathbf{y})}{\omega_q(\omega_{kn}-\lambda_q)}\int d^3x~\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)
+\frac{\partial\delta E}{\partial \psi_k^*} = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{\infty} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{nk}(\mathbf{y})}{\omega_q(\omega_{kn}-\lambda_q)}\int d^3x~\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)
 \end{equation}

### 2018-11-09 paper4pg.tex
incorporated tina's comments. main ready, sm ready, cover ready modulo reviewers 3,7

--- 2018-10-22 paper4pg.tex
@@
 
-The QED Hamiltonian is given by $H = H_{mat}+H_{em}+H_{int}$ where $H_{mat}$ describes the matter in the absence of the quantized electromagnetic field, $H_{em}$ describes the photons in the absence of the matter, and $H_{int}$ describes the coupling between light and matter.
+The QED Hamiltonian is given by $H = H_{\mathrm{mat}}+H_{\mathrm{em}}+H_{\mathrm{int}}$ where $H_{\mathrm{mat}}$ describes the matter in the absence of the quantized electromagnetic field, $H_{\mathrm{em}}$ describes the photons in the absence of the matter, and $H_{\mathrm{int}}$ describes the coupling between light and matter.
 The matter Hamiltonian takes the form:
 \begin{align}
-H_{el} &= \int d^3x ~\psi^{\dagger}(\mathbf{x})\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext}(\mathbf{x}) \right)\psi(\mathbf{x}) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(\mathbf{x})\psi^{\dagger}(\mathbf{x}')V(\mathbf{x}-\mathbf{x}')\psi(\mathbf{x}')\psi(\mathbf{x}),
+H_{\mathrm{el}} &= \int d^3x ~\psi^{\dagger}(\mathbf{x})\left(-\frac{\hbar^2\nabla^2}{2m} + v_{\mathrm{ext}}(\mathbf{x}) \right)\psi(\mathbf{x}) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(\mathbf{x})\psi^{\dagger}(\mathbf{x}')V(\mathbf{x}-\mathbf{x}')\psi(\mathbf{x}')\psi(\mathbf{x}),
 \end{align}
-where $v_{ext}$ is the one-body external potential, $V(\mathbf{x}-\mathbf{x}')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.
+where $v_{\mathrm{ext}}$ is the one-body external potential, $V(\mathbf{x}-\mathbf{x}')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.
 %\Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}
@@
 \begin{equation}
-H_{em} = \frac{\epsilon_0}{2}\int d^3x~ \epsilon (\partial_t \mathbf{A}(\mathbf{x}))^2 + \mathbf{A}(\mathbf{x})\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(\mathbf{x})),
+H_{\mathrm{em}} = \frac{\epsilon_0}{2}\int d^3x~ \epsilon (\partial_t \mathbf{A}(\mathbf{x}))^2 + \mathbf{A}(\mathbf{x})\cdot(\nabla\times\mu^{-1}\nabla\times\mathbf{A}(\mathbf{x})),
 \end{equation}
@@
 %\begin{align}
-%H_{int} = &\frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(\mathbf{x})(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(\mathbf{x}) + \nonumber \\ &\frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(\mathbf{x}).
+%H_{\mathrm{int}} = &\frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(\mathbf{x})(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(\mathbf{x}) + \nonumber \\ &\frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(\mathbf{x}).
 %\end{align}
 \begin{align}
-H_{int} &= \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(\mathbf{x})(\mathbf{A}(\mathbf{x})\cdot\nabla +  \nabla \cdot \mathbf{A}(\mathbf{x}))\psi(\mathbf{x}) \nonumber \\ &+ \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}(\mathbf{x})\psi(\mathbf{x})\mathbf{A}^2(\mathbf{x}).
+H_{\mathrm{int}} &= \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(\mathbf{x})(\mathbf{A}(\mathbf{x})\cdot\nabla +  \nabla \cdot \mathbf{A}(\mathbf{x}))\psi(\mathbf{x}) \nonumber \\ &+ \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}(\mathbf{x})\psi(\mathbf{x})\mathbf{A}^2(\mathbf{x}).
 \end{align}
@@
 &\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle \\
-&- \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right),\nonumber
+&- \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_i \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right),\nonumber
 \end{align}
@@
 \begin{align}
-&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\left(\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\right)\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
+&\left(\frac{\mathbf{p}^2}{2m}+v_{\mathrm{ext}}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\left(\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\right)\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
 \label{eq:mhf-electron}
@@
 The results of our variational method (blue) are compared to perturbation theory (orange), as well as numerical diagonalization (red) with the Fock space truncated to fifty cavity modes with no more than four photons.
-(Inset) Fourth and fifth energy levels shows a weak anti-crossing behavior which is well-reproduced by the variational theory.
+(Inset) The fourth and fifth energy levels show a weak anti-crossing behavior which is well-reproduced by the variational theory.
 Blue denotes variational while red denotes numerical.
@@
 \end{equation}
-where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained).
+where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, and$N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons.
 In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs.
@@
 \end{equation}
-with the emitter charge now expressed as $q$, $E, B, A$ being the electric field, magnetic field, and vector potential, and $S$ being a normalization area of the cavity in the $xy$ plane.
+with the emitter charge now expressed as $q$, $E, B$, and $A$ being the electric field, magnetic field, and vector potential, and $S$ being a normalization area of the cavity in the $xy$ plane.
 The fields can be expressed as a mode expansion, where for a cavity of length $L$, the modes are given by $F_n(z) = \sqrt{\frac{2}{L}}\sin\left(\frac{n\pi z}{L} \right)$ and the corresponding mode frequencies are $\omega_n = \frac{n\pi c}{L}$.
@@
 Thus, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot exist physically.
-It thus places a bound on how strong the effect of the $A\cdot p$ term can be.
+The TRK sum rule places a bound on how strong the effect of the $A\cdot p$ term can be.
 The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.
@@
 What is notable about the result of Eq.
-(\ref{eq:casimir}) is that our result says that the same logic about zero-point energy-differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
+(\ref{eq:casimir}) is it implies that the same logic about zero-point energy differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
 
@@
 These energy shifts, called Lamb shifts, depend on the particular position of the emitter in the photonic structure it is embedded in.
-These shifts not only lead to changes in the transition frequencies of the emitter, but the position dependence of these energy shifts also implies forces on the emitter, typically called Casimir-Polder forces.
-Such Casimir-Polder forces are often calculated using the celebrated Lifshitz theory \cite{lifshitz1956theory}, which is equivalent to a derivation that applies of second-order perturbation theory in the form of Eq.
-(\ref{eq:e-shift}) using bare atomic and photonic properties \cite{scheel2009macroscopic}.
-Thus our calculation of the energy shifts via Eq.
-(\ref{eq:e-shift}), which uses the interacting photon modes and frequencies (Eqs.
-(\ref{eq:cot}-\ref{eq:mode_normalization})), which differ greatly from the bare modes and frequencies in the non-perturbative regime, provide a compact, relatively simple, and semi-analytical extension of the theory of Lamb shifts and Casimir-Polder forces to the non-perturbative regime.
+These shifts not only lead to changes in the transition frequencies of the emitter, but the position dependence of these energy shifts also implies forces on the emitter, often called Casimir-Polder forces.
+Such forces are calculated by applying second-order perturbation theory in the form of Eq.
+(\ref{eq:e-shift}) using \textit{bare} atomic and photonic properties \cite{scheel2009macroscopic}.
+Our calculation of the Lamb shifts via Eq.
+(\ref{eq:e-shift}) says that the shifts result from virtual emission and re-absorption of the photonic quasiparticles (the \textit{interacting} modes), governed by (Eqs.
+(\ref{eq:cot}-\ref{eq:mode_normalization})), which are dependent on properties of the matter.
+As these interacting photon modes differ greatly from the bare modes and frequencies in the non-perturbative regime, Eq.
+(\ref{eq:e-shift}) using interacting modes provides a compact, and conceptually simple extension of the theory of Lamb shifts and Casimir-Polder forces to the non-perturbative regime.
 
-With the advent of ultra-strong coupling and deep-strong coupling in QED systems, the theory posed here, when applied to more complex systems, could form the basis for understanding Casimir phenomena in the ultra-strong coupling regime.
+With the advent of ultra-strong coupling and deep-strong coupling in QED systems, the theory posed here, when applied to more complex systems, could form the basis for understanding Lamb and Casimir phenomena in the ultra-strong coupling regime.
 Additionally, the results developed here could be extended to matter  or photon systems with infinitely many degrees of freedom, also allowing to capture effects like spontaneous emission in the non-perturbative regime.
@@
 %\begin{align}
-%&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) +  \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\left(\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) - \psi_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(\mathbf{x}) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(\mathbf{x}).
+%&\left(\frac{\mathbf{p}^2}{2m}+v_{\mathrm{ext}}(\mathbf{x}) \right)\psi_i(\mathbf{x}) +  \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\left(\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) - \psi_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(\mathbf{x}) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  = E_i\psi_i(\mathbf{x}).
 %\end{align}

### 2018-11-09 submit/supp_110918.tex
submit folder

--- 2018-11-08 supp2.tex
@@
 \end{equation}
-where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$ are transition current densities, $\omega_{mn} = \omega_m - \omega_n$ are transition frequencies, $N_{\sigma}$ is the number of occupied orbitals, \Jadd{$\textbf{F}_q$ is the cavity mode function, $\omega_q$ is the cavity frequency, and $\lambda_q$ a lagrange multiplier}.
+where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$ are transition current densities, $\omega_{mn} = \omega_m - \omega_n$ are transition frequencies, $N_{\sigma}$ is the number of occupied orbitals, the $\textbf{F}_q$ are the photon mode functions, the $\omega_q$ are the photon mode frequencies, and the $\lambda_q$ are Lagrange multipliers expressing the normalization of each photon mode.
+We note that the sum over photon modes $q$ is in principle over all modes but can be restricted to either a finite or countably infinite set of photon modes deemed relevant to the system at hand.
 In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (6) and (7) of the main text.
@@
 \begin{align}
-&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) +  \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\left(\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) - \psi_j^*(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n(\mathbf{x})|^2\psi_i(\mathbf{x}) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  \nonumber \\ &= E_i\psi_i(\mathbf{x}).
+&\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(\mathbf{x}) \right)\psi_i(\mathbf{x}) +  \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\left(\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) - \psi_j^*(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n(\mathbf{x})|^2\psi_i(\mathbf{x}) + \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{\infty} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{ni}(\mathbf{y})}{\omega_q(\omega_{in}-\lambda_q)}\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)  \nonumber \\ &= E_i\psi_i(\mathbf{x}).
 \end{align}
@@
 But for higher couplings, more photons in the numerical diagonalization are needed.
-For four photons and 100 cavity modes coupled to a four-level system, the dimension of the Hilbert space is  $4\times \left( 1 + 100 + 5050 + 171700 + 4421275 \right) = 18.392.504$.
+For four photons and 100 cavity modes coupled to a four-level system, the dimension of the Hilbert space is  $4\times \left( 1 + 100 + 5050 + 171,700 + 4,421,275 \right) = 18,392,504$.
 The four terms in the parentheses correspond to the dimension of the properly symmetrized zero-, one-, two-, three-, and four-photon Hilbert spaces respectively.

### 2018-11-11 paper4pg.tex
minor edits prior to submission.

@@
 Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of simple exchange-correlation functionals to describe the ground state energy, as well as other more involved observables, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the common use of the long-wavelength (dipole) approximation.
-A variational framework, as we shall show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
+A variational framework, as we show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
 Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom.
@@
 \end{equation}
-In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a "filled Fermi sea" for effectively non-interacting electrons, and $\left(\bigotimes_i|0_i\rangle\right)$ represents a "photonic vacuum" for effectively non-interacting photons (see Fig.
+In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a `filled Fermi sea' for effectively non-interacting electrons, and $\left(\bigotimes_i|0_i\rangle\right)$ represents a `photonic vacuum' for effectively non-interacting photons (see Fig.
 ~\ref{fig:ansatz}).
@@
 \end{equation}
-The corresponding "interacting" field mode profiles are given by
+The corresponding `interacting' field mode profiles are given by
 \begin{align}\label{eq:field_mode}
@@
 
-%We also acknowledge support from the STC Center for Integrated Quantum Materials NSF Grant No. DMR-1231319 and from the Harvard John A. Paulson School of Engineering and Applied Sciences.
-
 \bibliographystyle{apsrev4-1}

### 2018-11-14 submit/main.tex
Edits during submission.

--- 2018-11-11 paper4pg.tex
@@
 Although  \textit{ab initio} methods such as QEDFT are exact in principle and provide access to all observables, a number of practical difficulties arise related to: the lack of simple exchange-correlation functionals to describe the ground state energy, as well as other more involved observables, the difficulty of obtaining real-space information about the photons as they are affected by light-matter coupling, the difficulty of handling excited state energies, and the common use of the long-wavelength (dipole) approximation.
-A variational framework, as we show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
+A variational framework, as we shall show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
 Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom.
@@
 
-Immediately, we notice that the term in the interaction Hamiltonian linear in the vector potential (the "$A\cdot p$ term") makes no contribution to the expectation value of the ground state of the energy in this ansatz.
+Immediately, we notice that the term in the interaction Hamiltonian linear in the vector potential (the `$A\cdot p$ term') makes no contribution to the expectation value of the ground state of the energy in this ansatz.
 At second order in the $A\cdot p$ term, virtual photon processes arise, such as Lamb shifts, whose emitter-position-dependence gives rise to van der Waals and Casimir-Polder forces  \cite{scheel2009macroscopic}.
@@
 
-\section{Acknowledgements}
-We thank Prof.
+%\section{Acknowledgements} We thank Prof.
 Joel Yuen-Zhou (University of California San Diego), Prof.
@@
 
+%We also acknowledge support from the STC Center for Integrated Quantum Materials NSF Grant No. DMR-1231319 and from the Harvard John A. Paulson School of Engineering and Applied Sciences.
+
 \bibliographystyle{apsrev4-1}

### 2018-11-16 paper4pg.tex
trimmed, should be good now

@@
 \begin{abstract}
-The ability to achieve ultra-strong coupling between light and matter promises to bring about new means to control  material properties, new concepts for manipulating light at the atomic scale, and fundamentally new insights into quantum electrodynamics (QED).
+The ability to achieve ultra-strong coupling between light and matter promises to bring about new means to control  material properties, new concepts for manipulating light at the atomic scale, and new insights into quantum electrodynamics (QED).
 Thus, there is a need to develop quantitative theories of QED phenomena in complex electronic and photonic systems.
@@
 We find a compact semi-analytical formula which describes ground and excited state energies to less than 1\% error in all regimes of coupling parameters allowed by sum rules.
-Additionally, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
-Our method should give rise to highly accurate descriptions of phenomena in general QED systems, such as Casimir forces, Lamb shifts, spontaneous emission, and other fluctuational electrodynamical effects.
+Additionally, our formulation provides a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
+Our method should thus give rise to highly accurate non-perturbative descriptions of many other phenomena in general QED systems.
 \end{abstract}
@@
 Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED)~\cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,baranov2018}.
-Ultra-strong, or even deep-strong coupling is now regularly observed in systems involving electromagnetic cavities coupled to superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016,stranius2018,thomas2018}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
+Ultra-strong, and even deep-strong coupling has been observed in systems involving superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016,stranius2018,thomas2018}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
 Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect \cite{carmes2018non}, and many more.
-The proposed applications for ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,thomas2018, flick2017, herrera2016,feist2017,martinez2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
-Concomitantly with these exciting experimental developments are also theoretical developments in the study of QED systems \textit{ab initio}.
+Proposed applications of ultra- and deep-strong coupling of light and matter are similarly broad, including simulation of many-body systems~\cite{forn2018ultrastrong}, altering chemical reactivity~\cite{hutchison2012, thomas2016,thomas2018, flick2017, herrera2016,feist2017,martinez2017} and electronic transport properties~\cite{orgiu2015} and realizing analogues of nonlinear optical processes with vacuum fluctuations~\cite{kockum2017deterministic}.
+Concomitantly with these developments are also theoretical developments in the study of QED systems \textit{ab initio}.
 Through `reduced quantity theories' such as quantum electrodynamical density functional theory (QEDFT)~\cite{tokatly2013,ruggenthaler2014,flick2015,dimitrov2017,flick2018,flick2018b}, one is now able to calculate observables in large molecules coupled to realistic optical cavities~\cite{flick2017c, flick2018b,flick2018}.
@@
 
-With the Hamiltonian described, we now move to develop a variational theory of the ground state.
-In the variational theorem, we choose an ansatz $|\Omega\rangle$ for the ground state of $H$.
-The variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy.
-Parameterizing the ground state to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy for the chosen family of ground states.
+Given an ansatz $|\Omega\rangle$ for the ground state of $H$, the variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy.
+Parameterizing $|\Omega\rangle$ to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy within this family of ground states.
 We choose as our ansatz
@@
 \end{equation}
-In such an ansatz, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a `filled Fermi sea' for effectively non-interacting electrons, and $\left(\bigotimes_i|0_i\rangle\right)$ represents a `photonic vacuum' for effectively non-interacting photons (see Fig.
+where, $\prod\limits_n c_n^{\dagger}|0_n\rangle$ represents a `filled Fermi sea' for effectively non-interacting electrons, and $\left(\bigotimes_i|0_i\rangle\right)$ represents a `photonic vacuum' for effectively non-interacting photons (see Fig.
 ~\ref{fig:ansatz}).
-Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all the possibilities for the electron wavefunctions,  photon mode functions, and photon mode frequencies.
-Thus, we shall minimize the expectation value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$.
+Implicitly, this ansatz, once we take the expectation value $\langle \Omega|H|\Omega\rangle$, denotes a family of ansatzes labeled by all possibilities for the electron wavefunctions,  photon mode functions, and photon mode frequencies.
+Thus, we minimize the expectation value with respect to $\psi_n, \psi_n^*, \mathbf{F}_i, \mathbf{F}_i^*$, and $\omega_i$.
 We enforce that the matter and photon remain normalized by constructing the Lagrange function:
@@
 
-Immediately, we notice that the term in the interaction Hamiltonian linear in the vector potential (the "$A\cdot p$ term") makes no contribution to the expectation value of the ground state of the energy in this ansatz.
+Note that term in the interaction Hamiltonian linear in the vector potential (the "$A\cdot p$ term") makes no contribution to the expectation value of the ground state of the energy in this ansatz.
 At second order in the $A\cdot p$ term, virtual photon processes arise, such as Lamb shifts, whose emitter-position-dependence gives rise to van der Waals and Casimir-Polder forces  \cite{scheel2009macroscopic}.
-Thus, we seek to capture the effect of this term.
 Physically, this term will mix the factorizable ground state of Eq.~(\ref{eq:ansatz}) with states that simultaneously have virtual excitations of the matter and the electromagnetic field.
-The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of QED ground and excited states.
-We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations \footnote{We briefly note that this behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes.
+The resulting state is non-factorizable and thus, the term in the $A\cdot p$ term leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of QED ground and excited states.
+We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations \footnote{The behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes.
 }.
@@
 \includegraphics[width=8.5cm]{Figure2and3combined.pdf}
-\caption{\textbf{Variational theory of ground and excited states in the ultra-strong coupling regime of QED.} (a) Lowest few energy levels of a two (top), three (middle), and four (bottom) level system embedded in the middle of a one-dimensional cavity.
-The results of our variational method (blue) are compared to perturbation theory (orange), as well as numerical diagonalization (red) with the Fock space truncated to fifty cavity modes with no more than four photons.
+\caption{\textbf{Variational theory of ground and excited states in non-perturbative QED.} (a) Lowest few energy levels of a two (top), three (middle), and four (bottom) level system embedded in the middle of a one-dimensional cavity.
+The results of the variational method (blue) are compared to perturbation theory (orange), as well as numerical diagonalization (red) with the Fock space truncated to fifty cavity modes with no more than four photons.
 (Inset) The fourth and fifth energy levels show a weak anti-crossing behavior which is well-reproduced by the variational theory.
 Blue denotes variational while red denotes numerical.
-(b) Mechanism of overestimation of couplings and resonances in perturbation theory: the modes derived from the variational theorem are always suppressed in the vicinity of the emitter, and this self-consistently decreases the coupling between the emitter and the field.
+(b) Mechanism of overestimation of couplings and resonances in perturbation theory: modes derived from the variational theorem are suppressed in the vicinity of the emitter, which self-consistently decreases the field-emitter coupling.
 }
@@
 We capture the effect of correlations perturbatively.
-For the example of the ground state, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential.
+For the ground state, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential.
 That correction is given by
@@
 where $R$ is a constant with units of length representing roughly the difference in positions between sites.
-This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the Thomas-Reiche-Kuhn (TRK) sum rule is enforced.
-In other words: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.
-Since the TRK is based on a full electronic real-space description, this sum rule does not rigorously apply to a discrete-level system.
-However, the matrix elements and energy levels of a few-level approximated Hamiltonian  are derived from an underlying real-space (infinite dimensional) Hamiltonian.
+This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the Thomas-Reiche-Kuhn (TRK) sum rule is enforced: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.
+Since the sum rule is based on a full electronic real-space description, it does not rigorously apply to a discrete-level system.
+However, the matrix elements and energy levels of a few-level approximated Hamiltonian are derived from an underlying real-space Hamiltonian.
 Thus, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot exist physically.
@@
 The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.
-These considerations also imply that when we plot observables as a function of parameter, for fixed $R$, we vary the coupling by varying some external sum-rule independent measure such as the charge of the emitter, or the number of emitters collectively coupled to the mode.
-We choose the former.
+These considerations also imply that when we plot observables as a function of some coupling parameter, for fixed $R$, we must vary a parameter which does not explicitly appear in the sum rule such as the charge or the number of emitters in the cavity.
 
@@
 Here, we state the main results.
-Using an essentially one-dimensional version of Eq.
+Using a one-dimensional version of Eq.
 (\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations.
@@
 With the advent of ultra-strong coupling and deep-strong coupling in QED systems, the theory posed here, when applied to more complex systems, could form the basis for understanding Lamb and Casimir phenomena in the ultra-strong coupling regime.
-Additionally, the results developed here could be extended to matter  or photon systems with infinitely many degrees of freedom, also allowing to capture effects like spontaneous emission in the non-perturbative regime.
-Finally, one could use the non-perturbative real-space knowledge provided by the variational theory of how matter affects photons in order to design a photonic mode atom-by-atom.
+Additionally, the results developed here could be applied to capture spontaneous emission in the non-perturbative regime.
+Finally, one could use the non-perturbative real-space knowledge of how matter affects photons in order to design a photonic mode atom-by-atom.
 
@@
 
-\section{Acknowledgements}
-We thank Prof.
+%\section{Acknowledgements}
+%We thank Prof.
 Joel Yuen-Zhou (University of California San Diego), Prof.

### 2018-11-16 submit/main.tex
trimmed, should be good now

--- 2018-11-11 paper4pg.tex
@@
 We find a compact semi-analytical formula which describes ground and excited state energies to less than 1\% error in all regimes of coupling parameters allowed by sum rules.
-Additionally, our formulation provides essentially a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
-Our method should give rise to highly accurate descriptions of phenomena in general QED systems, such as Casimir forces, Lamb shifts, spontaneous emission, and other fluctuational electrodynamical effects.
+Additionally, our formulation provides a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
+Our method should thus enable highly accurate non-perturbative descriptions of many  phenomena in general QED systems.
 \end{abstract}

### 2019-01-22 paper4pg.tex
changes to main text. working on supp changes.

@@
 Recent years have brought an explosion of progress in the study of light-matter interactions in the non-perturbative regime of quantum electrodynamics (QED)~\cite{flick7strong,ruggenthaler2017b,forn2018ultrastrong,baranov2018}.
-Ultra-strong, and even deep-strong coupling has been observed in systems involving superconducting qubits~\cite{blais2004,wallraff2004,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016,stranius2018,thomas2018}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
+Ultra-strong, and even deep-strong coupling has been observed in systems involving superconducting qubits~\cite{blais2004,wallraff2004,forn2010observation,niemczyk2010circuit,yoshihara2017superconducting,forn2017ultrastrong}, large ensembles of molecules~\cite{hutchison2012,coles2014,coles2014b,shalabney2015coherent, thomas2016,ebbesen2016,stranius2018,thomas2018}, Landau level systems ~\cite{scalari2012ultrastrong,zhang2016collective},  quantum wells coupled to cavities ~\cite{todorov2010ultrastrong,geiser2012ultrastrong}, oscillators \cite{markovic2018demonstration}, and even in few-molecule systems ~\cite{benz2016,chikkaraddy2016}.
 Proposals for new platforms of ultra-strong coupling include emitters coupling to highly confined polaritons in metals and polar insulators \cite{rivera2016shrinking}, heavy ions coupled to optical media via the Cerenkov effect \cite{carmes2018non}, and many more.
@@
 Our findings also furnish a non-perturbative theory of the position-dependent energy (Lamb) shifts of ground and excited states that give rise to Casimir-Polder forces.
+The variational method put forth developed in this manuscript is particularly suited for dealing with QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds, and subsequently methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate.
+Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.
 
@@
 where $R$ is a constant with units of length representing roughly the difference in positions between sites.
-This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the Thomas-Reiche-Kuhn (TRK) sum rule is enforced: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states.
+This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the Thomas-Reiche-Kuhn (TRK) sum rule is enforced: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states \cite{cohen1992}.
 Since the sum rule is based on a full electronic real-space description, it does not rigorously apply to a discrete-level system.
@@
 \end{equation}
-where $E_a^{(0)}$ is the energy of the matter state in the absence of coupling, $\omega_n$ are found in our framework, $\omega_n^0 = \frac{n\pi c}{L}$.
+where $E_a^0$ is the energy of the matter state in the absence of coupling, $\omega_n$ are found in our framework, $\omega_n^0 = \frac{n\pi c}{L}$.
 Imposing continuity of the modes and discontinuity of their derivatives at $z=d$, the modes found in our framework have their frequencies given by
@@
 In all cases, the agreement between our variational approach and numerical diagonalization is excellent, suggesting that our variational method is sufficiently flexible to capture ground states and excited states both with and without photonic excitations.
+The accuracy as a function of number of levels suggests that the breakdown of gauge invariance associated with few-level systems is not crucial to the good agreement between variational and numerical results \cite{bernardis2018breakdown}.
 This is to be contrasted with perturbation theory in the bare matter and photon states, which can both strongly over- and underestimate the energies.

### 2019-01-22 supp2.tex
most of supp fixed added

@@
 
-\subsection{Model parameters for Figure 2}
+\subsubsection{Perturbation theory for the one-dimensional cavity model}
 
-Here, we note the parameters used in Fig. 2.
+As discussed in the main text, the variational energies, as well as the energies from numerical diagonalization, are compared against the energies calculated from perturbation theory.
+We provide additional details on the perturbative calculations here.
+What we calculate is the energy shift of a state $|a,0\rangle$, where $a$ is any matter state, and $0$ denotes the photonic vacuum, according to the perturbation Hamiltonian $\frac{q}{m}Ap + \frac{q^2}{2m}A^2$, where $p$ is the momentum operator for the matter defined in the main text, and $A$ is the vector potential operator of the one-dimensional cavity in the absence of the matter.
+In particular, $A(z) = \sum_n\sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L} \right) \left(a_n+a^{\dagger}_n\right)$, with $L$ the length of the cavity and $\omega_n = \frac{n\pi c}{L}$.
+Upon inspection, it is clear the the $A^2$ term leads to an energy shift at first-order in perturbation theory, while the $A\dot p$ term leads to an energy shift at second-order in perturbation theory.
+A straightforward application of perturbation theory leads to the following first-order energy shift from the $A^2$ term:
+\begin{equation}
+\delta E_a = \frac{q^2}{2m}\sum\limits_{n=1}^{N_p} \frac{\hbar}{\epsilon_0\omega_n L}\sin^2\left(\frac{n\pi z}{L}\right),
+\end{equation}
+where $z$ is the location of the matter in the cavity, and $N_p$ is the maximum photon number retained.
+Note that this value is dependent on the maximum photon number (although somewhat weakly) retained due to the logarithmically diverging nature of the sum, and so to compare variational, numerical, and perturbative calculations, the same number of modes are retained in each.
+A similarly straightforward application of perturbation theory leads to the energy shift coming from the $A\dot p$ term being given by:
+\begin{equation}
+\delta E_a = \frac{q^2}{m^2}\sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{\hbar}{\epsilon_0\omega_n L} \frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L} \right)}{\hbar(\omega_a - \omega_b - \omega_n)},
+\end{equation}
+where $b$ denotes a matter state different from $a$, and $p_{ba} = \langle b|p|a\rangle$.
+
+\subsection{Model parameters for Figure 2 and details of numerical diagonalization}
+
+Here, we note the parameters used in Fig. 2, as well as some details regarding the numerical diagonalization used to assess the validity of the variational approach advanced in the main text.
+
 \begin{enumerate}
@@
 
+Regarding the numerical diagonalization, we numerically implement the Hamiltonian of Eq.
+(9) of the main text by building the matter operators ($H_0$ and $p$ as defined in the main text) in the basis of states corresponding to the tensor product of any matter state, and any photon state having four photons or less .
+Details regarding the ordering convention for the properly symmetrized multi-photon states are provided in Ref.~ \cite{flick2017} for example.
+Photon operators ($a$, $a^{\dagger}$) are constructed in this basis, and used to construct field operators such as $A$ in the basis of bare cavity modes (the usual sine modes).
+In particular, $A(z) = \sum_n\sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L} \right) \left(a_n+a^{\dagger}_n\right)$, with $L$ the length of the cavity and $\omega_n = \frac{n\pi c}{L}$.
+Diagonalization is performed using standard sparse eigendecomposition routines (such as those implemented in MATLAB).
+
 \bibliographystyle{apsrev4-1}

### 2019-01-26 paper4pg.tex
reply to refs done. manuscript edits done.

@@
 Our findings also furnish a non-perturbative theory of the position-dependent energy (Lamb) shifts of ground and excited states that give rise to Casimir-Polder forces.
-The variational method put forth developed in this manuscript is particularly suited for dealing with QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds, and subsequently methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate.
-Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.
+\textcolor{blue}{The variational method put forth developed in this manuscript is particularly suited for dealing with QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds, and subsequently methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate.} Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.
 
@@
 By taking $m$ as an ansatz for an excited state, correlation corrections to excited states can also be found.
+\textcolor{blue}{In the Supplementary Information, we derive a set of equations for the matter orbitals and photonic mode functions which self-consistently takes into account the correlation energy associated with Eq.
+~\ref{eq:e-shift}.
+These equations take into account the spatially varying wavefunctions to the spatially varying mode functions, just like Eqs.
+\ref{eq:mhf-electron} and \ref{eq:mhf-photon}, and therefore do not assume the dipole approximation.}
 %\begin{figure}[t]
@@
 (\ref{eq:casimir}) is it implies that the same logic about zero-point energy differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
-
+\begin{figure}[t]
+\includegraphics[width=6cm]{figure3revision.pdf}
+\caption{\textcolor{blue}{\textbf{Expectation value of the correlated observable $\langle A\cdot p \rangle $ as a function of coupling.} Parameters are identical to those of the top panel of Fig.
+2a, and show that despite correlations being treated perturbatively, they are in excellent agreement with exact diagonalization, while in poor qualitative and quantitative agreement with perturbation theory in the bare photonic modes.}}
+\label{fig:correlated}
+\end{figure}
 In the presence of correlations  we must add to the energy a contribution of the form of Eq.
@@
 2(a) are listed in the SM.
+\textcolor{blue}{For the largest couplings considered here, the magnitude of the energy shift associated with the $A \cdot p$ term predicted from perturbation theory is larger than the energy separation between bare emitter levels, signaling the ultrastrong coupling regime.}
+
 In all cases, the agreement between our variational approach and numerical diagonalization is excellent, suggesting that our variational method is sufficiently flexible to capture ground states and excited states both with and without photonic excitations.
@@
 The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}).
-This is a so-called light-matter decoupling effect \cite{liberato2014}.
+This is a so-called light-matter decoupling effect, which was proposed in Ref.
+~\cite{liberato2014}.
+\textcolor{blue}{In Ref.
+~\cite{liberato2014}, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, deLiberato obtained "effective field mode profiles" with a strong dip in the location of the emitter, in qualitative agreement with what we report here.}
+
+\textcolor{blue}{This mechanism is also reflected in Fig.
+~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle Ap \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in Supplemental Materials).
+Such observables may play a role in correlated spectroscopies, as proposed in Ref.
+~\cite{flick2017}.
+We show predictions of the value of this observable based on the variational method, numerical diagonalization, and perturbation theory in the bare emitter and photon states.
+As can be seen, perturbation theory overestimates the magnitude of the energy shift, while both numerical and variational methods capture a saturation and then decrease of this expectation value, showing an apparent de-correlation between light and matter in the deep-strong coupling regime.
 The results of Fig.
-~\ref{fig:results} very clearly demonstrates not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime.
+~\ref{fig:results} and \ref{fig:correlated}  demonstrate not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime.
+The results of Fig.
+~\ref{fig:results} and \ref{fig:correlated} also show that despite correlations being treated perturbatively, it remains possible for correlated observables (and energies) to be predicted with high accuracy.}
 

### 2019-01-26 supp2.tex
reply to refs done. manuscript edits done.

@@
 
-As discussed in the main text, the variational energies, as well as the energies from numerical diagonalization, are compared against the energies calculated from perturbation theory.
+\textcolor{blue}{As discussed in the main text, the variational energies, as well as the energies from numerical diagonalization, are compared against the energies calculated from perturbation theory.
 We provide additional details on the perturbative calculations here.
@@
 \end{equation}
-where $b$ denotes a matter state different from $a$, and $p_{ba} = \langle b|p|a\rangle$.
+where $b$ denotes a matter state different from $a$, and $p_{ba} = \langle b|p|a\rangle$.}
+
+\textcolor{blue}{In the main text, we also considered expectation values such as $\langle A \cdot p \rangle$ in the ground state within perturbation theory.
+Calculating the shift of the ground-state wavefunction at first-order in perturbation theory, we immediately find that
+\begin{equation}
+\langle Ap \rangle = \frac{2q}{m} \text{Re } \sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{1}{2\epsilon_0L}\frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L}\right)}{\omega_n(\omega_a-\omega_b-\omega_n)}.
+\end{equation}}
+
+\textcolor{blue}{It is also of interest to consider "field-only" observables such as the number of (virtual) photons in the ground state.
+Calculating the wavefunction of the ground state at first-order in perturbation theory, and calculating the probability of a photon being in the ground state, one finds that
+\begin{equation}
+\langle N \rangle = \frac{q^2}{m^2} \sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{1}{2\hbar\epsilon_0L}\frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L}\right)}{\omega_n(\omega_a-\omega_b-\omega_n)^2}.
+\end{equation}
+In Fig.
+S1, we compare the expectation value of the number of photons calculated variationally, perturbatively, and with exact diagonalization.
+Unlike all other cases shown in this work (energy and $\langle A\cdot p \rangle$, the number of photons does not agree between variational and exact diagonalization.
+However, this behavior does not reflect any inconsistency with the variational method, but is instead a feature which is to be expected.
+In particular, the physics of the ansatz of Eq.
+4 of the main text is that a complicated ground state which has many virtual photons (see numerical diagonalization result) in it should be physically equivalent to a ground state with nearly no virtual photons (see variational result), but with different modes and frequencies.
+In other words, the reason the number of virtual photons differs is simply because they are being counted in the basis of "interacting modes" shown in Fig.
+2b of the main text.
+That said, our variational theory does not give a prescription to translate the interacting mode operators to the bare mode operators.}
+
+\begin{figure}[t]
+\includegraphics[width=6cm]{FigureS1.pdf}
+\caption{\textcolor{blue}{\textbf{Number of virtual photons in the ground state calculated variationally, numerically, and through perturbation theory.} Parameters are the same as in Fig.
+2 (top panel) of the main text.}}
+\label{fig:ansatz}
+\end{figure}
+
+\textcolor{blue}{We note that in all cases, when calculating these quantities in the variational theory, these expressions are still relevant, with the replacement of the bare photon frequencies and mode functions replaced by the ones that result from Eq.
+7 of the main text.}
 
@@
 
-Regarding the numerical diagonalization, we numerically implement the Hamiltonian of Eq.
+\textcolor{blue}{Regarding the numerical diagonalization, we numerically implement the Hamiltonian of Eq.
 (9) of the main text by building the matter operators ($H_0$ and $p$ as defined in the main text) in the basis of states corresponding to the tensor product of any matter state, and any photon state having four photons or less .
@@
 In particular, $A(z) = \sum_n\sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L} \right) \left(a_n+a^{\dagger}_n\right)$, with $L$ the length of the cavity and $\omega_n = \frac{n\pi c}{L}$.
-Diagonalization is performed using standard sparse eigendecomposition routines (such as those implemented in MATLAB).
+Diagonalization is performed using standard sparse eigendecomposition routines (such as those implemented in MATLAB).}
 

### 2019-01-27 supp2.tex
slight edit after johannes' comment

@@
 \begin{equation}
-\langle N \rangle = \frac{q^2}{m^2} \sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{1}{2\hbar\epsilon_0L}\frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L}\right)}{\omega_n(\omega_a-\omega_b-\omega_n)^2}.
+\langle N \rangle =\sum\limits_{n=1}^{N_p} \langle a^{\dagger}_{n,\text{bare}}a_{n,\text{bare}} \rangle = \frac{q^2}{m^2} \sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{1}{2\hbar\epsilon_0L}\frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L}\right)}{\omega_n(\omega_a-\omega_b-\omega_n)^2},
 \end{equation}
+where the "b" or "bare" subscript in the creation and annihilation operators denotes that these are creation and annihilation operators for photon modes in the cavity in the absence of interactions (e.g., the sine modes for the 1D cavity considered here).
 In Fig.
-S1, we compare the expectation value of the number of photons calculated variationally, perturbatively, and with exact diagonalization.
-Unlike all other cases shown in this work (energy and $\langle A\cdot p \rangle$, the number of photons does not agree between variational and exact diagonalization.
-However, this behavior does not reflect any inconsistency with the variational method, but is instead a feature which is to be expected.
+S1, we compare the expectation value of the number of "bare" photons calculated  perturbatively, and with exact diagonalization with the variationally calculated number of "interacting" photons, $\sum\limits_{n=1}^{N_p} \langle a^{\dagger}_{n,\text{int}}a_{n,\text{int}} \rangle$ (e.g., the modes shown in Fig.
+2b of the main text).
+The number of photons in the bare basis differs substantially from the number of photons in the interacting basis.}
+
+\textcolor{blue}{This behavior does not reflect any inconsistency with the variational method, but is instead a feature which is to be expected.
 In particular, the physics of the ansatz of Eq.
@@
 \includegraphics[width=6cm]{FigureS1.pdf}
-\caption{\textcolor{blue}{\textbf{Number of virtual photons in the ground state calculated variationally, numerically, and through perturbation theory.} Parameters are the same as in Fig.
+\caption{\textcolor{blue}{\textbf{Number of virtual photons (bare and interacting) in the ground state calculated variationally, numerically, and through perturbation theory.} Parameters are the same as in Fig.
 2 (top panel) of the main text.}}

### 2019-01-29 paper4pg.tex
modifications to text

@@
 Our findings also furnish a non-perturbative theory of the position-dependent energy (Lamb) shifts of ground and excited states that give rise to Casimir-Polder forces.
-\textcolor{blue}{The variational method put forth developed in this manuscript is particularly suited for dealing with QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds, and subsequently methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate.} Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.
+\textcolor{blue}{The variational method developed in this manuscript is particularly suited for dealing with QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds.
+Thus, methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate in this regime.} Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.
 
@@
 \textcolor{blue}{In Ref.
-~\cite{liberato2014}, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, deLiberato obtained "effective field mode profiles" with a strong dip in the location of the emitter, in qualitative agreement with what we report here.}
+~\cite{liberato2014}, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, the author obtained "effective field mode profiles" with a strong dip in the location of the emitter, in qualitative agreement with what we report here.}
 
 \textcolor{blue}{This mechanism is also reflected in Fig.
-~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle Ap \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in Supplemental Materials).
+~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle A\cdot p \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in Supplemental Materials).
 Such observables may play a role in correlated spectroscopies, as proposed in Ref.

### 2019-01-29 supp2.tex
small changes supp

@@
 In particular, $A(z) = \sum_n\sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L} \right) \left(a_n+a^{\dagger}_n\right)$, with $L$ the length of the cavity and $\omega_n = \frac{n\pi c}{L}$.
-Upon inspection, it is clear the the $A^2$ term leads to an energy shift at first-order in perturbation theory, while the $A\dot p$ term leads to an energy shift at second-order in perturbation theory.
+Upon inspection, it is clear that the $A^2$ term leads to an energy shift at first-order in perturbation theory, while the $A\cdot p$ term leads to an energy shift at second-order in perturbation theory.
 A straightforward application of perturbation theory leads to the following first-order energy shift from the $A^2$ term:
@@
 Note that this value is dependent on the maximum photon number (although somewhat weakly) retained due to the logarithmically diverging nature of the sum, and so to compare variational, numerical, and perturbative calculations, the same number of modes are retained in each.
-A similarly straightforward application of perturbation theory leads to the energy shift coming from the $A\dot p$ term being given by:
+A similarly straightforward application of perturbation theory leads to the energy shift coming from the $A\cdot p$ term being given by:
 \begin{equation}

### 2019-02-05 paper4pg.tex
shorter version just in case

--- 2019-01-26 paper4pg.tex
@@
 As a first step towards a new  \textit{ab initio} approach to ground and excited state energies in QED, we apply our ansatz to describe a multi-level emitter coupled to many optical modes, a system with no analytical solution.
-We find a compact semi-analytical formula which describes ground and excited state energies to less than 1\% error in all regimes of coupling parameters allowed by sum rules.
-Additionally, our formulation provides a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
+We find a compact semi-analytical formula which describes ground and excited state energies very well in all regimes of coupling parameters allowed by sum rules.
+Our formulation provides a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
 Our method should thus give rise to highly accurate non-perturbative descriptions of many other phenomena in general QED systems.
@@
 Our findings also furnish a non-perturbative theory of the position-dependent energy (Lamb) shifts of ground and excited states that give rise to Casimir-Polder forces.
-\textcolor{blue}{The variational method put forth developed in this manuscript is particularly suited for dealing with QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds, and subsequently methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate.} Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.
+\textcolor{blue}{The variational method developed in this manuscript is particularly suited for analyzing QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds, and subsequently methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate.} %Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.
 
@@
 
-The full Hamiltonian $H$, which depends on the fields $\psi$ and $\mathbf{A}$ can be parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$.
+The full Hamiltonian $H$, which depends on the fields $\psi$ and $\mathbf{A}$ is parameterized in terms of an orthonormal set of electron single-particle wavefunctions (orbitals) $\{\psi_n\}$, and in terms of a set of photonic mode functions (orbitals) $\{\mathbf{F}_i\}$.
 The electron field operator takes the form $\psi(\mathbf{x}) = \sum_n \psi_n(\mathbf{x})c_n$.
 The $c_n$ is an annihilation operator for an electron corresponding to state $n$.
-The electromagnetic field operator takes the form $\mathbf{A}(\mathbf{x}) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(\mathbf{x})a_i+\mathbf{F}^*_i(\mathbf{x})a^{\dagger}_i\right)$, where the $a_i^{(\dagger)}$ annihilate (create) a photon in mode $i$.
-In the electromagnetic field operator, we parameterize not only by the mode functions but also by mode frequencies.
+The electromagnetic field operator takes the form $\mathbf{A}(\mathbf{x}) = \sum_i\sqrt{\frac{\hbar}{2\epsilon_0\omega_i}} \left(\mathbf{F}_i(\mathbf{x})a_i+\mathbf{F}^*_i(\mathbf{x})a^{\dagger}_i\right)$, where the $a_i^{(\dagger)}$ annihilate (create) a photon in mode $i$.The electromagnetic field operator is parameterized by both mode functions and  frequencies.
 The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}$ \cite{joannopoulos2011photonic}.
@@
 Evaluating the expectation value of the Hamiltonian, and minimizing the Lagrange function  immediately yields:
-\begin{align}
-&\left(\frac{\mathbf{p}^2}{2m}+v_{\mathrm{ext}}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\left(\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\right)\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
+%\begin{align}
+%&\left(\frac{\mathbf{p}^2}{2m}+v_{\mathrm{ext}}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + \nonumber \\ &\sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x}')\psi_i(\mathbf{x}) \nonumber \\ & - \sum\limits_{j=1}^N \int d^3x' ~ V(\mathbf{x}-\mathbf{x}')\psi^*_j(\mathbf{x}')\psi_j(\mathbf{x})\psi_i(\mathbf{x}')  \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\left(\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\right)\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
+%\label{eq:mhf-electron}
+%\end{align}
+\begin{equation}
+\left(\frac{\mathbf{p}^2}{2m}+v_{\mathrm{ext}}(\mathbf{x}) \right)\psi_i(\mathbf{x}) + F[\{ \psi \}] + \frac{\hbar e^2}{4m\epsilon_0}\left(\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\right)\psi_i(\mathbf{x})   = E_i\psi_i(\mathbf{x}),
 \label{eq:mhf-electron}
-\end{align}
-for the electron orbitals and energies.
+\end{equation}
+for the electron orbitals and energies, where $F[\{ \psi \}]$ represent Hartree-Fock terms (see Supplementary Materials (SM)).
 We see that in addition to the one-body and Hartree-Fock terms for the electrons, the effect of the QED coupling is to add a one-body ponderomotive potential.
@@
 Note that term in the interaction Hamiltonian linear in the vector potential (the "$A\cdot p$ term") makes no contribution to the expectation value of the ground state of the energy in this ansatz.
-At second order in the $A\cdot p$ term, virtual photon processes arise, such as Lamb shifts, whose emitter-position-dependence gives rise to van der Waals and Casimir-Polder forces  \cite{scheel2009macroscopic}.
-Physically, this term will mix the factorizable ground state of Eq.~(\ref{eq:ansatz}) with states that simultaneously have virtual excitations of the matter and the electromagnetic field.
+Physically, this term will mix the factorizable ground state of Eq.~(\ref{eq:ansatz}) with states that  have virtual excitations of the matter and the electromagnetic field.
 The resulting state is non-factorizable and thus, the term in the $A\cdot p$ term leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of QED ground and excited states.
@@
 The results of the variational method (blue) are compared to perturbation theory (orange), as well as numerical diagonalization (red) with the Fock space truncated to fifty cavity modes with no more than four photons.
-(Inset) The fourth and fifth energy levels show a weak anti-crossing behavior which is well-reproduced by the variational theory.
-Blue denotes variational while red denotes numerical.
-(b) Mechanism of overestimation of couplings and resonances in perturbation theory: modes derived from the variational theorem are suppressed in the vicinity of the emitter, which self-consistently decreases the field-emitter coupling.
+(Inset) The fourth and fifth energy levels show a weak anti-crossing behavior which is reproduced by the variational theory.
+(b) Mechanism of overestimation of couplings and resonances in perturbation theory: modes derived from the variational theorem are suppressed in the vicinity of the emitter, self-consistently decreasing light-matter coupling.
 }
@@
 By taking $m$ as an ansatz for an excited state, correlation corrections to excited states can also be found.
-\textcolor{blue}{In the Supplementary Information, we derive a set of equations for the matter orbitals and photonic mode functions which self-consistently takes into account the correlation energy associated with Eq.
+\textcolor{blue}{In the SM, we derive a set of equations for the matter orbitals and photonic mode functions which self-consistently takes into account the correlation energy associated with Eq.
 ~\ref{eq:e-shift}.
@@
 The matter system we describe can thus be mapped to an $N_a$ site system, which be considered as a simplified model of a molecule within a tight-binding description.
-Thus we parameterize the general family of matter Hamiltonians as:
-\begin{equation}\label{eq:matter_hamiltonian}
-H_{\text{matter}} = \sum\limits_{i=1}^{{N_a-1}} V_i|i\rangle\langle i|+t(|i\rangle\langle i+1|+|i+1\rangle\langle i|).
-\end{equation}
-The momentum operator, we write as 
-\begin{equation}\label{eq:momentum_operator}
-p = \frac{-i\hbar}{R}\sum\limits_{i=1}^{N_a-1} \left(|i\rangle\langle i+1|-|i+1\rangle\langle i| \right),
-\end{equation}
+Thus we parameterize the general family of matter Hamiltonians as $H_{\text{matter}} = \sum\limits_{i=1}^{{N_a-1}} V_i|i\rangle\langle i|+t(|i\rangle\langle i+1|+|i+1\rangle\langle i|)$.
+The momentum operator, we write as $p = \frac{-i\hbar}{R}\sum\limits_{i=1}^{N_a-1} \left(|i\rangle\langle i+1|-|i+1\rangle\langle i| \right)$,
 where $R$ is a constant with units of length representing roughly the difference in positions between sites.
 This physical interpretation however is rough: it is also a function of the hopping elements $t$, because we choose $R$ in this work such that the Thomas-Reiche-Kuhn (TRK) sum rule is enforced: $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} = 1$, where $p_{ig} = \langle i|p|g\rangle$ are momentum matrix elements between different matter states \cite{cohen1992}.
-Since the sum rule is based on a full electronic real-space description, it does not rigorously apply to a discrete-level system.
-However, the matrix elements and energy levels of a few-level approximated Hamiltonian are derived from an underlying real-space Hamiltonian.
-Thus, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot exist physically.
+Although the sum rule is based on a full electronic real-space description, a discrete system which has  $\frac{2}{m}\sum\limits_{i=2}^{N_a}\frac{|p_{ig}|^2}{E_i - E_a} > 1$ cannot exist physically.
 The TRK sum rule places a bound on how strong the effect of the $A\cdot p$ term can be.
 The net effect is that the value of $R$ we choose is on the order of $\sqrt{\frac{\hbar}{2mt}}$.
-These considerations also imply that when we plot observables as a function of some coupling parameter, for fixed $R$, we must vary a parameter which does not explicitly appear in the sum rule such as the charge or the number of emitters in the cavity.
+%These considerations also imply that when we plot observables as a function of some coupling parameter, for fixed $R$, we must vary a parameter which does not explicitly appear in the sum rule such as the charge or the number of emitters in the cavity.
 
-The detailed derivations of the energies of states via the formalism introduced here are shown in the Supplementary Materials (SM).
-Here, we state the main results.
+Derivations of the energies of states via the formalism introduced here are shown in the SM.
+Here, we present the main results.
 Using a one-dimensional version of Eq.
@@
 \end{equation}
-The corresponding `interacting' field mode profiles are given by
-\begin{align}\label{eq:field_mode}
-N_n^{-1}F_n(z) =~ &\theta(z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\
--&\theta (z-d) \left(\frac{\cos\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\sin\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\ 
-+&\theta (d-z) \sin\left(\frac{\omega_n z}{c} \right),
-\end{align}
-with the normalization constant
-\begin{equation}\label{eq:mode_normalization}
-N_n = 2\sqrt{\frac{1}{\frac{c}{\omega_n}\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
-\end{equation}
+The corresponding `interacting' field mode profiles, shown in Fig. 2b, are given by compact expressions shown in the SM.
+%\begin{align}\label{eq:field_mode}
+%N_n^{-1}F_n(z) =~ &\theta(z-d) \left(\frac{\sin\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\cos\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\
+%-&\theta (z-d) \left(\frac{\cos\left(\frac{\omega_nL}{c}\right)\sin\left(\frac{\omega_nd}{c}\right)\sin\left(\frac{\omega_nz}{c}\right)}{\sin\left(\frac{\omega_n(L-d)}{c}\right)}\right) \nonumber \\ 
+%+&\theta (d-z) \sin\left(\frac{\omega_n z}{c} \right),
+%\end{align}
+%with the normalization constant
+%\begin{equation}\label{eq:mode_normalization}
+%N_n = 2\sqrt{\frac{1}{\frac{c}{\omega_n}\left(\frac{\omega_nL}{c}-\sin\left(\frac{\omega_nL}{c}\right) \right)\left(1+\frac{\sin^2\left(\frac{\omega_nd}{c}\right)}{\sin^2\left(\frac{\omega_n(L-d)}{c}\right)} \right)}}.
+%\end{equation}
 The result of Eq.
@@
 \begin{figure}[t]
-\includegraphics[width=6cm]{figure3revision.pdf}
+\includegraphics[width=4cm]{figure3revision.pdf}
 \caption{\textcolor{blue}{\textbf{Expectation value of the correlated observable $\langle A\cdot p \rangle $ as a function of coupling.} Parameters are identical to those of the top panel of Fig.
-2a, and show that despite correlations being treated perturbatively, they are in excellent agreement with exact diagonalization, while in poor qualitative and quantitative agreement with perturbation theory in the bare photonic modes.}}
+2a.
+Despite correlations being treated perturbatively, this observable is in excellent agreement with exact diagonalization, while in poor agreement with perturbation theory in the bare photonic modes.}}
 \label{fig:correlated}
@@
 (\ref{eq:e-shift}), specialized to the case of an emitter in a one-dimensional cavity.
-The interaction energy, given by Eqs.
-(\ref{eq:e-shift}) and (\ref{eq:casimir}) is semi-analytical once the bare emitter states are known, as it is fully specified by Eqs.
-(\ref{eq:cot}-\ref{eq:mode_normalization}) once the transcendental equation of Eq.
-(\ref{eq:cot}) is solved.
-We also apply the correlation correction to excited states as well, by using the second-order perturbation theory formula for the energy shift of excited states due to the $A \cdot p$ term, using the same electron and photon orbitals and frequencies as derived in Eqs.
-(\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}).
+We apply the correlation correction to excited states as well, calculating excited-state energy shifts within second-order perturbation theory.
 In Fig.
-~\ref{fig:results}(a), we show the result of this procedure when applied to calculate ground- and excited- state energies for two-, three-, and four-level systems coupled to a one-dimensional cavity.
+~\ref{fig:results}(a), we show the result of this procedure when applied to calculate ground- and excited- state energies for few-level systems coupled to a one-dimensional cavity.
 The relevant parameters for Fig.
@@
 
-Importantly, the reason perturbation theory fails for first excited state, much more so than for the ground state, is that the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ of nearly $2$ eV, which is far larger than the spacing of the bare emitter levels.
-On the other hand, the variational estimate from our formalism finds no such large negative energy shift, and leads to an energy gap between the first two levels which is similar to the bare gap, and in agreement with numerical diagonalization.
-The reason for this behavior is that the effect of the plasma term in Eq.
-(\ref{eq:mhf-photon}) is to blue-shift all of the photon frequencies.
-In particular, for the largest coupling considered in Figure~\ref{fig:results}, we find that the lowest photon frequency is shifted to 0.99 eV, and then becomes far off-resonance from the bare emitter transition.
-The plasma term, as shown in Fig.~\ref{fig:results}b, also strongly reduces the coupling between light and matter by a different mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field and the magnitude of the correlation term, as per Equation (\ref{eq:e-shift}).
-This is a so-called light-matter decoupling effect, which was proposed in Ref.
+Importantly, the reason perturbation theory fails for first excited state is that the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ of nearly $2$ eV, which is far larger than the spacing of the bare emitter levels.
+In contrast, no such near-resonance is found in the variational framework because the plasma term  in Eq.
+(\ref{eq:mhf-photon}) blue-shifts all of the photon frequencies.
+For the largest coupling considered in Figure~\ref{fig:results}, we find that the lowest photon frequency is shifted to 0.99 eV (from 0.62 eV),  far off-resonance from the bare emitter transition.
+The plasma term also strongly reduces the coupling between light and matter by a mechanism in which the field modes obtained from Equation (\ref{eq:mhf-photon}) are screened out of the emitter, thus self-consistently reducing the strength of the coupling between matter and field (see \ref{fig:results}b).
+This is a  light-matter decoupling effect, which was proposed in Ref.
 ~\cite{liberato2014}.
@@
 
-\textcolor{blue}{This mechanism is also reflected in Fig.
+\textcolor{blue}{This light-matter decoupling is also reflected in Fig.
+~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle Ap \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in Supplemental Materials).
+As shown in Fig.
+3 be seen, numerical and variational methods capture a saturation and then decrease of this expectation value.
+The results of Fig.
+~\ref{fig:results} and \ref{fig:correlated}  demonstrate not only the accuracy of our ansatz, but provides insight into the mechanisms by which light-matter coupling saturates in the nonperturbative QED regime.
+The results of Fig.
+~\ref{fig:results} and \ref{fig:correlated} also show that despite correlations being treated perturbatively, it remains possible for correlated observables (and energies) to be predicted with high accuracy.}
+
+%\textcolor{blue}{This mechanism is also reflected in Fig.
 ~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle Ap \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in Supplemental Materials).
@@
 Our calculation of the Lamb shifts via Eq.
-(\ref{eq:e-shift}) says that the shifts result from virtual emission and re-absorption of the photonic quasiparticles (the \textit{interacting} modes), governed by (Eqs.
-(\ref{eq:cot}-\ref{eq:mode_normalization})), which are dependent on properties of the matter.
+(\ref{eq:e-shift}) says that the shifts result from virtual emission and re-absorption of the photonic quasiparticles (the \textit{interacting} modes), which are dependent on properties of the matter.
 As these interacting photon modes differ greatly from the bare modes and frequencies in the non-perturbative regime, Eq.
@@
 
-With the advent of ultra-strong coupling and deep-strong coupling in QED systems, the theory posed here, when applied to more complex systems, could form the basis for understanding Lamb and Casimir phenomena in the ultra-strong coupling regime.
-Additionally, the results developed here could be applied to capture spontaneous emission in the non-perturbative regime.
-Finally, one could use the non-perturbative real-space knowledge of how matter affects photons in order to design a photonic mode atom-by-atom.
+The theory posed here could form the basis of an understanding of Lamb shift, Casimir forces, and potentially many other phenomena in the ultrastrong coupling regime for complex QED systems beyond current analytical and numerical methods.
 

### 2019-02-06 paper4pg.tex
small corrections and getting rid of few more words

@@
 Our formulation provides a non-perturbative theory of Lamb shifts and Casimir-Polder forces, as well as suggesting new physical concepts such as the Casimir energy of a single atom in a cavity.
-Our method should thus give rise to highly accurate non-perturbative descriptions of many other phenomena in general QED systems.
+Our method thus give rise to highly accurate non-perturbative descriptions of many other phenomena in general QED systems.
 \end{abstract}
@@
 A variational framework, as we show, flexibly allows a real-space description of the electrons and photons as they are modified by the coupling and also beyond the dipole approximation.
-Beyond these advantages, a variational framework also allows conceptual insights, as we shall show, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom.
+Beyond these advantages, a variational framework also allows conceptual insights, into a simple non-perturbative theory of Lamb shifts, into a quasiparticle description of QED systems, and into the notion of Casimir forces in the limit of one atom.
 A variational framework also allows compact semi-analytical formulae to describe complex systems which may assist the development of functionals for use in QEDFT.
@@
 
-The QED Hamiltonian is given by $H = H_{\mathrm{mat}}+H_{\mathrm{em}}+H_{\mathrm{int}}$ where $H_{\mathrm{mat}}$ describes the matter in the absence of the quantized electromagnetic field, $H_{\mathrm{em}}$ describes the photons in the absence of the matter, and $H_{\mathrm{int}}$ describes the coupling between light and matter.
+In general, the QED Hamiltonian is given by $H = H_{\mathrm{mat}}+H_{\mathrm{em}}+H_{\mathrm{int}}$ where $H_{\mathrm{mat}}$ describes the matter in the absence of the quantized electromagnetic field, $H_{\mathrm{em}}$ describes the photons in the absence of the matter, and $H_{\mathrm{int}}$ describes the coupling between light and matter.
 The matter Hamiltonian takes the form:
@@
 \end{equation}
-where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, and$N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons.
+where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, and $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons.
 In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Eqs.
@@
 \textcolor{blue}{In the SM, we derive a set of equations for the matter orbitals and photonic mode functions which self-consistently takes into account the correlation energy associated with Eq.
-~\ref{eq:e-shift}.
+(\ref{eq:e-shift}).
 These equations take into account the spatially varying wavefunctions to the spatially varying mode functions, just like Eqs.
-\ref{eq:mhf-electron} and \ref{eq:mhf-photon}, and therefore do not assume the dipole approximation.}
+(\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}), and therefore do not assume the dipole approximation.}
 %\begin{figure}[t]
@@
 This is a  light-matter decoupling effect, which was proposed in Ref.
-~\cite{liberato2014}.
-\textcolor{blue}{In Ref.
-~\cite{liberato2014}, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, deLiberato obtained "effective field mode profiles" with a strong dip in the location of the emitter, in qualitative agreement with what we report here.}
+~\cite{liberato2014}, \textcolor{blue}{where, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, "effective field mode profiles" are obtained with a strong dip in the location of the emitter, in qualitative agreement with what we report here.}
 
 \textcolor{blue}{This light-matter decoupling is also reflected in Fig.
-~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle Ap \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in Supplemental Materials).
+~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle A\cdot p \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in SM).
 As shown in Fig.
-3 be seen, numerical and variational methods capture a saturation and then decrease of this expectation value.
+3, numerical and variational methods capture a saturation and then decrease of this expectation value.
 The results of Fig.

### 2019-02-13 paper4pg.tex
removed 148 words
3866 - 148 = 3718 < 3750

@@
 \begin{figure}[t]
-\includegraphics[width=8.5cm]{conceptfigure.pdf}
+\includegraphics[width=8cm]{conceptfigure.pdf}
 \caption{\textbf{Ground-state ansatz applied to matter in a cavity: effectively decoupled matter and photons.} (Left) Bare description of the coupled light-matter ground state in terms of many virtual excitations of the emitter state and the bare cavity photons.
@@
 Given an ansatz $|\Omega\rangle$ for the ground state of $H$, the variational theorem ensures that $\langle \Omega|H|\Omega\rangle$ is an upper bound for the ground state energy.
-Parameterizing $|\Omega\rangle$ to generate a family of ground states, and minimizing $\langle \Omega|H|\Omega\rangle$ with respect to the introduced parameters gives the best upper bound for the ground state energy within this family of ground states.
 We choose as our ansatz
@@
 for the electron orbitals and energies, where $F[\{ \psi \}]$ represent Hartree-Fock terms (see Supplementary Materials (SM)).
-We see that in addition to the one-body and Hartree-Fock terms for the electrons, the effect of the QED coupling is to add a one-body ponderomotive potential.
+Here the effect of the QED coupling is to add a one-body ponderomotive potential.
 %We note that in the often used dipole approximation, the ponderomotive potential has no effect as it becomes independent of position.
@@
 Physically, this term will mix the factorizable ground state of Eq.~(\ref{eq:ansatz}) with states that  have virtual excitations of the matter and the electromagnetic field.
-The resulting state is non-factorizable and thus, the term in the $A\cdot p$ term leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of QED ground and excited states.
-We note that correlations can also be included in the energy shifts of excited emitter states, as well as states that already have photonic excitations \footnote{The behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes.
+The resulting state is non-factorizable and thus, the $A\cdot p$ term leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of QED ground and excited states \footnote{The behavior of the $A\cdot p$ and $A^2$-term is similar to the ${r}\cdot D$ and $r^2$ term in the length-gauge reported in recent work on the optimized effective potential~\cite{pellegrini2015,flick2017c} method for QEDFT including one-photon processes.
 }.
 \begin{figure}[t]
-\includegraphics[width=8.5cm]{Figure2and3combined.pdf}
+\includegraphics[width=7.5cm]{Figure2and3combined.pdf}
 \caption{\textbf{Variational theory of ground and excited states in non-perturbative QED.} (a) Lowest few energy levels of a two (top), three (middle), and four (bottom) level system embedded in the middle of a one-dimensional cavity.
@@
 We capture the effect of correlations perturbatively.
-For the ground state, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential.
+For the ground state, we consider the second-order correction $\delta E$ to the ground state energy arising from the $A\cdot p$ term.
 That correction is given by
@@
 (\ref{eq:e-shift}).
-By taking $m$ as an ansatz for an excited state, correlation corrections to excited states can also be found.
+By considering an ansatz for an excited state, correlation corrections to excited states can also be found.
 \textcolor{blue}{In the SM, we derive a set of equations for the matter orbitals and photonic mode functions which self-consistently takes into account the correlation energy associated with Eq.
@@
 (\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}), we calculate the electron orbitals, photon orbitals, and photon frequencies in the absence of correlations.
-In the absence of correlations, we found for example that the energy of any matter state $a$ with no photonic quasiparticles is given by:
+In the absence of correlations, we found that the energy of any matter state $a$ with no photonic quasiparticles is given by:
 \begin{equation}
@@
 where $E_a^0$ is the energy of the matter state in the absence of coupling, $\omega_n$ are found in our framework, $\omega_n^0 = \frac{n\pi c}{L}$.
-Imposing continuity of the modes and discontinuity of their derivatives at $z=d$, the modes found in our framework have their frequencies given by
+The modes found in our framework have their frequencies given by
 \begin{equation}
@@
 In particular, it has long been known that when two conducting plates are placed near each other, there is a Casimir energy associated with the fact that the zero-point energy of the nearby plates is different than the zero-point energy of plates infinitely apart.
-This is because the electromagnetic mode structure of two nearby plates is different from that of two infinitely separated plates.
 This Casimir energy is simply the difference between the interacting and non-interacting zero-point energies \cite{casimir1948attraction,lifshitz1956theory}.
@@
 What is notable about the result of Eq.
-(\ref{eq:casimir}) is it implies that the same logic about zero-point energy differences can be applied to find the interaction energy case of a \textit{single atom} placed near a cavity, even though a single emitter is very far from the limit of a macroscopic polarizable object.
+(\ref{eq:casimir}) is it implies that the same logic about zero-point energy differences can be applied to find the interaction energy  of a \textit{single atom} placed near a cavity.
 \begin{figure}[t]
@@
 
-In all cases, the agreement between our variational approach and numerical diagonalization is excellent, suggesting that our variational method is sufficiently flexible to capture ground states and excited states both with and without photonic excitations.
+In all cases, the agreement between our variational approach and numerical diagonalization is excellent, suggesting that our variational method is sufficiently flexible to capture ground states and excited states.
 The accuracy as a function of number of levels suggests that the breakdown of gauge invariance associated with few-level systems is not crucial to the good agreement between variational and numerical results \cite{bernardis2018breakdown}.
-This is to be contrasted with perturbation theory in the bare matter and photon states, which can both strongly over- and underestimate the energies.
-The most interesting case of disagreement arises in the case of the two-level system (top panel).
-For the two-level system considered here, the variational result agrees very well with numerical diagonalization, while perturbation theory predicts an energy which evolves with coupling in the wrong direction and is off from the true energy by over 100\%.
+Perturbation theory in the bare matter and photon states can both strongly over- and underestimate the energies.
+Strong disagreement arises in the case of the two-level system (top panel).
+For the two-level system considered here, the variational result agrees very well with numerical diagonalization, while perturbation theory predicts an energy which evolves with coupling in the wrong direction and is off from the true energy by over 100\% for the largest coupling.
 
-Importantly, the reason perturbation theory fails for first excited state is that the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ of nearly $2$ eV, which is far larger than the spacing of the bare emitter levels.
+Perturbation theory fails for first excited state because the first bare cavity mode is nearly resonant with the transition between ground and excited emitter states, leading to a very large negative contribution from the $A\cdot p$ term of nearly $2$ eV, which is far larger than the spacing of the bare emitter levels.
 In contrast, no such near-resonance is found in the variational framework because the plasma term  in Eq.

### 2019-03-26 paper4pg.tex
manuscript edits

@@
 Our findings also furnish a non-perturbative theory of the position-dependent energy (Lamb) shifts of ground and excited states that give rise to Casimir-Polder forces.
-\textcolor{blue}{The variational method developed in this manuscript is particularly suited for analyzing QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds, and subsequently methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate.} %Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.
+The variational method developed in this manuscript is particularly suited for analyzing QED systems in the ultrastrong coupling regime, in which the rotating-wave approximation no longer holds, and subsequently methods based on the Jaynes-Cummings model such as dressed state approaches \cite{cohen1992} are no longer accurate.
+%Moreover, our approach is particularly suited to scalably dealing with systems with many emitter levels, as well as many cavity modes which may have large virtual photon occupation numbers.
 
@@
 The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}$ \cite{joannopoulos2011photonic}.
+\textcolor{blue}{We comment that in writing the Hamiltonian this way, we make a few assumptions.
+For one, that the interaction of matter with the quantized electromagnetic field is mediated solely by the vector potential.
+Moreover, we are assuming that the effect of free-space modes, which an emitter weakly interacts with, is taken into account into renormalization of the electron mass and charge \cite{milonni1976semiclassical}, and so we focus only on the interaction of the emitter with low-energy cavity modes where strong interaction is possible.
+For these cavity modes, a natural cutoff emerges based on the transparency of the cavity.}
 
@@
 Excited states in this framework can be identified with matter and photon quasiparticle excitations.
+\textcolor{blue}{Taking the divergence of this equation, we see that $\nabla\cdot(1-\frac{\omega_p^2(\mathbf{x})}{\omega^2})\mathbf{F}(\mathbf{x})=0$, which is a generalized Coulomb gauge condition on the modes.
+The generalized Coulomb gauge is such that $\nabla\cdot\epsilon(\mathbf{x})\mathbf{A}(\mathbf{x})=0$ for some dielectric permittivity $\epsilon(\mathbf{x})$ and is the most common gauge for analyzing phenomena in the discipline of macroscopic QED, in which one seeks to analyze the interaction of charges with fields in dielectric media \cite{glauber1991quantum}.
+The reason a particular gauge was chosen was that in the Hamiltonian, we have already assumed that all interaction with the quantized modes is manifested only through the vector potential, and that there is no scalar potential contribution to the interaction: relaxing this assumption allows more general gauge freedom.
+We also remark that if we performed a frame transformation, for example to the length gauge, the energy would be the same as that of our ansatz in the velocity gauge, although the state would be markedly different, and would be a highly correlated state.}
 
@@
 By considering an ansatz for an excited state, correlation corrections to excited states can also be found.
-\textcolor{blue}{In the SM, we derive a set of equations for the matter orbitals and photonic mode functions which self-consistently takes into account the correlation energy associated with Eq.
+In the SM, we derive a set of equations for the matter orbitals and photonic mode functions which self-consistently takes into account the correlation energy associated with Eq.
 (\ref{eq:e-shift}).
 These equations take into account the spatially varying wavefunctions to the spatially varying mode functions, just like Eqs.
-(\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}), and therefore do not assume the dipole approximation.}
+(\ref{eq:mhf-electron}) and (\ref{eq:mhf-photon}), and therefore do not assume the dipole approximation.
 %\begin{figure}[t]
@@
 \includegraphics[width=4cm]{figure3revision.pdf}
-\caption{\textcolor{blue}{\textbf{Expectation value of the correlated observable $\langle A\cdot p \rangle $ as a function of coupling.} Parameters are identical to those of the top panel of Fig.
+\caption{\textbf{Expectation value of the correlated observable $\langle A\cdot p \rangle $ as a function of coupling.} Parameters are identical to those of the top panel of Fig.
 2a.
-Despite correlations being treated perturbatively, this observable is in excellent agreement with exact diagonalization, while in poor agreement with perturbation theory in the bare photonic modes.}}
+Despite correlations being treated perturbatively, this observable is in excellent agreement with exact diagonalization, while in poor agreement with perturbation theory in the bare photonic modes.}
 \label{fig:correlated}
@@
 2(a) are listed in the SM.
-\textcolor{blue}{For the largest couplings considered here, the magnitude of the energy shift associated with the $A \cdot p$ term predicted from perturbation theory is larger than the energy separation between bare emitter levels, signaling the ultrastrong coupling regime.}
+For the largest couplings considered here, the magnitude of the energy shift associated with the $A \cdot p$ term predicted from perturbation theory is larger than the energy separation between bare emitter levels, signaling the ultrastrong coupling regime.
 
@@
 This is a  light-matter decoupling effect, which was proposed in Ref.
-~\cite{liberato2014}, \textcolor{blue}{where, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, "effective field mode profiles" are obtained with a strong dip in the location of the emitter, in qualitative agreement with what we report here.}
+~\cite{liberato2014}, where, on the basis of photodetection probabilities for exactly-obtained excited polaritonic eigenstates in a Hopfield model, "effective field mode profiles" are obtained with a strong dip in the location of the emitter, in qualitative agreement with what we report here.
 
-\textcolor{blue}{This light-matter decoupling is also reflected in Fig.
+This light-matter decoupling is also reflected in Fig.
 ~\ref{fig:correlated}, where we calculate a correlated ground state observable such as $\langle A\cdot p \rangle$, which is a measure of entanglement between the ground state and excitations of the photon and matter (details shown in SM).
@@
 The results of Fig.
-~\ref{fig:results} and \ref{fig:correlated} also show that despite correlations being treated perturbatively, it remains possible for correlated observables (and energies) to be predicted with high accuracy.}
+~\ref{fig:results} and \ref{fig:correlated} also show that despite correlations being treated perturbatively, it remains possible for correlated observables (and energies) to be predicted with high accuracy.
 

### 2019-03-26 paper4pg.tex
small changes to the text

@@
 For one, that the interaction of matter with the quantized electromagnetic field is mediated solely by the vector potential.
-Moreover, we are assuming that the effect of free-space modes, which an emitter weakly interacts with, is taken into account into renormalization of the electron mass and charge \cite{milonni1976semiclassical}, and so we focus only on the interaction of the emitter with low-energy cavity modes where strong interaction is possible.
+Moreover, we are assuming that the effect of free-space modes, which an emitter weakly interacts with, is taken into account {\color{red}by} renormalization of the electron mass and charge \cite{milonni1976semiclassical}{\color{red}.
+As a consequence}, we focus only on the interaction of the emitter with low-energy cavity modes where strong interaction is possible.
 For these cavity modes, a natural cutoff emerges based on the transparency of the cavity.}
@@
 Excited states in this framework can be identified with matter and photon quasiparticle excitations.
-\textcolor{blue}{Taking the divergence of this equation, we see that $\nabla\cdot(1-\frac{\omega_p^2(\mathbf{x})}{\omega^2})\mathbf{F}(\mathbf{x})=0$, which is a generalized Coulomb gauge condition on the modes.
-The generalized Coulomb gauge is such that $\nabla\cdot\epsilon(\mathbf{x})\mathbf{A}(\mathbf{x})=0$ for some dielectric permittivity $\epsilon(\mathbf{x})$ and is the most common gauge for analyzing phenomena in the discipline of macroscopic QED, in which one seeks to analyze the interaction of charges with fields in dielectric media \cite{glauber1991quantum}.
+\textcolor{blue}{Taking the divergence of {\color{red}Eq.~\ref{eq:mhf-photon}}, we see that $\nabla\cdot(1-\frac{\omega_p^2(\mathbf{x})}{\omega^2})\mathbf{F}(\mathbf{x})=0$, which is a generalized Coulomb gauge condition on the modes.
+The generalized Coulomb gauge is {\color{red}defined by} $\nabla\cdot\epsilon(\mathbf{x})\mathbf{A}(\mathbf{x})=0$ for some dielectric permittivity $\epsilon(\mathbf{x})${\color{red}.
+It is } the most common gauge for analyzing phenomena in the discipline of macroscopic QED, in which one seeks to analyze the interaction of charges with fields in dielectric media \cite{glauber1991quantum}.
 The reason a particular gauge was chosen was that in the Hamiltonian, we have already assumed that all interaction with the quantized modes is manifested only through the vector potential, and that there is no scalar potential contribution to the interaction: relaxing this assumption allows more general gauge freedom.
-We also remark that if we performed a frame transformation, for example to the length gauge, the energy would be the same as that of our ansatz in the velocity gauge, although the state would be markedly different, and would be a highly correlated state.}
+We also remark that if we performed a frame transformation, for example to the length gauge, the energy would be the same as that of our ansatz in the velocity gauge, although the state would be markedly different, and would be a highly correlated state {\color{red}in the new frame}.}
 

### 2019-03-26 paper4pg.tex
updated according to small text changes

--- 2019-03-26 paper4pg.tex
@@
 For one, that the interaction of matter with the quantized electromagnetic field is mediated solely by the vector potential.
-Moreover, we are assuming that the effect of free-space modes, which an emitter weakly interacts with, is taken into account into renormalization of the electron mass and charge \cite{milonni1976semiclassical}, and so we focus only on the interaction of the emitter with low-energy cavity modes where strong interaction is possible.
+Moreover, we are assuming that the effect of free-space modes, which an emitter weakly interacts with, is taken into account by renormalization of the electron mass and charge \cite{milonni1976semiclassical}.
+We focus only on the interaction of the emitter with low-energy cavity modes where strong interaction is possible.
 For these cavity modes, a natural cutoff emerges based on the transparency of the cavity.}
@@
 Excited states in this framework can be identified with matter and photon quasiparticle excitations.
-\textcolor{blue}{Taking the divergence of this equation, we see that $\nabla\cdot(1-\frac{\omega_p^2(\mathbf{x})}{\omega^2})\mathbf{F}(\mathbf{x})=0$, which is a generalized Coulomb gauge condition on the modes.
-The generalized Coulomb gauge is such that $\nabla\cdot\epsilon(\mathbf{x})\mathbf{A}(\mathbf{x})=0$ for some dielectric permittivity $\epsilon(\mathbf{x})$ and is the most common gauge for analyzing phenomena in the discipline of macroscopic QED, in which one seeks to analyze the interaction of charges with fields in dielectric media \cite{glauber1991quantum}.
+\textcolor{blue}{Taking the divergence of Eq.~\ref{eq:mhf-photon}, we see that $\nabla\cdot(1-\frac{\omega_p^2(\mathbf{x})}{\omega^2})\mathbf{F}(\mathbf{x})=0$, which is a generalized Coulomb gauge condition on the modes.
+The generalized Coulomb gauge, $\nabla\cdot\epsilon(\mathbf{x})\mathbf{A}(\mathbf{x})=0$ for some dielectric permittivity $\epsilon(\mathbf{x}), is the most common gauge for analyzing phenomena in the discipline of macroscopic QED, in which one seeks to analyze the interaction of charges with fields in dielectric media \cite{glauber1991quantum}.
 The reason a particular gauge was chosen was that in the Hamiltonian, we have already assumed that all interaction with the quantized modes is manifested only through the vector potential, and that there is no scalar potential contribution to the interaction: relaxing this assumption allows more general gauge freedom.

### 2019-04-01 paper4pg.tex
moved discussion to supp, i think we're ok now.

@@
 The normalization chosen for the electron wavefunctions is $\int d^3x~ \psi_m^*\psi_n = \delta_{mn}$ while for the photon mode functions, it is $\int d^3x~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_j = \delta_{ij}$ \cite{joannopoulos2011photonic}.
-\textcolor{blue}{We comment that in writing the Hamiltonian this way, we make a few assumptions.
+\textcolor{blue}{Assumptions behind the form of the Hamiltonian are stated in Supplementary Materials, page 2.}
+
+%\textcolor{blue}{In writing the Hamiltonian this way, we make a few assumptions.
 For one, that the interaction of matter with the quantized electromagnetic field is mediated solely by the vector potential.
@@
 Excited states in this framework can be identified with matter and photon quasiparticle excitations.
-\textcolor{blue}{Taking the divergence of Eq.~\ref{eq:mhf-photon}, we see that $\nabla\cdot(1-\frac{\omega_p^2(\mathbf{x})}{\omega^2})\mathbf{F}(\mathbf{x})=0$, which is a generalized Coulomb gauge condition on the modes.
-The generalized Coulomb gauge, $\nabla\cdot\epsilon(\mathbf{x})\mathbf{A}(\mathbf{x})=0$ for some dielectric permittivity $\epsilon(\mathbf{x}), is the most common gauge for analyzing phenomena in the discipline of macroscopic QED, in which one seeks to analyze the interaction of charges with fields in dielectric media \cite{glauber1991quantum}.
-The reason a particular gauge was chosen was that in the Hamiltonian, we have already assumed that all interaction with the quantized modes is manifested only through the vector potential, and that there is no scalar potential contribution to the interaction: relaxing this assumption allows more general gauge freedom.
-We also remark that if we performed a frame transformation, for example to the length gauge, the energy would be the same as that of our ansatz in the velocity gauge, although the state would be markedly different, and would be a highly correlated state.}
+\textcolor{blue}{Taking the divergence of Eq.~\ref{eq:mhf-photon}, we see that $\nabla\cdot(1-\frac{\omega_p^2(\mathbf{x})}{\omega^2})\mathbf{F}(\mathbf{x})=0$, which is a generalized Coulomb gauge condition on the modes \cite{glauber1991quantum}.
+For more discussion, see Supplementary Information, page 2.}
 

### 2019-04-01 supp2.tex
moved discussion to supp, i think we're ok now.

@@
 2 of the main text.
+Before doing this, we briefly describe the assumptions made in our framework, regarding our assumptions on gauge, as well as high-frequency free-space modes.
+
+\subsection{Assumptions behind the Hamiltonian of Equations (1-3)}
+
+\textcolor{blue}{In writing the Hamiltonian of the main text in the form of Equations (1-3), with the particular form of the field operators, we make a few assumptions.
+For one, that the interaction of matter with the quantized electromagnetic field is mediated solely by the vector potential.
+Moreover, we are assuming that the effect of free-space modes, which an emitter weakly interacts with, is taken into account by renormalization of the electron mass and charge \cite{milonni1976semiclassical}.
+We focus only on the interaction of the emitter with low-energy cavity modes where strong interaction is possible.
+For these cavity modes, a natural cutoff emerges based on the transparency of the cavity.}
+
+\subsection{Generalized Coulomb gauge condition of Equation (7) of the main text}
+
+\textcolor{blue}{In the main text, we arrive at the generalized Coulomb gauge for the modes.
+The generalized Coulomb gauge, $\nabla\cdot\epsilon(\mathbf{x})\mathbf{A}(\mathbf{x})=0$ for some dielectric permittivity $\epsilon(\mathbf{x})$, is the most common gauge for analyzing phenomena in the discipline of macroscopic QED, in which one seeks to analyze the interaction of charges with fields in dielectric media \cite{glauber1991quantum}.
+The reason a particular gauge was chosen was that in the Hamiltonian, we have already assumed that all interaction with the quantized modes is manifested only through the vector potential, and that there is no scalar potential contribution to the interaction: relaxing this assumption allows more general gauge freedom.
+We also remark that if we performed a frame transformation, for example to the length gauge, the energy would be the same as that of our ansatz in the velocity gauge, although the state would be markedly different, and would be a highly correlated state.}
 
@@
 
-\textcolor{blue}{As discussed in the main text, the variational energies, as well as the energies from numerical diagonalization, are compared against the energies calculated from perturbation theory.
+As discussed in the main text, the variational energies, as well as the energies from numerical diagonalization, are compared against the energies calculated from perturbation theory.
 We provide additional details on the perturbative calculations here.
@@
 \end{equation}
-where $b$ denotes a matter state different from $a$, and $p_{ba} = \langle b|p|a\rangle$.}
+where $b$ denotes a matter state different from $a$, and $p_{ba} = \langle b|p|a\rangle$.
 
-\textcolor{blue}{In the main text, we also considered expectation values such as $\langle A \cdot p \rangle$ in the ground state within perturbation theory.
+In the main text, we also considered expectation values such as $\langle A \cdot p \rangle$ in the ground state within perturbation theory.
 Calculating the shift of the ground-state wavefunction at first-order in perturbation theory, we immediately find that
@@
 \langle Ap \rangle = \frac{2q}{m} \text{Re } \sum\limits_{b\neq a}\sum\limits_{n=1}^{N_p} \frac{1}{2\epsilon_0L}\frac{|p_{ba}|^2\sin^2\left(\frac{n\pi z}{L}\right)}{\omega_n(\omega_a-\omega_b-\omega_n)}.
-\end{equation}}
+\end{equation}
 
-\textcolor{blue}{It is also of interest to consider "field-only" observables such as the number of (virtual) photons in the ground state.
+It is also of interest to consider "field-only" observables such as the number of (virtual) photons in the ground state.
 Calculating the wavefunction of the ground state at first-order in perturbation theory, and calculating the probability of a photon being in the ground state, one finds that
@@
 2b of the main text).
-The number of photons in the bare basis differs substantially from the number of photons in the interacting basis.}
+The number of photons in the bare basis differs substantially from the number of photons in the interacting basis.
 
-\textcolor{blue}{This behavior does not reflect any inconsistency with the variational method, but is instead a feature which is to be expected.
+This behavior does not reflect any inconsistency with the variational method, but is instead a feature which is to be expected.
 In particular, the physics of the ansatz of Eq.
@@
 2b of the main text.
-That said, our variational theory does not give a prescription to translate the interacting mode operators to the bare mode operators.}
+That said, our variational theory does not give a prescription to translate the interacting mode operators to the bare mode operators.
 
@@
 \includegraphics[width=6cm]{FigureS1.pdf}
-\caption{\textcolor{blue}{\textbf{Number of virtual photons (bare and interacting) in the ground state calculated variationally, numerically, and through perturbation theory.} Parameters are the same as in Fig.
-2 (top panel) of the main text.}}
+\caption{\textbf{Number of virtual photons (bare and interacting) in the ground state calculated variationally, numerically, and through perturbation theory.} Parameters are the same as in Fig.
+2 (top panel) of the main text.}
 \label{fig:ansatz}
@@
 
-\textcolor{blue}{We note that in all cases, when calculating these quantities in the variational theory, these expressions are still relevant, with the replacement of the bare photon frequencies and mode functions replaced by the ones that result from Eq.
-7 of the main text.}
+We note that in all cases, when calculating these quantities in the variational theory, these expressions are still relevant, with the replacement of the bare photon frequencies and mode functions replaced by the ones that result from Eq.
+7 of the main text.
 
@@
 
-\textcolor{blue}{Regarding the numerical diagonalization, we numerically implement the Hamiltonian of Eq.
+Regarding the numerical diagonalization, we numerically implement the Hamiltonian of Eq.
 (9) of the main text by building the matter operators ($H_0$ and $p$ as defined in the main text) in the basis of states corresponding to the tensor product of any matter state, and any photon state having four photons or less .
@@
 In particular, $A(z) = \sum_n\sqrt{\frac{\hbar}{\epsilon_0\omega_n L}}\sin\left(\frac{n\pi z}{L} \right) \left(a_n+a^{\dagger}_n\right)$, with $L$ the length of the cavity and $\omega_n = \frac{n\pi c}{L}$.
-Diagonalization is performed using standard sparse eigendecomposition routines (such as those implemented in MATLAB).}
+Diagonalization is performed using standard sparse eigendecomposition routines (such as those implemented in MATLAB).
 
